# Bayesian Causal Inference

```{r}
#| include: false
library(tidyverse)
library(BART)
library(bcf)
library(broom)
```

```{r}
#| eval: false
library(tidyverse)
library(BART)
library(bcf)
```

The earlier chapters mostly compute point estimates and standard errors.
Bayesian methods instead put priors on unknown quantities and return a
posterior distribution. For causal inference this is useful when we want
uncertainty about a function of the model, such as an individual treatment
effect or a policy value, not just one coefficient.

I use two models:

1. **BART for causal inference** (Hill 2011), where BART is used as a
   flexible outcome model.
2. **Bayesian Causal Forests** (BCF; Hahn, Murray, & Carvalho 2020), which
   separates the baseline outcome model from the treatment-effect model.

> **Related reading**: The chapter [Using Numpyro](https://xiangao.github.io/blog_book/numpyro.html)
> in *Topics on Econometrics and Causal Inference* demonstrates a
> Bayesian hierarchical model with Numpyro. For an MCMC-from-scratch
> perspective on causal inference, that companion piece is the natural
> next step after this chapter.

## Setup

I use the same simulation as in the [Heterogeneous Treatment Effects](heterogeneous-effects.qmd)
chapter. The true treatment effect varies with `X1`.

```{r}
#| label: dgp
#| cache: true

set.seed(42)
n <- 2000
p <- 5

X <- matrix(runif(n * p), n, p)
colnames(X) <- paste0("X", 1:p)
# Treatment depends only on observed covariates, so unconfoundedness given X
# holds and the "True ATE" comparison measures estimator performance (not bias).
ps <- plogis(-0.3 + 1.5 * X[, 2])
D  <- rbinom(n, 1, ps)
tau <- 1 + 2 * X[, 1]
Y0  <- 0.5 * X[, 2] + rnorm(n)
Y1  <- Y0 + tau
Y   <- ifelse(D == 1, Y1, Y0)

df <- data.frame(Y = Y, D = D, X)
cat(sprintf("True ATE = %.3f\n", mean(tau)))
```

## BART for causal inference (Hill 2011)

BART is a sum of many small regression trees. The prior keeps each tree
weak, and the sum of trees gives flexibility.

For causal inference, the recipe is:

1. Fit BART for $\mu(d,x)=\mathbb{E}[Y \mid D=d,X=x]$.
2. For each unit $i$, predict counterfactuals $\mu(1, x_i)$ and $\mu(0, x_i)$.
3. The individual treatment effect (ITE) is $\hat\tau_i = \mu(1, x_i) - \mu(0, x_i)$.
4. Average the posterior draws to get the posterior for ATE or ITEs.

```{r}
#| label: bart-fit
#| cache: true

# BART::wbart fits a continuous-outcome BART model
# Combine D and X as the design matrix
X_train <- cbind(D = D, X)
set.seed(42)
bart_fit <- wbart(
  x.train = X_train,
  y.train = Y,
  ndpost = 1000,    # posterior samples after burn-in
  nskip  = 200,     # burn-in
  printevery = 1000
)

# Counterfactual predictions
X_treat   <- cbind(D = rep(1, n), X)
X_control <- cbind(D = rep(0, n), X)

mu1_post <- predict(bart_fit, newdata = X_treat)   # 1000 x n posterior draws
mu0_post <- predict(bart_fit, newdata = X_control)

# Posterior of individual treatment effects
tau_post <- mu1_post - mu0_post   # 1000 x n posterior draws of ITEs

# Posterior summaries
tau_mean <- rowMeans(tau_post)    # ATE in each draw
ate_post_mean   <- mean(tau_mean)
ate_post_sd     <- sd(tau_mean)
ate_post_ci     <- quantile(tau_mean, c(0.025, 0.975))

cat(sprintf("BART ATE posterior:\n"))
cat(sprintf("  Mean:    %.3f\n", ate_post_mean))
cat(sprintf("  SD:      %.3f\n", ate_post_sd))
cat(sprintf("  95%% CrI: [%.3f, %.3f]\n", ate_post_ci[1], ate_post_ci[2]))
cat(sprintf("  True ATE: %.3f\n", mean(tau)))
```

The posterior draws give the uncertainty in the ATE directly.

### Individual-level credible intervals

BART also gives a posterior for each unit's treatment effect. Here I plot
the posterior mean and pointwise credible intervals against the true
$\tau_i$.

```{r}
#| label: bart-ite
#| cache: true
#| fig-cap: "BART-estimated individual treatment effects (mean of posterior) vs true τ(x) = 1 + 2 X₁. Shaded band shows pointwise 95% credible intervals."
#| fig-width: 8
#| fig-height: 5

ite_mean <- colMeans(tau_post)
ite_lo   <- apply(tau_post, 2, quantile, 0.025)
ite_hi   <- apply(tau_post, 2, quantile, 0.975)

df_ite <- tibble(
  X1     = X[, 1],
  true   = tau,
  est    = ite_mean,
  lo     = ite_lo,
  hi     = ite_hi
)

ggplot(df_ite, aes(x = X1)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.2,
              fill = "steelblue") +
  geom_point(aes(y = est), size = 0.4, alpha = 0.3, colour = "steelblue") +
  geom_abline(aes(intercept = 1, slope = 2), colour = "firebrick",
              linetype = "dashed", linewidth = 1) +
  labs(x = "X1", y = "Individual treatment effect",
       caption = "Red dashed = true τ(x). Blue band = pointwise 95% credible interval.") +
  theme_minimal()
```

The blue ribbon is the pointwise 95% credible interval. The red dashed line
is the true treatment-effect function.

## Bayesian Causal Forests (BCF)

BART models $\mathbb{E}[Y \mid D,X]$ directly. That means the same trees
have to learn the baseline outcome and the treatment-effect heterogeneity.
When strong predictors of $Y(0)$ are also related to treatment, this can
create noisy treatment-effect estimates.

BCF writes the conditional mean as

$$
\mathbb{E}[Y \mid D, X] = \mu(X) + \tau(X)\, D,
$$ {#eq-bayesian-causal-1}

and puts separate priors on $\mu(X)$ and $\tau(X)$. The treatment-effect
part is regularized more strongly, which helps avoid overfitting
heterogeneity.

```{r}
#| label: bcf-fit
#| cache: true

# BCF requires a propensity score as an input ("piHat")
ps_fit <- glm(D ~ X1 + X2 + X3 + X4 + X5, data = df, family = binomial)
piHat  <- predict(ps_fit, type = "response")

# BCF prints MCMC progress in a way that conflicts with knitr's encoding.
# Its `verbose=FALSE`/`no_output=TRUE` arguments do not fully suppress this
# within the SAME R session (the progress output comes from C++ code, not
# from R's own message()/cat() machinery that sink()/suppressMessages()
# can intercept) -- so the only reliable fix is running bcf() in its own R
# process and only reading back the saved result, never its console output.
# Use SEPARATE input and output files so a failed subprocess can never leave
# the input list sitting where we expect the result, and check the exit
# status and object names before trusting it.
tmp_in  <- tempfile(fileext = ".rds")
tmp_out <- tempfile(fileext = ".rds")
tmp_err <- tempfile(fileext = ".log")
saveRDS(list(Y = df$Y, D = df$D, X = X, piHat = piHat), tmp_in)
status <- system(paste0("Rscript -e '",
  "args <- readRDS(\"", tmp_in, "\"); ",
  "suppressMessages(library(bcf)); ",
  "f <- bcf(y=args$Y, z=args$D, x_control=args$X, x_moderate=args$X, ",
  "pihat=args$piHat, nburn=200, nsim=1000, n_chains=1, no_output=TRUE, ",
  "verbose=FALSE); ",
  "saveRDS(list(tau=f$tau), \"", tmp_out, "\")'"),
  ignore.stdout = TRUE, ignore.stderr = FALSE)
# Surface the real cause instead of failing far downstream on a missing field.
if (status != 0L || !file.exists(tmp_out)) {
  stop("BCF subprocess failed (exit status ", status, "). ",
       "Check that the 'bcf' package is installed in the worker session.")
}
bcf_fit <- readRDS(tmp_out)
if (is.null(bcf_fit$tau)) {
  stop("BCF subprocess produced no 'tau' draws; aborting.")
}
unlink(c(tmp_in, tmp_out, tmp_err))
```

Now extract the posterior treatment effects:

```{r}
#| label: bcf-summary
#| cache: true

tau_post_bcf <- bcf_fit$tau    # nsim x n matrix of posterior ITEs

bcf_ate_post <- rowMeans(tau_post_bcf)
bcf_ite_mean <- colMeans(tau_post_bcf)

cat(sprintf("BCF ATE posterior:\n"))
cat(sprintf("  Mean:    %.3f\n", mean(bcf_ate_post)))
cat(sprintf("  SD:      %.3f\n", sd(bcf_ate_post)))
cat(sprintf("  95%% CrI: [%.3f, %.3f]\n",
            quantile(bcf_ate_post, 0.025),
            quantile(bcf_ate_post, 0.975)))
cat(sprintf("  True ATE: %.3f\n", mean(tau)))
```

### BART vs BCF: individual treatment effects

Compare the estimated CATEs:

```{r}
#| label: compare-ite
#| cache: true
#| fig-cap: "BCF (left) vs BART (right) individual treatment effect estimates. BCF's separation prior gives tighter, less noisy CATE estimates."
#| fig-width: 9
#| fig-height: 4

df_compare <- bind_rows(
  tibble(X1 = X[, 1], est = bcf_ite_mean, true = tau, method = "BCF"),
  tibble(X1 = X[, 1], est = ite_mean,     true = tau, method = "BART")
)

ggplot(df_compare, aes(X1, est)) +
  geom_point(alpha = 0.2, size = 0.3, colour = "steelblue") +
  geom_smooth(method = "loess", se = FALSE, colour = "darkblue",
              linewidth = 1) +
  geom_abline(intercept = 1, slope = 2, linetype = "dashed",
              colour = "firebrick") +
  facet_wrap(~ method) +
  labs(x = "X1", y = "Estimated ITE",
       caption = "Red dashed = truth = 1 + 2 X1") +
  theme_minimal()
```

BCF is usually smoother because it separates the prognostic and treatment
effect parts of the model.

## When to use Bayesian methods

| Reason to use Bayesian | Reason to use Frequentist |
|---|---|
| Prior information matters | Asymptotic theory is well-understood |
| Need full posterior (e.g. for downstream decisions) | Computational cost matters (BART is slow on large $n$) |
| Hierarchical / multilevel data structure | Effects are simple averages |
| Small sample, weak identification | Effects are well-identified |
| Want individual-level CrIs | Existing frequentist toolkit is sufficient |

Bayesian methods are most useful when the posterior itself will be used:
for individual treatment decisions, policy values, or hierarchical models.
If the target is just a well-identified ATE in a large sample, the
frequentist methods in earlier chapters are often simpler.

## Practical guidance

- Start with BCF if the goal is heterogeneous effects.
- Check MCMC diagnostics. A bad chain makes the posterior summaries useless.
- For BCF, pay attention to `piHat`; it is part of the model input.
- Report posterior mean, posterior SD, and credible intervals for ATE and
  meaningful CATE summaries.

## Connections

- The [Heterogeneous Treatment Effects](heterogeneous-effects.qmd)
  chapter covers frequentist meta-learners (S/T/X/R/DR) and causal
  forests via `grf`. BCF can be viewed as the Bayesian counterpart to
  causal forests: both use ensembles of trees, but BCF puts priors on
  the trees while `grf` uses honest sample-splitting.
- The [Sensitivity Analysis](sensitivity-analysis.qmd) chapter's
  Cinelli-Hazlett bounds are frequentist; the Bayesian analog is a
  prior on the unmeasured-confounder strength (cf. McCandless et al.
  2007).
- For Bayesian *Numpyro*-style hierarchical models, see the [companion
  blog chapter](https://xiangao.github.io/blog_book/numpyro.html).

## Summary

- BART estimates counterfactual outcomes by fitting a flexible outcome model.
- BCF separates baseline outcome prediction from treatment-effect
  heterogeneity.
- The main output is the posterior, especially for individual effects or
  policy decisions.
- Computation is the main cost. For very large data, frequentist HTE methods
  are usually easier.
