# Five Estimands on One DGP

```{r}
#| include: false
library(ggplot2)
library(dplyr)
library(gridExtra)
```

```{r}
#| eval: false
library(ggplot2)
library(dplyr)
library(gridExtra)
```

Before choosing an estimator, we need to know the estimand. ATE, ATT,
LATE, CATE and QTE are not five ways to estimate the same object. They are
five different objects. Sometimes they are numerically close, but that is a
feature of the data-generating process, not a general result.

Here I use one simulated data set so we can calculate all of them. The
simulation is useful because we observe both potential outcomes. In real
data we do not, which is exactly why identification assumptions matter.

## The data-generating process

We draw $n = 20{,}000$ people. Each has an observed covariate
$X_i \sim U(0,1)$, an unobserved type $U_i \sim N(0,1)$, and a randomly
assigned offer $Z_i \sim \text{Bernoulli}(0.5)$. The offer is the instrument.
It raises the probability of treatment but does not enter the outcome. The
treatment probabilities are $P(D=1 \mid Z=0, U) = 0.10 + 0.20U$ and
$P(D=1 \mid Z=1, U) = 0.10 + 0.20U + 0.50$, both clipped to $[0,1]$. One
uniform draw $V_i$ is compared against both thresholds, so $D_i(1) \ge D_i(0)$
for every unit and there are no defiers. The potential outcomes are
$Y_i(0) = 0.5X_i + 0.3U_i + \varepsilon_i$ with $\varepsilon_i \sim N(0,1)$,
and $Y_i(1) = Y_i(0) + \tau(X_i)$ with $\tau(x) = 1 + 2x$. We observe
$D_i = Z_i D_i(1) + (1-Z_i) D_i(0)$ and $Y_i = D_i Y_i(1) + (1-D_i) Y_i(0)$.

Two features of this design drive everything below. Selection into treatment
runs through $U$ and $V$, never through $X$. The treatment effect runs through
$X$ alone. And $X$ is drawn independently of $U$, $V$ and $Z$.

```{r}
#| label: dgp
#| cache: true

set.seed(42)
n <- 20000

X <- runif(n, 0, 1)        # observed covariate, in [0,1]
U <- rnorm(n)              # latent type, unobserved
Z <- rbinom(n, 1, 0.5)    # random instrument

# Treatment assignment: depends on Z (the instrument) and U (selection on type)
# Higher U → more likely to take treatment regardless of Z (always-takers)
# Lower  U → less likely to take treatment regardless of Z (never-takers)
# Middle U → responsive to Z (compliers)
pD0 <- pmin(pmax(0.10 + 0.20 * U,        0), 1)  # P(D=1 | Z=0, U)
pD1 <- pmin(pmax(0.10 + 0.20 * U + 0.50, 0), 1)  # P(D=1 | Z=1, U) — adds 0.50

# One latent draw V per person, thresholded against both probabilities.
# Since pD1 >= pD0, V < pD0 implies V < pD1, so D1 >= D0 for every unit:
# monotonicity (no defiers) holds by construction, as the LATE theorem requires.
V  <- runif(n)
D0 <- as.integer(V < pD0)   # potential treatment if not offered slot
D1 <- as.integer(V < pD1)   # potential treatment if offered slot
D  <- ifelse(Z == 1, D1, D0)

# Heterogeneous treatment effect: depends on X
# Y(d) = baseline + d * tau(X) + 0.3 * U + noise
tau_fn <- function(x) 1.0 + 2.0 * x   # the true individual treatment effect

Y0 <- 0.5 * X + 0.3 * U + rnorm(n)
Y1 <- Y0 + tau_fn(X)
Y  <- ifelse(D == 1, Y1, Y0)

df <- data.frame(X = X, Z = Z, D = D, Y = Y)
head(df)
```

The printed rows are the data an analyst would receive: $X$, $Z$, $D$ and $Y$.
We keep `Y0` and `Y1` aside, and every "true" number in this chapter is
computed from them. With real data each observation reveals only one of the
two.

## The five estimands

### ATE — average treatment effect

$$
\text{ATE} = \mathbb{E}[Y(1) - Y(0)]
$$ {#eq-causal-estimands-1}

ATE is the mean effect in the whole population. It compares the mean outcome
if everyone were treated with the mean outcome if nobody were treated. Here it
is the average of $\tau(x) = 1+2x$ over $X \sim U(0,1)$, which is exactly 2.
We average the individual effects in the sample and compare.

```{r}
#| label: ate
#| cache: true

ate_true <- mean(Y1 - Y0)
cat(sprintf("ATE (population mean of Y1 - Y0) = %.3f\n", ate_true))
# Theoretical ATE: integral of (1 + 2x) over Uniform[0,1] = 1 + 1 = 2
cat("Theoretical ATE = integral(1 + 2x)dx on [0,1] = 2.000\n")
```

The sample average is 1.996 against a population value of 2. The 0.004 gap is
simulation noise at $n = 20{,}000$.

### ATT — average treatment effect on the treated

$$
\text{ATT} = \mathbb{E}[Y(1) - Y(0) \mid D = 1]
$$ {#eq-causal-estimands-2}

ATT is the mean effect for the units that actually took treatment. It differs
from ATE when treatment selection is related to treatment-effect
heterogeneity. We average the individual effects over the treated units only.

```{r}
#| label: att
#| cache: true

att_true <- mean(Y1[D == 1] - Y0[D == 1])
cat(sprintf("ATT (mean of Y1-Y0 conditional on D=1) = %.3f\n", att_true))
```

ATT is 1.995 against an ATE of 1.996. The two are analytically identical in
this DGP, exactly $2.0$ in the population, so the small difference is
simulation noise and not a true gap between the estimands.
The reason is mechanical: the individual treatment effect $\tau_i = 1+2X_i$
depends only on $X_i$, while selection into treatment depends on $U_i$ (and
the instrument $Z_i$ below), and $X_i$ is generated independently of $U_i$
and $Z_i$. Since treatment-effect heterogeneity is unrelated to who selects
into treatment or who complies with the instrument, averaging $\tau_i$ over
the treated subpopulation or over compliers gives the same population
average as over everyone -- hence ATE = ATT = LATE = $2.0$ exactly here.
If selection (or compliance) also depended on $X$, these estimands would
diverge, since ATE, ATT, and LATE are different averages of $\tau_i$
across different, non-identical subpopulations.

### LATE — local average treatment effect (Imbens-Angrist)

$$
\text{LATE} = \mathbb{E}[Y(1) - Y(0) \mid D(1) > D(0)]
$$ {#eq-causal-estimands-3}

LATE is the mean effect for compliers, the units whose treatment status is
changed by the instrument. Under the usual IV assumptions, the Wald estimator
identifies this effect, not the ATE. Identifying LATE by the Wald ratio
additionally requires monotonicity (no defiers): $D(1) \ge D(0)$ for all $i$.
This DGP imposes it by generating both potential treatments from one shared
latent draw $V_i$ with ordered thresholds ($D(z) = \mathbb{1}\{V_i < p_{Dz}\}$
with $p_{D1} \ge p_{D0}$), so no unit is pushed *out* of treatment by the offer.
Raising the *probability* alone would not be enough: with independent draws for
$D(0)$ and $D(1)$, roughly 3% of units would be defiers and the Wald ratio would
no longer equal the complier mean.

We know $D_i(0)$ and $D_i(1)$ for every unit, so we can label the compliers and
average their individual effects directly. We then compute the Wald ratio from
$Y$, $D$ and $Z$ alone, which is all an analyst would have, and compare.

```{r}
#| label: late
#| cache: true

compliers <- (D1 == 1) & (D0 == 0)
late_true <- mean(Y1[compliers] - Y0[compliers])
cat(sprintf("LATE (mean of Y1-Y0 conditional on complier status) = %.3f\n", late_true))
cat(sprintf("Share of compliers: %.3f\n", mean(compliers)))

# Wald estimator from the data alone
wald <- (mean(Y[Z == 1]) - mean(Y[Z == 0])) /
        (mean(D[Z == 1]) - mean(D[Z == 0]))
cat(sprintf("Wald IV estimate (should match LATE): %.3f\n", wald))
```

Compliers are 45.6% of the sample and their mean effect is 1.999, again the
population value of 2. The Wald ratio gives 1.949. The two agree up to
sampling noise: the ratio divides a difference in mean outcomes by a treatment
rate difference of 0.456, which magnifies the noise in the numerator, and the
0.05 gap is about one standard error of the ratio. With heterogeneous effects
IV averages over the people moved by the instrument, and here that group has
the same distribution of $X$ as everyone else.

### CATE — conditional average treatment effect

$$
\text{CATE}(x) = \mathbb{E}[Y(1) - Y(0) \mid X = x]
$$ {#eq-causal-estimands-4}

CATE keeps $X$ in the estimand. In this DGP, CATE($x$) is simply
$\tau(x)=1+2x$. We cannot condition on a single point of a continuous $X$ in a
finite sample, so we cut $[0,1]$ into 20 equal-width bins, average the
individual effects within each bin, and print two of them next to the true
$\tau$ at the bin centre.

```{r}
#| label: cate
#| cache: true

nbins <- 20
breaks <- seq(0, 1, length.out = nbins + 1)
centers <- (breaks[-1] + breaks[-(nbins + 1)]) / 2
bin_idx <- cut(X, breaks = breaks, include.lowest = TRUE, labels = FALSE)

cate_est <- tapply(Y1 - Y0, bin_idx, mean)

cat(sprintf("CATE at x=%.3f: estimated %.2f, true %.2f\n",
            centers[4], cate_est[4], tau_fn(centers[4])))
cat(sprintf("CATE at x=%.3f: estimated %.2f, true %.2f\n",
            centers[16], cate_est[16], tau_fn(centers[16])))
```

The binned averages are 1.35 at $x = 0.175$ and 2.55 at $x = 0.775$, matching
$\tau$ at both bin centres to two decimals. The effect at $x=0.775$ is nearly
twice the effect at $x=0.175$, and no single scalar reports that.

### QTE — quantile treatment effect

$$
\text{QTE}(q) = F^{-1}_{Y(1)}(q) - F^{-1}_{Y(0)}(q)
$$ {#eq-causal-estimands-5}

QTE compares the two marginal outcome distributions. It is the difference
between a quantile under treatment and the same quantile under control. It
does not follow the same person across the two potential outcomes. We take the
quantiles of $Y(1)$ and of $Y(0)$ on a grid from 0.05 to 0.95 and difference
them, printing the 10th, 50th and 90th percentiles.

```{r}
#| label: qte
#| cache: true

qte_grid <- seq(0.05, 0.95, by = 0.05)
qte_est  <- quantile(Y1, qte_grid) - quantile(Y0, qte_grid)

cat(sprintf("QTE(0.10) = %.3f\n", quantile(Y1, 0.10) - quantile(Y0, 0.10)))
cat(sprintf("QTE(0.50) = %.3f\n", quantile(Y1, 0.50) - quantile(Y0, 0.50)))
cat(sprintf("QTE(0.90) = %.3f\n", quantile(Y1, 0.90) - quantile(Y0, 0.90)))
```

QTE rises from 1.709 at the 10th percentile to 2.009 at the median and 2.283
at the 90th. The reason is that $\tau(X) = 1+2X$ is largest where $X$ is
large, and $Y(0) = 0.5X + 0.3U + \varepsilon$ is also increasing in $X$, so
treatment adds most to the units that were already high in the control
distribution. Treatment stretches the distribution rather than shifting it.
The median QTE happens to sit near the ATE here, and there is no general
reason for that.

## All five estimands on two plots

The left panel plots the binned CATE against the true $\tau(x)$, with ATE, ATT
and LATE drawn as horizontal lines so the collapse to a scalar is visible. The
right panel plots QTE($q$) across the grid with the ATE marked for reference.

```{r}
#| label: estimand-plot
#| cache: true
#| fig-cap: "Five views of the same DGP. The horizontal lines for ATE/ATT/LATE collapse the effect to a single scalar. CATE(x) shows how the effect varies with X. QTE(q) shows how the effect varies across the outcome distribution. Each curve answers a different question."
#| fig-width: 10
#| fig-height: 4

df_cate <- data.frame(x = centers, cate = as.numeric(cate_est),
                      true_tau = tau_fn(centers))
df_qte  <- data.frame(q = qte_grid, qte = as.numeric(qte_est))

p1 <- ggplot(df_cate, aes(x = x)) +
  geom_line(aes(y = cate, colour = "CATE(x)"), linewidth = 1.2) +
  geom_line(aes(y = true_tau, colour = "True τ(x)"),
            linetype = "dotted", linewidth = 1) +
  geom_hline(aes(yintercept = ate_true,  colour = "ATE"),  linetype = "dashed") +
  geom_hline(aes(yintercept = att_true,  colour = "ATT"),  linetype = "dashed") +
  geom_hline(aes(yintercept = late_true, colour = "LATE"), linetype = "dashed") +
  scale_colour_manual(values = c(
    "CATE(x)" = "#c0392b", "True τ(x)" = "#e74c3c",
    "ATE" = "#2c3e50", "ATT" = "#2980b9", "LATE" = "#27ae60"
  )) +
  labs(x = "X (covariate)", y = "Effect", colour = NULL,
       title = "CATE(x) vs scalar estimands") +
  theme_minimal() +
  theme(legend.position = "bottom")

p2 <- ggplot(df_qte, aes(x = q, y = qte)) +
  geom_line(colour = "#8e44ad", linewidth = 1.2) +
  geom_hline(yintercept = ate_true, linetype = "dashed", colour = "#2c3e50") +
  annotate("text", x = 0.07, y = ate_true + 0.06, label = "ATE",
           colour = "#2c3e50", size = 3) +
  labs(x = "Quantile q", y = "QTE(q)",
       title = "QTE — distribution-level effect") +
  theme_minimal()

grid.arrange(p1, p2, ncol = 2)
```

The left panel shows why the scalar estimands can differ. ATE averages
$\tau(x)$ over the whole population. ATT averages it over the treated.
LATE averages it over compliers. CATE does not collapse the curve.

QTE, in the right panel, is different again. It describes how the treatment
changes the outcome distribution. It is not a conditional treatment effect
as a function of $X$.

## When the estimands differ

The three scalar estimands coincided because selection had nothing to do with
the effect modifier. We now break that, and see how far apart they move.

The second DGP is the first one with a single change: $X$ enters the treatment
probability, $P(D=1 \mid Z=0, U, X) = 0.10 + 0.20U + 0.6X$, clipped to $[0,1]$,
with the offer adding 0.50 as before. Everything else is unchanged --- a fresh
sample of $n = 20{,}000$, $X \sim U(0,1)$, $U \sim N(0,1)$,
$Z \sim \text{Bernoulli}(0.5)$, one shared uniform draw for the two potential
treatments, $Y(0) = 0.5X + 0.3U + \varepsilon$ and $\tau(x) = 1 + 2x$.

```{r}
#| label: dgp-selective
#| cache: true

set.seed(7)
n2 <- 20000
X2 <- runif(n2, 0, 1)
U2 <- rnorm(n2)
Z2 <- rbinom(n2, 1, 0.5)

# High X → much more likely to take treatment (selection on X, the modifier)
pD0_2 <- pmin(pmax(0.10 + 0.20 * U2 + 0.6 * X2, 0), 1)
pD1_2 <- pmin(pmax(pD0_2 + 0.50,                 0), 1)
V2    <- runif(n2)                  # shared draw → D1_2 >= D0_2, no defiers
D0_2  <- as.integer(V2 < pD0_2)
D1_2  <- as.integer(V2 < pD1_2)
D2    <- ifelse(Z2 == 1, D1_2, D0_2)
Y0_2  <- 0.5 * X2 + 0.3 * U2 + rnorm(n2)
Y1_2  <- Y0_2 + tau_fn(X2)
Y2    <- ifelse(D2 == 1, Y1_2, Y0_2)

ate2  <- mean(Y1_2 - Y0_2)
att2  <- mean(Y1_2[D2 == 1] - Y0_2[D2 == 1])
compliers2 <- (D1_2 == 1) & (D0_2 == 0)
late2 <- mean(Y1_2[compliers2] - Y0_2[compliers2])

cat("Selection-on-X DGP:\n")
cat(sprintf("  ATE  = %.3f\n", ate2))
cat(sprintf("  ATT  = %.3f  (now higher: treated have higher X → larger τ)\n", att2))
cat(sprintf("  LATE = %.3f  (compliers' mean X drives this)\n", late2))
```

ATE is still 2.001, because the distribution of $X$ has not moved. ATT rises
to 2.124 and LATE falls to 1.926. Since $\tau = 1 + 2X$, every subgroup
average is $1 + 2\,\mathbb{E}[X \mid \text{subgroup}]$, so the numbers report
the subgroups directly: the treated have mean $X$ of 0.562 and the compliers
have mean $X$ of 0.463, against 0.5 in the population.

The treated are high-$X$ because high $X$ raises the treatment probability.
The compliers are low-$X$ for a less obvious reason. The offer adds 0.50 to a
probability capped at 1, so a unit with $p_{D0}$ above 0.5 has a complier
window of width $1 - p_{D0}$, narrower than 0.5. High-$X$ units are the ones
the cap bites on: mean window width falls from 0.49 in the bottom quartile of
$X$ to 0.34 in the top, and the complier rate falls with it, from 0.48 to
0.34. So the instrument moves mostly low-$X$ units, and LATE lands below ATE.

Here the choice between ATE, ATT and LATE changes the substantive answer.

## Which estimand should you choose?

The estimand should come from the research question:

| Policy question | Right estimand |
|---|---|
| "What if we treated everyone?" | ATE |
| "What did treating the currently-treated achieve?" | ATT |
| "What can the instrument tell us?" | LATE |
| "Who benefits most?" | CATE($x$) |
| "Does the effect vary across the outcome distribution?" | QTE($q$) |
| "What is the distribution of individual effects?" | Often unidentifiable; bounds required |

It is fine for one paper to report more than one estimand, as long as each is
named correctly. What is not fine is to report the coefficient that is easiest
to estimate and call it "the causal effect."

## Summary

- ATE, ATT and LATE are different averages of the individual treatment
  effect.
- CATE($x$) keeps heterogeneity by covariates. QTE($q$) describes changes
  in the outcome distribution.
- With heterogeneous effects, IV estimates LATE. It should not be described
  as ATE unless extra assumptions justify that interpretation. This is the
  clean case with no covariates. Once we add covariates linearly to 2SLS,
  even LATE needs a *rich covariates* condition that is rarely defended.
  See @blandhol-et-al-2025 and the IV chapter section "When 2SLS with
  Covariates Is *Actually* LATE".
- Estimator choice comes after the estimand is fixed.
