# Five Estimands on One DGP

```{r}
#| include: false
library(ggplot2)
library(dplyr)
library(gridExtra)
```

```{r}
#| eval: false
library(ggplot2)
library(dplyr)
library(gridExtra)
```

Before choosing an estimator, we need to know the estimand. ATE, ATT,
LATE, CATE and QTE are not five ways to estimate the same object. They are
five different objects. Sometimes they are numerically close, but that is a
feature of the data-generating process, not a general result.

Here I use one simulated data set so we can calculate all of them. The
simulation is useful because we observe both potential outcomes. In real
data we do not, which is exactly why identification assumptions matter.

## The data-generating process

Each person has an observed covariate $X_i$, an unobserved type $U_i$, and
a binary treatment $D_i$. The instrument $Z_i$ is a random offer. It changes
the probability of treatment, but it has no direct effect on the outcome.

```{r}
#| label: dgp
#| cache: true

set.seed(42)
n <- 20000

X <- runif(n, 0, 1)        # observed covariate, in [0,1]
U <- rnorm(n)              # latent type, unobserved
Z <- rbinom(n, 1, 0.5)    # random instrument

# Treatment assignment: depends on Z (the instrument) and U (selection on type)
# Higher U → more likely to take treatment regardless of Z (always-takers)
# Lower  U → less likely to take treatment regardless of Z (never-takers)
# Middle U → responsive to Z (compliers)
pD0 <- pmin(pmax(0.10 + 0.20 * U,        0), 1)  # P(D=1 | Z=0, U)
pD1 <- pmin(pmax(0.10 + 0.20 * U + 0.50, 0), 1)  # P(D=1 | Z=1, U) — adds 0.50

D0 <- rbinom(n, 1, pD0)   # potential treatment if not offered slot
D1 <- rbinom(n, 1, pD1)   # potential treatment if offered slot
D  <- ifelse(Z == 1, D1, D0)

# Heterogeneous treatment effect: depends on X
# Y(d) = baseline + d * tau(X) + 0.3 * U + noise
tau_fn <- function(x) 1.0 + 2.0 * x   # the true individual treatment effect

Y0 <- 0.5 * X + 0.3 * U + rnorm(n)
Y1 <- Y0 + tau_fn(X)
Y  <- ifelse(D == 1, Y1, Y0)

df <- data.frame(X = X, Z = Z, D = D, Y = Y)
head(df)
```

In the simulated data we have `Y0` and `Y1` for every observation. That lets
us calculate the true estimands directly. With real data, each observation
only reveals one potential outcome.

## The five estimands

### ATE — average treatment effect

$$
\text{ATE} = \mathbb{E}[Y(1) - Y(0)]
$$

ATE is the mean effect in the whole population. It compares the mean outcome
if everyone were treated with the mean outcome if nobody were treated.

```{r}
#| label: ate
#| cache: true

ate_true <- mean(Y1 - Y0)
cat(sprintf("ATE (population mean of Y1 - Y0) = %.3f\n", ate_true))
# Theoretical ATE: integral of (1 + 2x) over Uniform[0,1] = 1 + 1 = 2
cat("Theoretical ATE = integral(1 + 2x)dx on [0,1] = 2.000\n")
```

### ATT — average treatment effect on the treated

$$
\text{ATT} = \mathbb{E}[Y(1) - Y(0) \mid D = 1]
$$

ATT is the mean effect for the units that actually took treatment. It differs
from ATE when treatment selection is related to treatment-effect
heterogeneity.

```{r}
#| label: att
#| cache: true

att_true <- mean(Y1[D == 1] - Y0[D == 1])
cat(sprintf("ATT (mean of Y1-Y0 conditional on D=1) = %.3f\n", att_true))
```

Here ATT is close to ATE. The reason is mechanical: treatment effects depend
on $X$, while selection into treatment depends on $U$, and $X$ is independent
of $U$. If selection also depended on $X$, ATT would move away from ATE.

### LATE — local average treatment effect (Imbens-Angrist)

$$
\text{LATE} = \mathbb{E}[Y(1) - Y(0) \mid D(1) > D(0)]
$$

LATE is the mean effect for compliers, the units whose treatment status is
changed by the instrument. Under the usual IV assumptions, the Wald estimator
identifies this effect, not the ATE.

```{r}
#| label: late
#| cache: true

compliers <- (D1 == 1) & (D0 == 0)
late_true <- mean(Y1[compliers] - Y0[compliers])
cat(sprintf("LATE (mean of Y1-Y0 conditional on complier status) = %.3f\n", late_true))
cat(sprintf("Share of compliers: %.3f\n", mean(compliers)))

# Wald estimator from the data alone
wald <- (mean(Y[Z == 1]) - mean(Y[Z == 0])) /
        (mean(D[Z == 1]) - mean(D[Z == 0]))
cat(sprintf("Wald IV estimate (should match LATE): %.3f\n", wald))
```

With heterogeneous effects, IV averages over the people moved by the
instrument. In this example the Wald estimate matches the complier mean
because $Z$ changes treatment only for that group.

### CATE — conditional average treatment effect

$$
\text{CATE}(x) = \mathbb{E}[Y(1) - Y(0) \mid X = x]
$$

CATE keeps $X$ in the estimand. In this DGP, CATE($x$) is simply
$\tau(x)=1+2x$.

```{r}
#| label: cate
#| cache: true

nbins <- 20
breaks <- seq(0, 1, length.out = nbins + 1)
centers <- (breaks[-1] + breaks[-(nbins + 1)]) / 2
bin_idx <- cut(X, breaks = breaks, include.lowest = TRUE, labels = FALSE)

cate_est <- tapply(Y1 - Y0, bin_idx, mean)

cat(sprintf("CATE at x=0.2: estimated %.2f, true %.2f\n",
            cate_est[4], tau_fn(0.2)))
cat(sprintf("CATE at x=0.8: estimated %.2f, true %.2f\n",
            cate_est[16], tau_fn(0.8)))
```

### QTE — quantile treatment effect

$$
\text{QTE}(\tau) = F^{-1}_{Y(1)}(\tau) - F^{-1}_{Y(0)}(\tau)
$$

QTE compares the two marginal outcome distributions. It is the difference
between a quantile under treatment and the same quantile under control. It
does not follow the same person across the two potential outcomes.

```{r}
#| label: qte
#| cache: true

qte_grid <- seq(0.05, 0.95, by = 0.05)
qte_est  <- quantile(Y1, qte_grid) - quantile(Y0, qte_grid)

cat(sprintf("QTE(0.10) = %.3f\n", quantile(Y1, 0.10) - quantile(Y0, 0.10)))
cat(sprintf("QTE(0.50) = %.3f\n", quantile(Y1, 0.50) - quantile(Y0, 0.50)))
cat(sprintf("QTE(0.90) = %.3f\n", quantile(Y1, 0.90) - quantile(Y0, 0.90)))
```

## All five estimands on one plot

```{r}
#| label: estimand-plot
#| cache: true
#| fig-cap: "Five views of the same DGP. The horizontal lines for ATE/ATT/LATE collapse the effect to a single scalar. CATE(x) shows how the effect varies with X. QTE(τ) shows how the effect varies across the outcome distribution. Each curve answers a different question."
#| fig-width: 10
#| fig-height: 4

df_cate <- data.frame(x = centers, cate = as.numeric(cate_est),
                      true_tau = tau_fn(centers))
df_qte  <- data.frame(tau = qte_grid, qte = as.numeric(qte_est))

p1 <- ggplot(df_cate, aes(x = x)) +
  geom_line(aes(y = cate, colour = "CATE(x)"), linewidth = 1.2) +
  geom_line(aes(y = true_tau, colour = "True τ(x)"),
            linetype = "dotted", linewidth = 1) +
  geom_hline(aes(yintercept = ate_true,  colour = "ATE"),  linetype = "dashed") +
  geom_hline(aes(yintercept = att_true,  colour = "ATT"),  linetype = "dashed") +
  geom_hline(aes(yintercept = late_true, colour = "LATE"), linetype = "dashed") +
  scale_colour_manual(values = c(
    "CATE(x)" = "#c0392b", "True τ(x)" = "#e74c3c",
    "ATE" = "#2c3e50", "ATT" = "#2980b9", "LATE" = "#27ae60"
  )) +
  labs(x = "X (covariate)", y = "Effect", colour = NULL,
       title = "CATE(x) vs scalar estimands") +
  theme_minimal() +
  theme(legend.position = "bottom")

p2 <- ggplot(df_qte, aes(x = tau, y = qte)) +
  geom_line(colour = "#8e44ad", linewidth = 1.2) +
  geom_hline(yintercept = ate_true, linetype = "dashed", colour = "#2c3e50") +
  annotate("text", x = 0.07, y = ate_true + 0.06, label = "ATE",
           colour = "#2c3e50", size = 3) +
  labs(x = "Quantile τ", y = "QTE(τ)",
       title = "QTE — distribution-level effect") +
  theme_minimal()

grid.arrange(p1, p2, ncol = 2)
```

The left panel shows why the scalar estimands can differ. ATE averages
$\tau(x)$ over the whole population. ATT averages it over the treated.
LATE averages it over compliers. CATE does not collapse the curve.

QTE, in the right panel, is different again. It describes how the treatment
changes the outcome distribution. It is not a conditional treatment effect
as a function of $X$.

## When the estimands differ

In the first DGP the scalar estimands are close because:

- $X$ is uniform on [0, 1] (so ATE = average of τ over uniform $X$ = 2)
- Treatment selection is on $U$ (unobserved type), not $X$ (which drives τ),
  so ATT ≈ ATE
- The instrument shifts compliers uniformly across $X$, so LATE ≈ ATE

If treatment selection depends on the effect modifier, the numbers separate.
Here I make high-$X$ units more likely to take treatment:

```{r}
#| label: dgp-selective
#| cache: true

set.seed(7)
n2 <- 20000
X2 <- runif(n2, 0, 1)
U2 <- rnorm(n2)
Z2 <- rbinom(n2, 1, 0.5)

# High X → much more likely to take treatment (selection on X, the modifier)
pD0_2 <- pmin(pmax(0.10 + 0.20 * U2 + 0.6 * X2, 0), 1)
pD1_2 <- pmin(pmax(pD0_2 + 0.50,                 0), 1)
D0_2  <- rbinom(n2, 1, pD0_2)
D1_2  <- rbinom(n2, 1, pD1_2)
D2    <- ifelse(Z2 == 1, D1_2, D0_2)
Y0_2  <- 0.5 * X2 + 0.3 * U2 + rnorm(n2)
Y1_2  <- Y0_2 + tau_fn(X2)
Y2    <- ifelse(D2 == 1, Y1_2, Y0_2)

ate2  <- mean(Y1_2 - Y0_2)
att2  <- mean(Y1_2[D2 == 1] - Y0_2[D2 == 1])
compliers2 <- (D1_2 == 1) & (D0_2 == 0)
late2 <- mean(Y1_2[compliers2] - Y0_2[compliers2])

cat("Selection-on-X DGP:\n")
cat(sprintf("  ATE  = %.3f\n", ate2))
cat(sprintf("  ATT  = %.3f  (now higher: treated have higher X → larger τ)\n", att2))
cat(sprintf("  LATE = %.3f  (compliers' mean X drives this)\n", late2))
```

Now ATT is larger than ATE. The treated group has higher $X$, and high $X$
means a larger treatment effect. In this case reporting ATE or ATT changes
the substantive answer.

## Which estimand should you choose?

The estimand should come from the research question:

| Policy question | Right estimand |
|---|---|
| "What if we treated everyone?" | ATE |
| "What did treating the currently-treated achieve?" | ATT |
| "What can the instrument tell us?" | LATE |
| "Who benefits most?" | CATE($x$) |
| "Does the effect vary across the outcome distribution?" | QTE($\tau$) |
| "What is the distribution of individual effects?" | Often unidentifiable; bounds required |

It is fine for one paper to report more than one estimand, as long as each is
named correctly. What is not fine is to report the coefficient that is easiest
to estimate and call it "the causal effect."

## Summary

- ATE, ATT and LATE are different averages of the individual treatment
  effect.
- CATE($x$) keeps heterogeneity by covariates. QTE($\tau$) describes changes
  in the outcome distribution.
- With heterogeneous effects, IV estimates LATE. It should not be described
  as ATE unless extra assumptions justify that interpretation.
- Estimator choice comes after the estimand is fixed.
