# Continuous and Multivalued Treatments

```{r}
#| include: false
library(tidyverse)
library(WeightIt)
library(causaldrf)
library(lmtp)
library(SuperLearner)
library(broom)
```

```{r}
#| eval: false
library(tidyverse)
library(WeightIt)
library(causaldrf)
library(lmtp)
library(SuperLearner)
```

So far most examples use a binary treatment. Many treatments are not binary:
dose of a drug, hours in a program, dollars spent, or a treatment with
several levels. In those cases "the treatment effect" is not one number. We
usually want a dose-response curve,

$$
\mu(d) = \mathbb{E}[Y(d)], \quad d \in \mathcal{D}.
$$

I use three approaches:

1. **Generalized propensity score** (GPS; Hirano-Imbens 2004), which models
   the conditional density of treatment given covariates.
2. **Doubly-robust dose-response estimation** [in the spirit of
   @kennedy-2017-continuous], which combines a GPS with an outcome model.
3. **Longitudinal modified treatment policies** [LMTP; @diaz-2023-lmtp],
   which are useful when treatments are time-varying or when fixed treatment
   levels create positivity problems.

> **Related reading**: The LMTP material below is condensed from the
> longer treatment in [Longitudinal modified treatment policy (LMTP)](https://xiangao.github.io/blog_book/lmtp.html)
> in *Topics on Econometrics and Causal Inference*, which follows
> Nicholas Williams's tutorials at
> [beyondtheate.com](https://beyondtheate.com/).

## The continuous-treatment setup

Start with a simple simulation. Treatment is a dose, and the dose depends on
covariates:

```{r}
#| label: dgp
#| cache: true

set.seed(42)
n  <- 2000
X1 <- rnorm(n)
X2 <- rnorm(n)
# Continuous treatment, depends on X1, X2
D  <- 2 + 0.5 * X1 + 0.5 * X2 + rnorm(n, sd = 1.0)
# Dose-response: nonlinear in D
true_mu <- function(d) 1.0 + 2.0 * d - 0.15 * d^2
Y  <- true_mu(D) + 0.3 * X1 - 0.3 * X2 + rnorm(n)

df <- tibble(Y = Y, D = D, X1 = X1, X2 = X2)
cat(sprintf("Treatment range: [%.2f, %.2f]\n", min(D), max(D)))
cat(sprintf("True μ(d=2) = %.2f, μ(d=4) = %.2f, μ(d=6) = %.2f\n",
            true_mu(2), true_mu(4), true_mu(6)))
```

The true dose-response curve is hump-shaped. It increases at low doses and
falls at high doses. The estimation problem is to recover this shape when
the observed dose is confounded by $X$.

## Naive regression vs adjusted regression

A linear regression of $Y$ on $D$ has two problems: it ignores confounding
by $X$, and it imposes a straight line on a nonlinear dose-response curve.

```{r}
#| label: naive
#| cache: true

# Linear fit, ignoring covariates
linear_naive <- lm(Y ~ D, data = df)
# Linear fit with covariate adjustment
linear_adj   <- lm(Y ~ D + X1 + X2, data = df)
# Nonlinear fit with covariate adjustment
poly_adj     <- lm(Y ~ poly(D, 2) + X1 + X2, data = df)

ds <- seq(0, 10, by = 0.5)

predict_mu <- function(fit, ds) {
  data.frame(d = ds, X1 = 0, X2 = 0) |>
    rename(D = d) -> nd
  predict(fit, newdata = nd)
}

# True dose-response
df_true <- tibble(d = ds, mu = true_mu(ds))

df_fits <- bind_rows(
  tibble(d = ds, mu = predict_mu(linear_naive, ds), method = "Linear (naive)"),
  tibble(d = ds, mu = predict_mu(linear_adj,   ds), method = "Linear (adj X)"),
  tibble(d = ds, mu = predict_mu(poly_adj,     ds), method = "Quadratic (adj X)")
)
```

```{r}
#| label: naive-plot
#| fig-cap: "Fitted vs true dose-response. Linear fits miss the curvature; the quadratic-adjusted fit comes closer."
#| fig-width: 7
#| fig-height: 4

ggplot() +
  geom_line(data = df_fits, aes(d, mu, colour = method), linewidth = 1.0) +
  geom_line(data = df_true, aes(d, mu), colour = "black",
            linetype = "dashed", linewidth = 1.0) +
  labs(x = "Dose D", y = "Expected outcome E[Y(d)]",
       colour = "Method",
       caption = "Black dashed line = true μ(d) = 1 + 2d - 0.15 d²") +
  theme_minimal()
```

The quadratic adjusted model is better in this simulation because the true
curve is quadratic. In real applications, the functional form is usually not
known.

## Generalized propensity score (GPS)

The **generalized propensity score** is the conditional density of the
treatment given covariates:

$$
r(d, x) = f_{D \mid X}(d \mid x).
$$

Under unconfoundedness, the GPS plays the same role as the propensity score
for binary treatment. In practice we often start with a normal model for
$D \mid X$:
$D \mid X \sim \mathcal{N}(\beta_0+\beta'X,\sigma^2)$.

```{r}
#| label: gps
#| cache: true

# Fit GPS model: D ~ X with Normal residuals
gps_model <- lm(D ~ X1 + X2, data = df)
d_pred    <- predict(gps_model)
sigma_e   <- sigma(gps_model)

# GPS for each unit's observed treatment
gps <- dnorm(df$D, mean = d_pred, sd = sigma_e)
df$gps <- gps

# IPW-style weights for continuous treatment: 1 / GPS
# Stabilise by multiplying by the marginal density f_D(D)
# NOTE: sw_cont is computed only as an overlap/positivity diagnostic.
# The Hirano-Imbens estimator below does NOT weight by it: the GPS
# enters as a regressor in the outcome model.
marginal_D <- density(df$D)
fD <- approx(marginal_D$x, marginal_D$y, xout = df$D)$y
df$sw_cont <- pmin(fD / gps, quantile(fD / gps, 0.99))

cat(sprintf("Weight summary: min=%.3f, mean=%.3f, max=%.3f\n",
            min(df$sw_cont), mean(df$sw_cont), max(df$sw_cont)))
```

The Hirano-Imbens estimator has two steps. First regress $Y$ on flexible
functions of $D$ and the estimated GPS. Then, for each dose $d$, average the
predicted outcome over the sample.

```{r}
#| label: hi-est
#| cache: true

# Stage 2: regress Y on flexible function of (D, GPS)
hi_fit <- lm(Y ~ poly(D, 2) + poly(gps, 2) + I(D * gps), data = df)

# Dose-response at each d: average over the empirical distribution of GPS
# at that d.
estimate_dose <- function(d) {
  # Predict GPS at this d for each unit
  gps_at_d <- dnorm(d, mean = d_pred, sd = sigma_e)
  nd <- data.frame(D = d, gps = gps_at_d)
  mean(predict(hi_fit, newdata = nd))
}

ds <- seq(0, 10, by = 0.5)
hi_dose <- sapply(ds, estimate_dose)

df_hi <- tibble(d = ds, mu = hi_dose, method = "Hirano-Imbens GPS")

ggplot() +
  geom_line(data = df_hi, aes(d, mu, colour = method), linewidth = 1.2) +
  geom_line(data = df_true, aes(d, mu), colour = "black",
            linetype = "dashed", linewidth = 1.0) +
  labs(x = "Dose D", y = "E[Y(d)]",
       caption = "Dashed black = truth; coloured = Hirano-Imbens GPS estimator") +
  theme_minimal()
```

In this example the GPS estimator tracks the nonlinear curve much better
than the naive linear regression.

## Doubly-robust dose-response estimation

A kernel-localized AIPW estimator — in the spirit of
@kennedy-2017-continuous, who develop a more elaborate pseudo-outcome
version with formal guarantees — is the continuous-treatment analogue of
AIPW. It combines:

- An outcome model $\hat\mu(d, x) = \mathbb{E}[Y \mid D = d, X = x]$
- A generalised propensity score $\hat r(d \mid x)$

Then the dose-response at $d$ is estimated by

$$
\hat\mu_{DR}(d) = \frac{1}{n} \sum_i \left[ \hat\mu(d, X_i) + K_h(D_i - d) \cdot \frac{Y_i - \hat\mu(D_i, X_i)}{\hat r(D_i \mid X_i)} \right],
$$

where $K_h$ is a kernel around dose $d$. The bandwidth $h$ controls how much
nearby observed doses contribute to the estimate. (The code below uses the
weight-normalized — Hájek — version of the correction term, dividing the
weights by their mean; the population limit is unchanged.)

Using the GPS already estimated above, the implementation is short:

```{r}
#| label: drf
#| cache: true

# Outcome model μ̂(D, X): flexible polynomial
mu_fit <- lm(Y ~ D + I(D^2) + X1 + X2 + I(D * X1) + I(D * X2), data = df)
df$mu_at_obs <- predict(mu_fit)

# DR estimator at dose d with bandwidth h
dr_estimate <- function(d, h = 0.5) {
  # Counterfactual prediction at D = d for each unit
  mu_at_d <- predict(mu_fit, newdata = transform(df, D = d))
  # Kernel weight on observed treatment near d
  kernel <- dnorm(df$D - d, sd = h)
  weight <- kernel / df$gps
  weight <- weight / mean(weight)  # normalise
  # DR correction
  correction <- weight * (df$Y - df$mu_at_obs)
  mean(mu_at_d) + mean(correction)
}

ds <- seq(0, 10, by = 0.5)
dr_dose <- sapply(ds, dr_estimate)

df_drf <- tibble(d = ds, mu = dr_dose, method = "DR (kernel AIPW)")

ggplot() +
  geom_line(data = df_hi,  aes(d, mu, colour = method), linewidth = 1.0) +
  geom_line(data = df_drf, aes(d, mu, colour = method), linewidth = 1.0) +
  geom_line(data = df_true, aes(d, mu), colour = "black",
            linetype = "dashed", linewidth = 1.0) +
  labs(x = "Dose D", y = "E[Y(d)]",
       caption = "Black dashed = truth") +
  theme_minimal()
```

Read this as an illustrative kernel-AIPW construction rather than a
turnkey doubly robust theorem. For continuous treatments, a formal
doubly robust dose-response result -- consistency if either the
outcome model or the GPS is correct -- requires additional bandwidth,
nuisance-convergence-rate, smoothness, and orthogonalization conditions
(see Kennedy et al. on continuous-treatment DR estimators). The
self-normalized kernel correction shown here conveys the intuition but
does not by itself establish those guarantees. As usual, the bandwidth
trades bias against variance.

The `causaldrf` package provides ready-made versions of these estimators,
including `hi_est`, `iptw_est`, `aipwee_est`, `gam_est`, `nw_est`, and
`bart_est`.

## Longitudinal Modified Treatment Policies (LMTP)

Fixed-dose interventions can be unrealistic. For example, there may be
almost no observations near $D=10$, so estimating $Y(10)$ requires
extrapolation. LMTP avoids this by defining interventions as changes to the
observed treatment, such as "shift everyone's dose down by one unit." This
keeps the counterfactual treatment closer to the observed support.

### A two-period example

```{r}
#| label: lmtp-sim
#| cache: true

set.seed(1)
N    <- 1500
L_1  <- runif(N, 0, 1)
A_1  <- rbinom(N, 1, 0.5)
L_2  <- rnorm(N, 0.25 * L_1, 0.5)
A_2  <- rbinom(N, 1, plogis(0.5 - 0.2 * A_1 + 0.1 * L_2))
Y    <- rnorm(N, A_2 + L_2, 0.3)

lmtp_data <- data.frame(L_1, A_1, L_2, A_2, Y)
head(lmtp_data)
```

For simplicity the worked example uses *static* policies (set every
treatment to 0, then to 1) on a binary treatment — `lmtp` handles these in
the same framework. A genuinely *modified* policy on a continuous dose
would instead return a shift of the observed value, e.g.
`function(data, trt) data[[trt]] - 1`, which is what keeps the
counterfactual on the observed support. Here is the set-to-zero policy:

```{r}
#| label: shift-fn

shift_zero <- function(data, trt) {
  # Set treatment to 0 everywhere — for comparison to "never treat"
  a <- data[[trt]]
  rep(0, length(a))
}
```

Then fit the LMTP estimator with `lmtp::lmtp_tmle`:

```{r}
#| label: lmtp-fit
#| cache: true
#| warning: false

# TMLE-based LMTP estimator with super-learners for nuisance functions
# (using SL.glm to keep it fast; in practice use richer SL libraries)
lmtp_fit <- lmtp_tmle(
  data       = lmtp_data,
  trt        = c("A_1", "A_2"),
  outcome    = "Y",
  baseline   = NULL,
  time_vary  = list(c("L_1"), c("L_2")),
  shift      = shift_zero,
  outcome_type = "continuous",
  learners_outcome = c("SL.glm"),
  learners_trt     = c("SL.glm"),
  folds = 2
)

# tidy() gives the estimate, SE, and CI as a data frame
# (the styled console print does not render cleanly in the book)
tidy(lmtp_fit)
```

The useful part of `lmtp` is that it handles the time-ordering of
covariates, treatments, and outcomes. The output gives the TMLE estimate
with its standard error and confidence interval.

### Contrasts between policies

We can also compare two policies, such as "set to 0" versus "set to 1":

```{r}
#| label: lmtp-contrast
#| cache: true
#| warning: false

shift_one <- function(data, trt) {
  a <- data[[trt]]
  rep(1, length(a))
}

lmtp_one <- lmtp_tmle(
  data       = lmtp_data,
  trt        = c("A_1", "A_2"),
  outcome    = "Y",
  baseline   = NULL,
  time_vary  = list(c("L_1"), c("L_2")),
  shift      = shift_one,
  outcome_type = "continuous",
  learners_outcome = c("SL.glm"),
  learners_trt     = c("SL.glm"),
  folds = 2
)

contrast <- lmtp_contrast(lmtp_one, ref = lmtp_fit)
as.data.frame(contrast$estimates)
```

The contrast is the difference between the two policy-specific mean
outcomes.

## Choosing among the three

| Method | Strengths | Weaknesses |
|---|---|---|
| **GPS (Hirano-Imbens)** | Conceptually transparent; widely understood | Parametric assumptions on the GPS and the outcome-GPS regression; sensitive to their misspecification |
| **Doubly-robust** (kernel AIPW) | Robust to misspecification of one nuisance; attains nonparametric oracle rates | More complex; nuisance models need careful tuning; sensitive to bandwidth |
| **LMTP** | Handles continuous + multivalued + time-varying uniformly; respects positivity | More setup; requires defining a shift function |

In practice:

- For a simple cross-sectional continuous treatment with good overlap, GPS
  or the Kennedy DR estimator is a natural start.
- For time-varying treatments or policy shifts, use LMTP.
- For multivalued treatments, use a multinomial propensity model and the
  same adjustment logic.

## When to use these methods

These methods are useful when:

1. Treatment has more than two values.
2. The research question is about the shape of the dose-response curve.
3. The intervention is better described as shifting treatment than setting
   treatment to a fixed value.

For strictly binary, single-time-period problems, the standard
[Estimation](estimation.qmd) and [Heterogeneous Effects](heterogeneous-effects.qmd)
methods are sufficient.

## Summary

- Continuous and multivalued treatments usually require a dose-response
  curve, not a single coefficient.
- GPS extends propensity-score logic to continuous treatments by modeling
  the density of treatment given covariates.
- Kennedy's DR estimator adds an outcome model to the GPS.
- LMTP is useful when the intervention is a policy shift or when treatment is
  time-varying.
