# Continuous and Multivalued Treatments

```{r}
#| include: false
library(tidyverse)
library(WeightIt)
library(causaldrf)
library(lmtp)
library(SuperLearner)
library(broom)
```

```{r}
#| eval: false
library(tidyverse)
library(WeightIt)
library(causaldrf)
library(lmtp)
library(SuperLearner)
```

The previous chapters treated $D$ as binary — a single yes/no decision.
Real treatments are often *continuous* (dose of a medication, dollars
spent, hours of exposure) or *multivalued* (one of several treatment
arms). For these, we cannot speak of "the effect of treatment" as a
single scalar; instead we estimate the entire **dose-response curve**

$$
\mu(d) = \mathbb{E}[Y(d)], \quad d \in \mathcal{D}.
$$

This chapter covers three approaches:

1. **Generalized propensity score** (GPS; Hirano-Imbens 2004) — extends
   propensity score methods to continuous treatments via the
   density-of-treatment-given-covariates.
2. **Doubly-robust dose-response estimation** (Kennedy 2017) — combines
   GPS with an outcome model for efficiency and robustness.
3. **Longitudinal modified treatment policies** (LMTP; Diaz-Hejazi 2020)
   — a unified framework for continuous, multivalued, and time-varying
   treatments, with native handling of positivity violations.

> **Related reading**: The LMTP material below is condensed from the
> longer treatment in [Longitudinal modified treatment policy (LMTP)](https://xiangao.github.io/blog_book/lmtp.html)
> in *Topics on Econometrics and Causal Inference*, which follows
> Nicholas Williams's tutorials at
> [beyondtheate.com](https://beyondtheate.com/).

## The continuous-treatment setup

A simulation makes the conceptual issue clear. Suppose treatment is a
dose $D \in [0, 10]$ assigned via a generalized propensity score that
depends on covariates:

```{r}
#| label: dgp
#| cache: true

set.seed(42)
n  <- 2000
X1 <- rnorm(n)
X2 <- rnorm(n)
# Continuous treatment, depends on X1, X2
D  <- 2 + 0.5 * X1 + 0.5 * X2 + rnorm(n, sd = 1.0)
# Dose-response: nonlinear in D
true_mu <- function(d) 1.0 + 2.0 * d - 0.15 * d^2
Y  <- true_mu(D) + 0.3 * X1 - 0.3 * X2 + rnorm(n)

df <- tibble(Y = Y, D = D, X1 = X1, X2 = X2)
cat(sprintf("Treatment range: [%.2f, %.2f]\n", min(D), max(D)))
cat(sprintf("True μ(d=2) = %.2f, μ(d=4) = %.2f, μ(d=6) = %.2f\n",
            true_mu(2), true_mu(4), true_mu(6)))
```

The dose-response curve $\mu(d) = 1 + 2d - 0.15 d^2$ is hump-shaped: it
increases at low doses and decreases at high doses. The challenge is to
recover this shape from confounded observational data.

## Naive regression vs adjusted regression

A linear regression of $Y$ on $D$ alone is doubly biased: confounding
through $X$ plus misspecification of the nonlinearity:

```{r}
#| label: naive
#| cache: true

# Linear fit, ignoring covariates
linear_naive <- lm(Y ~ D, data = df)
# Linear fit with covariate adjustment
linear_adj   <- lm(Y ~ D + X1 + X2, data = df)
# Nonlinear fit with covariate adjustment
poly_adj     <- lm(Y ~ poly(D, 2) + X1 + X2, data = df)

ds <- seq(0, 10, by = 0.5)

predict_mu <- function(fit, ds) {
  data.frame(d = ds, X1 = 0, X2 = 0) |>
    rename(D = d) -> nd
  predict(fit, newdata = nd)
}

# True dose-response
df_true <- tibble(d = ds, mu = true_mu(ds))

df_fits <- bind_rows(
  tibble(d = ds, mu = predict_mu(linear_naive, ds), method = "Linear (naive)"),
  tibble(d = ds, mu = predict_mu(linear_adj,   ds), method = "Linear (adj X)"),
  tibble(d = ds, mu = predict_mu(poly_adj,     ds), method = "Quadratic (adj X)")
)
```

```{r}
#| label: naive-plot
#| fig-cap: "Fitted vs true dose-response. Linear fits miss the curvature; the quadratic-adjusted fit comes closer."
#| fig-width: 7
#| fig-height: 4

ggplot() +
  geom_line(data = df_fits, aes(d, mu, colour = method), linewidth = 1.0) +
  geom_line(data = df_true, aes(d, mu), colour = "black",
            linetype = "dashed", linewidth = 1.0) +
  labs(x = "Dose D", y = "Expected outcome E[Y(d)]",
       colour = "Method",
       caption = "Black dashed line = true μ(d) = 1 + 2d - 0.15 d²") +
  theme_minimal()
```

Even the quadratic adjustment leaves bias if the true dose-response is
not quadratic or if the covariate adjustment is misspecified.

## Generalized propensity score (GPS)

The **generalized propensity score** is the conditional density of the
treatment given covariates:

$$
r(d, x) = f_{D \mid X}(d \mid x).
$$

Under unconfoundedness, conditioning on $r(D, X)$ alone is sufficient to
identify the dose-response curve (Hirano-Imbens 2004). In practice the
GPS is estimated assuming, say, $D \mid X \sim \mathcal{N}(\beta_0 + \beta'X, \sigma^2)$
or by quantile binning.

```{r}
#| label: gps
#| cache: true

# Fit GPS model: D ~ X with Normal residuals
gps_model <- lm(D ~ X1 + X2, data = df)
d_pred    <- predict(gps_model)
sigma_e   <- sigma(gps_model)

# GPS for each unit's observed treatment
gps <- dnorm(df$D, mean = d_pred, sd = sigma_e)
df$gps <- gps

# IPW-style weights for continuous treatment: 1 / GPS
# Stabilise by multiplying by the marginal density f_D(D)
marginal_D <- density(df$D)
fD <- approx(marginal_D$x, marginal_D$y, xout = df$D)$y
df$sw_cont <- pmin(fD / gps, quantile(fD / gps, 0.99))

cat(sprintf("Weight summary: min=%.3f, mean=%.3f, max=%.3f\n",
            min(df$sw_cont), mean(df$sw_cont), max(df$sw_cont)))
```

The Hirano-Imbens estimator regresses $Y$ on a flexible function of $(D,
\hat r(D, X))$ and then averages across the GPS distribution at each
fixed $D$:

```{r}
#| label: hi-est
#| cache: true

# Stage 2: regress Y on flexible function of (D, GPS)
hi_fit <- lm(Y ~ poly(D, 2) + poly(gps, 2) + I(D * gps), data = df)

# Dose-response at each d: average over the empirical distribution of GPS
# at that d.
estimate_dose <- function(d) {
  # Predict GPS at this d for each unit
  gps_at_d <- dnorm(d, mean = d_pred, sd = sigma_e)
  nd <- data.frame(D = d, gps = gps_at_d)
  mean(predict(hi_fit, newdata = nd))
}

ds <- seq(0, 10, by = 0.5)
hi_dose <- sapply(ds, estimate_dose)

df_hi <- tibble(d = ds, mu = hi_dose, method = "Hirano-Imbens GPS")

ggplot() +
  geom_line(data = df_hi, aes(d, mu, colour = method), linewidth = 1.2) +
  geom_line(data = df_true, aes(d, mu), colour = "black",
            linetype = "dashed", linewidth = 1.0) +
  labs(x = "Dose D", y = "E[Y(d)]",
       caption = "Dashed black = truth; coloured = Hirano-Imbens GPS estimator") +
  theme_minimal()
```

The Hirano-Imbens estimator recovers the curvature missing from naive
regression because it does not impose a parametric form on the
dose-response curve.

## Doubly-robust dose-response (Kennedy 2017)

Kennedy (2017) extended AIPW to continuous treatments. The estimator
combines:

- An outcome model $\hat\mu(d, x) = \mathbb{E}[Y \mid D = d, X = x]$
- A generalised propensity score $\hat r(d \mid x)$

Then the dose-response at $d$ is estimated by

$$
\hat\mu_{DR}(d) = \frac{1}{n} \sum_i \left[ \hat\mu(d, X_i) + K_h(D_i - d) \cdot \frac{Y_i - \hat\mu(D_i, X_i)}{\hat r(D_i \mid X_i)} \right],
$$

where $K_h$ is a kernel that smooths over a bandwidth $h$ near $d$. The
result is doubly robust and asymptotically normal.

The implementation is a few lines using a kernel for smoothing near the
target dose $d$ and the GPS already estimated above:

```{r}
#| label: drf
#| cache: true

# Outcome model μ̂(D, X): flexible polynomial
mu_fit <- lm(Y ~ D + I(D^2) + X1 + X2 + I(D * X1) + I(D * X2), data = df)
df$mu_at_obs <- predict(mu_fit)

# DR estimator at dose d with bandwidth h
dr_estimate <- function(d, h = 0.5) {
  # Counterfactual prediction at D = d for each unit
  mu_at_d <- predict(mu_fit, newdata = transform(df, D = d))
  # Kernel weight on observed treatment near d
  kernel <- dnorm(df$D - d, sd = h)
  weight <- kernel / df$gps
  weight <- weight / mean(weight)  # normalise
  # DR correction
  correction <- weight * (df$Y - df$mu_at_obs)
  mean(mu_at_d) + mean(correction)
}

ds <- seq(0, 10, by = 0.5)
dr_dose <- sapply(ds, dr_estimate)

df_drf <- tibble(d = ds, mu = dr_dose, method = "DR (Kennedy 2017)")

ggplot() +
  geom_line(data = df_hi,  aes(d, mu, colour = method), linewidth = 1.0) +
  geom_line(data = df_drf, aes(d, mu, colour = method), linewidth = 1.0) +
  geom_line(data = df_true, aes(d, mu), colour = "black",
            linetype = "dashed", linewidth = 1.0) +
  labs(x = "Dose D", y = "E[Y(d)]",
       caption = "Black dashed = truth") +
  theme_minimal()
```

The DR estimator is consistent if *either* the outcome model
$\hat\mu(d, x)$ or the GPS $\hat r(d \mid x)$ is correctly specified, and
efficient when both are. The kernel bandwidth $h$ controls the
bias-variance trade-off near each evaluation dose.

The `causaldrf` package provides several related estimators (`hi_est`,
`iptw_est`, `aipwee_est`, `gam_est`, `nw_est`, `bart_est`) with
pre-packaged inference and standard errors — useful for production work
once the manual version above clarifies the underlying logic.

## Longitudinal Modified Treatment Policies (LMTP)

For continuous, multivalued, and time-varying treatments — and especially
when **positivity** is a concern (e.g. the dose-response at $d = 10$
might have zero observations to learn from) — the LMTP framework
(Diaz-Hejazi 2020) is the modern workhorse.

LMTP replaces hypothetical interventions like "everyone gets $D = 4$"
with **modified treatment policies** like "shift everyone's $D$ down by
1 unit." The shifted-treatment distribution stays close to the observed
data, preserving positivity.

### A two-period example

```{r}
#| label: lmtp-sim
#| cache: true

set.seed(1)
N    <- 1500
L_1  <- runif(N, 0, 1)
A_1  <- rbinom(N, 1, 0.5)
L_2  <- rnorm(N, 0.25 * L_1, 0.5)
A_2  <- rbinom(N, 1, plogis(0.5 - 0.2 * A_1 + 0.1 * L_2))
Y    <- rnorm(N, A_2 + L_2, 0.3)

lmtp_data <- data.frame(L_1, A_1, L_2, A_2, Y)
head(lmtp_data)
```

A modified treatment policy might shift treatments downward. Define the
shift function:

```{r}
#| label: shift-fn

shift_zero <- function(data, trt) {
  # Set treatment to 0 everywhere — for comparison to "never treat"
  a <- data[[trt]]
  rep(0, length(a))
}
```

Then fit the LMTP estimator with `lmtp::lmtp_tmle`:

```{r}
#| label: lmtp-fit
#| cache: true
#| warning: false

# TMLE-based LMTP estimator with super-learners for nuisance functions
# (using SL.glm to keep it fast; in practice use richer SL libraries)
lmtp_fit <- lmtp_tmle(
  data       = lmtp_data,
  trt        = c("A_1", "A_2"),
  outcome    = "Y",
  baseline   = NULL,
  time_vary  = list(c("L_1"), c("L_2")),
  shift      = shift_zero,
  outcome_type = "continuous",
  learners_outcome = c("SL.glm"),
  learners_trt     = c("SL.glm"),
  folds = 2
)

lmtp_fit
```

The `lmtp` package handles the bookkeeping for time-varying covariates
and treatments transparently. Its output includes the TMLE point
estimate, IPW estimate, and bootstrap-style confidence intervals.

### Contrasts between policies

LMTP shines when comparing two policies (e.g. "set to 0" vs "set to 1"):

```{r}
#| label: lmtp-contrast
#| cache: true
#| warning: false

shift_one <- function(data, trt) {
  a <- data[[trt]]
  rep(1, length(a))
}

lmtp_one <- lmtp_tmle(
  data       = lmtp_data,
  trt        = c("A_1", "A_2"),
  outcome    = "Y",
  baseline   = NULL,
  time_vary  = list(c("L_1"), c("L_2")),
  shift      = shift_one,
  outcome_type = "continuous",
  learners_outcome = c("SL.glm"),
  learners_trt     = c("SL.glm"),
  folds = 2
)

contrast <- lmtp_contrast(lmtp_one, ref = lmtp_fit)
contrast
```

The contrast estimate corresponds to the ATE under "always treat" vs
"never treat" — but computed within the LMTP framework, with valid
inference and explicit positivity handling.

## Choosing among the three

| Method | Strengths | Weaknesses |
|---|---|---|
| **GPS (Hirano-Imbens)** | Conceptually transparent; widely understood | Parametric assumptions on GPS; sensitive to bandwidth |
| **Doubly-robust** (Kennedy 2017) | Robust to misspecification; asymptotically efficient | More complex; nuisance models need careful tuning |
| **LMTP** | Handles continuous + multivalued + time-varying uniformly; respects positivity | More setup; requires defining a shift function |

In practice:

- For **simple cross-sectional continuous treatment** with good overlap,
  use Hirano-Imbens GPS or the Kennedy DR estimator.
- For **time-varying or modified-policy** questions, use LMTP.
- For **multivalued** treatments, generalize the propensity score to a
  multinomial model and use the same DR machinery.

## When to use these methods

These methods are needed whenever:

1. The treatment variable takes on more than two values (continuous,
   ordinal, or categorical with > 2 categories).
2. The research question concerns the *shape* of the dose-response
   curve, not just a single contrast.
3. The hypothetical intervention is best expressed as a *shift* in the
   distribution of treatment rather than fixed assignment.

For strictly binary, single-time-period problems, the standard
[Estimation](estimation.qmd) and [Heterogeneous Effects](heterogeneous-effects.qmd)
methods are sufficient.

## Summary

- **Continuous and multivalued treatments** require estimating an entire
  dose-response curve $\mu(d) = \mathbb{E}[Y(d)]$, not a single scalar.
- **Generalized propensity score (GPS)** extends propensity-score methods
  to continuous $D$; Hirano-Imbens (2004) is the canonical reference.
- **Doubly-robust dose-response** (Kennedy 2017) combines GPS with an
  outcome model; `causaldrf::bart_est` provides a BART-based
  implementation with native uncertainty quantification.
- **LMTP** (Diaz-Hejazi 2020, `lmtp` package) is the modern framework for
  continuous, multivalued, and time-varying treatments under modified
  policy interventions; respects positivity by construction.
- See the [companion blog chapter on LMTP](https://xiangao.github.io/blog_book/lmtp.html)
  for an extended treatment with worked time-varying examples.

## References

- Hirano, K., & Imbens, G. W. (2004). The propensity score with continuous treatments. *Applied Bayesian Modeling and Causal Inference from Incomplete-Data Perspectives*, 226164, 73–84.
- Kennedy, E. H., Ma, Z., McHugh, M. D., & Small, D. S. (2017). Non-parametric methods for doubly robust estimation of continuous treatment effects. *JRSS-B*, 79(4), 1229–1245.
- Díaz, I., Williams, N., Hoffman, K. L., & Schenck, E. J. (2023). Nonparametric causal effects based on longitudinal modified treatment policies. *JASA*, 118(542), 846–857.
- Imbens, G. W. (2000). The role of the propensity score in estimating dose-response functions. *Biometrika*, 87(3), 706–710.
