# Estimation: RA, IPW, AIPW and IPWRA

```{r}
#| include: false
suppressPackageStartupMessages({
  library(tidyverse)
  library(haven)
  library(hdm)
  library(lmtest)
  library(sandwich)
  library(broom)
  library(knitr)
})
```

Once the causal effect is identified, we need an estimator. This chapter covers four: regression adjustment (RA), inverse probability weighting (IPW), augmented IPW (AIPW), and inverse-probability-weighted regression adjustment (IPWRA).

## Experimental Data

With experimental data the assumptions hold by design. A difference-in-means estimator suffices (under randomization ATE = ATT = ATU): $$ \hat{\tau} = \bar Y_1 - \bar Y_0 $$ {#eq-estimation-1}

A $t$-test or a regression of $Y$ on $W$ and $X$ both recover the same estimate.

## Adjustment Formula

The adjustment formula connects causal and statistical quantities: $$
\begin{aligned}
E[Y(1)-Y(0)] &= E[Y(1)] - E[Y(0)] \\
&=E_X[E[Y(1)|X]] - E_X[E[Y(0)|X]] \\
&=E_X[E[Y(1)|W=1, X]] - E_X[E[Y(0)|W=0, X]] \\
&=E_X[E[Y|W=1, X]] - E_X[E[Y|W=0, X]]
\end{aligned}
$$ {#eq-estimation-2}  The steps are: linearity of expectation, iterated expectations, unconfoundedness plus overlap, and consistency.

## Regression Adjustment

We estimate $E[Y \mid W=w, X=x]$ by regression.

We define a conditional mean function:

  $$
\begin{aligned}
\mu_0(X) &\equiv E[Y|W=0, X] \\
&= E[Y(0)|W=0, X]  \\
&= E[Y(0)|W=1, X]
\end{aligned}
$$ {#eq-estimation-3}  

The RA estimators are:

$$
\hat \tau_{ATT}
= \frac{1}{n_1}\sum_{i:W_i=1}\{Y_i-\hat\mu_0(X_i)\}
$$ {#eq-estimation-4}

$$
\hat \tau_{ATE}
= \frac{1}{n}\sum_{i=1}^n\{\hat\mu_1(X_i)-\hat\mu_0(X_i)\}
$$ {#eq-estimation-5}

where $\hat \mu_0(X_i)$ is the predicted value of $Y_i$ from a regression of $Y$ on $X$ for the control group.

Target:   $$
\begin{aligned}
\tau_{ATT} &= E[Y|W=1] - E[Y(0)|W=1] \\
&= E[Y|W=1] - E[\mu_0(X) | W=1]  \\
& = E[Y|W=1] - (\alpha_0 + E[X|W=1] \beta_0)
\end{aligned}
$$ {#eq-estimation-6}


The last term is a linear form of $\mu_0(X)$. We can specify other forms, but the idea is to model it with some functional form. For parametric forms, we need to make sure extrapolation does not go out of control.

## Linear RA

Regression Adjustment is basically an imputation estimator. While we observe $E[Y|W=1,X]$, we model $E[Y(0)|W=1]$, based on unconfoundedness and a functional form (say linear form). We estimate $\beta_0$ on the control sample, then get the expected values for the treated sample, for each value of $X$.

Implementation of linear RA is easy. We regress $Y$ on $X$ and $W$ and their interaction. It's shown that we need to de-mean $X$ to get the effect correct.

The data are the 401(k) file distributed with `hdm`, from the 1991 Survey of
Income and Program Participation. There are 9,915 households. The outcome
`net_tfa` is net total financial assets in dollars, the treatment `p401` is
participation in a 401(k) plan, taken by 2,594 households, and we condition on
household income `inc` and the reference person's `age`. Whether those two
covariates suffice for unconfoundedness is doubtful, and the point here is the
mechanics of the estimator rather than the substantive effect.

We de-mean income and age at their treated-group means, then regress net
financial assets on participation, the two de-meaned covariates, and their
interactions with participation.

```{r ra1, warning=FALSE, cache=TRUE, message=FALSE, echo=TRUE}
data(pension)
#  for ATE, de-mean X by deducting the mean of X in the whole sample:
#  data <- pension %>% mutate(inc_dm=inc-mean(inc), age_dm=age-mean(age))
#  for ATT (used here), de-mean X by deducting the mean of X in the treated group:
data <- pension %>%
  # Reference the local (in-mutate) inc/age rather than pension$inc/pension$age
  # directly: hardcoding the parent object name here would silently keep using
  # the ORIGINAL full-sample means even if this code were reused on a
  # resampled/bootstrapped copy of the data assigned to `data`, producing
  # wrong point estimates and standard errors for that use case.
  mutate(inc_dm=inc-mean(inc[p401==1]), age_dm=age-mean(age[p401==1]))
lm_ra <- lm(net_tfa ~ p401*(inc_dm + age_dm),data=data)
summary(lm_ra)
#coeftest(lm_ra, vcov=vcovHC(lm_ra, type='HC2'))[2,]
```

The coefficient on `p401` is 14,160 with a standard error of 1,397. Because we
de-meaned at the treated-group means, that coefficient is the ATT: participants
hold about \$14,000 more in net financial assets than they would have without a
401(k), under the assumption that income and age are the only confounders.
De-meaning at the whole-sample means instead would give the ATE.

Both interaction terms are large and significant, so the effect varies with the
covariates. A household \$10,000 above the treated-group mean income has an
effect larger by \$3,300, and each extra year of age adds \$815. This is why
the interactions are needed at all.

### Questions on RA

1.  Why do we have to do interaction of $W$ and $X$? What if we don't do it?
2.  What if we don't de-mean $X$?

Answers:

1.  It has been shown [@sloczynski-2022] that with heterogeneous treatment effect, and unequal divide between treated and control, we need to do the interaction term. Otherwise, we'll get biased estimates, for $\tau_{ATT}$ or $\tau_{ATE}$. The intuition is that we need to model the difference between treated and control, and the difference is not constant across $X$.
2.  If we don't de-mean $X$, then the coefficient on $W$ is not the average treatment effect. We can still retrieve the ATT or ATE by calculating the marginal effect of $W$ though.

## IPW

Under unconfoundedness and overlap,

 $$
\begin{aligned}
E[Y(w)] &= E[E[Y|W=w,X]] \\
&= \sum_x (\sum_y y P(y|w,x)) P(x) \\
&= \sum_x \sum_y y P(y|w,x) P(x) {\frac{P(w|x)}{P(w|x)}} \\
&= \sum_x \sum_y y P(y,w,x) {\frac{1}{P(w|x)}} \\
&= \sum_x E[\mathbb{1}(W=w,X=x)Y] {\frac{1}{P(w|x)}} \\
&= E[\frac{\mathbb{1}(W=w)Y} {P(w|X)}]
\end{aligned}
$$ {#eq-estimation-7}


The IPW equation means that the expected value of the potential outcome is the weighted average of the observed outcome, where the weight is the inverse of the propensity score $P(w|x)$. $\mathbb{1}(W=w)$ is an indicator function, which is 1 if $W=w$ and 0 otherwise.

 $$
\begin{aligned}
\tau_{ATE} &= E[Y(1)] - E[Y(0)] \\
&= E[\frac{W Y} {\pi(X)}] - E[\frac{(1-W) Y} {1-\pi(X)}] \\
\end{aligned}
$$ {#eq-estimation-8}


The sample estimators are:

 $$
\begin{aligned}
\hat \tau_{ATE, IPW} &= \frac{1}{N} \sum_i \frac{W_i Y_i} {\hat\pi(X_i)} - \frac{1}{N} \sum_i \frac{(1-W_i) Y_i} {1-\hat \pi(X_i)}
\end{aligned}
$$ {#eq-estimation-9}


 $$
\begin{aligned}
\hat \tau_{ATT, IPW} &= \bar Y_1 - \frac{1}{N_1} \sum_i \frac{\hat\pi(X_i)}{1-\hat \pi(X_i)} (1-W_i) Y_i
\end{aligned}
$$ {#eq-estimation-10}


For the ATT, control units are weighted by the odds $\hat\pi(X_i)/(1-\hat\pi(X_i))$
rather than $1/(1-\hat\pi(X_i))$: this reweights the controls to the covariate
distribution of the treated (one can check
${\rm E}[\frac{\pi(X)}{1-\pi(X)}(1-W)Y] = P(W=1)\, {\rm E}[Y(0)|W=1]$).

### IPW Intuition

IPW removes confounding $X$ by creating a pseudo-population in which $X$ is independent of $W$. The intuition is that we can create a pseudo-population in which $X$ is independent of $W$, by weighting the observations by the inverse of the propensity score. Once $X$ is independent of $W$, $X$ is not a confounder anymore. We can then estimate the effect of $W$ on $Y$ by comparing the weighted average of $Y$ for $W=1$ and $W=0$.

Suppose there are two types of people, one type has high probability to be treated, the other one has low probability to be treated. Without adjusting for the propensity score, the treated group will be dominated by the first type, and the control group will be dominated by the second type. The difference between the treated and control group will be due to the difference in the composition of the two groups, not due to the treatment effect. By weighting the observations by the inverse of the propensity score, we can create a pseudo-population in which the two groups have the same composition. The difference between the treated and control group will be due to the treatment effect.

## Doubly Robust Estimators

We can model both outcome and treatment assignment, and use both models to estimate the treatment effect. The estimator is called doubly robust because it is consistent if *either* the outcome regression *or* the propensity/treatment model is correctly specified -- we only need one of the two to be right, not both. (Note the precise statement: it is *not* consistent when both are misspecified, so "robust to either being misspecified" is the wrong reading.)

### AIPW

$$ \psi_1^{AIPW} = E[\frac{W(Y-\mu_1(X))}{\pi(X)} + \mu_1(X)] $$ {#eq-estimation-11}

$$ \psi_0^{AIPW} = E[\frac{(1-W)(Y-\mu_0(X))}{1-\pi(X)} + \mu_0(X)] $$ {#eq-estimation-12}

Here $\mu_w(X)=E[Y|W=w,X]$ is the outcome-model nuisance function, while the
scalar $\psi_w^{AIPW}$ equals $E[Y(w)]$ as long as either the outcome model
or the propensity score is correctly specified; the ATE is
$\psi_1^{AIPW}-\psi_0^{AIPW}$.

We switch to the Cattaneo (2010) birth-weight data used in the Stata manuals:
4,642 births, outcome `bweight` in grams, treatment `mbsmoke` for maternal
smoking during pregnancy (864 mothers), and covariates `prenatal1` (a prenatal
visit in the first trimester), `mmarried`, `mage` (mother's age), `fbaby`
(first baby) and `medu` (mother's education). Smokers' babies weigh 3,137.7 g
on average and non-smokers' 3,412.9 g, a raw gap of $-275.3$ g.

The outcome model is a linear regression of birth weight on smoking interacted
with prenatal care, marital status, mother's age and first-baby status. The
propensity model is a logit of smoking on marital status, mother's age,
first-baby status and education. We form the AIPW summand for each arm and
average the difference.

```{r aipw1, warning=FALSE, cache=TRUE, message=FALSE, echo=TRUE}
data <- read_dta(file="https://www.stata-press.com/data/r18/cattaneo2.dta")
mu <- lm(bweight ~ mbsmoke*(prenatal1 + mmarried + mage + fbaby), data=data)
pi <- glm( mbsmoke ~ mmarried + mage + fbaby + medu, data=data, family=binomial(link="logit"))
mm1 <- predict(mu, newdata=data %>% mutate(mbsmoke=1))
mm0 <- predict(mu, newdata=data %>% mutate(mbsmoke=0))
pi1 <- predict(pi, newdata=data, type="response")
data <- data %>%
  mutate(w=mbsmoke, y=bweight, m1=mm1, m0=mm0, pi1=pi1)
data2 <- data %>%
  mutate(mu1=(w*(y-m1)/pi1 + m1), mu0=((1-w)*(y-m0)/(1-pi1) + m0)) %>%
  mutate(tau=mu1-mu0)
data2 %>% summarise(mean(tau))
```

AIPW gives $-234$ g. Adjustment moves the estimate about 41 g toward zero from
the raw $-275$ g, so part of the raw gap is composition: smokers differ from
non-smokers in age, education and marital status.

### IPWRA

The idea of IPWRA is to combine RA and IPW. Basically RA with IPW weights.

Implementation:

-   Estimate propensity score. Get predicted probability of treated for each observation.
-   Estimate outcome model with the IPW weights: two separate equations, one for treated (weights $1/\hat \pi(X)$) and one for control (weights $1/(1-\hat \pi(X))$). Equivalently, one weighted regression with full treatment–covariate interaction — the interacted design matrix is block-diagonal across the two groups, so the weighted normal equations decouple and the two routes give identical fits.
-   Get predicted values for setting everyone treated. Get predicted values for setting everyone control. These are the two potential outcomes.
-   Take difference. The average is the ATE.

We run this on the same birth-weight data and the same propensity model as
AIPW, so the two estimates differ only in how the outcome and treatment models
are combined.

```{r ipwra1, warning=FALSE, cache=TRUE, message=FALSE, echo=TRUE}
data <- read_dta(file="https://www.stata-press.com/data/r18/cattaneo2.dta")
pi <- glm( mbsmoke ~ mmarried + mage + fbaby + medu, data=data, family=binomial(link="logit"))
pi1 <- predict(pi, newdata=data, type="response")
data <- data %>%
  mutate(pi1=pi1)
data1 <- data %>% filter(mbsmoke==1)
data0 <- data %>% filter(mbsmoke==0)
mu1 <- lm(bweight ~ prenatal1 + mmarried + mage + fbaby, data=data1, weights=1/data1$pi1)
mu0 <- lm(bweight ~ prenatal1 + mmarried + mage + fbaby, data=data0, weights=1/(1-data0$pi1))
mm1 <- predict(mu1, newdata=data %>% mutate(mbsmoke=1))
mm0 <- predict(mu0, newdata=data %>% mutate(mbsmoke=0))
data <- data %>%
  mutate(w=mbsmoke, y=bweight, m1=mm1, m0=mm0, pi1=pi1)
data2 <- data %>%
  mutate(tau=m1-m0)
data2 %>% summarise(mean(tau))
```

IPWRA gives $-233$ g, within a gram of the AIPW estimate. The agreement is not
automatic. Both are doubly robust and both use the same two nuisance models
here, so they can only differ in how the models are combined. When the outcome
and propensity models point in different directions, the two estimators
separate.
