# Estimation: RA, IPW, AIPW

```{r}
#| include: false
library(npcausal)
library(boot)
library(SuperLearner)
library(tmle)
library(tidyverse)
library(ranger)
library(hdm)
library(lmtest)
library(sandwich)
library(haven)
library(grf)
library(rpart)
library(glmnet)
library(splines)
library(ggplot2)
library(reshape2)
library(stringr)
library(causaldata)
library(fixest)
```

## Experimental Data

If we have experimental data, then we have the assumptions satisfied. We can directly compare the treated and control groups. We can use a difference-in-means estimator. $$ \hat{\tau_{ATT}} = \bar Y_1 - \bar Y_0 $$

In experimental setting, we can do a t test for the difference in means. We can also do a regression of $Y$ on $W$ and $X$.

## Adjustment Formula

From causal to estimation: $$ \small \begin{align}
E[Y(1)-Y(0)] &= E[Y(1)] - E[Y(0)] \\
&=E_X[E[Y(1)|X]] - E_X[E[Y(0)|X]] \\
&=E_X[E[Y(1)|W=1, X]] - E_X[E[Y(0)|W=0, X]] \\
&=E_X[E[Y|W=1, X]] - E_X[E[Y|W=0, X]]
\end{align} $$ where the first step is linear expectation. Second step is law of iterated expectation. Third step is unconfoundedness and overlap. Fourth step is consistency.

## Regression Adjustment

How do we get $E[Y|W=w, X=x]$? We can use regression adjustment.

We define a conditional mean function:

$$  \begin{align}
\mu_0(X) &\equiv E[Y|W=0, X] \\
&= E[Y(0)|W=0, X]  \\
&= E[Y(0)|W=1, X]
\end{align}  $$

The RA estimators are:

$$ \hat \tau_{ATT} = \sum_{i=1}^n (W_i(Y_i-\hat \mu_0(X_i)) /n_1$$ $$ \hat \tau_{ATE} = \sum_{i=1}^n (W_i(Y_i-\hat \mu_0(X_i)) + (1-W_i)(\hat \mu_1(X_i) - Y_i)) /n$$

where $\hat \mu_0(X_i)$ is the predicted value of $Y_i$ from a regression of $Y$ on $X$ for the control group.

Target: $$  \begin{align}
\tau_{ATT} &= E[Y|W=1] - E[Y(0)|W=1] \\
&= E[Y|W=1, X] - \mu_0(X)  \\
& = E[Y|W=1, X] - (\alpha_0 + E(X) \beta_0)
\end{align}
$$

The last term is a linear form of $\mu_0(X)$. We can specify other forms, but the idea is to model it with some functional form. For parametric forms, we need to make sure extrapolation does not go out of control.

## Linear RA

Regression Adjustment is basically an imputation estimator. While we observe $E[Y|W=1,X]$, we model $E[Y(0)|W=1]$, based on unconfoundedness and a functional form (say linear form). We estimate $\beta_0$ on the control sample, then get the expected values for the treated sample, for each value of $X$.

Implementation of linear RA is easy. We regress $Y$ on $X$ and $W$ and their interaction. It's shown that we need to de-mean $X$ to get the effect correct.

```{r ra1, warning=FALSE, cache=TRUE, message=FALSE, echo=TRUE}
data(pension)
data <- pension %>%
  mutate(inc_dm=inc-mean(inc), age_dm=age-mean(age))

data <- pension %>%
  mutate(inc_dm=inc-mean(pension$inc[pension$p401==1]), age_dm=age- mean(pension$age[pension$p401==1]))
#  for ATT, we de-mean X by deducting the mean of X in the treated group
#  for ATE, we de-mean X by deducting the mean of X in the whole sample
lm_ra <- lm(net_tfa ~ p401*(inc_dm + age_dm),data=data)
summary(lm_ra)
#coeftest(lm_ra, vcov=vcovHC(lm_ra, type='HC2'))[2,]
```

### Questions on RA

1.  Why do we have to do interaction of $W$ and $X$? What if we don't do it?
2.  What if we don't de-mean $X$?

Answers:

1.  It has been shown (Słoczyński (2022, REStat)) that with heterogeneous treatment effect, and unequal divide between treated and control, we need to do the interaction term. Otherwise, we'll get biased estimates, for $\tau_{ATT}$ or $\tau_{ATE}$. The intuition is that we need to model the difference between treated and control, and the difference is not constant across $X$.
2.  If we don't de-mean $X$, then the coefficient on $W$ is not the average treatment effect. We can still retrieve the ATT or ATE by calculating the marginal effect of $W$ though.

## IPW

Under unconfoundedness and overlap,

$$ \small \begin{align}
E[Y(w)] &= E[E[y|W=w,X]] \\
&= \sum_x (\sum_y P(y|w,x)) P(x) \\
&= \sum_x \sum_y P(y|w,x)) P(x) {\frac{P(w|x)}{P(w|x)}} \\
&= \sum_x \sum_y y P(y,w,x) {\frac{1}{P(w|x)}} \\
&= \sum_x E[\mathbb{1}(W=w,X=x)Y] {\frac{1}{P(w|x)}} \\
&= E[\frac{\mathbb{1}(W=w)Y} {P(w|X)}]
\end{align}
$$

The IPW equation means that the expected value of the potential outcome is the weighted average of the observed outcome, where the weight is the inverse of the propensity score $P(w|x)$. $\mathbb{1}(W=w)$ is an indicator function, which is 1 if $W=w$ and 0 otherwise.

$$ \begin{align}
\tau_{ATE} &= E[Y(1)] - E[Y(0)] \\
&= E[\frac{W Y} {\pi_1(X)}] - E[\frac{(1-W) Y} {1-\pi_1(X)}] \\
\end{align}
$$

The sample estimators are:

$$ \begin{align}
\hat \tau_{ATE, IPW} &= \frac{1}{N} \sum_i \frac{W_i Y_i} {\hat\pi(X_i)} - \frac{1}{N} \sum_i \frac{(1-W_i) Y_i} {1-\hat \pi(X_i)}
\end{align}
$$

$$ \begin{align}
\hat \tau_{ATT, IPW} &= \bar Y_1 - \frac{1}{N_1} \sum_i \frac{(1-W_i) Y_i} {1-\hat \pi(X_i)}
\end{align}
$$

### IPW Intuition

IPW removes confounding $X$ by creating a pseudo-population in which $X$ is independent of $W$. The intuition is that we can create a pseudo-population in which $X$ is independent of $W$, by weighting the observations by the inverse of the propensity score. Once $X$ is independent of $W$, $X$ is not a confounder anymore. We can then estimate the effect of $W$ on $Y$ by comparing the weighted average of $Y$ for $W=1$ and $W=0$.

Suppose there are two types of people, one type has high probability to be treated, the other one has low probability to be treated. Without adjusting for the propensity score, the treated group will be dominated by the first type, and the control group will be dominated by the second type. The difference between the treated and control group will be due to the difference in the composition of the two groups, not due to the treatment effect. By weighting the observations by the inverse of the propensity score, we can create a pseudo-population in which the two groups have the same composition. The difference between the treated and control group will be due to the treatment effect.

## Doubly Robust Estimators

We can model both outcome and treatment assignment, and use both models to estimate the treatment effect. The intuition is that if either model is correct, the estimator will be consistent. The estimator is called doubly robust because it is robust to either model being misspecified.

### AIPW

$$ \mu_1^{AIPW} = E[\frac{W(Y-\mu_1(X))}{\pi(X)} + \mu_1(X)] $$

$$ \mu_0^{AIPW} = E[\frac{(1-W)(Y-\mu_0(X))}{1-\pi(X)} + \mu_0(X)] $$

```{r aipw1, warning=FALSE, cache=TRUE, message=FALSE, echo=TRUE}
data <- read_dta(file="https://www.stata-press.com/data/r18/cattaneo2.dta")
mu <- lm(bweight ~ mbsmoke*(prenatal1 + mmarried + mage + fbaby), data=data)
pi <- glm( mbsmoke ~ mmarried + mage + fbaby + medu, data=data, family=binomial(link="logit"))
mm1 <- predict(mu, newdata=data %>% mutate(mbsmoke=1))
mm0 <- predict(mu, newdata=data %>% mutate(mbsmoke=0))
pi1 <- predict(pi, newdata=data, type="response")
data <- data %>%
  mutate(w=mbsmoke, y=bweight, m1=mm1, m0=mm0, pi1=pi1)
data2 <- data %>%
  mutate(mu1=(w*(y-m1)/pi1 + m1), mu0=((1-w)*(y-m0)/(1-pi1) + m0)) %>%
  mutate(tau=mu1-mu0)
data2 %>% summarise(mean(tau))
```

### IPWRA

The idea of IPWRA is to combine RA and IPW. Basically RA with IPW weights.

Implementation:

-   Estimate propensity score. Get predicted probability of treated for each observation.
-   Estimate outcome model. But cannot use fully interaction of treatment and covariates. Instead, estimate two equations, one for treated and one for control. Use $1/\hat \pi(X)$ as weights for treated group and $1/(1-\hat \pi(X))$ as weights for control group.
-   Get predicted values for setting everyone treated. Get predicted values for setting everyone control. These are the two potential outcomes.
-   Take difference. The average is the ATE.

```{r ipwra1, warning=FALSE, cache=TRUE, message=FALSE, echo=TRUE}
data <- read_dta(file="https://www.stata-press.com/data/r18/cattaneo2.dta")
pi <- glm( mbsmoke ~ mmarried + mage + fbaby + medu, data=data, family=binomial(link="logit"))
pi1 <- predict(pi, newdata=data, type="response")
data <- data %>%
  mutate(pi1=pi1)
data1 <- data %>% filter(mbsmoke==1)
data0 <- data %>% filter(mbsmoke==0)
mu1 <- lm(bweight ~ prenatal1 + mmarried + mage + fbaby, data=data1, weights=1/data1$pi1)
mu0 <- lm(bweight ~ prenatal1 + mmarried + mage + fbaby, data=data0, weights=1/(1-data0$pi1))
mm1 <- predict(mu1, newdata=data %>% mutate(mbsmoke=1))
mm0 <- predict(mu0, newdata=data %>% mutate(mbsmoke=0))
data <- data %>%
  mutate(w=mbsmoke, y=bweight, m1=mm1, m0=mm0, pi1=pi1)
data2 <- data %>%
  mutate(tau=m1-m0)
data2 %>% summarise(mean(tau))
```
