# G-Methods for Time-Varying Treatments

```{r}
#| include: false
library(tidyverse)
library(ltmle)
library(ipw)
library(survey)
library(broom)
```

```{r}
#| eval: false
library(tidyverse)
library(ltmle)
library(ipw)
library(survey)
library(broom)
```

All the methods covered so far have one feature in common: a single
treatment decision at a single point in time. Real policy questions are
often more complex. A patient receives a medication at multiple visits.
A worker enrols in training programmes over several years. A student
takes one course after another. The treatment evolves; covariates that
respond to past treatment then influence future treatment. **Standard
regression adjustment fails in this setting** — and the failure is not
subtle.

This chapter covers the **g-methods** developed by Robins and collaborators
to handle time-varying treatments and time-varying confounders:

1. **The g-formula** (parametric g-computation) — simulate the outcome
   under a hypothetical treatment regime by recursively integrating over
   the time-varying covariates.
2. **Inverse probability of treatment weighting (IPTW)** for time-varying
   treatments — reweight to a population where treatment is unrelated to
   past covariates.
3. **Marginal structural models (MSMs)** — regression-style models for the
   IPTW-weighted population.
4. **Longitudinal TMLE** — efficient doubly-robust estimation combining
   g-computation and IPTW.

## The time-varying confounding problem

A two-period example makes the failure of standard regression vivid.
Consider:

- $L_0$: baseline covariate (e.g. health status at study entry)
- $A_0$: treatment at time 0 (e.g. medication)
- $L_1$: covariate at time 1, influenced by $A_0$ (e.g. health status after
  first dose)
- $A_1$: treatment at time 1, influenced by $L_1$
- $Y$: outcome at the end of follow-up (e.g. mortality)

The DAG:

```
L0 → A0 → L1 → A1 → Y
      ↘   ↑    ↑    ↑
       ↘ /     |    |
       ↘×      |    |
        L1 ← ⌣ ↩     (L1 affects A1, both affect Y)
```

$L_1$ is a **time-varying confounder**: it confounds the $A_1 \to Y$
relationship (so we need to condition on it), but it is also a mediator
of the $A_0 \to Y$ effect (so conditioning on it blocks the path we want
to measure). Standard regression of $Y$ on $(A_0, A_1)$ with adjustment
for $L_0, L_1$ over-controls. Regression without $L_1$ leaves bias.

### Simulating the failure of regression

```{r}
#| label: dgp
#| cache: true

set.seed(1)
n <- 5000

# Baseline covariate
L0 <- rnorm(n)
# Treatment at t=0 depends on L0
A0 <- rbinom(n, 1, plogis(-0.5 + 0.8 * L0))
# Time-varying confounder L1: affected by A0 (and L0)
L1 <- 0.5 * L0 + 1.0 * A0 + rnorm(n)
# Treatment at t=1 depends on L1
A1 <- rbinom(n, 1, plogis(-0.5 + 0.8 * L1))
# Outcome: true effect of EACH treatment is 0.5 → cumulative effect = 1.0
# But L1 also affects Y directly
Y <- 0.5 * A0 + 0.5 * A1 + 0.5 * L1 + 0.5 * L0 + rnorm(n)

df <- tibble(L0, A0, L1, A1, Y)

cat("True total effect of always-treated vs never-treated: 1.0\n\n")

# Naive regression 1: ignore L1
naive_no_L1 <- lm(Y ~ A0 + A1 + L0, data = df)
cat(sprintf("Naive (no L1): A0=%.3f, A1=%.3f, total=%.3f\n",
            coef(naive_no_L1)["A0"], coef(naive_no_L1)["A1"],
            coef(naive_no_L1)["A0"] + coef(naive_no_L1)["A1"]))

# Naive regression 2: include L1
naive_with_L1 <- lm(Y ~ A0 + A1 + L0 + L1, data = df)
cat(sprintf("Naive (with L1): A0=%.3f, A1=%.3f, total=%.3f\n",
            coef(naive_with_L1)["A0"], coef(naive_with_L1)["A1"],
            coef(naive_with_L1)["A0"] + coef(naive_with_L1)["A1"]))
```

Neither regression recovers the true total effect of 1.0:

- Excluding $L_1$ leaves confounding from $L_1 \to A_1$ that biases $A_1$'s
  coefficient.
- Including $L_1$ blocks the $A_0 \to L_1 \to Y$ pathway, so $A_0$'s
  coefficient now measures only the direct effect, not the total effect.

The g-methods solve this.

## Parametric g-formula

The g-formula says:

$$
\mathbb{E}[Y(\bar a)] = \sum_{\bar l} \mathbb{E}[Y \mid \bar a, \bar l]
\prod_{t=0}^{K} f(l_t \mid \bar a_{t-1}, \bar l_{t-1}),
$$

where $\bar a = (a_0, a_1, \ldots, a_K)$ is a treatment regime, $\bar l$
the covariate history, and the product runs over all time points. In
practice, we fit models for the conditional distributions of each
$L_t$ and $Y$, then simulate the joint distribution under the
counterfactual regime $\bar a$.

```{r}
#| label: gformula
#| cache: true

# Step 1: Fit models for each time-varying variable
# Model L1 | L0, A0
mod_L1 <- lm(L1 ~ L0 + A0, data = df)
# Model Y | L0, A0, L1, A1
mod_Y  <- lm(Y  ~ L0 + A0 + L1 + A1, data = df)

# Step 2: Simulate counterfactual outcomes under each treatment regime
# Regime 1: always treat (A0=1, A1=1)
# Regime 2: never treat (A0=0, A1=0)

g_compute <- function(a0, a1) {
  L1_sim <- predict(mod_L1, newdata = transform(df, A0 = a0)) +
            rnorm(n, 0, sigma(mod_L1))   # simulate from L1 | L0, A0=a0
  Y_sim  <- predict(mod_Y, newdata = transform(df, A0 = a0, L1 = L1_sim, A1 = a1))
  mean(Y_sim)
}

# Monte Carlo: simulate many times to average over L1 distribution
set.seed(99)
B <- 100
E_Y11 <- mean(replicate(B, g_compute(1, 1)))
E_Y00 <- mean(replicate(B, g_compute(0, 0)))

g_effect <- E_Y11 - E_Y00
cat(sprintf("G-formula estimate (always treat vs never treat): %.3f\n", g_effect))
cat("True total effect: 1.0\n")
```

The g-formula recovers the true effect by simulating $L_1$ under the
counterfactual regime, rather than conditioning on the observed $L_1$
(which would block the desired pathway).

### What if we wanted the effect of treating only at t=0?

The g-formula handles any regime — that is its main strength. To estimate
the effect of $A_0 = 1, A_1 = 0$ vs $A_0 = 0, A_1 = 0$:

```{r}
#| label: gformula-partial
#| cache: true

set.seed(99)
E_Y10 <- mean(replicate(B, g_compute(1, 0)))
cat(sprintf("Effect of treating only at t=0:  %.3f  (true = 0.5)\n",
            E_Y10 - E_Y00))
```

This recovers the true effect of 0.5 from treating only at $t = 0$.

## IPTW for time-varying treatments

The IPTW approach reweights observations by the inverse of the probability
of their *observed* treatment history, given past covariates. The
**stabilized weight** for a unit with history $(\bar A, \bar L)$ is:

$$
SW_i = \prod_{t=0}^{K} \frac{f(A_t \mid \bar A_{t-1})}{f(A_t \mid \bar A_{t-1}, \bar L_{t-1})}.
$$

The numerator uses past treatment history (but not covariates); the
denominator uses both. Stabilisation keeps the weights bounded and reduces
variance compared to unstabilized weights.

```{r}
#| label: iptw
#| cache: true

# Estimate the propensity-score components at each time
# Numerator: P(A0) and P(A1 | A0)
n0_fit <- glm(A0 ~ 1, data = df, family = binomial)
n1_fit <- glm(A1 ~ A0, data = df, family = binomial)

# Denominator: P(A0 | L0) and P(A1 | A0, L0, L1)
d0_fit <- glm(A0 ~ L0, data = df, family = binomial)
d1_fit <- glm(A1 ~ A0 + L0 + L1, data = df, family = binomial)

# Probability of observed A at each time
p_n0 <- predict(n0_fit, type = "response")
p_n1 <- predict(n1_fit, type = "response")
p_d0 <- predict(d0_fit, type = "response")
p_d1 <- predict(d1_fit, type = "response")

w_t0_num <- ifelse(df$A0 == 1, p_n0,            1 - p_n0)
w_t0_den <- ifelse(df$A0 == 1, p_d0,            1 - p_d0)
w_t1_num <- ifelse(df$A1 == 1, p_n1,            1 - p_n1)
w_t1_den <- ifelse(df$A1 == 1, p_d1,            1 - p_d1)

# Stabilized weight (product over time)
sw <- (w_t0_num * w_t1_num) / (w_t0_den * w_t1_den)

cat(sprintf("Weight summary: min=%.3f, mean=%.3f, max=%.3f\n",
            min(sw), mean(sw), max(sw)))

# Trim extreme weights (standard practice)
sw_trimmed <- pmin(sw, quantile(sw, 0.99))
```

## Marginal structural models

With the stabilised weights in hand, the **marginal structural model** is
a simple regression of $Y$ on treatment history, *weighted* by $SW$:

```{r}
#| label: msm
#| cache: true

# MSM: linear in cumulative treatment
df$sw <- sw_trimmed
design <- svydesign(ids = ~1, weights = ~sw, data = df)
msm_fit <- svyglm(Y ~ A0 + A1, design = design)

tidy(msm_fit) |> knitr::kable(digits = 3,
                              caption = "MSM coefficients with stabilized weights")

cat(sprintf("\nMSM total effect (always vs never): %.3f  (true = 1.0)\n",
            coef(msm_fit)["A0"] + coef(msm_fit)["A1"]))
```

The MSM recovers the total effect because the IPTW reweighting breaks the
$L_1 \to A_1$ arrow in the reweighted population.

## Longitudinal TMLE

The `ltmle` package implements **longitudinal TMLE**, which combines
g-computation and IPTW to produce a doubly-robust efficient estimator. It
also automatically reports TMLE, IPTW, and g-comp estimates side by side.

```{r}
#| label: ltmle
#| cache: true
#| warning: false

# ltmle expects a specific column order: covariates, treatments, outcome
ltmle_data <- df[, c("L0", "A0", "L1", "A1", "Y")]

ltmle_fit <- ltmle(
  data    = ltmle_data,
  Anodes  = c("A0", "A1"),
  Lnodes  = c("L0", "L1"),
  Ynodes  = "Y",
  abar    = c(1, 1),       # the regime: always treated
  estimate.time = FALSE
)

summary(ltmle_fit)
```

`ltmle` returns four estimators of $\mathbb{E}[Y(\bar a = 1)]$ side by side: TMLE
(efficient, doubly robust), IPTW (Horvitz-Thompson reweighting), G-comp
(parametric g-formula), and IC-based standard errors throughout.

To estimate a *contrast* — the always-treated vs never-treated effect —
specify both regimes:

```{r}
#| label: ltmle-contrast
#| cache: true
#| warning: false

ltmle_contrast <- ltmle(
  data    = ltmle_data,
  Anodes  = c("A0", "A1"),
  Lnodes  = c("L0", "L1"),
  Ynodes  = "Y",
  abar    = list(treatment = c(1, 1), control = c(0, 0)),
  estimate.time = FALSE
)

summary(ltmle_contrast)
```

The output contains the additive treatment effect of "always treat" vs
"never treat" along with valid confidence intervals. In this simulation,
the LTMLE estimate should be close to 1.0.

## Choosing among the four

| Estimator | Strengths | Weaknesses |
|---|---|---|
| **G-formula** | Handles arbitrary regimes; transparent | Sensitive to model misspecification on $L_t$ and $Y$ |
| **IPTW + MSM** | Simple, intuitive; valid with correct propensity model | Variance can be large with extreme weights |
| **LTMLE** | Doubly robust; efficient | More complex; needs care with super-learners |

In practice:

- Use the **g-formula** when you have good models for both the time-varying
  confounders and the outcome.
- Use **IPTW + MSM** when the propensity score model is more credible than
  the outcome model.
- Use **LTMLE** as the modern default — it is doubly robust and efficient.

## When to reach for these methods

G-methods are needed whenever:

1. There is a time-varying confounder $L_t$ that is affected by past
   treatment $A_{t-1}$ and itself affects subsequent treatment $A_t$.
2. The research question concerns a *sustained* or *dynamic* treatment
   regime, not a single decision at a single time.

This includes most clinical follow-up studies, longitudinal observational
studies of education and labour market policies, and any setting where
treatment evolves with the unit's response.

For a one-shot treatment, the methods from the [Estimation](estimation.qmd)
and [Nonparametric Causal Methods](nonparametric.qmd) chapters are
sufficient. The added complexity of g-methods is only justified when the
time-varying confounder problem is present.

## Summary

- **Standard regression fails** with time-varying confounders that are
  themselves affected by past treatment.
- **G-formula** (parametric g-computation): simulate the outcome under a
  hypothetical regime by integrating over the time-varying covariates.
- **IPTW + MSM**: reweight the population so treatment is independent of
  past covariates; then fit a regression on the weighted sample.
- **LTMLE** (`ltmle`): a doubly-robust efficient estimator that combines
  g-computation and IPTW.
- **Reporting practice**: when in doubt, report multiple estimators (TMLE,
  IPTW, G-comp). Disagreement between them is informative — it usually
  indicates model misspecification.

## References

- Robins, J. M. (1986). A new approach to causal inference in mortality studies with sustained exposure periods. *Mathematical Modelling*, 7, 1393–1512.
- Robins, J. M., Hernan, M. A., & Brumback, B. (2000). Marginal structural models and causal inference in epidemiology. *Epidemiology*, 11(5), 550–560.
- van der Laan, M. J., & Gruber, S. (2012). Targeted minimum loss based estimation of causal effects of multiple time point interventions. *International Journal of Biostatistics*, 8(1).
- Hernán, M. A., & Robins, J. M. (2020). *Causal Inference: What If.* Boca Raton: Chapman & Hall/CRC.
- Petersen, M., Schwab, J., Gruber, S., Blaser, N., Schomaker, M., & van der Laan, M. (2014). Targeted maximum likelihood estimation for dynamic and static longitudinal marginal structural working models. *Journal of Causal Inference*, 2(2), 147–185.
