# From Graph to Estimate

```{r}
#| include: false
library(dagitty)
library(igraph)
library(causaleffect)
library(ggplot2)
library(dplyr)
```

```{r}
#| eval: false
library(dagitty)
library(igraph)
library(causaleffect)
library(ggplot2)
```

The previous chapters covered discovery: algorithms that learn a graph from
data. But a graph is not an estimate. After we have a graph, we still need
to ask two questions:

1. **Identification**: is the target effect a function of the observed-data
   distribution?
2. **Estimation**: once the identifying formula is known, estimate it from
   the data.

In R, `dagitty` handles identification via the backdoor and front-door
criteria; `causaleffect` implements the full Pearl–Shpitser ID algorithm for
acyclic directed mixed graphs (ADMGs) with hidden confounders. Estimation
is done directly from the identifying formula — a step that is explicit in
R, making the connection between the formula and the estimator transparent.

## Backdoor: standard adjustment

Start with the simple case. $X$ is a common cause of treatment $A$ and
outcome $Y$, and $X$ is observed. Backdoor adjustment applies.

```{r}
#| label: backdoor-graph
#| cache: true

# X is a common cause of A and Y
g_bd <- dagitty("dag { X -> A; X -> Y; A -> Y }")
exposures(g_bd) <- "A"
outcomes(g_bd)  <- "Y"

sets <- adjustmentSets(g_bd, exposure = "A", outcome = "Y")
cat("Adjustment sets:\n")
print(sets)
```

`dagitty` finds `{X}` as the minimal adjustment set. Adjusting on $X$
blocks the only open backdoor path $A \leftarrow X \rightarrow Y$.

We simulate data matching that graph. We draw $n = 2000$ observations with
$X \sim N(0,1)$, treatment $A \sim \text{Bernoulli}(\Lambda(X))$, and outcome
$Y = 2A + X + 0.5\varepsilon$ with $\varepsilon \sim N(0,1)$. The effect is
constant, so the true ACE is 2. We then estimate it with AIPW using the
adjustment set the graph chose, and take a 95% interval from 500 bootstrap
resamples.

```{r}
#| label: backdoor-sim
#| cache: true

set.seed(1)
n     <- 2000
X     <- rnorm(n)
A     <- as.numeric(runif(n) < 1 / (1 + exp(-X)))  # logistic: A depends on X
Y     <- 2 * A + X + 0.5 * rnorm(n)                 # true ACE = 2
data_bd <- data.frame(X = X, A = A, Y = Y)

# AIPW (doubly-robust) estimator for the backdoor-adjusted ACE
# Outcome model: E[Y | A, X]
mu1 <- predict(lm(Y ~ A + X, data = data_bd), newdata = transform(data_bd, A = 1))
mu0 <- predict(lm(Y ~ A + X, data = data_bd), newdata = transform(data_bd, A = 0))

# Propensity score: P(A = 1 | X)
ps  <- predict(glm(A ~ X, data = data_bd, family = binomial), type = "response")

# AIPW influence function
aipw <- mean(
  (data_bd$A / ps) * (Y - mu1) - ((1 - data_bd$A) / (1 - ps)) * (Y - mu0) +
  mu1 - mu0
)

cat(sprintf("Backdoor AIPW ACE (true = 2.0): %.3f\n", aipw))

# Bootstrap 95% CI
set.seed(99)
boot_bd <- replicate(500, {
  idx  <- sample.int(n, replace = TRUE)
  bd   <- data_bd[idx, ]
  m1   <- predict(lm(Y ~ A + X, data = bd), newdata = transform(bd, A = 1))
  m0   <- predict(lm(Y ~ A + X, data = bd), newdata = transform(bd, A = 0))
  p    <- predict(glm(A ~ X, data = bd, family = binomial), type = "response")
  mean((bd$A / p) * (bd$Y - m1) - ((1 - bd$A) / (1 - p)) * (bd$Y - m0) + m1 - m0)
})
ci_bd <- quantile(boot_bd, c(0.025, 0.975))
cat(sprintf("95%% CI: [%.3f, %.3f]\n", ci_bd[1], ci_bd[2]))
```

AIPW returns 1.990 with a 95% bootstrap interval of $[1.942, 2.043]$, covering
the true 2. The AIPW estimator is consistent if either the outcome model or the
propensity score is correctly specified; here both are, since the DGP is linear
in the outcome and logistic in the treatment.

## Front-door: p-fixable effects

Now suppose $A$ and $Y$ have an unmeasured common cause $U$, but there is a
measured mediator $M$ on the path $A \rightarrow M \rightarrow Y$. Backdoor
adjustment fails because $U$ is unobserved. The front-door formula can still
identify the effect.

In `dagitty`, the bidirected edge $A \leftrightarrow Y$ (representing the
latent $U$) appears as `A <-> Y`. For `causaleffect`, the same edge is
encoded as two directed edges `A -> Y` and `Y -> A` both marked with
`description = "U"` in the igraph object.

```{r}
#| label: frontdoor-graph
#| cache: true

# dagitty for display and adjustment-set check
g_fd <- dagitty("dag { A -> M; M -> Y; A <-> Y }")
exposures(g_fd) <- "A"
outcomes(g_fd)  <- "Y"

# An empty dagitty adjustment-set list prints nothing at all, so report the
# count rather than calling print() on it and appearing to produce no output.
sets_fd <- adjustmentSets(g_fd, exposure = "A", outcome = "Y")
cat("Valid backdoor adjustment sets:", length(sets_fd), "\n")

# causaleffect for ID algorithm: bidirected A <-> Y = two directed edges + description "U"
g_fd_ig <- graph_from_literal(A -+ M, M -+ Y, A -+ Y, Y -+ A)
# Mark the A<->Y bidirected edge by endpoint name rather than a hardcoded
# edge index: igraph happens to sort edges by source vertex today (giving
# order A->M(1), A->Y(2), M->Y(3), Y->A(4)), but that ordering isn't part of
# igraph's documented contract, so indices like c(2,4) can silently break if
# igraph's internals or this graph's construction ever changes.
edge_ends <- ends(g_fd_ig, E(g_fd_ig))
confounded <- (edge_ends[, 1] == "A" & edge_ends[, 2] == "Y") |
              (edge_ends[, 1] == "Y" & edge_ends[, 2] == "A")
g_fd_ig <- set_edge_attr(g_fd_ig, "description", index = which(confounded), value = "U")

expr_fd <- causal.effect(y = "Y", x = "A", G = g_fd_ig, simp = TRUE)
cat("\nIdentifying expression:\n")
cat(expr_fd, "\n")
```

There are no valid backdoor adjustment sets, as expected: $U$ is unobserved, so
no set of measured variables blocks the path it opens. And the expression
`causal.effect` returns is the front-door formula,
$\sum_M P(M \mid A) \sum_{A'} P(Y \mid A', M)\, P(A')$.

We now simulate data for it. We draw $n = 3000$ observations with an unmeasured
confounder $U \sim N(0,1)$, treatment $A \sim \text{Bernoulli}(\Lambda(U))$,
mediator $M \sim \text{Bernoulli}(\Lambda(0.5 + 2A))$, and outcome
$Y = 1.5M + 2U + 0.5\varepsilon$ with $\varepsilon \sim N(0,1)$. The mediator
depends on $A$ alone and the outcome depends on $M$ and $U$ but not directly on
$A$, which is what the front-door criterion requires.

The truth follows from the DGP. Intervening on $A$ changes $Y$ only through
$M$, so the ACE is $1.5 \times [P(M=1 \mid A=1) - P(M=1 \mid A=0)]
= 1.5 \times [\Lambda(2.5) - \Lambda(0.5)] = 1.5 \times 0.302 = 0.453$.

```{r}
#| label: frontdoor-sim
#| cache: true

set.seed(2)
n     <- 3000
U     <- rnorm(n)                                     # unmeasured confounder
A_fd  <- as.numeric(runif(n) < 1 / (1 + exp(-U)))     # A depends on U
M_fd  <- as.numeric(runif(n) < 1 / (1 + exp(-(0.5 + 2 * A_fd))))  # M depends on A only
Y_fd  <- 1.5 * M_fd + 2 * U + 0.5 * rnorm(n)          # Y depends on M and U

# True ACE: E[Y | do(A=1)] - E[Y | do(A=0)]
# = 1.5 * (E[M | A=1] - E[M | A=0])
true_EMA1  <- 1 / (1 + exp(-(0.5 + 2)))   # scalar P(M=1 | A=1)
true_EMA0  <- 1 / (1 + exp(-0.5))         # scalar P(M=1 | A=0)
true_ace_fd <- 1.5 * (true_EMA1 - true_EMA0)
cat(sprintf("True front-door ACE = %.3f\n", true_ace_fd))

data_fd <- data.frame(A = A_fd, M = M_fd, Y = Y_fd)
```

Estimate the front-door functional using the plug-in formula:

```{r}
#| label: frontdoor-est
#| cache: true

front_door_ace <- function(data) {
  pA <- mean(data$A)
  pM_given_A <- prop.table(table(data$A, data$M), margin = 1)
  pY_given_AM <- with(data, tapply(Y, list(A, M), mean))

  ey_do <- function(a) {
    sum(sapply(c(0, 1), function(m) {
      p_m_given_a <- pM_given_A[as.character(a), as.character(m)]
      inner <- sum(sapply(c(0, 1), function(ap)
        pY_given_AM[as.character(ap), as.character(m)] *
          ifelse(ap == 1, pA, 1 - pA)))
      p_m_given_a * inner
    }))
  }
  ey_do(1) - ey_do(0)
}

est_fd <- front_door_ace(data_fd)

set.seed(77)
boot_fd <- replicate(500, {
  idx <- sample.int(nrow(data_fd), replace = TRUE)
  front_door_ace(data_fd[idx, ])
})
ci_fd  <- quantile(boot_fd, c(0.025, 0.975))

cat(sprintf("Front-door ACE: %.3f  [95%% CI: %.3f, %.3f]\n",
            est_fd, ci_fd[1], ci_fd[2]))

# Naïve mean difference for comparison
naive_ace <- mean(Y_fd[A_fd == 1]) - mean(Y_fd[A_fd == 0])
cat(sprintf("Naïve mean difference (biased!): %.3f\n", naive_ace))
cat(sprintf("True ACE:                        %.3f\n", true_ace_fd))
```

The front-door estimate is 0.499 with a 95% bootstrap interval of
$[0.432, 0.577]$, which covers the true 0.453. The naive mean difference is
2.083 --- more than four times the truth, and outside the front-door interval
by a factor of four. The formula recovers the effect even though $U$ is never
measured; the naive comparison is picking up the path through $U$, which enters
the outcome with coefficient 2 and the treatment through the logit.

The bias here is far larger than in the front-door example of the
[graphs chapter](graphs-identification-estimation.qmd), where the naive
contrast was 0.14 against a truth of 0.055. The difference is the strength of
the confounding: $U$ enters $Y$ with coefficient 2 here against 0.7 there.
Nothing about the front-door formula changes; what changes is how much work it
is doing.

## When identification fails

Not every graph identifies the effect. If $A$ and $Y$ have a hidden common
cause and there is no front-door variable, the effect is not identified. The
graph below is the **bow**: a directed edge $A \to Y$ plus a bidirected
$A \leftrightarrow Y$, and nothing else. Remove the mediator $M$ from the
front-door graph and this is what is left.

```{r}
#| label: not-identified
#| cache: true

# A → Y with unmeasured confounding (A <-> Y) and no mediator: the "bow"
# graph needs THREE igraph edges — the directed A → Y plus the
# bidirected pair (encoded as two opposite edges marked "U").
# graph_from_literal() silently collapses a duplicated A -+ Y, so we
# build the multigraph with make_graph().
g_unident_ig <- make_graph(c("A","Y",  "A","Y",  "Y","A"), directed = TRUE)
# make_graph() preserves the literal edge order given above (unlike
# graph_from_literal(), which re-sorts), so index 2:3 is reliable as long as
# the c(...) vector above isn't reordered -- mark by endpoint pair anyway so
# a future edit to that vector can't silently mismark the real A->Y edge.
edge_ends <- ends(g_unident_ig, E(g_unident_ig))
# The first A->Y edge is the real causal edge; the second (a duplicate,
# per `duplicated()`) together with Y->A forms the bidirected U pair.
u_edges <- which(((edge_ends[,1]=="A" & edge_ends[,2]=="Y") & duplicated(edge_ends)) |
                  (edge_ends[,1]=="Y" & edge_ends[,2]=="A"))
g_unident_ig <- set_edge_attr(g_unident_ig, "description",
                               index = u_edges, value = "U")

result <- tryCatch(
  causal.effect(y = "Y", x = "A", G = g_unident_ig, simp = TRUE),
  error = function(e) conditionMessage(e)
)
cat("Result from causal.effect:\n", result, "\n")
```

`causaleffect` raises an error, printed above as "Not identifiable.", which the
`tryCatch` captures. This is a useful failure. The algorithm did not return a
formula that would quietly estimate the wrong thing; it reported that no
function of the observed distribution equals the causal effect under this
graph. No amount of data helps, because the obstacle is the graph and not the
sample size.

The honest response in applied work is to report the
non-identifiability, then either:

1. Find additional variables that close the unmeasured-confounding paths
2. Apply sensitivity analysis to bound the effect under maintained assumptions
3. Find an instrumental variable for a LATE-style identification (IV chapter)

Saying that the hidden confounding is probably small is sensitivity analysis,
not identification.

## End-to-end: discover, identify, estimate

Discovery algorithms return graphs that can be passed to the identification
step:

```{r}
#| eval: false
#| echo: true

library(pcalg)

# Step 1: discover the graph (chapters on causal discovery)
suffStat <- list(C = cor(data_obs), n = nrow(data_obs))
pc_fit   <- pc(suffStat, indepTest = gaussCItest,
               alpha = 0.05, p = ncol(data_obs))

# Step 2: check for backdoor adjustment set using dagitty
# (convert pcalg CPDAG → dagitty for convenience).
# NOTE the t(): as(graphNEL, "matrix") is from→to, but pcalg2dagitty
# expects the amat.cpdag coding, which is its transpose — without it
# every directed edge is silently reversed.
g_dag <- pcalg2dagitty(t(as(pc_fit@graph, "matrix")),
                        colnames(data_obs), type = "cpdag")
sets  <- adjustmentSets(g_dag, exposure = "A", outcome = "Y")

# Step 3: if backdoor set found, estimate with AIPW
# Step 4: if not, call causaleffect::causal.effect() for the ID expression
# Step 5: estimate the returned functional by plug-in or regression
```

In applied work, I would not trust the discovered graph blindly. A more
reasonable workflow is:

1. **Hypothesise** a graph based on subject-matter knowledge
2. **Discover** a data-driven graph using PC/GES/FCI
3. **Compare** the two — disagreement points to substantive questions
4. **Identify** under both graphs separately
5. **Estimate** under both, and report the range as part of the result

If the estimate changes a lot across plausible graphs, the data alone is not
settling the question.

## Why this matters

The graph-identification-estimation sequence matters because:

- **Backdoor adjustment is not always available.** Many real DAGs have
  hidden confounders. Front-door, nested, and general ID algorithms are
  the recourse, but they are too complex to derive by hand for non-trivial
  graphs.
- **The right estimator depends on the identification strategy.**
  AIPW-for-backdoor uses a different formula than the front-door plug-in.
  Matching the estimator to the identification result avoids the common
  mistake of plugging a graph into the wrong estimator.
- **Reporting becomes more disciplined.** Once the workflow forces you to
  specify a graph, it becomes harder to wave hands about "unconfoundedness"
  — you have to say exactly which variables are doing the unconfounding
  work.

## Summary

- A graph is not an estimate. Identification is the bridge.
- `dagitty` handles backdoor and front-door checks; `causaleffect` handles
  general ADMG identification.
- Backdoor effects can be estimated with AIPW.
- Front-door effects require estimating the front-door functional.
- If the effect is not identified, report that instead of forcing a biased
  regression estimate.
