# From Graph to Estimate

```{r}
#| include: false
library(dagitty)
library(igraph)
library(causaleffect)
library(ggplot2)
library(dplyr)
```

```{r}
#| eval: false
library(dagitty)
library(igraph)
library(causaleffect)
library(ggplot2)
```

The two previous chapters covered **discovery** — algorithms that learn a
graph from data. Discovery returns a graph, but a graph is not yet an
estimate. To go from the graph to an actionable number, we need two more
steps:

1. **Identification** — given the graph and a target effect, is the effect a
   function of the observed-data distribution? If yes, derive the
   identifying functional. If no, declare non-identification (or place
   bounds).
2. **Estimation** — fit the identifying functional to data, ideally with a
   doubly-robust or efficient estimator that yields a confidence interval.

In R, `dagitty` handles identification via the backdoor and front-door
criteria; `causaleffect` implements the full Pearl–Shpitser ID algorithm for
acyclic directed mixed graphs (ADMGs) with hidden confounders. Estimation
is done directly from the identifying formula — a step that is explicit in
R, making the connection between the formula and the estimator transparent.

## Backdoor: standard adjustment

The simplest case. Treatment $A$ has measured parents, and there is no
unmeasured common cause between $A$ and $Y$ along any backdoor path.
Standard backdoor adjustment applies.

```{r}
#| label: backdoor-graph
#| cache: true

# X is a common cause of A and Y
g_bd <- dagitty("dag { X -> A; X -> Y; A -> Y }")
exposures(g_bd) <- "A"
outcomes(g_bd)  <- "Y"

sets <- adjustmentSets(g_bd, exposure = "A", outcome = "Y")
cat("Adjustment sets:\n")
print(sets)
```

`dagitty` finds `{X}` as the minimal adjustment set. Adjusting on $X$
blocks the only open backdoor path $A \leftarrow X \rightarrow Y$.

```{r}
#| label: backdoor-sim
#| cache: true

set.seed(1)
n     <- 2000
X     <- rnorm(n)
A     <- as.numeric(runif(n) < 1 / (1 + exp(-X)))  # logistic: A depends on X
Y     <- 2 * A + X + 0.5 * rnorm(n)                 # true ACE = 2
data_bd <- data.frame(X = X, A = A, Y = Y)

# AIPW (doubly-robust) estimator for the backdoor-adjusted ACE
# Outcome model: E[Y | A, X]
mu1 <- predict(lm(Y ~ A + X, data = data_bd), newdata = transform(data_bd, A = 1))
mu0 <- predict(lm(Y ~ A + X, data = data_bd), newdata = transform(data_bd, A = 0))

# Propensity score: P(A = 1 | X)
ps  <- predict(glm(A ~ X, data = data_bd, family = binomial), type = "response")

# AIPW influence function
aipw <- mean(
  (data_bd$A / ps) * (Y - mu1) - ((1 - data_bd$A) / (1 - ps)) * (Y - mu0) +
  mu1 - mu0
)

cat(sprintf("Backdoor AIPW ACE (true = 2.0): %.3f\n", aipw))

# Bootstrap 95% CI
set.seed(99)
boot_bd <- replicate(500, {
  idx  <- sample.int(n, replace = TRUE)
  bd   <- data_bd[idx, ]
  m1   <- predict(lm(Y ~ A + X, data = bd), newdata = transform(bd, A = 1))
  m0   <- predict(lm(Y ~ A + X, data = bd), newdata = transform(bd, A = 0))
  p    <- predict(glm(A ~ X, data = bd, family = binomial), type = "response")
  mean((bd$A / p) * (bd$Y - m1) - ((1 - bd$A) / (1 - p)) * (bd$Y - m0) + m1 - m0)
})
ci_bd <- quantile(boot_bd, c(0.025, 0.975))
cat(sprintf("95%% CI: [%.3f, %.3f]\n", ci_bd[1], ci_bd[2]))
```

The AIPW estimator is doubly robust: it is consistent if either the outcome
model or the propensity score is correctly specified, and efficient when
both are.

## Front-door: p-fixable effects

Now suppose there is an *unmeasured* common cause $U$ of $A$ and $Y$, but a
measured mediator $M$ lies on the causal path $A \rightarrow M \rightarrow Y$,
with no arrow from $U$ to $M$ except through $A$. Standard backdoor
adjustment fails because $U$ is unobserved, but the front-door criterion
identifies the effect.

In `dagitty`, the bidirected edge $A \leftrightarrow Y$ (representing the
latent $U$) appears as `A <-> Y`. For `causaleffect`, the same edge is
encoded as two directed edges `A -> Y` and `Y -> A` both marked with
`description = "U"` in the igraph object.

```{r}
#| label: frontdoor-graph
#| cache: true

# dagitty for display and adjustment-set check
g_fd <- dagitty("dag { A -> M; M -> Y; A <-> Y }")
exposures(g_fd) <- "A"
outcomes(g_fd)  <- "Y"

cat("Backdoor adjustment sets (should be empty):\n")
print(adjustmentSets(g_fd, exposure = "A", outcome = "Y"))

# causaleffect for ID algorithm: bidirected A <-> Y = two directed edges + description "U"
g_fd_ig <- graph_from_literal(A -+ M, M -+ Y, A -+ Y, Y -+ A)
# igraph sorts edges by source vertex: A-edges come before M-edges before Y-edges.
# Resulting order: A->M(1), A->Y(2), M->Y(3), Y->A(4). Bidirected pair = edges 2 & 4.
g_fd_ig <- set_edge_attr(g_fd_ig, "description", index = c(2, 4), value = "U")

expr_fd <- causal.effect(y = "Y", x = "A", G = g_fd_ig, simp = TRUE)
cat("\nIdentifying expression:\n")
cat(expr_fd, "\n")
```

The expression returned is the front-door formula:
$\sum_M P(M \mid A) \sum_{A'} P(Y \mid A', M)\, P(A')$. We now estimate it directly.

```{r}
#| label: frontdoor-sim
#| cache: true

set.seed(2)
n     <- 3000
U     <- rnorm(n)                                     # unmeasured confounder
A_fd  <- as.numeric(runif(n) < 1 / (1 + exp(-U)))     # A depends on U
M_fd  <- as.numeric(runif(n) < 1 / (1 + exp(-(0.5 + 2 * A_fd))))  # M depends on A only
Y_fd  <- 1.5 * M_fd + 2 * U + 0.5 * rnorm(n)          # Y depends on M and U

# True ACE: E[Y | do(A=1)] - E[Y | do(A=0)]
# = 1.5 * (E[M | A=1] - E[M | A=0])
true_EMA1  <- mean(1 / (1 + exp(-(0.5 + 2))))
true_EMA0  <- mean(1 / (1 + exp(-0.5)))
true_ace_fd <- 1.5 * (true_EMA1 - true_EMA0)
cat(sprintf("True front-door ACE = %.3f\n", true_ace_fd))

data_fd <- data.frame(A = A_fd, M = M_fd, Y = Y_fd)
```

Estimate the front-door functional using the plug-in formula:

```{r}
#| label: frontdoor-est
#| cache: true

front_door_ace <- function(data) {
  pA <- mean(data$A)
  pM_given_A <- prop.table(table(data$A, data$M), margin = 1)
  pY_given_AM <- with(data, tapply(Y, list(A, M), mean))

  ey_do <- function(a) {
    sum(sapply(c(0, 1), function(m) {
      p_m_given_a <- pM_given_A[as.character(a), as.character(m)]
      inner <- sum(sapply(c(0, 1), function(ap)
        pY_given_AM[as.character(ap), as.character(m)] *
          ifelse(ap == 1, pA, 1 - pA)))
      p_m_given_a * inner
    }))
  }
  ey_do(1) - ey_do(0)
}

est_fd <- front_door_ace(data_fd)

set.seed(77)
boot_fd <- replicate(500, {
  idx <- sample.int(nrow(data_fd), replace = TRUE)
  front_door_ace(data_fd[idx, ])
})
ci_fd  <- quantile(boot_fd, c(0.025, 0.975))

cat(sprintf("Front-door ACE: %.3f  [95%% CI: %.3f, %.3f]\n",
            est_fd, ci_fd[1], ci_fd[2]))

# Naïve mean difference for comparison
naive_ace <- mean(Y_fd[A_fd == 1]) - mean(Y_fd[A_fd == 0])
cat(sprintf("Naïve mean difference (biased!): %.3f\n", naive_ace))
cat(sprintf("True ACE:                        %.3f\n", true_ace_fd))
```

Despite $U$ being unobserved, the front-door formula recovers the causal
effect. The naïve difference is contaminated by the $U \rightarrow A$ and
$U \rightarrow Y$ paths that the front-door route closes.

## When identification fails

Not every ADMG yields an identified effect. Suppose $A$ and $Y$ have a
hidden common cause **and** there is no front-door mediator. The effect is
**not identified**.

```{r}
#| label: not-identified
#| cache: true

# A → Y with unmeasured confounding (A <-> Y) and no mediator
g_unident_ig <- graph_from_literal(A -+ Y, Y -+ A)
g_unident_ig <- set_edge_attr(g_unident_ig, "description",
                               index = 1:2, value = "U")

result <- tryCatch(
  causal.effect(y = "Y", x = "A", G = g_unident_ig, simp = TRUE),
  error = function(e) conditionMessage(e)
)
cat("Result from causal.effect:\n", result, "\n")
```

`causaleffect` returns an error or `FALSE`, indicating the effect is not
identifiable. The honest response in applied work is to report the
non-identifiability, then either:

1. Find additional variables that close the unmeasured-confounding paths
2. Apply sensitivity analysis to bound the effect under maintained assumptions
3. Find an instrumental variable for a LATE-style identification (IV chapter)

A hidden assumption that everyone tacitly makes — "of course the
confounding is small" — is *not* identification. Either the graph supports
the effect or it does not.

## End-to-end: discover, identify, estimate

The previous chapters' discovery algorithms (PC, GES, FCI, RFCI) return
graphs that can be passed directly to the identification step. The full
workflow is:

```{r}
#| eval: false
#| echo: true

library(pcalg)

# Step 1: discover the graph (chapters on causal discovery)
suffStat <- list(C = cor(data_obs), n = nrow(data_obs))
pc_fit   <- pc(suffStat, indepTest = gaussCItest,
               alpha = 0.05, p = ncol(data_obs))

# Step 2: check for backdoor adjustment set using dagitty
# (convert pcalg CPDAG → dagitty for convenience)
g_dag <- pcalg2dagitty(as(pc_fit@graph, "matrix"),
                        colnames(data_obs), type = "cpdag")
sets  <- adjustmentSets(g_dag, exposure = "A", outcome = "Y")

# Step 3: if backdoor set found, estimate with AIPW
# Step 4: if not, call causaleffect::causal.effect() for the ID expression
# Step 5: estimate the returned functional by plug-in or regression
```

In practice, the discovered graph is almost never trusted blindly. The
applied-research workflow is:

1. **Hypothesise** a graph based on subject-matter knowledge
2. **Discover** a data-driven graph using PC/GES/FCI
3. **Compare** the two — disagreement points to substantive questions
4. **Identify** under both graphs separately
5. **Estimate** under both, and report the range as part of the result

If the estimate is robust to the choice of graph, the conclusion is strong.
If it flips sign depending on which graph you use, the data alone is not
enough to answer the question.

## Why this matters

The "graph → identification → estimation" pipeline closes the loop between
causal-discovery papers and applied empirical work. Practically:

- **Backdoor adjustment is not always available.** Many real DAGs have
  hidden confounders. Front-door, nested, and general ID algorithms are
  the recourse, but they are too complex to derive by hand for non-trivial
  graphs.
- **The right estimator depends on the identification strategy.**
  AIPW-for-backdoor uses a different formula than the front-door plug-in.
  Matching the estimator to the identification result avoids the common
  mistake of plugging a graph into the wrong estimator.
- **Reporting becomes more disciplined.** Once the workflow forces you to
  specify a graph, it becomes harder to wave hands about "unconfoundedness"
  — you have to say exactly which variables are doing the unconfounding
  work.

## Summary

- **Discovery returns a graph; estimation returns a number.** The bridge is
  the **identification** step.
- `dagitty` implements backdoor and front-door criteria; `causaleffect`
  implements the full Pearl–Shpitser ID algorithm for ADMGs with bidirected
  edges (hidden confounders).
- **Backdoor case**: `adjustmentSets()` + AIPW estimation.
- **Front-door case**: `causal.effect()` returns the functional; estimate
  it by direct plug-in or regression.
- **Non-identified case**: `causal.effect()` errors — report
  non-identifiability, not a biased point estimate.
- **End-to-end workflow**: hypothesise a graph, run discovery as a check,
  identify, estimate, and report the range across plausible graphs.

