# DAGs, ADMGs, Identification, and Estimation

```{r}
#| include: false
suppressPackageStartupMessages({
  library(dagitty)
  library(ggdag)
  library(ggplot2)
  library(igraph)
  library(causaleffect)
  library(dplyr)
  library(knitr)
})
```

The potential-outcomes chapter asked when an observed data set can tell us about $E[Y(1)-Y(0)]$. Graphs give one way to state the assumptions behind the answer. A graph records which variables are allowed to cause other variables, and whether some observed variables may share unobserved common causes. Once the graph is stated, identification is a graph problem. Estimation comes after that: estimate the functional that the graph identified.

I use `dagitty` for graphs and backdoor adjustment, `causaleffect` for the general Pearl-Shpitser ID algorithm, and small direct estimators for the examples. The goal is to keep the order clear: graph first, identified functional second, estimator third.

## DAGs and ADMGs

A directed acyclic graph, or DAG, has directed arrows such as `X -> A`. In causal work, an arrow means a direct causal relation is allowed by the model. Absence of an arrow is also a substantive assumption.

An acyclic directed mixed graph, or ADMG, adds bidirected arrows such as `A <-> Y`. A bidirected edge represents an unobserved common cause. This is useful because many econometric applications have variables we cannot measure: ability, preferences, latent health, firm quality, and so on.

The graph is not an estimator. It is a compact statement of assumptions. The workflow is:

1. write down a DAG or ADMG;
2. ask whether the target effect is identified;
3. estimate the identified functional from data.

## A DAG: Backdoor Adjustment

Start with a simple observational study. `A` is treatment, `Y` is the outcome, and `X` is an observed confounder:

```{r}
g_dag <- dagitty("dag {
  X -> A
  X -> Y
  A -> Y
}")
coordinates(g_dag) <- list(
  x = c(X = 1, A = 0, Y = 2),
  y = c(X = 1, A = 0, Y = 0)
)
```

```{r}
#| label: fig-dag-backdoor
#| fig-cap: "Backdoor DAG: X is an observed confounder of A → Y."
#| fig-width: 5
#| fig-height: 3
ggdag(g_dag) + theme_dag_blank()
```

The graph says `X` affects both treatment and outcome. It also says there is no unobserved common cause between `A` and `Y` after we condition on `X`. In this case the treatment effect is identifiable by **backdoor adjustment**.

```{r}
adj_sets <- adjustmentSets(g_dag, exposure = "A", outcome = "Y")
adj_sets
```

The adjustment set is $\{X\}$. Under this graph,

$$
E[Y(a)] = E_X\{E(Y \mid A=a, X)\}.
$$

So the graph has translated a causal query into an observed-data functional.

## Estimating the DAG Effect with AIPW

Here is a small simulated data set where `X` confounds the treatment–outcome association.

```{r}
#| warning: false
#| message: false
#| cache: true
set.seed(2026)

logistic <- function(x) 1 / (1 + exp(-x))

n  <- 400
X  <- rnorm(n)
pA <- logistic(-0.2 + 0.8 * X)
A  <- as.numeric(runif(n) < pA)
Y  <- 1 + 1.5 * A + 0.7 * X + 0.3 * rnorm(n)

df_dag <- data.frame(X = X, A = A, Y = Y)
head(df_dag)
```

Because the graph is backdoor-identified, use AIPW with the adjustment set the graph chose. The function below implements the AIPW point estimate and an EIF-based standard error.

```{r}
aipw_backdoor <- function(data, treatment, outcome, adjust) {
  fmla_y <- as.formula(paste0(outcome, " ~ ", treatment, " * (",
                              paste(adjust, collapse = " + "), ")"))
  fmla_a <- as.formula(paste0(treatment, " ~ ",
                              paste(adjust, collapse = " + ")))

  mu_fit <- lm(fmla_y, data = data)
  pi_fit <- glm(fmla_a, data = data, family = binomial())

  d1 <- data; d1[[treatment]] <- 1
  d0 <- data; d0[[treatment]] <- 0
  mu1 <- predict(mu_fit, newdata = d1)
  mu0 <- predict(mu_fit, newdata = d0)
  pi1 <- predict(pi_fit, newdata = data, type = "response")

  w  <- data[[treatment]]; y <- data[[outcome]]
  if1 <- w     * (y - mu1) / pi1       + mu1
  if0 <- (1-w) * (y - mu0) / (1 - pi1) + mu0
  psi  <- if1 - if0

  est  <- mean(psi)
  se   <- sd(psi) / sqrt(length(psi))
  ci   <- est + c(-1, 1) * qnorm(0.975) * se
  data.frame(estimand = "Backdoor ACE",
             estimate = round(est, 4),
             lower_95 = round(ci[1], 4),
             upper_95 = round(ci[2], 4),
             se       = round(se, 4))
}

aipw_backdoor(df_dag, "A", "Y", adjust = "X") |> kable()
```

The graph chose the adjustment set; the estimator used that set. The point estimate recovers the true ACE of 1.5 within sampling noise.

::: {.callout-note}
The standard error here, `sd(psi) / sqrt(n)`, is the influence-function
standard error of the AIPW estimator. It is asymptotically valid when
the nuisance functions ($\mu$ and $\pi$) are regular parametric fits,
as they are here. With flexible machine-learning nuisances it is **not**
valid as written: you then need cross-fitting (sample splitting) so the
nuisance estimates are independent of the fold being evaluated, and the
variance should still be computed from the efficient influence function
(or by bootstrap). Treating a full-sample ML AIPW estimate with this
plug-in SE will generally understate uncertainty.
:::

## An ADMG: Front-Door Identification

Now suppose treatment and outcome share an unobserved common cause. A simple adjustment argument no longer works. The front-door ADMG has a mediator `M` that transmits the effect of `A` to `Y`, while `A` and `Y` are hidden-confounded:

```{r}
g_fd <- dagitty("dag {
  A -> M
  M -> Y
  A <-> Y
}")
coordinates(g_fd) <- list(
  x = c(A = 0, M = 1, Y = 2),
  y = c(A = 0, M = 0, Y = 0)
)
```

```{r}
#| label: fig-front-door
#| fig-cap: "Front-door ADMG: A and Y share an unobserved common cause (curved bidirected edge), but the effect propagates through the mediator M."
#| fig-width: 5
#| fig-height: 3
ggdag(g_fd) + theme_dag_blank()
```

The bidirected edge `A <-> Y` means that ordinary backdoor adjustment is not available. Indeed `dagitty` reports no adjustment set:

```{r}
adjustmentSets(g_fd, exposure = "A", outcome = "Y")
```

But the effect is still identified by the front-door criterion (Pearl, 1995). I call the Pearl-Shpitser ID algorithm in `causaleffect` to obtain the symbolic expression. The function wants the graph as an `igraph` object with bidirected edges encoded as two directed edges marked `description = "U"`:

```{r}
g_fd_ig <- graph_from_literal(A -+ M, M -+ Y, A -+ Y, Y -+ A)
g_fd_ig <- set_edge_attr(g_fd_ig, "description",
                         index = c(2, 4),  # A->Y and Y->A
                         value = "U")
expr_fd <- causal.effect(y = "Y", x = "A", G = g_fd_ig, simp = TRUE)
cat(expr_fd, "\n")
```

The expression $\sum_M P(M\mid A)\sum_{A'} P(Y\mid A', M) P(A')$ is precisely the front-door formula.

## Estimating the ADMG Effect with the Front-Door Formula

Simulate a front-door setting. The latent `U` is not included in the observed data; it is represented in the graph by `A <-> Y`.

```{r}
#| warning: false
#| message: false
#| cache: true
set.seed(2027)

n     <- 400
U     <- rnorm(n)
A_fd  <- as.numeric(runif(n) < logistic(-0.1 + 0.8 * U))
M     <- as.numeric(runif(n) < logistic(-0.4 + 1.2 * A_fd))
Y_fd  <- as.numeric(runif(n) < logistic(-1.0 + 0.9 * M + 0.7 * U))

df_fd <- data.frame(A = A_fd, M = M, Y = Y_fd)
head(df_fd)
```

The front-door point estimator is a direct evaluation of the identification expression:

$$
E[Y(a)] = \sum_m P(M=m\mid A=a)\sum_{a'} P(Y=1\mid A=a', M=m) P(A=a').
$$

For binary $A$, $M$, $Y$ we plug in sample proportions:

```{r}
front_door_ace <- function(data) {
  pA <- mean(data$A)
  p_marginal_A <- c("0" = 1 - pA, "1" = pA)

  # P(M | A): rows indexed by A, columns by M
  pM_given_A <- prop.table(table(data$A, data$M), margin = 1)
  # P(Y=1 | A, M)
  pY_given_AM <- with(data, tapply(Y, list(A, M), mean))

  # E[Y(a)] for a in {0, 1}
  ey_do <- function(a) {
    m_levels <- as.numeric(colnames(pM_given_A))
    sum(sapply(m_levels, function(m) {
      mar <- pM_given_A[as.character(a), as.character(m)]
      inner <- sum(sapply(c(0, 1), function(ap)
        pY_given_AM[as.character(ap), as.character(m)] *
          p_marginal_A[as.character(ap)]))
      mar * inner
    }))
  }
  ey_do(1) - ey_do(0)
}

# Bootstrap for SE
set.seed(99)
B   <- 500
boot_ace <- replicate(B, {
  idx <- sample.int(nrow(df_fd), replace = TRUE)
  front_door_ace(df_fd[idx, ])
})

est_fd <- front_door_ace(df_fd)
se_fd  <- sd(boot_ace)
ci_fd  <- est_fd + c(-1, 1) * qnorm(0.975) * se_fd

data.frame(estimand = "Front-door ACE",
           estimate = round(est_fd, 4),
           lower_95 = round(ci_fd[1], 4),
           upper_95 = round(ci_fd[2], 4),
           se       = round(se_fd, 4)) |> kable()
```

This is the practical advantage of separating identification from estimation. State the graph first. Then use the estimator that matches the identified functional.

## General ID Algorithm

The named routes above are useful because they are easy to estimate: backdoor adjustment and the front-door formula. ADMGs can be more general. The Pearl-Shpitser ID algorithm asks the broader question:

```text
Can P(Y | do(A)) be written using only the observed joint law P(V)?
```

The algorithm works recursively:

1. remove variables that are not ancestors of the outcome;
2. add interventions that are irrelevant after graph surgery;
3. split the remaining graph into bidirected districts;
4. identify each district as a factor of the observed law, if possible;
5. return a hedge witness when no such reduction is possible.

For the front-door graph, the general ID algorithm succeeds and returns the same front-door formula we obtained above. For more complex ADMGs it may yield novel functionals that no named estimator covers — this is where the plug-in approach is essential.

## When the Graph Says No

The **bow graph** has both a direct causal edge and unobserved confounding between the same two variables:

```{r}
g_bow <- dagitty("dag {
  A -> Y
  A <-> Y
}")
coordinates(g_bow) <- list(
  x = c(A = 0, Y = 1),
  y = c(A = 0, Y = 0)
)
```

```{r}
#| label: fig-bow
#| fig-cap: "Bow ADMG: a direct edge A → Y plus a bidirected A ↔ Y. The effect is not identified."
#| fig-width: 4
#| fig-height: 2.5
ggdag(g_bow) + theme_dag_blank()
```

The effect of `A` on `Y` is not identified in this ADMG. The observed association mixes the direct causal effect with the latent common cause.

```{r}
g_bow_ig <- graph_from_literal(A -+ Y, Y -+ A)
g_bow_ig <- set_edge_attr(g_bow_ig, "description",
                          index = c(1, 2), value = "U")
g_bow_ig <- add_edges(g_bow_ig, c("A", "Y"))   # add the direct A -> Y

result <- tryCatch(
  causal.effect(y = "Y", x = "A", G = g_bow_ig, simp = TRUE),
  error = function(e) paste("Not identifiable:", conditionMessage(e))
)
result
```

`causaleffect` raises an error message indicating the effect is not identifiable. This is a useful failure mode. If the graph does not justify the causal effect, the estimator should not manufacture one.

## Practical Workflow

For applied work, a useful workflow is:

1. write the graph before looking at estimates;
2. use bidirected edges for latent common causes;
3. ask `dagitty::adjustmentSets(graph, exposure, outcome)` — if non-empty, you have a backdoor route;
4. if no adjustment set exists, call `causaleffect::causal.effect(y, x, G)` — if it returns an expression, the effect is identified by a more general route; if it errors, the effect is not identified in this ADMG;
5. estimate using AIPW for backdoor effects, the front-door formula when the algorithm returns that pattern, or a plug-in evaluation of the symbolic expression for general ID functionals;
6. report the graph and the identification route along with the estimate.
