# Heterogeneous Treatment Effects with Machine Learning

```{r}
#| include: false
library(tidyverse)
library(grf)
library(policytree)
library(ggplot2)
```

```{r}
#| eval: false
library(tidyverse)
library(grf)
library(policytree)
library(ggplot2)
```

The [estimands chapter](causal-estimands.qmd) defined the conditional average
treatment effect

$$
\text{CATE}(x) = \mathbb{E}[Y(1) - Y(0) \mid X = x]
$$

as the effect of treatment for individuals with covariates $X = x$. There,
we computed CATE from known potential outcomes — a luxury only simulation
affords. In real data, $Y(0)$ and $Y(1)$ are never both observed, and CATE
must be *estimated* from $(X, D, Y)$ without strong functional-form
assumptions.

This chapter covers the modern toolkit for that problem:

- **Meta-learners** — recipes that turn any ML regressor into a CATE
  estimator: S-, T-, X-, R-, and DR-learners.
- **Causal forests** — Wager-Athey (2018) trees designed specifically for
  CATE estimation, with honest sample-splitting and valid pointwise
  confidence intervals.
- **Best linear projection (BLP) and CLAN** — Chernozhukov-Demirer-Duflo-
  Fernandez-Val (2018) tools for summarising and testing heterogeneity.
- **Policy trees** — Athey-Wager (2021) trees that recommend who to treat,
  given a CATE estimate.

## The data-generating process

Reuse the DGP from the estimands chapter, where the true CATE is
$\tau(x) = 1 + 2x$:

```{r}
#| label: dgp
#| cache: true

set.seed(42)
n <- 5000
p <- 5      # 5 covariates: only X1 is the effect modifier; X2..X5 are noise

X <- matrix(runif(n * p), n, p)
colnames(X) <- paste0("X", 1:p)
U <- rnorm(n)
# Treatment depends on a noise covariate (X2) and U — backdoor confounding
ps  <- plogis(-0.3 + 1.5 * X[, 2] + 0.5 * U)
D   <- rbinom(n, 1, ps)
tau <- 1 + 2 * X[, 1]          # CATE varies along X1 only
Y0  <- 0.5 * X[, 2] + 0.3 * U + rnorm(n)
Y1  <- Y0 + tau
Y   <- ifelse(D == 1, Y1, Y0)

cat(sprintf("n = %d, true ATE = %.3f\n", n, mean(tau)))
cat(sprintf("True CATE at X1 = 0.2: %.2f\n", 1 + 2 * 0.2))
cat(sprintf("True CATE at X1 = 0.8: %.2f\n", 1 + 2 * 0.8))
```

The treatment is confounded through $X_2$ and the unobserved $U$. Including
$X_2$ in the conditioning set restores identification; we will not condition
on $U$ (it is unobserved).

## Meta-learners

A meta-learner is a recipe that turns a *generic* regression estimator (any
black-box ML method) into a CATE estimator. The simplest meta-learners are:

### S-learner ("single")

Fit one regression of $Y$ on $(D, X)$, then predict the difference between
$D=1$ and $D=0$ at each $x$:

$$
\hat\tau^S(x) = \hat\mu(1, x) - \hat\mu(0, x).
$$

```{r}
#| label: s-learner
#| cache: true

# Use a flexible learner — random forest from grf::regression_forest
XD_train <- cbind(D = D, X)
fit_S <- regression_forest(XD_train, Y, num.trees = 500)

# Counterfactual predictions at D = 1 and D = 0 for every X
mu1 <- predict(fit_S, cbind(D = rep(1, n), X))$predictions
mu0 <- predict(fit_S, cbind(D = rep(0, n), X))$predictions
tau_S <- mu1 - mu0

cor_S <- cor(tau_S, tau)
cat(sprintf("S-learner CATE correlation with truth: %.3f\n", cor_S))
```

The S-learner is simple but biased toward zero when the treatment effect is
small relative to outcome variation, because the regression "ignores" the
$D$ variable if it has little predictive value.

### T-learner ("two")

Fit two separate regressions: $\hat\mu_1$ on treated units, $\hat\mu_0$ on
controls. Then $\hat\tau^T(x) = \hat\mu_1(x) - \hat\mu_0(x)$.

```{r}
#| label: t-learner
#| cache: true

fit_T1 <- regression_forest(X[D == 1, ], Y[D == 1], num.trees = 500)
fit_T0 <- regression_forest(X[D == 0, ], Y[D == 0], num.trees = 500)

mu1_T <- predict(fit_T1, X)$predictions
mu0_T <- predict(fit_T0, X)$predictions
tau_T <- mu1_T - mu0_T

cor_T <- cor(tau_T, tau)
cat(sprintf("T-learner CATE correlation with truth: %.3f\n", cor_T))
```

The T-learner avoids the bias-toward-zero of the S-learner but suffers
when the treated and control groups have very different propensity
distributions: extrapolation errors compound when subtracting the two
fitted functions.

### X-learner (Künzel et al. 2019)

The X-learner imputes the missing potential outcomes using the T-learner
fits, then regresses these "pseudo-outcomes" on $X$ separately within each
arm. Finally, the two CATE estimates are combined with weights from the
propensity score.

```{r}
#| label: x-learner
#| cache: true

# Step 1: T-learner fits (already done above)
# Step 2: Pseudo-outcomes
D1_pseudo <- Y[D == 1] - predict(fit_T0, X[D == 1, ])$predictions   # observed - imputed Y(0)
D0_pseudo <- predict(fit_T1, X[D == 0, ])$predictions - Y[D == 0]   # imputed Y(1) - observed

# Step 3: Regress pseudo-outcomes on X in each arm
fit_X1 <- regression_forest(X[D == 1, ], D1_pseudo, num.trees = 500)
fit_X0 <- regression_forest(X[D == 0, ], D0_pseudo, num.trees = 500)

tau_X1 <- predict(fit_X1, X)$predictions
tau_X0 <- predict(fit_X0, X)$predictions

# Step 4: Weighted combination — use propensity score as the weight
ps_fit <- regression_forest(X, D, num.trees = 500)
ps_hat <- predict(ps_fit, X)$predictions
ps_hat <- pmin(pmax(ps_hat, 0.02), 0.98)   # trim

tau_X <- ps_hat * tau_X0 + (1 - ps_hat) * tau_X1

cor_X <- cor(tau_X, tau)
cat(sprintf("X-learner CATE correlation with truth: %.3f\n", cor_X))
```

The X-learner is particularly effective when the propensity score is far
from 0.5 and one treatment arm is much smaller than the other — common in
observational studies.

### R-learner (Nie & Wager 2021)

The R-learner uses Robinson's (1988) partialled-out residual approach.
Define pseudo-outcomes that, in expectation, equal $\tau(X)$:

$$
\tilde Y_i = \frac{Y_i - \hat m(X_i)}{D_i - \hat e(X_i)}, \quad
\text{weights} = (D_i - \hat e(X_i))^2,
$$

where $\hat m(x) = \mathbb{E}[Y \mid X = x]$ and $\hat e(x) = \mathbb{E}[D \mid X = x]$
are cross-fitted nuisance estimates. Regressing $\tilde Y$ on $X$ with the
indicated weights yields the CATE.

```{r}
#| label: r-learner
#| cache: true

# Cross-fitting: split sample into 2 folds, fit nuisances on one, predict on the other
set.seed(7)
folds <- sample(rep(1:2, length.out = n))

m_hat <- numeric(n)
e_hat <- numeric(n)
for (k in 1:2) {
  train <- folds != k
  test  <- folds == k
  m_fit <- regression_forest(X[train, ], Y[train], num.trees = 500)
  e_fit <- regression_forest(X[train, ], D[train], num.trees = 500)
  m_hat[test] <- predict(m_fit, X[test, ])$predictions
  e_hat[test] <- predict(e_fit, X[test, ])$predictions
}
e_hat <- pmin(pmax(e_hat, 0.02), 0.98)

# R-learner pseudo-outcome and weights
pseudo_R <- (Y - m_hat) / (D - e_hat)
weights_R <- (D - e_hat)^2

# Regress pseudo-outcomes on X with weights — use a regression forest
fit_R <- regression_forest(X, pseudo_R, sample.weights = weights_R, num.trees = 500)
tau_R <- predict(fit_R, X)$predictions

cor_R <- cor(tau_R, tau)
cat(sprintf("R-learner CATE correlation with truth: %.3f\n", cor_R))
```

The R-learner is doubly robust in a Neyman-orthogonality sense and tends to
be the most efficient meta-learner in moderately well-overlapping data.

### DR-learner (Kennedy 2020)

The DR-learner constructs pseudo-outcomes via the AIPW formula and regresses
them on $X$:

$$
\tilde Y_i^{DR} = \hat\mu_1(X_i) - \hat\mu_0(X_i) +
\frac{D_i (Y_i - \hat\mu_1(X_i))}{\hat e(X_i)} -
\frac{(1 - D_i) (Y_i - \hat\mu_0(X_i))}{1 - \hat e(X_i)}.
$$

```{r}
#| label: dr-learner
#| cache: true

mu1_cf <- numeric(n); mu0_cf <- numeric(n); e_cf <- numeric(n)
for (k in 1:2) {
  train <- folds != k
  test  <- folds == k
  mu1_fit <- regression_forest(X[train & D == 1, , drop = FALSE], Y[train & D == 1],
                               num.trees = 500)
  mu0_fit <- regression_forest(X[train & D == 0, , drop = FALSE], Y[train & D == 0],
                               num.trees = 500)
  e_fit   <- regression_forest(X[train, ], D[train], num.trees = 500)
  mu1_cf[test] <- predict(mu1_fit, X[test, ])$predictions
  mu0_cf[test] <- predict(mu0_fit, X[test, ])$predictions
  e_cf[test]   <- predict(e_fit,   X[test, ])$predictions
}
e_cf <- pmin(pmax(e_cf, 0.02), 0.98)

pseudo_DR <- (mu1_cf - mu0_cf) +
             D * (Y - mu1_cf) / e_cf -
             (1 - D) * (Y - mu0_cf) / (1 - e_cf)

fit_DR <- regression_forest(X, pseudo_DR, num.trees = 500)
tau_DR <- predict(fit_DR, X)$predictions

cor_DR <- cor(tau_DR, tau)
cat(sprintf("DR-learner CATE correlation with truth: %.3f\n", cor_DR))
```

## Causal forests

Wager-Athey (2018) causal forests are tailored for CATE estimation. They
use a splitting criterion that maximises heterogeneity in treatment effects
(rather than outcome variance), and they use *honest* sample-splitting so
that each tree's leaf estimates are valid for inference at any test point.

```{r}
#| label: causal-forest
#| cache: true

cf <- causal_forest(X = X, Y = Y, W = D, num.trees = 2000)

tau_cf <- predict(cf)$predictions
cor_cf <- cor(tau_cf, tau)
cat(sprintf("Causal forest CATE correlation with truth: %.3f\n", cor_cf))

# Doubly robust ATE estimate
ate_cf <- average_treatment_effect(cf)
cat(sprintf("\nCausal forest AIPW ATE: %.3f  (SE = %.3f, true = %.3f)\n",
            ate_cf["estimate"], ate_cf["std.err"], mean(tau)))
```

The pointwise variance estimates `predict(cf, estimate.variance = TRUE)`
give asymptotically valid confidence intervals for each CATE prediction.

### Comparing all estimators

```{r}
#| label: compare-plot
#| cache: true
#| fig-cap: "Estimated CATE vs true τ(x) = 1 + 2*X1 for each estimator. Closer to the dashed line = better."
#| fig-width: 10
#| fig-height: 6

comparison <- tibble(
  X1     = X[, 1],
  true   = tau,
  S      = tau_S,
  Tlearn = tau_T,
  X      = tau_X,
  R      = tau_R,
  DR     = tau_DR,
  CF     = tau_cf
) |>
  pivot_longer(cols = c(S, Tlearn, X, R, DR, CF),
               names_to = "Estimator", values_to = "Predicted")

ggplot(comparison, aes(x = X1, y = Predicted)) +
  geom_point(alpha = 0.1, size = 0.4) +
  geom_smooth(method = "loess", se = FALSE, colour = "steelblue", linewidth = 1) +
  geom_abline(aes(intercept = 1, slope = 2), linetype = "dashed",
              colour = "firebrick") +
  facet_wrap(~ Estimator, ncol = 3) +
  labs(x = expression(X[1]),
       y = "Estimated CATE",
       caption = "Red dashed line = true τ(x) = 1 + 2*X1") +
  theme_minimal()
```

## Best linear projection (BLP)

Even when CATE is high-dimensional, we often want a simple summary: how
much does the treatment effect change with each covariate? The
**best linear projection** of CATE onto $X$ is the OLS projection

$$
\text{BLP}(X) = \arg\min_{\beta} \mathbb{E}\left[ (\tau(X) - X'\beta)^2 \right],
$$

which `grf` estimates with valid standard errors:

```{r}
#| label: blp
#| cache: true

blp <- best_linear_projection(cf, A = X)
print(blp)
```

The BLP coefficients on $X_1$ should be close to 2 (the true slope of the
CATE) and significant; coefficients on $X_2$ through $X_5$ should be close
to 0 and insignificant. This is a clean test for heterogeneity along
specific covariates.

## Variable importance

`grf` provides a variable-importance measure based on how often each
covariate is used for splits in the causal-effect splits (vs the
outcome-prediction splits):

```{r}
#| label: var-imp
#| cache: true

vi <- variable_importance(cf)
vi_df <- tibble(variable = paste0("X", 1:p), importance = as.numeric(vi)) |>
  arrange(desc(importance))
print(vi_df)
```

$X_1$ should top the list. In real applications, this is a useful guide
to which covariates carry heterogeneity.

## Policy trees: who should we treat?

A CATE estimate tells us the effect of treating each individual. But the
**policy question** is often different: given a budget or a cost-benefit
trade-off, *which* individuals should we treat? Athey-Wager (2021) propose
**policy trees** — shallow decision trees that maximise welfare given a
CATE estimate and a treatment cost.

```{r}
#| label: policy-tree
#| cache: true

# Compute doubly-robust scores (used for policy learning)
dr_scores <- double_robust_scores(cf)
cost <- 0.5     # cost of treating one unit, in outcome units

# Net benefit of treatment vs control
gain <- cbind(0, dr_scores - cost)   # column 1 = no treat (gain 0), col 2 = treat (gain τ - cost)

# Fit a depth-2 policy tree (shallow for interpretability)
tree <- policy_tree(X, gain, depth = 2)
print(tree)

# Recommended action for each unit
actions <- predict(tree, X)
cat(sprintf("\nTreatment rate under policy tree: %.1f%%\n",
            mean(actions == 2) * 100))
cat(sprintf("Estimated welfare gain over treat-everyone: %.3f units per person\n",
            mean((actions == 2) * (dr_scores - cost) - (dr_scores - cost))))
```

The policy tree provides an interpretable assignment rule. Real-world
policies are often required to be interpretable (e.g. fair-housing laws,
medical decision rules), and shallow trees are the natural compromise
between flexibility and explainability.

For an extended walkthrough of policy trees with IPW and AIPW losses,
see the [companion blog chapter on `policytree`](https://xiangao.github.io/blog_book/policytree.html).
For a comparison of CATE estimators across software ecosystems
(including Stata 19's new `cate` command), see the
[Stata CATE blog chapter](https://xiangao.github.io/blog_book/stata-cate.html).

## CLAN: testing for heterogeneity

A formal test of heterogeneity is the **CLAN** (Classification Analysis)
proposed by Chernozhukov-Demirer-Duflo-Fernandez-Val (2018). The idea: bin
the population by predicted CATE, then estimate the average treatment
effect in each bin. If the bin-specific ATEs are statistically different,
there is heterogeneity.

```{r}
#| label: clan
#| cache: true

# Sort by predicted CATE, split into 5 quintiles
nq <- 5
quintile <- cut(tau_cf, breaks = quantile(tau_cf, probs = seq(0, 1, 1/nq)),
                include.lowest = TRUE, labels = FALSE)

# Compute AIPW ATE within each quintile
get_ate <- function(idx) {
  ate <- average_treatment_effect(cf, subset = idx)
  c(est = ate["estimate"], se = ate["std.err"])
}

clan <- t(sapply(1:nq, function(q) get_ate(quintile == q)))
clan_df <- as_tibble(clan) |>
  mutate(quintile = 1:nq,
         lo = est.estimate - 1.96 * se.std.err,
         hi = est.estimate + 1.96 * se.std.err)

knitr::kable(clan_df[, c("quintile", "est.estimate", "se.std.err", "lo", "hi")],
             col.names = c("Quintile", "ATE", "SE", "95% LB", "95% UB"),
             digits = 3,
             caption = "Quintile-specific ATEs from CLAN: heterogeneity is evident if quintile estimates differ.")
```

The bottom and top quintiles of predicted CATE should give significantly
different ATEs — a clean test for whether the heterogeneity is real
(rather than estimation noise).

## Causal forests with panel data

When the data are a panel — repeated observations on the same units —
and the unobserved unit effects are correlated with treatment or
covariates, the cross-sectional CATE methods above can be biased. Two
practical solutions:

1. **Demean by fixed effects** before running the causal forest.
2. **Pass a `cluster` argument** to `causal_forest` so that observations
   from the same unit are kept in the same training fold during honest
   sample-splitting.

```{r}
#| label: panel-dgp
#| cache: true

set.seed(2024)
n_firms <- 200
n_t     <- 5
n_total <- n_firms * n_t

# Panel data with a unit-specific effect correlated with V1 (the modifier)
firm_id  <- rep(1:n_firms, each = n_t)
unit_fe  <- rnorm(n_firms, sd = 1.5)[firm_id]
V1_firm  <- rnorm(n_firms, sd = 1.0)[firm_id]   # firm-level covariate
V1       <- V1_firm + rnorm(n_total, sd = 0.3)  # within-firm variation
W        <- rbinom(n_total, 1, plogis(0.5 + 0.3 * V1_firm))
# True CATE varies linearly with V1
tau_panel <- 0.5 + 1.0 * V1
Y_panel  <- unit_fe + V1 + W + tau_panel * (W - mean(W)) + rnorm(n_total)

df_panel <- tibble(firm = firm_id, V1 = V1, W = W, Y = Y_panel)
glimpse(df_panel)
```

A naive causal forest on the panel data ignores the unit effects:

```{r}
#| label: cf-naive
#| cache: true

X_panel <- as.matrix(df_panel[, "V1", drop = FALSE])

cf_naive <- causal_forest(X = X_panel, Y = df_panel$Y, W = df_panel$W,
                          num.trees = 1000)
ate_naive <- average_treatment_effect(cf_naive)

cat(sprintf("Naive panel CF ATE: %.3f  (SE = %.3f)\n",
            ate_naive["estimate"], ate_naive["std.err"]))
cat(sprintf("True ATE: %.3f\n", mean(tau_panel)))
```

Adding `clusters = firm`: `grf` keeps each firm's observations in the
same training/honest sample, eliminating the leakage that would
otherwise bias CATE estimates:

```{r}
#| label: cf-clustered
#| cache: true

cf_clust <- causal_forest(X = X_panel, Y = df_panel$Y, W = df_panel$W,
                          clusters = df_panel$firm,
                          num.trees = 1000)
ate_clust <- average_treatment_effect(cf_clust)

cat(sprintf("Clustered panel CF ATE: %.3f  (SE = %.3f)\n",
            ate_clust["estimate"], ate_clust["std.err"]))
```

For the strongest panel-data identification, demean both $Y$ and $W$ by
firm before fitting (within transformation):

```{r}
#| label: cf-demean
#| cache: true

df_dm <- df_panel |>
  group_by(firm) |>
  mutate(Y_dm = Y - mean(Y),
         W_dm = W - mean(W),
         V1_dm = V1 - mean(V1)) |>
  ungroup()

cf_fe <- causal_forest(X = as.matrix(df_dm[, "V1_dm"]),
                       Y = df_dm$Y_dm,
                       W = df_dm$W_dm,
                       clusters = df_dm$firm,
                       num.trees = 1000)
ate_fe <- average_treatment_effect(cf_fe)
cat(sprintf("Within-transform CF ATE: %.3f  (SE = %.3f)\n",
            ate_fe["estimate"], ate_fe["std.err"]))
```

The clustered + within-transform causal forest is the cleanest analog of
a panel fixed-effects regression for heterogeneous treatment effects.
See the [companion blog chapter on causal forests in panel data](https://xiangao.github.io/blog_book/causal-forest-panel.html)
for an extended simulation study comparing the panel CF to fixed-effect
regression and random-effect regression under linear and nonlinear DGPs.

## Summary

- **Meta-learners** (S, T, X, R, DR) turn any regression algorithm into a
  CATE estimator. R-learner and DR-learner exploit Neyman orthogonality and
  cross-fitting for valid inference; X-learner helps with imbalanced
  treatment arms.
- **Causal forests** (`grf`) provide a purpose-built CATE estimator with
  honest sample splitting, pointwise confidence intervals, and convenient
  diagnostics (BLP, variable importance, CLAN).
- **Best Linear Projection (BLP)** gives a simple regression-style summary
  of CATE — coefficients with valid standard errors on how the effect
  varies with each covariate.
- **CLAN** (quintile-binned ATEs) provides a formal test for heterogeneity
  and a clean way to communicate it.
- **Policy trees** (`policytree`) translate a CATE into an interpretable
  treatment-assignment rule, balancing welfare against cost.
- For applied work: estimate CATE with at least *two* methods (e.g.
  causal forest + R-learner), report BLP coefficients and quintile-binned
  ATEs, and use a shallow policy tree when treatment-assignment
  recommendations are needed.

## References

- Wager, S., & Athey, S. (2018). Estimation and inference of heterogeneous treatment effects using random forests. *JASA*, 113(523), 1228–1242.
- Künzel, S. R., Sekhon, J. S., Bickel, P. J., & Yu, B. (2019). Metalearners for estimating heterogeneous treatment effects using machine learning. *PNAS*, 116(10), 4156–4165.
- Nie, X., & Wager, S. (2021). Quasi-oracle estimation of heterogeneous treatment effects. *Biometrika*, 108(2), 299–319.
- Kennedy, E. H. (2020). Towards optimal doubly robust estimation of heterogeneous causal effects. *arXiv:2004.14497*.
- Athey, S., & Wager, S. (2021). Policy learning with observational data. *Econometrica*, 89(1), 133–161.
- Chernozhukov, V., Demirer, M., Duflo, E., & Fernandez-Val, I. (2018). Generic machine learning inference on heterogeneous treatment effects in randomized experiments. *NBER Working Paper 24678*.
