# Heterogeneous Treatment Effects with Machine Learning

```{r}
#| include: false
library(tidyverse)
library(grf)
library(policytree)
library(ggplot2)
```

```{r}
#| eval: false
library(tidyverse)
library(grf)
library(policytree)
library(ggplot2)
```

The [estimands chapter](causal-estimands.qmd) defined the conditional average
treatment effect

$$
\text{CATE}(x) = \mathbb{E}[Y(1) - Y(0) \mid X = x]
$$

as the effect of treatment for units with covariates $X=x$. In the
simulation we could compute it from known potential outcomes. In real data
we do not observe both $Y(0)$ and $Y(1)$, so CATE has to be estimated from
$(X,D,Y)$.

I use four kinds of tools:

- **Meta-learners**: S-, T-, X-, R-, and DR-learners.
- **Causal forests**: random forests designed for CATE estimation.
- **BLP and CLAN**: simple summaries and tests of heterogeneity.
- **Policy trees**: rules for deciding who should be treated.

## The data-generating process

I use the same CATE function as the estimands chapter,
$\tau(x)=1+2x_1$ (the rest of the DGP differs: five covariates and
confounding through the observed $X_2$).

```{r}
#| label: dgp
#| cache: true

set.seed(42)
n <- 5000
p <- 5      # 5 covariates: only X1 is the effect modifier; X2..X5 are noise

X <- matrix(runif(n * p), n, p)
colnames(X) <- paste0("X", 1:p)
# Treatment depends on the OBSERVED covariate X2 (a backdoor confounder we
# adjust for), so unconfoundedness given X holds and the estimators are
# consistent for the true ATE/CATE.
ps  <- plogis(-0.3 + 1.5 * X[, 2])
D   <- rbinom(n, 1, ps)
tau <- 1 + 2 * X[, 1]          # CATE varies along X1 only
Y0  <- 0.5 * X[, 2] + rnorm(n)
Y1  <- Y0 + tau
Y   <- ifelse(D == 1, Y1, Y0)

cat(sprintf("n = %d, true ATE = %.3f\n", n, mean(tau)))
cat(sprintf("True CATE at X1 = 0.2: %.2f\n", 1 + 2 * 0.2))
cat(sprintf("True CATE at X1 = 0.8: %.2f\n", 1 + 2 * 0.8))
```

Treatment is related to the observed covariate $X_2$, which also affects the
outcome — a backdoor confounder. Because $X_2$ is observed and included in the
conditioning set, unconfoundedness holds and the meta-learners and causal
forest below are consistent for the true CATE/ATE.

## Meta-learners

A meta-learner is a recipe that turns a regression method into a CATE
estimator. The regression method can be a random forest, boosting, neural
network, or a simple linear model.

### S-learner ("single")

Fit one regression of $Y$ on $(D, X)$, then predict the difference between
$D=1$ and $D=0$ at each $x$:

$$
\hat\tau^S(x) = \hat\mu(1, x) - \hat\mu(0, x).
$$

```{r}
#| label: s-learner
#| cache: true

# Use a flexible learner — random forest from grf::regression_forest
XD_train <- cbind(D = D, X)
fit_S <- regression_forest(XD_train, Y, num.trees = 500)

# Counterfactual predictions at D = 1 and D = 0 for every X
mu1 <- predict(fit_S, cbind(D = rep(1, n), X))$predictions
mu0 <- predict(fit_S, cbind(D = rep(0, n), X))$predictions
tau_S <- mu1 - mu0

cor_S <- cor(tau_S, tau)
cat(sprintf("S-learner CATE correlation with truth: %.3f\n", cor_S))
```

The S-learner is simple. Its weakness is that the model may treat $D$ as a
minor predictor of $Y$ and shrink the treatment effect toward zero.

### T-learner ("two")

Fit two separate regressions: $\hat\mu_1$ on treated units, $\hat\mu_0$ on
controls. Then $\hat\tau^T(x) = \hat\mu_1(x) - \hat\mu_0(x)$.

```{r}
#| label: t-learner
#| cache: true

fit_T1 <- regression_forest(X[D == 1, ], Y[D == 1], num.trees = 500)
fit_T0 <- regression_forest(X[D == 0, ], Y[D == 0], num.trees = 500)

mu1_T <- predict(fit_T1, X)$predictions
mu0_T <- predict(fit_T0, X)$predictions
tau_T <- mu1_T - mu0_T

cor_T <- cor(tau_T, tau)
cat(sprintf("T-learner CATE correlation with truth: %.3f\n", cor_T))
```

The T-learner gives treatment and control separate outcome models. That can
help with heterogeneity, but it can also extrapolate badly when the treated
and control covariate distributions do not overlap.

### X-learner [@kunzel-2019-metalearners]

The X-learner starts from the T-learner. It imputes missing potential
outcomes, creates pseudo-treatment effects, and then smooths those
pseudo-effects as functions of $X$.

```{r}
#| label: x-learner
#| cache: true

# Step 1: T-learner fits (already done above)
# Step 2: Pseudo-outcomes
D1_pseudo <- Y[D == 1] - predict(fit_T0, X[D == 1, ])$predictions   # observed - imputed Y(0)
D0_pseudo <- predict(fit_T1, X[D == 0, ])$predictions - Y[D == 0]   # imputed Y(1) - observed

# Step 3: Regress pseudo-outcomes on X in each arm
fit_X1 <- regression_forest(X[D == 1, ], D1_pseudo, num.trees = 500)
fit_X0 <- regression_forest(X[D == 0, ], D0_pseudo, num.trees = 500)

tau_X1 <- predict(fit_X1, X)$predictions
tau_X0 <- predict(fit_X0, X)$predictions

# Step 4: Weighted combination — use propensity score as the weight
ps_fit <- regression_forest(X, D, num.trees = 500)
ps_hat <- predict(ps_fit, X)$predictions
ps_hat <- pmin(pmax(ps_hat, 0.02), 0.98)   # trim

tau_X <- ps_hat * tau_X0 + (1 - ps_hat) * tau_X1

cor_X <- cor(tau_X, tau)
cat(sprintf("X-learner CATE correlation with truth: %.3f\n", cor_X))
```

The X-learner is useful when one treatment arm is much smaller than the
other.

### R-learner [@nie-wager-2021]

The R-learner partials out the main effects of $X$ from both $Y$ and $D$.
It then estimates how the residualized treatment predicts the residualized
outcome:

$$
\tilde Y_i = \frac{Y_i - \hat m(X_i)}{D_i - \hat e(X_i)}, \quad
\text{weights} = (D_i - \hat e(X_i))^2,
$$

where $\hat m(x) = \mathbb{E}[Y \mid X = x]$ and $\hat e(x) = \mathbb{E}[D \mid X = x]$
are cross-fitted nuisance estimates. Regressing $\tilde Y$ on $X$ with these
weights gives the CATE estimate.

```{r}
#| label: r-learner
#| cache: true

# Cross-fitting: split sample into 2 folds, fit nuisances on one, predict on the other
set.seed(7)
folds <- sample(rep(1:2, length.out = n))

m_hat <- numeric(n)
e_hat <- numeric(n)
for (k in 1:2) {
  train <- folds != k
  test  <- folds == k
  m_fit <- regression_forest(X[train, ], Y[train], num.trees = 500)
  e_fit <- regression_forest(X[train, ], D[train], num.trees = 500)
  m_hat[test] <- predict(m_fit, X[test, ])$predictions
  e_hat[test] <- predict(e_fit, X[test, ])$predictions
}
e_hat <- pmin(pmax(e_hat, 0.02), 0.98)

# R-learner pseudo-outcome and weights
pseudo_R <- (Y - m_hat) / (D - e_hat)
weights_R <- (D - e_hat)^2

# Regress pseudo-outcomes on X with weights — use a regression forest
fit_R <- regression_forest(X, pseudo_R, sample.weights = weights_R, num.trees = 500)
tau_R <- predict(fit_R, X)$predictions

cor_R <- cor(tau_R, tau)
cat(sprintf("R-learner CATE correlation with truth: %.3f\n", cor_R))
```

The R-learner is often stable when overlap is reasonable because errors in
the nuisance models have only second-order effects on the target.

### DR-learner [@kennedy-2020-optimal-dr]

The DR-learner uses an AIPW-style pseudo-outcome and then regresses it on
$X$:

$$
\tilde Y_i^{DR} = \hat\mu_1(X_i) - \hat\mu_0(X_i) +
\frac{D_i (Y_i - \hat\mu_1(X_i))}{\hat e(X_i)} -
\frac{(1 - D_i) (Y_i - \hat\mu_0(X_i))}{1 - \hat e(X_i)}.
$$

```{r}
#| label: dr-learner
#| cache: true

mu1_cf <- numeric(n); mu0_cf <- numeric(n); e_cf <- numeric(n)
for (k in 1:2) {
  train <- folds != k
  test  <- folds == k
  mu1_fit <- regression_forest(X[train & D == 1, , drop = FALSE], Y[train & D == 1],
                               num.trees = 500)
  mu0_fit <- regression_forest(X[train & D == 0, , drop = FALSE], Y[train & D == 0],
                               num.trees = 500)
  e_fit   <- regression_forest(X[train, ], D[train], num.trees = 500)
  mu1_cf[test] <- predict(mu1_fit, X[test, ])$predictions
  mu0_cf[test] <- predict(mu0_fit, X[test, ])$predictions
  e_cf[test]   <- predict(e_fit,   X[test, ])$predictions
}
e_cf <- pmin(pmax(e_cf, 0.02), 0.98)

pseudo_DR <- (mu1_cf - mu0_cf) +
             D * (Y - mu1_cf) / e_cf -
             (1 - D) * (Y - mu0_cf) / (1 - e_cf)

fit_DR <- regression_forest(X, pseudo_DR, num.trees = 500)
tau_DR <- predict(fit_DR, X)$predictions

cor_DR <- cor(tau_DR, tau)
cat(sprintf("DR-learner CATE correlation with truth: %.3f\n", cor_DR))
```

## Causal forests

Causal forests are random forests built for treatment-effect heterogeneity.
They split the data to find treatment-effect differences, not just outcome
differences. The `grf` implementation also uses honest sample splitting for
inference.

```{r}
#| label: causal-forest
#| cache: true

cf <- causal_forest(X = X, Y = Y, W = D, num.trees = 2000)

tau_cf <- predict(cf)$predictions
cor_cf <- cor(tau_cf, tau)
cat(sprintf("Causal forest CATE correlation with truth: %.3f\n", cor_cf))

# Doubly robust ATE estimate
ate_cf <- average_treatment_effect(cf)
cat(sprintf("\nCausal forest AIPW ATE: %.3f  (SE = %.3f, true = %.3f)\n",
            ate_cf["estimate"], ate_cf["std.err"], mean(tau)))
```

Use `predict(cf, estimate.variance = TRUE)` when pointwise confidence
intervals are needed.

### Comparing all estimators

```{r}
#| label: compare-plot
#| cache: true
#| fig-cap: "Estimated CATE vs true τ(x) = 1 + 2*X1 for each estimator. Closer to the dashed line = better."
#| fig-width: 10
#| fig-height: 6

comparison <- tibble(
  X1     = X[, 1],
  true   = tau,
  S      = tau_S,
  Tlearn = tau_T,
  X      = tau_X,
  R      = tau_R,
  DR     = tau_DR,
  CF     = tau_cf
) |>
  pivot_longer(cols = c(S, Tlearn, X, R, DR, CF),
               names_to = "Estimator", values_to = "Predicted")

ggplot(comparison, aes(x = X1, y = Predicted)) +
  geom_point(alpha = 0.1, size = 0.4) +
  geom_smooth(method = "loess", se = FALSE, colour = "steelblue", linewidth = 1) +
  geom_abline(aes(intercept = 1, slope = 2), linetype = "dashed",
              colour = "firebrick") +
  facet_wrap(~ Estimator, ncol = 3) +
  labs(x = expression(X[1]),
       y = "Estimated CATE",
       caption = "Red dashed line = true τ(x) = 1 + 2*X1") +
  theme_minimal()
```

## Best linear projection (BLP)

Even if CATE is nonlinear, we often want a regression-style summary: which
covariates are associated with larger effects? The best linear projection
of CATE onto $X$ is

$$
\text{BLP}(X) = \arg\min_{\beta_0, \beta} \mathbb{E}\left[ (\tau(X) - \beta_0 - X'\beta)^2 \right],
$$

which `grf` estimates with valid standard errors:

```{r}
#| label: blp
#| cache: true

blp <- best_linear_projection(cf, A = X)
print(blp)
```

Here the coefficient on $X_1$ should be close to 2, and the coefficients on
the other variables should be close to 0.

## Variable importance

`grf` also reports variable importance. It is a rough diagnostic for which
variables the forest uses to find heterogeneity.

```{r}
#| label: var-imp
#| cache: true

vi <- variable_importance(cf)
vi_df <- tibble(variable = paste0("X", 1:p), importance = as.numeric(vi)) |>
  arrange(desc(importance))
print(vi_df)
```

$X_1$ should be the most important variable in this simulation.

## Policy trees: who should we treat?

A CATE estimate is not yet a policy. If treatment has a cost, the policy
question is who should be treated. A policy tree turns estimated treatment
effects into a simple treatment rule.

```{r}
#| label: policy-tree
#| cache: true

# double_robust_scores() returns an n x 2 matrix of AIPW scores,
# one column per action: Gamma_i(control), Gamma_i(treated)
Gamma <- double_robust_scores(cf)
cost <- 2     # cost of treating one unit, in outcome units

# Net value of each action: control stays as is, treating pays the cost.
# With tau(x) = 1 + 2*x1 in [1, 3], the optimal rule is: treat iff x1 > 0.5.
Gamma[, "treated"] <- Gamma[, "treated"] - cost

# Fit a depth-2 policy tree (shallow for interpretability)
tree <- policy_tree(X, Gamma, depth = 2)
print(tree)

# Recommended action for each unit (1 = control, 2 = treated)
actions <- predict(tree, X)
cat(sprintf("\nTreatment rate under policy tree: %.1f%% (optimal: %.1f%%)\n",
            mean(actions == 2) * 100, mean(X[, 1] > 0.5) * 100))

# Value of the learned rule vs treating everyone, both evaluated with
# the same doubly-robust scores
welfare_tree <- mean(ifelse(actions == 2, Gamma[, "treated"], Gamma[, "control"]))
welfare_all  <- mean(Gamma[, "treated"])
cat(sprintf("Estimated welfare: policy tree %.3f vs treat-everyone %.3f (gain %.3f)\n",
            welfare_tree, welfare_all, welfare_tree - welfare_all))
```

The point of the shallow tree is interpretability. It gives up some
flexibility in exchange for a rule that can be inspected. Here the tree
splits on $X_1$ — the true effect modifier — close to the optimal
threshold of $0.5$, and the learned rule beats treat-everyone because
units with $\tau(x) < 2$ are not worth the cost.

For an extended walkthrough of policy trees with IPW and AIPW losses,
see the [companion blog chapter on `policytree`](https://xiangao.github.io/blog_book/policytree.html).
For a comparison of CATE estimators across software ecosystems
(including Stata 19's new `cate` command), see the
[Stata CATE blog chapter](https://xiangao.github.io/blog_book/stata-cate.html).

## GATES: sorted group ATEs

GATES (group average treatment effects) is a simple way to report
heterogeneity [@chernozhukov-2018-generic-ml]. Sort observations by
predicted CATE, split them into bins, and estimate the ATE in each bin. If
the bin ATEs differ, the heterogeneity is not just a graph. (The same
paper's CLAN — classification analysis — is the natural companion: compare
average *covariates* between the most- and least-affected bins; here that
comparison would show high $X_1$ in the top quintile and low $X_1$ in the
bottom one.)

```{r}
#| label: clan
#| cache: true

# Sort by predicted CATE, split into 5 quintiles
nq <- 5
quintile <- cut(tau_cf, breaks = quantile(tau_cf, probs = seq(0, 1, 1/nq)),
                include.lowest = TRUE, labels = FALSE)

# Compute AIPW ATE within each quintile
get_ate <- function(idx) {
  ate <- average_treatment_effect(cf, subset = idx)
  c(est = ate["estimate"], se = ate["std.err"])
}

clan <- t(sapply(1:nq, function(q) get_ate(quintile == q)))
clan_df <- as_tibble(clan) |>
  mutate(quintile = 1:nq,
         lo = est.estimate - 1.96 * se.std.err,
         hi = est.estimate + 1.96 * se.std.err)

knitr::kable(clan_df[, c("quintile", "est.estimate", "se.std.err", "lo", "hi")],
             col.names = c("Quintile", "ATE", "SE", "95% LB", "95% UB"),
             digits = 3,
             caption = "GATES: quintile-specific ATEs sorted by predicted CATE. Heterogeneity is evident if quintile estimates differ.")
```

The bottom and top quintiles should have different estimated ATEs in this
simulation. One caveat: the quintiles are formed from the same forest's
(out-of-bag) predictions used to estimate the group ATEs, which mitigates
but does not fully remove the bias from grouping and estimating on the same
data; for formal inference use a held-out fold or
`grf::rank_average_treatment_effect()`.

## Causal forests with panel data

With panel data, unit effects can be correlated with treatment and
covariates. A cross-sectional causal forest can then be biased. Two
practical fixes are:

1. **Demean by fixed effects** before running the causal forest.
2. **Pass a `clusters` argument** to `causal_forest` so that observations
   from the same unit are kept in the same training fold during honest
   sample-splitting.

```{r}
#| label: panel-dgp
#| cache: true

set.seed(2024)
n_firms <- 200
n_t     <- 5
n_total <- n_firms * n_t

# Panel data with a unit effect that drives BOTH treatment and the outcome
# (classic fixed-effect confounding; unit_fe is unobserved to the forest)
firm_id  <- rep(1:n_firms, each = n_t)
unit_fe  <- rnorm(n_firms, sd = 1.5)[firm_id]
V1_firm  <- rnorm(n_firms, sd = 1.0)[firm_id]   # firm-level covariate
V1       <- V1_firm + rnorm(n_total, sd = 0.3)  # within-firm variation
W        <- rbinom(n_total, 1, plogis(0.3 * V1_firm + 0.5 * unit_fe))
# True CATE varies linearly with V1
tau_panel <- 0.5 + 1.0 * V1
Y_panel  <- unit_fe + V1 + tau_panel * W + rnorm(n_total)

df_panel <- tibble(firm = firm_id, V1 = V1, W = W, Y = Y_panel)
glimpse(df_panel)
```

A naive causal forest ignores the firm effects. Because `unit_fe` raises
both the treatment probability and the outcome, and is not in $X$, the
naive estimate is badly biased upward:

```{r}
#| label: cf-naive
#| cache: true

X_panel <- as.matrix(df_panel[, "V1", drop = FALSE])

cf_naive <- causal_forest(X = X_panel, Y = df_panel$Y, W = df_panel$W,
                          num.trees = 1000)
ate_naive <- average_treatment_effect(cf_naive)

cat(sprintf("Naive panel CF ATE: %.3f  (SE = %.3f)\n",
            ate_naive["estimate"], ate_naive["std.err"]))
cat(sprintf("True ATE: %.3f\n", mean(tau_panel)))
```

Adding `clusters = firm` keeps a firm's observations together when `grf`
does the honest sample split. Note what this does and does not do: it makes
the honest splitting and the standard errors respect the panel structure,
but it does *not* remove the unobserved-unit-effect confounding — the
estimate stays close to the naive one:

```{r}
#| label: cf-clustered
#| cache: true

cf_clust <- causal_forest(X = X_panel, Y = df_panel$Y, W = df_panel$W,
                          clusters = df_panel$firm,
                          num.trees = 1000)
ate_clust <- average_treatment_effect(cf_clust)

cat(sprintf("Clustered panel CF ATE: %.3f  (SE = %.3f)\n",
            ate_clust["estimate"], ate_clust["std.err"]))
```

What removes the confounding is the fixed-effect style adjustment: demean
$Y$, $W$, and $V_1$ by firm before fitting, which sweeps out `unit_fe`:

```{r}
#| label: cf-demean
#| cache: true

df_dm <- df_panel |>
  group_by(firm) |>
  mutate(Y_dm = Y - mean(Y),
         W_dm = W - mean(W),
         V1_dm = V1 - mean(V1)) |>
  ungroup()

cf_fe <- causal_forest(X = as.matrix(df_dm[, "V1_dm"]),
                       Y = df_dm$Y_dm,
                       W = df_dm$W_dm,
                       clusters = df_dm$firm,
                       num.trees = 1000)
ate_fe <- average_treatment_effect(cf_fe)
cat(sprintf("Within-transform CF ATE: %.3f  (SE = %.3f)\n",
            ate_fe["estimate"], ate_fe["std.err"]))
```

The clustered, within-transformed causal forest is the closest analogue to a
fixed-effect regression with heterogeneous effects.
See the [companion blog chapter on causal forests in panel data](https://xiangao.github.io/blog_book/causal-forest-panel.html)
for an extended simulation study comparing the panel CF to fixed-effect
regression and random-effect regression under linear and nonlinear DGPs.

## Summary

- Meta-learners turn outcome-regression tools into CATE estimators.
- Causal forests are the main `grf` estimator for heterogeneous effects.
- BLP and GATES are useful because they summarize heterogeneity in a way
  that looks like familiar regression output.
- Policy trees translate CATE estimates into treatment rules.
- In applied work, compare at least two CATE estimators and always report
  simple heterogeneity summaries, not only a CATE scatterplot.
