# IV & Regression Discontinuity

```{r}
#| include: false
library(ggplot2)
library(ggdag)
library(sem)
library(MASS)
library(rdrobust)
library(causaldata)
library(fixest)
library(tidyverse)
```

If we have an unobserved confounder, unconfoundedness fails. Even when we have random experiment, we could have noncompliance. In that case, subjects are randomly assigned to treatment or control, but some of them do not comply with the assignment. In other words, there is selection bias in the sense that the treated group is self-selected. This selection could be correlated with the outcome.

## Instrumental Variables

### Uncontrolled Confounder

```{r dag2, warning=FALSE, cache=TRUE, message=FALSE, echo=FALSE}
library(ggplot2)
library(ggdag)
g <- dagify(
  y ~ w,
  w ~ c,
  y ~ c,
  exposure = "w",
  outcome = "y",
  coords = list(x = c(w = 2, c = 3, y = 4),
                y = c(w = 2, c = 3, y = 2))
)
ggdag(g)
```

### Why Does IV Work?

```{r dag3, warning=FALSE, cache=TRUE, message=FALSE, echo=FALSE}
g <- dagify(
  y ~ w,
  w ~ c,
  y ~ c,
  w ~ z,
  exposure = "w",
  outcome = "y",
  coords = list(x = c(w = 2, c = 3, y = 4, z=0),
                y = c(w = 2, c = 3, y = 2, z=2))
)
ggdag(g)
```

### IV Under Potential Outcomes

-   W has two potential outcomes W(1) and W(0), as a function of Z.
-   Y has two potential outcomes Y(1) and Y(0), as a function of W.

Intention to Treat (ITT) is the average treatment effect of the treatment assignment Z. It is the difference between the average potential outcome under treatment and the average potential outcome under control. It is the average treatment effect of the treatment assignment Z, regardless of whether the subject complies with the assignment.

There are four possible groups of subjects in the case of binary treatment and binary instrument. For example treatment assignment Z and treated status W could be:

-   Z=0, W=0: never-taker
-   Z=0, W=1: complier
-   Z=1, W=0: defier
-   Z=1, W=1: always-taker

|        | W(1)=0 | W(1)=1 |
|:-------|:-------|:-------|
| W(0)=0 | N      | C      |
| W(0)=1 | D      | A      |

: compliance groups

These four groups (stratum) have different causal mechanisms. For example, the compliers are the only group that is affected by the treatment assignment Z in the same direction. The defiers are the only group that is affected by the treatment assignment Z in the opposite direction. The always-takers and never-takers are not affected by the treatment assignment Z.

Unfortunately we cannot identify the compliance group by looking at the data.

| Z   | W   | G    |
|:----|:----|:-----|
| 0   | 0   | C, N |
| 0   | 1   | A, D |
| 1   | 0   | N, D |
| 1   | 1   | C, A |

### Assumptions

-   Exclusion restriction: $Y(w, 1) = Y(w,0)$
-   Exogeneity of instrument: $Z \perp [W(0),W(1), Y(0), Y(1)]$
-   Monotonicity (no defiers): $W(0) \leq W(1)$

### LATE Identification

Since we excluded defiers, and the other two groups (always-takers and never-takers) are not affected by the treatment assignment Z. The only effect we can identify is the treatment effect on the compliers. This is called the Local Average Treatment Effect (LATE).

$$ \small
\begin{aligned}
& E[Y(Z=1) - Y(Z=0) | G = C] \\
&= \frac{E[Y(Z=1) - Y(Z=0)]}{P(W(1)=1, W(0)=0)} \\
&= \frac{E[Y|Z=1] - E[Y|Z=0]}{1-P(W=0|Z=1)-P(W=1|Z=0)} \text{ (rule out groups A and N) } \\
&= \frac{E[Y|Z=1] - E[Y|Z=0]}{P(W=1|Z=1)-P(W=1|Z=0)} \\
&= \frac{E[Y|Z=1] - E[Y|Z=0]}{E[W|Z=1)-E[W|Z=0]}
\end{aligned}
$$

### Parametric Models

If we assume linearity, then this becomes the ratio of two OLS coefficients, sometimes called the Wald estimator.

$$
\begin{aligned}
\tau_{ATE} &= \frac{E[Y|Z=1] - E[Y|Z=0]}{E[W|Z=1)-E[W|Z=0]} \\
&= \frac{Cov(Y,Z)}{Cov(W,Z)}
\end{aligned}
$$

Why do covariates help? Because we hope even if we don't have exogeneity or exclusion assumption held, we would have those assumptions held within each level of $X$.

### Example

```{r iv1, warning=FALSE, cache=TRUE, message=FALSE, echo=TRUE}
## DGP: data$y <- data$x + data$z + data$u
library(sem)
set.seed(66)
nobs=10000
nDim = 3
sdww = 1
sdzz=1
sdmm=1
## here we have three variables x,z,w.
## m is the omitted variable,x and m are correlated,  z is the instrument, which is correlated with x, but not m.  u is independent of everything else.
crwm=.6
crmz=0
crwz=.8
covarMat = matrix( c(sdww^2, crwm^2, crwz^2, crwm^2, sdmm^2, crmz^2,  crwz^2, crmz^2, sdzz^2 ) , nrow=nDim , ncol=nDim )
data  = data.frame(MASS::mvrnorm(n=nobs, mu=rep(0,nDim), Sigma=covarMat ))
names(data) <- c('w','m','z')
data$u <- rnorm(nobs,0,1)
# dgp
data$y <- data$w +  data$m + data$u
lm <- lm(y~w, data=data)
lm.full <- lm(y~ w + m, data=data)
tsls.model <- sem::tsls(y ~ w , ~ z , data=data)
```

```{r iv2, warning=FALSE, cache=TRUE, message=FALSE, echo=TRUE}
# lm is biased
summary(lm)$coefficients
# lm.full is good
summary(lm.full)$coefficients
# tsls is good.
summary(tsls.model)$coefficients
```

### Control Function Approach

Two-stage least squares can be implemented equivalently via the **control function (CF) approach**. Understanding why makes the extension to nonlinear models (next chapter) transparent.

**Algebraic motivation.** Because `w` is endogenous, decompose the structural error:

$$\varepsilon = \rho v + \eta, \qquad v \equiv w - E[w \mid z], \qquad E(z\eta) = 0,\; E(zv) = 0.$$

If we *knew* $v$, controlling for it directly would eliminate the confounding channel. We don't know $v$, but the first-stage residual $\hat v = w - \hat w$ is a consistent estimate. Substituting into the structural equation:

$$y = \alpha + \beta w + \rho\,\hat v + \text{error}.$$

OLS on this **augmented** second stage gives a consistent $\hat\beta$ — identical to 2SLS by the Frisch–Waugh–Lovell (FWL) theorem. The coefficient $\hat\rho$ on $\hat v$ doubles as a **Durbin–Wu–Hausman (DWH) endogeneity test**: under $H_0$ (w is exogenous), $\rho = 0$.

```{r iv_cf, warning=FALSE, cache=TRUE, message=FALSE, echo=TRUE}
# Manual 2SLS, explicitly
stage1     <- lm(w ~ z, data = data)
data$w_hat <- fitted(stage1)
tsls_manual <- lm(y ~ w_hat, data = data)

# Control function: first-stage residual v_hat in the augmented regression
data$v_hat <- residuals(stage1)
lm_cf      <- lm(y ~ w + v_hat, data = data)

# Compare β̂_w from manual 2SLS and CF — should be identical by FWL
coef_tsls <- coef(tsls_manual)["w_hat"]
coef_cf   <- coef(lm_cf)["w"]
sprintf("Manual 2SLS   beta_w = %.8f", coef_tsls)
sprintf("Control funct beta_w = %.8f", coef_cf)

# DWH endogeneity test: t-stat on v_hat
rho_summary <- summary(lm_cf)$coefficients["v_hat", ]
sprintf("DWH test on v_hat: coef = %.4f  se = %.4f  t = %.3f  p = %.4g",
        rho_summary["Estimate"], rho_summary["Std. Error"],
        rho_summary["t value"], rho_summary["Pr(>|t|)"])
```

The two $\hat\beta_w$ values agree to machine precision, confirming FWL. The large $|t|$ on $\hat v$ rejects $H_0\!:\rho = 0$, confirming that `w` is endogenous.

::: {.callout-note}
**Why FWL fails in nonlinear models.** FWL relies on the residual-maker matrix being idempotent and the second stage being linear — both fail once we replace OLS with Poisson or Probit. In a nonlinear second stage, $\hat v$ must enter *inside* the link function (e.g. $\exp(\mathbf{x}\boldsymbol\beta + \rho \hat v)$), which requires a structural assumption beyond instrument exogeneity. This is the subject of the next chapter.
:::

## When TSLS-with-Covariates Is *Actually* LATE

The clean LATE identification result above assumed a binary treatment, a binary instrument, and **no covariates**. In that world the Wald estimator $E[Y|Z=1]-E[Y|Z=0]$ over $E[W|Z=1]-E[W|Z=0]$ identifies the LATE under Imbens-Angrist assumptions.

In practice almost no one uses IV that way. We add covariates to the second stage to defend conditional exogeneity:

$$
Y \;=\; \beta\,W \;+\; \gamma^\top X \;+\; U, \qquad
W \;=\; \pi\,Z \;+\; \delta^\top X \;+\; V.
$$

It is widely taught — following Angrist & Pischke (2009) — that this linear-IV specification still recovers a non-negatively weighted average of covariate-specific LATEs. Blandhol, Bonney, Mogstad and Torgovitsky (2025) show this is generally **false**. Understanding why, and what to do about it, is the goal of this section.

### The Hidden Functional-Form Assumption

By the Frisch-Waugh-Lovell theorem, the IV estimand can be written

$$
\beta_{iv} \;=\; \frac{E[\,Y\,\tilde Z\,]}{E[\,W\,\tilde Z\,]},
\qquad \tilde Z \;\equiv\; Z - L[Z\mid X],
$$

where $L[Z\mid X]$ is the **linear** projection of $Z$ on $X$ — i.e., the fitted values from a linear regression of $Z$ on $X$. Blandhol et al. prove that $\beta_{iv}$ is a non-negatively weighted average of conditional LATEs **if and only if** the *rich covariates* condition holds:

$$
L[Z\mid X] \;=\; E[Z\mid X].
$$

In words: the linear projection of $Z$ on $X$ must coincide with the true conditional mean of $Z$ given $X$. This is **not** a property of the data generating process — it is a parametric assumption about the functional form you put into the regression. Under rich covariates, the linear first stage is correctly specified. Without it, $L[Z\mid X]$ omits curvature and interactions, and $\beta_{iv}$ generally mixes complier effects with always-/never-taker effects, with some always-taker terms entering with **negative weights**.

::: {.callout-warning}
"Two-stage least squares with covariates estimates a LATE" is a claim that requires either (i) a saturated specification (one dummy per cell of $X$), or (ii) the implicit, untested assertion that $E[Z\mid X]$ happens to be linear in $X$. The widespread shorthand omits this caveat.
:::

### A Simulation: When Linear TSLS Misses

We construct a DGP where conditional exogeneity holds given $X$, but $E[Z\mid X]$ is genuinely nonlinear. A linear-in-$X$ TSLS specification will therefore violate rich covariates. We compare four estimators against the true LATE.

```{r tsls-late-dgp, warning=FALSE, message=FALSE, cache=TRUE}
library(ivreg)          # ivreg()
library(DoubleML)       # PLIV / debiased ML
library(mlr3)
library(mlr3learners)
library(lmtest)         # resettest()

set.seed(20260522)
n <- 8000

# Single covariate on [-1, 1]
X <- runif(n, -1, 1)

# True conditional mean of Z is wildly nonlinear in X.
# A linear projection L[Z|X] cannot reproduce this.
pZ_true <- plogis(0.3 + 1.0 * X + 2.0 * sin(2.5 * pi * X))
Z       <- rbinom(n, 1, pZ_true)

# Potential treatment states with monotonicity (T1 >= T0).
# X enters the propensity, so X confounds W <-> Y.
U   <- runif(n)
p0  <- plogis(-1.6 + 1.0 * X)            # P(always-taker | X)
p1  <- pmin(plogis(-0.4 + 1.0 * X) + 0.2, 0.95)  # P(AT or CP | X)
T0  <- as.integer(U < p0)
T1  <- as.integer(U < p1)
W   <- ifelse(Z == 1, T1, T0)            # observed treatment

G   <- ifelse(T1 == 1 & T0 == 1, "AT",
        ifelse(T1 == 0 & T0 == 0, "NT", "CP"))

# Heterogeneous treatment effect — varies with X
tau <- 0.5 + 1.5 * X
Y0  <- 1.0 + 2.0 * X + 0.8 * X^2 + rnorm(n, 0, 1)
Y1  <- Y0 + tau
Y   <- ifelse(W == 1, Y1, Y0)

# Truth (only knowable in simulation)
true_LATE <- mean(tau[G == "CP"])
prop.table(table(G))
cat(sprintf("True unconditional LATE = %.3f\n", true_LATE))
```

Now we compare four estimators.

```{r tsls-late-estimators, warning=FALSE, message=FALSE, cache=TRUE}
# (a) Wald — biased: Z is not unconditionally exogenous because X confounds.
wald <- (mean(Y[Z == 1]) - mean(Y[Z == 0])) /
        (mean(W[Z == 1]) - mean(W[Z == 0]))

# (b) TSLS with X entered linearly — the textbook spec.
tsls_lin  <- ivreg(Y ~ W + X | Z + X)
b_lin     <- coef(tsls_lin)["W"]

# (c) TSLS with a rich basis (polynomial up to degree 7) — closer to saturation.
tsls_poly <- ivreg(Y ~ W + poly(X, 7) | Z + poly(X, 7))
b_poly    <- coef(tsls_poly)["W"]

# (d) DDML PLIV via the DoubleML package with random-forest nuisance learners.
lgr::get_logger("mlr3")$set_threshold("warn")
dml_df   <- data.frame(Y = Y, W = W, Z = Z, X = X)
dml_data <- DoubleMLData$new(dml_df, y_col = "Y", d_cols = "W",
                             x_cols = "X", z_cols = "Z")
rf <- lrn("regr.ranger", num.trees = 500, max.depth = 5, min.node.size = 2)
set.seed(20260522)
dml_pliv <- DoubleMLPLIV$new(dml_data,
                             ml_l = rf$clone(),
                             ml_m = rf$clone(),
                             ml_r = rf$clone(),
                             n_folds = 5)
dml_pliv$fit()
b_pliv <- as.numeric(dml_pliv$coef)

results <- data.frame(
  Estimator = c("True LATE",
                "Wald (no covariates)",
                "TSLS, X linear",
                "TSLS, poly(X, 7)",
                "DDML PLIV (DoubleML + ranger)"),
  Estimate  = c(true_LATE, wald, b_lin, b_poly, b_pliv)
)
print(results, row.names = FALSE, digits = 3)
```

The linear-in-$X$ TSLS does not recover the true LATE. The polynomial specification, by adding flexibility, gets closer — and the cross-fitted nonparametric PLIV is closer still. The Wald estimator is badly off because $Z$ is not unconditionally exogenous.

::: {.callout-note}
The DDML PLIV estimand $\beta_{\text{rich}}$ is a **weighted average of conditional LATEs**, with weights determined by $\text{Var}(Z\mid X)\cdot P(\text{complier}\mid X)$. This is non-negatively weighted (so weakly causal) but is generally **not** equal to the unconditional LATE. To target the unconditional LATE itself, see the IPSW approach below.
:::

### A Diagnostic: RESET on `Z ~ X`

The null hypothesis of rich covariates is equivalent to the null that the linear regression $Z = \alpha + \delta^\top X$ has no omitted nonlinear terms — exactly what Ramsey's (1969) RESET test was designed to detect. If RESET rejects, the linear-in-$X$ TSLS specification is not weakly causal in general, and the LATE interpretation does not apply.

```{r tsls-late-reset, warning=FALSE, message=FALSE, cache=TRUE}
reset_z <- resettest(lm(Z ~ X), power = 2:3, type = "regressor")
reset_z
```

In our DGP the test rejects strongly, flagging the rich-covariates failure that we know is there by construction. The same test on real data is a cheap, principled check before reporting a LATE interpretation of a 2SLS coefficient.

### What to Do Instead

Blandhol et al. (2025) offer four practical steps for empirical work.

1. **Reconsider the role of covariates.** If $Z$ is unconditionally exogenous, drop the covariates. Adding them for precision is fine; a kitchen-sink set of controls makes rich covariates *less* likely to hold.
2. **Run RESET on `Z ~ X`** as above. If it does not reject, the linear-IV-as-LATE interpretation is defensible (subject to the usual IV assumptions).
3. **If RESET rejects, report a partially-linear IV estimate alongside TSLS.** This is what DDML PLIV delivers. R packages: `DoubleML` (used above), `ddml`. Stata: `ddml` (Ahrens et al. 2023).
4. **For a binary instrument, also estimate the unconditional LATE directly.** This can be done with DDML or with instrument propensity score weighting (IPSW; Słoczyński et al. 2024):

   $$
   \hat\beta_{\text{late}}
   \;=\;
   \frac{\sum_i Y_i \big[Z_i / \hat p(X_i) - (1-Z_i) / (1 - \hat p(X_i))\big]}
        {\sum_i W_i \big[Z_i / \hat p(X_i) - (1-Z_i) / (1 - \hat p(X_i))\big]}
   $$

   where $\hat p(X)$ is a nonparametric estimate of $P(Z = 1\mid X)$. Stata's `ipsw` package implements this; in R it is straightforward to construct point estimates by hand once $\hat p(X)$ is in place.

### Why This Matters for the FRDD Example Below

The fuzzy-RDD application at the end of this chapter instruments the treatment with an above-cutoff indicator while controlling for race, state of birth, and quarter-of-birth fixed effects. The instrument's conditional mean given those covariates is unlikely to be linear in them — and indeed the `feols` 2SLS estimate and the local `rdrobust` estimate differ markedly. The original commentary attributes that gap to "sensitivity to model selection." Rich-covariates failure is at least as plausible an explanation, and the RESET diagnostic above is the right first thing to run on that specification.

### Bottom Line

- "TSLS with covariates estimates a LATE" is true only under the rich-covariates condition, which is a parametric assumption on the conditional mean of the instrument.
- That assumption is rarely defended in published work and is often rejected by a simple RESET test.
- The practical alternatives — DDML PLIV for $\beta_{\text{rich}}$, IPSW or DDML for the unconditional LATE — are available in mature R and Stata packages.
- The criterion "non-negatively weighted average of conditional LATEs" is itself a weak one. Even when it holds, the estimand can differ substantially from the unconditional LATE that practitioners usually have in mind.

> **Reference.** Blandhol, C., Bonney, J., Mogstad, M., & Torgovitsky, A. (2025). *When is TSLS Actually LATE?* Working paper.

## Regression Discontinuity Design

RDD (regression discontinuity design) is a quasi-experimental design that is used to estimate causal effects of interventions when assignment to the intervention is determined by whether a subject's value on an observed covariate exceeds a threshold. The idea is that the assignment is as good as random, so we can estimate the causal effect of the intervention by comparing the outcomes of subjects who are just above and just below the threshold.

### Sharp RDD

We have a continuous variable $X$, called the running variable, which determines the binary treatment $W$. The treatment is assigned according to a threshold $c$. The outcome variable $Y$ is a function of $X$ and $W$.

$$ W=\mathbb{1}(X>c)$$

Lee (2008) studies the effect of incumbency advantage in elections. His identification strategy is based on the discontinuity generated by the rule that the party with a majority vote share wins. The forcing variable $X_i$ is the difference in vote share between the Democratic and Republican parties in one election, with the threshold $c = 0$. The outcome variable $Y_i$ is vote share at the second election.

```{r rdd1, warning=FALSE, cache=TRUE, message=FALSE, echo=TRUE}
library(rdrobust)
data(rdrobust_RDsenate)
rdplot(x=rdrobust_RDsenate$margin, y=rdrobust_RDsenate$vote)
```

### Identification of SRDD

$$ \tau_{RD} = \lim_{x \to c^+} E[Y|X=x] - \lim_{x \to c^-} E[Y|X=x] $$

In words, the treatment effect is the difference in the outcome from the right side and from the left side. This is the same as the difference in the outcome at the threshold, as if the treatment is assigned randomly.

### Estimation of SRDD

```{r rdd2, warning=FALSE, cache=TRUE, message=FALSE, echo=TRUE}
GPA <- runif(1000, 0, 4)
future_success <- 10 + 1.5 * GPA + 2 * (GPA <= 3) + rnorm(1000)
rdplot(x=GPA, y=future_success, c=3)
```

```{r rdd3, warning=FALSE, cache=TRUE, message=FALSE, echo=TRUE}
# estimate the sharp RDD model
rdd_gpa <- rdrobust(x=GPA, y=future_success, c=3)
summary(rdd_gpa)
```

### Nonparametric Estimation

```{r rdd4, warning=FALSE, cache=TRUE, message=FALSE, echo=TRUE}
rdd_house <- rdrobust(x=rdrobust_RDsenate$margin, y=rdrobust_RDsenate$vote, c=0, all=TRUE)
summary(rdd_house)
```

### Fuzzy RDD

In fuzzy RDD, the treatment is assigned according to a threshold $c$, but there exist non-compliance. This is similar to IV case.

$$ \tau_{FRD} = \frac{\lim_{x \to c^+} E[Y|X=x] - \lim_{x \to c^-} E[Y|X=x]}{\lim_{x \to c^+} E[W|X=x] - \lim_{x \to c^-} E[W|X=x]} $$

For example, if food stamp eligibility is given to all households below a certain income, but not all households receive the food stamps. In other words, income does not solely determine the assignment. Cutoff point increases the probability of treatment but doesn't completely determine treatment.

FRDD is IV.

### FRDD Example

Fetter (2013)'s main question of interest is how much of the increase in the home ownership rate in the midcentury US was due to mortgage subsidies given out by the government.

We're using the running variable quarter of birth (qob), which has been centered on the quarter of birth you'd need to be to be eligible for a mortgage subsidy for fighting in the Korean War (qob_minus_kw). This determines whether you were a veteran of either the Korean War or World War II (vet_wwko).

```{r rdd5, warning=FALSE, cache=TRUE, message=FALSE, echo=TRUE}
vet <- causaldata::mortgages
# Create an "above-cutoff" variable as the instrument
vet <- vet %>% mutate(above = qob_minus_kw > 0)
# Impose a bandwidth of 12 quarters on either side
vet <- vet %>%  filter(abs(qob_minus_kw) < 12)
m <- feols(home_ownership ~
    nonwhite  | # Control for race
    bpl + qob | # fixed effect controls
    qob_minus_kw*vet_wwko ~ # Instrument our standard RDD
    qob_minus_kw*above, # with being above the cutoff
    se = 'hetero', # heteroskedasticity-robust SEs
    data = vet)
summary(m)
```

```{r rdd6, warning=FALSE, cache=TRUE, message=FALSE, echo=TRUE}
controls <- vet %>%
    select(nonwhite, bpl, qob) %>%
    mutate(qob = factor(qob))
conmatrix <- model.matrix(~., data = controls)
m <- rdrobust(vet$home_ownership,
              vet$qob_minus_kw,
              fuzzy = vet$vet_wwko,
              c = 0,
              covs = conmatrix)
m$Estimate
```

These two results are very different, which is one of RDD's problems. It can be sensitive to model selection. For example, in the case of 2SLS, it's a linear model. In the case of "rdrobust", it's a local regression, which can be sensitive to bandwidth selection.
