# IV & Regression Discontinuity

```{r}
#| include: false
library(ggplot2)
library(ggdag)
library(sem)
library(MASS)
library(rdrobust)
library(causaldata)
library(fixest)
library(tidyverse)
```

If we have an unobserved confounder, unconfoundedness fails. Even when we have random experiment, we could have noncompliance. In that case, subjects are randomly assigned to treatment or control, but some of them do not comply with the assignment. In other words, there is selection bias in the sense that the treated group is self-selected. This selection could be correlated with the outcome.

## Instrumental Variables

### Uncontrolled Confounder

```{r dag2, warning=FALSE, cache=TRUE, message=FALSE, echo=FALSE}
library(ggplot2)
library(ggdag)
g <- dagify(
  y ~ w,
  w ~ c,
  y ~ c,
  exposure = "w",
  outcome = "y",
  coords = list(x = c(w = 2, c = 3, y = 4),
                y = c(w = 2, c = 3, y = 2))
)
ggdag(g)
```

### Why Does IV Work?

```{r dag3, warning=FALSE, cache=TRUE, message=FALSE, echo=FALSE}
g <- dagify(
  y ~ w,
  w ~ c,
  y ~ c,
  w ~ z,
  exposure = "w",
  outcome = "y",
  coords = list(x = c(w = 2, c = 3, y = 4, z=0),
                y = c(w = 2, c = 3, y = 2, z=2))
)
ggdag(g)
```

### IV Under Potential Outcomes

-   W has two potential outcomes W(1) and W(0), as a function of Z.
-   Y has two potential outcomes Y(1) and Y(0), as a function of W.

Intention to Treat (ITT) is the average treatment effect of the treatment assignment Z. It is the difference between the average potential outcome under treatment and the average potential outcome under control. It is the average treatment effect of the treatment assignment Z, regardless of whether the subject complies with the assignment.

There are four possible groups of subjects in the case of binary treatment and binary instrument. The groups are defined by the *potential* treatment status $(W(0), W(1))$ — how the subject would take treatment under each value of the instrument — not by the observed $(Z, W)$:

-   $W(0)=0, W(1)=0$: never-taker
-   $W(0)=0, W(1)=1$: complier
-   $W(0)=1, W(1)=0$: defier
-   $W(0)=1, W(1)=1$: always-taker

|        | W(1)=0 | W(1)=1 |
|:-------|:-------|:-------|
| W(0)=0 | N      | C      |
| W(0)=1 | D      | A      |

: compliance groups

These four groups (strata) have different causal mechanisms. The compliers and defiers are the groups whose *treatment status* responds to the assignment $Z$ — compliers in the same direction as the assignment, defiers in the opposite direction. The always-takers and never-takers take the same treatment regardless of $Z$.

Unfortunately we cannot identify the compliance group by looking at the data.

| Z   | W   | G    |
|:----|:----|:-----|
| 0   | 0   | C, N |
| 0   | 1   | A, D |
| 1   | 0   | N, D |
| 1   | 1   | C, A |

### Assumptions

-   Exclusion restriction: $Y(w, 1) = Y(w,0)$
-   Exogeneity of instrument: $Z \perp [W(0),W(1), Y(0), Y(1)]$
-   Monotonicity (no defiers): $W(0) \leq W(1)$

### LATE Identification

Since we excluded defiers, and the other two groups (always-takers and never-takers) are not affected by the treatment assignment Z. The only effect we can identify is the treatment effect on the compliers. This is called the Local Average Treatment Effect (LATE).

Below we write potential outcomes indexed by the instrument, $Y(Z=z)$, rather than by treatment status, $Y(d)$. The exclusion restriction is exactly what licenses this: since $Y(w,1)=Y(w,0)$, $Y$ depends on $Z$ only through $W$, so $Y(Z=z) = Y(W(z), z) = Y(W(z))$ in the usual treatment-indexed notation. For a complier ($W(1)=1, W(0)=0$), this means $Y(Z=1) = Y(1)$ and $Y(Z=0) = Y(0)$, so $E[Y(Z=1)-Y(Z=0) \mid G=C]$ below is exactly $E[Y(1)-Y(0) \mid G=C]$, the LATE in treatment-indexed notation.

$$ \small
\begin{aligned}
& E[Y(Z=1) - Y(Z=0) | G = C] \\
&= \frac{E[Y(Z=1) - Y(Z=0)]}{P(W(1)=1, W(0)=0)} \\
&= \frac{E[Y|Z=1] - E[Y|Z=0]}{1-P(W=0|Z=1)-P(W=1|Z=0)} \text{ (monotonicity rules out D; the A and N shares subtract) } \\
&= \frac{E[Y|Z=1] - E[Y|Z=0]}{P(W=1|Z=1)-P(W=1|Z=0)} \\
&= \frac{E[Y|Z=1] - E[Y|Z=0]}{E[W|Z=1]-E[W|Z=0]}
\end{aligned}
$$ {#eq-iv-rdd-1}

### Parametric Models

With a binary instrument this is exactly the ratio of two OLS coefficients — no linearity assumption is needed, since a regression on a binary regressor just computes the two group means. This is sometimes called the Wald estimator.

$$
\begin{aligned}
\tau_{LATE} &= \frac{E[Y|Z=1] - E[Y|Z=0]}{E[W|Z=1]-E[W|Z=0]} \\
&= \frac{Cov(Y,Z)}{Cov(W,Z)}
\end{aligned}
$$ {#eq-iv-rdd-2}

Why do covariates help? It is important to separate two assumptions
here. Conditioning on covariates $X$ can make the *as-if-random*
(conditional independence) assumption more credible: the instrument
may be unconfounded only within levels of $X$. But the *exclusion
restriction* -- that the instrument has no direct effect on the
outcome -- is a substantive structural assumption that covariates do
not generally repair. Adding $X$ removes a direct $Z \to Y$ path only
if that path runs entirely through $X$, or if exclusion is conditional
by design. So covariates help with conditional independence, not with
exclusion per se.

### Example

```{r iv1, warning=FALSE, cache=TRUE, message=FALSE, echo=TRUE}
## DGP: data$y <- data$w + data$m + data$u
library(sem)
set.seed(66)
nobs=10000
nDim = 3
sdww = 1
sdzz=1
sdmm=1
## here we have three variables w, m, z.
## m is the omitted variable; w and m are correlated; z is the instrument, which is correlated with w, but not m.  u is independent of everything else.
## with unit variances, the covariances below ARE the correlations
crwm=.36
crmz=0
crwz=.64
covarMat = matrix( c(sdww^2, crwm, crwz, crwm, sdmm^2, crmz,  crwz, crmz, sdzz^2 ) , nrow=nDim , ncol=nDim )
data  = data.frame(MASS::mvrnorm(n=nobs, mu=rep(0,nDim), Sigma=covarMat ))
names(data) <- c('w','m','z')
data$u <- rnorm(nobs,0,1)
# dgp
data$y <- data$w +  data$m + data$u
lm <- lm(y~w, data=data)
lm.full <- lm(y~ w + m, data=data)
tsls.model <- sem::tsls(y ~ w , ~ z , data=data)
```

```{r iv2, warning=FALSE, cache=TRUE, message=FALSE, echo=TRUE}
# lm is biased
summary(lm)$coefficients
# lm.full is good
summary(lm.full)$coefficients
# tsls is good.
summary(tsls.model)$coefficients
```

### Control Function Approach

Two-stage least squares can be implemented equivalently via the **control function (CF) approach**. Understanding why makes the extension to nonlinear models (next chapter) transparent.

**Algebraic motivation.** Because `w` is endogenous, decompose the structural error:

$$\varepsilon = \rho v + \eta, \qquad v \equiv w - E[w \mid z], \qquad E(z\eta) = 0,\; E(zv) = 0.$$ {#eq-iv-rdd-3}

If we *knew* $v$, controlling for it directly would eliminate the confounding channel. We don't know $v$, but the first-stage residual $\hat v = w - \hat w$ is a consistent estimate. Substituting into the structural equation:

$$y = \alpha + \beta w + \rho\,\hat v + \text{error}.$$ {#eq-iv-rdd-4}

OLS on this **augmented** second stage gives a consistent $\hat\beta$ — identical to 2SLS by the Frisch–Waugh–Lovell (FWL) theorem. The coefficient $\hat\rho$ on $\hat v$ doubles as a **Durbin–Wu–Hausman (DWH) endogeneity test**: under $H_0$ (w is exogenous), $\rho = 0$.

```{r iv_cf, warning=FALSE, cache=TRUE, message=FALSE, echo=TRUE}
# Manual 2SLS, explicitly
stage1     <- lm(w ~ z, data = data)
data$w_hat <- fitted(stage1)
tsls_manual <- lm(y ~ w_hat, data = data)

# Control function: first-stage residual v_hat in the augmented regression
data$v_hat <- residuals(stage1)
lm_cf      <- lm(y ~ w + v_hat, data = data)

# Compare β̂_w from manual 2SLS and CF — should be identical by FWL
coef_tsls <- coef(tsls_manual)["w_hat"]
coef_cf   <- coef(lm_cf)["w"]
sprintf("Manual 2SLS   beta_w = %.8f", coef_tsls)
sprintf("Control funct beta_w = %.8f", coef_cf)

# DWH endogeneity test: t-stat on v_hat
rho_summary <- summary(lm_cf)$coefficients["v_hat", ]
sprintf("DWH test on v_hat: coef = %.4f  se = %.4f  t = %.3f  p = %.4g",
        rho_summary["Estimate"], rho_summary["Std. Error"],
        rho_summary["t value"], rho_summary["Pr(>|t|)"])
```

The two $\hat\beta_w$ values agree to machine precision, confirming FWL. The large $|t|$ on $\hat v$ rejects $H_0\!:\rho = 0$, confirming that `w` is endogenous. One caveat: the OLS standard errors from the augmented regression are not valid for $\hat\beta$ ($\hat v$ is a generated regressor; the DWH test is fine because under $H_0$ the first-stage estimation error does not matter). Use 2SLS standard errors, or bootstrap the two stages jointly as we do in the Poisson IV chapter.

::: {.callout-note}
**Why FWL fails in nonlinear models.** FWL is a property of linear projections; once we replace OLS with Poisson or Probit, the estimator is no longer a linear projection and the partialling-out logic does not apply. In a nonlinear second stage, $\hat v$ must enter *inside* the link function (e.g. $\exp(\mathbf{x}\boldsymbol\beta + \rho \hat v)$), which requires a structural assumption beyond instrument exogeneity. This is the subject of the next chapter.
:::

## When 2SLS with Covariates Is *Actually* LATE

The LATE identification above is clean because we have no covariates. The Wald estimator gives us LATE, full stop. But in practice we usually add covariates to defend conditional exogeneity:

$$
Y = \beta W + \gamma^\top X + U, \qquad W = \pi Z + \delta^\top X + V.
$$ {#eq-iv-rdd-5}

Does 2SLS on this still give us a LATE? @angrist-pischke-2009 say it gives a weighted average of covariate-specific LATEs. @blandhol-et-al-2025 show that this is generally not true.

### Why the Linear Specification Hides an Assumption

By Frisch-Waugh-Lovell, the IV estimand is

$$
\beta_{iv} = \frac{E[Y \tilde Z]}{E[W \tilde Z]}, \qquad \tilde Z = Z - L[Z \mid X],
$$ {#eq-iv-rdd-6}

where $L[Z \mid X]$ is the **linear** projection of $Z$ on $X$. So $\tilde Z$ is the part of $Z$ that a linear regression on $X$ cannot explain.

What if the true $E[Z \mid X]$ is not linear in $X$? Then $L[Z \mid X] \neq E[Z \mid X]$, and the residual instrument $\tilde Z$ still picks up some of the conditional mean. The IV estimand then mixes treatment effects across compliance groups. @blandhol-et-al-2025 prove that $\beta_{iv}$ is a non-negatively weighted average of conditional LATEs only if the rich covariates condition holds:

$$
L[Z \mid X] = E[Z \mid X].
$$ {#eq-iv-rdd-7}

This is a parametric assumption about $E[Z \mid X]$. It is implicit in the linear-in-$X$ specification, and it is rarely defended. When it fails, $\beta_{iv}$ picks up treatment effects for always-takers as well as compliers, and some always-taker terms enter with negative weight.

::: {.callout-warning}
"2SLS with covariates estimates LATE" is true only if $X$ enters saturated (one dummy per cell), or the true $E[Z \mid X]$ happens to be linear in $X$.
:::

### A Simulation

We build a DGP where $E[Z \mid X]$ is wildly nonlinear in $X$, so the linear-in-$X$ specification has to fail. Then we compare four estimators against the true LATE.

```{r tsls-late-dgp, warning=FALSE, message=FALSE, cache=TRUE}
library(ivreg)          # ivreg()
library(DoubleML)       # PLIV / debiased ML
library(mlr3)
library(mlr3learners)
library(lmtest)         # resettest()

set.seed(20260522)
n <- 8000

# Single covariate on [-1, 1]
X <- runif(n, -1, 1)

# True conditional mean of Z is wildly nonlinear in X.
# A linear projection L[Z|X] cannot reproduce this.
pZ_true <- plogis(0.3 + 1.0 * X + 2.0 * sin(2.5 * pi * X))
Z       <- rbinom(n, 1, pZ_true)

# Potential treatment states with monotonicity (T1 >= T0).
# X enters the propensity, so X confounds W <-> Y.
U   <- runif(n)
p0  <- plogis(-1.6 + 1.0 * X)            # P(always-taker | X)
p1  <- pmin(plogis(-0.4 + 1.0 * X) + 0.2, 0.95)  # P(AT or CP | X)
T0  <- as.integer(U < p0)
T1  <- as.integer(U < p1)
W   <- ifelse(Z == 1, T1, T0)            # observed treatment

G   <- ifelse(T1 == 1 & T0 == 1, "AT",
        ifelse(T1 == 0 & T0 == 0, "NT", "CP"))

# Heterogeneous treatment effect — varies with X
tau <- 0.5 + 1.5 * X
Y0  <- 1.0 + 2.0 * X + 0.8 * X^2 + rnorm(n, 0, 1)
Y1  <- Y0 + tau
Y   <- ifelse(W == 1, Y1, Y0)

# Truth (only knowable in simulation)
true_LATE <- mean(tau[G == "CP"])
prop.table(table(G))
cat(sprintf("True unconditional LATE = %.3f\n", true_LATE))
```

Now we compare four estimators.

```{r tsls-late-estimators, warning=FALSE, message=FALSE, cache=TRUE}
# (a) Wald — biased: Z is not unconditionally exogenous because X confounds.
wald <- (mean(Y[Z == 1]) - mean(Y[Z == 0])) /
        (mean(W[Z == 1]) - mean(W[Z == 0]))

# (b) TSLS with X entered linearly — the textbook spec.
tsls_lin  <- ivreg(Y ~ W + X | Z + X)
b_lin     <- coef(tsls_lin)["W"]

# (c) TSLS with a rich basis (polynomial up to degree 7) — closer to saturation.
tsls_poly <- ivreg(Y ~ W + poly(X, 7) | Z + poly(X, 7))
b_poly    <- coef(tsls_poly)["W"]

# (d) DDML PLIV via the DoubleML package with random-forest nuisance learners.
lgr::get_logger("mlr3")$set_threshold("warn")
dml_df   <- data.frame(Y = Y, W = W, Z = Z, X = X)
dml_data <- DoubleMLData$new(dml_df, y_col = "Y", d_cols = "W",
                             x_cols = "X", z_cols = "Z")
rf <- lrn("regr.ranger", num.trees = 500, max.depth = 5, min.node.size = 2)
set.seed(20260522)
dml_pliv <- DoubleMLPLIV$new(dml_data,
                             ml_l = rf$clone(),
                             ml_m = rf$clone(),
                             ml_r = rf$clone(),
                             n_folds = 5)
dml_pliv$fit()
b_pliv <- as.numeric(dml_pliv$coef)

se_lin  <- summary(tsls_lin)$coefficients["W", "Std. Error"]
se_poly <- summary(tsls_poly)$coefficients["W", "Std. Error"]
se_pliv <- as.numeric(dml_pliv$se)

results <- data.frame(
  Estimator = c("True LATE",
                "Wald (no covariates)",
                "TSLS, X linear",
                "TSLS, poly(X, 7)",
                "DDML PLIV (DoubleML + ranger)"),
  Estimate  = c(true_LATE, wald, b_lin, b_poly, b_pliv),
  SE        = c(NA, NA, se_lin, se_poly, se_pliv)
)
print(results, row.names = FALSE, digits = 3)
```

The linear-in-$X$ 2SLS does not recover the true LATE — and neither do the more flexible versions. Two things are going on, and the standard errors in the table help separate them. First, the estimands differ: for this DGP the linear-projection 2SLS converges to 0.55, and the polynomial 2SLS and DDML PLIV converge to approximately $\beta_{rich} = 0.58$ (computable by integrating the weight formula below), all below the population LATE of 0.60 — a real but modest gap that persists at any sample size. Second, sampling noise: with SEs near 0.06, this particular draw happens to land about 1.5 SEs below its own estimand, which is why the estimates cluster near 0.49–0.51 rather than 0.55–0.58. The Wald estimator is far off for a different reason entirely: $Z$ is not unconditionally exogenous, and Wald has no way to use $X$.

What are the polynomial 2SLS and DDML PLIV estimating, if not LATE? PLIV targets $\beta_{rich}$, a weighted average of conditional LATEs with weights proportional to $\text{Var}(Z \mid X) \cdot P(\text{complier} \mid X)$; the polynomial 2SLS targets nearly the same quantity (a degree-7 polynomial cannot fully reproduce this $E[Z \mid X]$, so its estimand is a misspecified-projection variant sitting just below $\beta_{rich}$). These are non-negatively weighted, so weakly causal. But they are not the unconditional LATE we usually want.

### Diagnostic: RESET on Z ~ X

Rich covariates says $E[Z \mid X]$ is linear in $X$. That is exactly the null of Ramsey's RESET test [@ramsey-1969]. We can apply it to a regression of $Z$ on $X$ and see whether higher-order terms of $X$ are jointly significant.

```{r tsls-late-reset, warning=FALSE, message=FALSE, cache=TRUE}
reset_z <- resettest(lm(Z ~ X), power = 2:3, type = "regressor")
reset_z
```

In our DGP it rejects strongly. The same test on real data is cheap, and it tells us whether the LATE interpretation is defensible before we report a 2SLS coefficient.

### What to Do Instead

@blandhol-et-al-2025 give four steps for empirical work.

1. **Reconsider the covariates.** If $Z$ is unconditionally exogenous, drop them. A kitchen-sink set of controls makes rich covariates *less* likely.
2. **Run RESET on `Z ~ X`.** If it does not reject, the linear-IV-as-LATE interpretation is defensible.
3. **If RESET rejects, report DDML PLIV alongside 2SLS.** That is what the `DoubleML` chunk above does. DDML PLIV targets $\beta_{rich}$, a non-negatively weighted average of conditional LATEs.
4. **For a binary instrument, also estimate the unconditional LATE.** With instrument propensity score weighting [@sloczynski-2024],

   $$
   \hat\beta_{late} = \frac{\sum_i Y_i [Z_i/\hat p(X_i) - (1-Z_i)/(1-\hat p(X_i))]}{\sum_i W_i [Z_i/\hat p(X_i) - (1-Z_i)/(1-\hat p(X_i))]},
   $$ {#eq-iv-rdd-8}

   where $\hat p(X) = P(Z = 1 \mid X)$ is estimated nonparametrically. Stata has `kappalate` (Słoczyński, Uysal and Wooldridge), which implements this and related weighting estimators of the LATE. In R it is straightforward to build once $\hat p(X)$ is in hand.

### A Note on the FRDD Example

The fuzzy-RDD example at the end of this chapter is 2SLS with race, state of birth, and quarter-of-birth fixed effects entered as covariates. The `feols` 2SLS estimate and the local `rdrobust` estimate differ. We called that "sensitivity to model selection" earlier. The rich-covariates story is at least as plausible an explanation, and the RESET test is the right first thing to run.

### Bottom Line

"2SLS with covariates estimates LATE" needs the rich covariates condition. That condition is a parametric assumption on $E[Z \mid X]$ that researchers rarely defend, and a simple RESET test often rejects it. The honest alternatives are DDML PLIV for $\beta_{rich}$ and IPSW (or DDML) for the unconditional LATE. Even when rich covariates holds, $\beta_{rich}$ can be quite different from the unconditional LATE we usually care about.

## Regression Discontinuity Design

RDD (regression discontinuity design) is a quasi-experimental design that is used to estimate causal effects of interventions when assignment to the intervention is determined by whether a subject's value on an observed covariate exceeds a threshold. The idea is that the assignment is as good as random, so we can estimate the causal effect of the intervention by comparing the outcomes of subjects who are just above and just below the threshold.

### Sharp RDD

We have a continuous variable $X$, called the running variable, which determines the binary treatment $W$. The treatment is assigned according to a threshold $c$. The outcome variable $Y$ is a function of $X$ and $W$.

$$ W=\mathbb{1}(X>c)$$ {#eq-iv-rdd-9}

Lee (2008) studies the effect of incumbency advantage in elections. His identification strategy is based on the discontinuity generated by the rule that the party with a majority vote share wins. The forcing variable $X_i$ is the difference in vote share between the Democratic and Republican parties in one election, with the threshold $c = 0$. The outcome variable $Y_i$ is vote share at the second election.

```{r rdd1, warning=FALSE, cache=TRUE, message=FALSE, echo=TRUE}
library(rdrobust)
data(rdrobust_RDsenate)
rdplot(x=rdrobust_RDsenate$margin, y=rdrobust_RDsenate$vote)
```

### Identification of SRDD

$$ \tau_{RD} = \lim_{x \to c^+} E[Y|X=x] - \lim_{x \to c^-} E[Y|X=x] $$ {#eq-iv-rdd-10}

In words, the treatment effect is the difference in the outcome from the right side and from the left side. This is the same as the difference in the outcome at the threshold, as if the treatment is assigned randomly.

### Estimation of SRDD

```{r rdd2, warning=FALSE, cache=TRUE, message=FALSE, echo=TRUE}
GPA <- runif(1000, 0, 4)
# Treatment is W = 1(GPA > 3), so the +2 jump must sit on the right
# (above-cutoff) side to match the right-minus-left estimand above.
future_success <- 10 + 1.5 * GPA + 2 * (GPA > 3) + rnorm(1000)
rdplot(x=GPA, y=future_success, c=3)
```

```{r rdd3, warning=FALSE, cache=TRUE, message=FALSE, echo=TRUE}
# estimate the sharp RDD model
rdd_gpa <- rdrobust(x=GPA, y=future_success, c=3)
summary(rdd_gpa)
```

### Nonparametric Estimation

```{r rdd4, warning=FALSE, cache=TRUE, message=FALSE, echo=TRUE}
rdd_house <- rdrobust(x=rdrobust_RDsenate$margin, y=rdrobust_RDsenate$vote, c=0)
summary(rdd_house)
```

### Fuzzy RDD

In fuzzy RDD, the treatment is assigned according to a threshold $c$, but there exist non-compliance. This is similar to IV case.

$$ \tau_{FRD} = \frac{\lim_{x \to c^+} E[Y|X=x] - \lim_{x \to c^-} E[Y|X=x]}{\lim_{x \to c^+} E[W|X=x] - \lim_{x \to c^-} E[W|X=x]} $$ {#eq-iv-rdd-11}

For example, if food stamp eligibility is given to all households below a certain income, but not all households receive the food stamps. In other words, income does not solely determine the assignment. Cutoff point increases the probability of treatment but doesn't completely determine treatment.

FRDD is IV.

### FRDD Example

Fetter (2013)'s main question of interest is how much of the increase in the home ownership rate in the midcentury US was due to mortgage subsidies given out by the government.

We're using the running variable quarter of birth (qob), which has been centered on the quarter of birth you'd need to be to be eligible for a mortgage subsidy for fighting in the Korean War (qob_minus_kw). This determines whether you were a veteran of either the Korean War or World War II (vet_wwko).

```{r rdd5, warning=FALSE, cache=TRUE, message=FALSE, echo=TRUE}
vet <- causaldata::mortgages
# Create an "above-cutoff" variable as the instrument
vet <- vet %>% mutate(above = qob_minus_kw > 0)
# Impose a bandwidth of 12 quarters on either side
vet <- vet %>%  filter(abs(qob_minus_kw) < 12)
# Note: the running variable qob_minus_kw is exogenous, so it goes in the
# control block (not the endogenous block). Listing it in both the
# endogenous and instrument sets makes newer fixest error ("endogenous
# variable fully explained"). The endogenous regressors are veteran status
# and its interaction with the running variable, instrumented by being above
# the cutoff and that interaction.
m <- feols(home_ownership ~
    nonwhite + qob_minus_kw | # exogenous controls incl. the running variable
    bpl + qob | # fixed effect controls
    vet_wwko + qob_minus_kw:vet_wwko ~ # endogenous: treatment and its RDD slope interaction
    above + qob_minus_kw:above, # instruments: above-cutoff and its interaction
    se = 'hetero', # heteroskedasticity-robust SEs
    data = vet)
summary(m)
```

```{r rdd6, warning=FALSE, cache=TRUE, message=FALSE, echo=TRUE}
controls <- vet %>%
    select(nonwhite, bpl, qob) %>%
    mutate(qob = factor(qob))
conmatrix <- model.matrix(~., data = controls)
m <- rdrobust(vet$home_ownership,
              vet$qob_minus_kw,
              fuzzy = vet$vet_wwko,
              c = 0,
              covs = conmatrix)
m$Estimate
```

These two results are very different, which is one of RDD's problems. It can be sensitive to model selection. For example, in the case of 2SLS, it's a linear model. In the case of "rdrobust", it's a local regression, which can be sensitive to bandwidth selection.
