# Matching Estimators

```{r}
#| include: false
library(tidyverse)
library(MatchIt)
library(cobalt)
library(WeightIt)
library(marginaleffects)
library(broom)
```

```{r}
#| eval: false
library(tidyverse)
library(MatchIt)
library(cobalt)
library(WeightIt)
library(marginaleffects)
```

Matching is the oldest and most-taught approach to observational causal
inference. The idea is simple: for each treated unit, find a "twin"
control unit with similar covariates, then compare outcomes within
matched pairs. With enough good matches, the comparison approximates
what we would have learned from a randomised experiment.

The implementation has many variants — nearest-neighbour matching,
optimal matching, full matching, coarsened exact matching (CEM) — and
the choice has real consequences. This chapter walks through the modern
matching workflow using R's `MatchIt` and companion `cobalt` packages,
which together represent the de facto standard in applied econometrics
and epidemiology.

> **Related reading**: This chapter is based on the longer treatment in
> [Matching and Weighting Part 1](https://xiangao.github.io/blog_book/matching-part1.html)
> of *Topics on Econometrics and Causal Inference*, which credits Noah
> Greifer and coauthors (the `MatchIt`/`WeightIt` package authors).

## Assumptions

Matching identifies a causal effect under three assumptions:

1. **SUTVA** — no interference, no hidden treatment versions.
2. **Ignorability** (unconfoundedness) — conditional on $X$, the
   treatment $D$ is independent of the potential outcomes $(Y(0), Y(1))$.
3. **Overlap** (positivity) — every value of $X$ has positive probability
   of being both treated and untreated.

The motivating problem: even under ignorability, the *observed* distribution
of $X$ usually differs between treated and control. Regression adjustment
implicitly extrapolates a parametric model across that imbalance; matching
restricts the comparison to the region of overlap, which is more robust
to model misspecification.

## The Lalonde example

The canonical test case is the Lalonde (1986) job-training experiment. We
use the observational version included in `MatchIt`, where the goal is to
recover the experimental estimate of the effect of a training programme
on 1978 earnings:

```{r}
#| label: data
#| cache: true

data("lalonde", package = "MatchIt")
glimpse(lalonde)
```

The treatment is `treat`; the outcome is `re78` (1978 earnings); the
covariates are `age`, `educ`, `race`, `married`, `nodegree`, `re74`,
`re75` (pre-treatment earnings).

Before matching, treated and control groups are badly imbalanced:

```{r}
#| label: pre-balance
#| cache: true

# Construct a pre-match MatchIt object for balance reporting
m.pre <- matchit(treat ~ age + educ + race + married + nodegree + re74 + re75,
                 data = lalonde, method = NULL,
                 distance = "glm")
bal.tab(m.pre, thresholds = c(m = 0.1))
```

The standardised mean differences ("Diff.Adj") on several covariates
exceed 0.1 — the conventional threshold for adequate balance.

## Distance measures

Matching needs a way to define "close." Three common choices:

1. **Propensity score**: estimate $\hat e(x) = P(D = 1 \mid X = x)$ and
   use $|\hat e(x_i) - \hat e(x_j)|$ as the distance. Reduces a
   multidimensional matching problem to a one-dimensional one.
2. **Mahalanobis distance**: $(x_i - x_j)' \Sigma^{-1} (x_i - x_j)$,
   computed on the raw covariates. Sensitive to scale; works well in
   low dimensions.
3. **Hybrid**: use Mahalanobis distance *within* propensity score
   calipers — the Rubin-Stuart "best of both" approach.

## Matching methods

### Nearest-neighbour matching on a propensity score

```{r}
#| label: nn-match
#| cache: true

m.nn <- matchit(treat ~ age + educ + race + married + nodegree + re74 + re75,
                data = lalonde,
                method = "nearest",
                distance = "glm",
                link = "linear.logit",
                ratio = 1)

m.nn
```

After matching, the treated and control groups should be balanced:

```{r}
#| label: nn-balance
#| cache: true

bal.tab(m.nn, thresholds = c(m = 0.1))
```

Several covariates still exceed the 0.1 threshold — nearest-neighbour
matching on a logistic propensity score isn't always enough. Try a more
flexible distance:

### Full matching

Full matching creates subclasses of one treated unit with multiple
controls, or one control with multiple treated, weighted to minimise
total within-pair distance. It uses *all* observations and typically
achieves better balance:

```{r}
#| label: full-match
#| cache: true

m.full <- matchit(treat ~ age + educ + race + married + nodegree + re74 + re75,
                  data = lalonde,
                  method = "full",
                  distance = "glm",
                  link = "probit")

bal.tab(m.full, thresholds = c(m = 0.1))
```

Full matching usually achieves much better balance than 1:1 nearest-neighbour.

### Mahalanobis matching

For low-dimensional matching problems, Mahalanobis distance on the raw
covariates is a good alternative:

```{r}
#| label: maha-match
#| cache: true

m.maha <- matchit(treat ~ age + educ + race + married + nodegree + re74 + re75,
                  data = lalonde,
                  distance = "mahalanobis")
bal.tab(m.maha, thresholds = c(m = 0.1))
```

### Coarsened Exact Matching (CEM)

CEM (Iacus, King, Porro 2012) coarsens each covariate into bins, then
matches *exactly* on the coarsened values. The bin widths are typically
chosen automatically. CEM has the advantage that the resulting matched
sample has *guaranteed* covariate balance up to the coarsening level.

```{r}
#| label: cem-match
#| cache: true

m.cem <- matchit(treat ~ age + educ + race + married + nodegree + re74 + re75,
                 data = lalonde,
                 method = "cem")
bal.tab(m.cem, thresholds = c(m = 0.1))
```

CEM often drops treated units that have no exact match — a feature, not a
bug. It is transparent about lack of overlap.

## Estimation after matching

Once a matched dataset is in hand, the treatment effect is estimated as
a *weighted* regression on the matched data, using the weights from
`MatchIt`. Standard errors should be clustered by the subclass:

```{r}
#| label: estimate
#| cache: true

m.data <- match_data(m.full)

fit <- lm(re78 ~ treat * (age + educ + race + married + nodegree + re74 + re75),
          data    = m.data,
          weights = weights)

# Use marginaleffects for the average treatment effect with cluster-by-subclass SEs
avg_comparisons(fit,
                variables = "treat",
                vcov = ~subclass,
                newdata = subset(m.data, treat == 1))   # ATT
```

The estimate from full matching should be much closer to the experimental
benchmark of about \$1,800 (Dehejia & Wahba 1999) than a naive comparison
of means.

## Balance plots

`cobalt::love.plot` produces the standard "love plot" diagnostic showing
absolute standardised mean differences before and after matching:

```{r}
#| label: love-plot
#| cache: true
#| fig-cap: "Balance improvement from nearest-neighbour and full matching."
#| fig-width: 8
#| fig-height: 5

love.plot(m.full,
          stats = "mean.diffs",
          thresholds = c(m = 0.1),
          abs = TRUE,
          var.order = "unadjusted",
          title = "Balance plot — Full matching")
```

A successful matching produces balance below 0.1 (or, more conservatively,
0.05) for every covariate. The love plot makes this immediate.

## Matching with replacement vs without

By default, `matchit(..., replace = FALSE)` uses each control unit at most
once. When the treated group is much larger or the propensity distributions
diverge sharply, matching *with* replacement (`replace = TRUE`) can
improve balance at the cost of effective sample size:

```{r}
#| label: replace-match
#| cache: true

m.repl <- matchit(treat ~ age + educ + race + married + nodegree + re74 + re75,
                  data = lalonde,
                  method = "nearest",
                  distance = "glm",
                  replace = TRUE)
bal.tab(m.repl, thresholds = c(m = 0.1))
```

Matching with replacement is recommended when overlap is poor and balance
is otherwise unachievable.

## ATE, ATT, ATC — choose carefully

The default `MatchIt` estimand is ATT (effect on the treated). To target
the ATE, use `estimand = "ATE"` and a matching method that doesn't drop
units (full matching or weighting). To target ATC (effect on the controls),
use `estimand = "ATC"`. The three estimands answer different policy
questions — see the [Five Estimands chapter](causal-estimands.qmd) for
the framing.

## When matching fails

Matching has known limitations:

- **Curse of dimensionality**: high-dimensional matching is hard. Use
  propensity scores or coarsened exact matching to reduce dimensions.
- **No overlap**: if certain regions of $X$ have no treated (or no control)
  units, no matching method can rescue them. CEM exposes this honestly by
  dropping such units; nearest-neighbour matching hides the problem in
  imbalanced pairs.
- **Sensitivity to specification**: the propensity score model is itself
  a regression that can be misspecified. The [Sensitivity Analysis chapter](sensitivity-analysis.qmd)
  covers Rosenbaum bounds, which specifically address matched-sample
  hidden bias.

## Matching vs weighting

Inverse propensity weighting (IPW) and matching are close cousins.
Matching subsets to comparable units; weighting reweights to make the
treated and control distributions match. Modern theory increasingly
favours weighting for two reasons:

1. **No information loss** — IPW uses every observation; matching may
   drop unmatched units.
2. **Smooth estimands** — IPW's reweighted distribution is exact, while
   matching introduces noise from the matching algorithm.

In practice, the difference is small when overlap is good. When overlap
is poor, both methods break down, and the right response is either to
restrict the population (CEM-style) or to use a doubly-robust estimator
([AIPW](estimation.qmd), [TMLE](nonparametric.qmd)) that combines a
weighting model with an outcome model and is robust to misspecification
of either.

## Modern weighting alternatives

The `WeightIt` package implements modern weighting methods that often
outperform plain IPW, particularly when overlap is marginal. The
companion [blog chapter on weighting](https://xiangao.github.io/blog_book/weighting-part2.html)
covers these in more detail; here we summarise the four most useful
variants and apply them to the Lalonde data already in this chapter.

### Inverse probability of treatment weighting (IPW)

The baseline: estimate the propensity score by logistic regression,
weight treated units by $1/\hat e(X)$ and controls by $1/(1 - \hat e(X))$:

```{r}
#| label: ipw-weighting
#| cache: true

w_ipw <- weightit(treat ~ age + educ + race + married + nodegree + re74 + re75,
                  data = lalonde, estimand = "ATT", method = "glm")
bal.tab(w_ipw, stats = c("m"), thresholds = c(m = 0.05))
```

### Covariate balancing propensity score (CBPS)

CBPS (Imai-Ratkovic 2014) estimates the propensity score under
*balance constraints*, ensuring that the weighted covariate means are
balanced even when the logistic model is misspecified:

```{r}
#| label: cbps
#| cache: true

w_cbps <- weightit(treat ~ age + educ + race + married + nodegree + re74 + re75,
                   data = lalonde, estimand = "ATT", method = "cbps")
bal.tab(w_cbps, stats = c("m"), thresholds = c(m = 0.05))
```

### Entropy balancing

Entropy balancing (Hainmueller 2012) solves a convex optimisation
problem that yields weights with *exact* mean balance on the included
covariates. It is doubly robust for the ATT and attains the
semiparametric efficiency bound when both nuisance models are correct.

```{r}
#| label: ebal
#| cache: true

w_ebal <- weightit(treat ~ age + educ + race + married + nodegree + re74 + re75,
                   data = lalonde, estimand = "ATT", method = "ebal")
bal.tab(w_ebal, stats = c("m"), thresholds = c(m = 0.05))
```

Notice that every standardised mean difference is now essentially zero
— entropy balancing *guarantees* this by construction.

### Energy balancing

Energy balancing (Huling & Mak 2024) optimises a weighted energy
distance between treated and control covariate distributions. Unlike
entropy balancing — which balances *means* — energy balancing balances
the entire *distribution* of every covariate.

```{r}
#| label: energy
#| cache: true

w_energy <- weightit(treat ~ age + educ + race + married + nodegree + re74 + re75,
                     data = lalonde, estimand = "ATT", method = "energy")
bal.tab(w_energy, stats = c("m"), thresholds = c(m = 0.05))
```

### Estimation with weighted-aware regression

Once weights are computed, use `lm_weightit()` (rather than plain `lm`
with a `weights` argument) so that the weighting uncertainty is
propagated into the treatment-effect standard error:

```{r}
#| label: lm-weightit
#| cache: true

fit_ebal <- lm_weightit(
  re78 ~ treat * (age + educ + race + married + nodegree + re74 + re75),
  data     = lalonde,
  weightit = w_ebal
)

avg_comparisons(fit_ebal, variables = "treat",
                newdata = subset(lalonde, treat == 1))   # ATT
```

The result is the entropy-balanced ATT with valid sandwich-style
standard errors that account for the estimated weights.

### When to use each weighting method

| Method | Strengths | Weaknesses |
|---|---|---|
| **IPW** (glm) | Familiar; well-studied | Sensitive to PS misspecification; extreme weights |
| **CBPS** | Balance-constrained PS; robust to model misspecification | Slower; can fail with many covariates |
| **Entropy balancing** | Exact mean balance; doubly robust for ATT | Balances means only, not distributions |
| **Energy balancing** | Balances entire distribution | Computationally heavier; less mature theory |

For applied work, **entropy balancing** is the practical default — it
typically achieves better balance than IPW with fewer extreme weights
and faster runtime than CBPS. **Energy balancing** is the modern
state-of-the-art when full distributional balance matters.

### Comparing matching to weighting

The matching examples earlier in this chapter and the weighting examples
here both target the ATT on the Lalonde data. In practice, an applied
researcher should try several methods, report all balance diagnostics
and ATT estimates, and choose based on which method achieves the best
covariate balance. When matching, weighting, and modern weighting
methods all converge to similar estimates, the identification strategy
is on firm ground. When they diverge sharply, the overlap is poor and
the right response is to restrict the population or use a doubly-robust
estimator.

For an extended treatment of matching with `MatchIt` (including a
comparison with Stata's `teffects` commands), see the
[companion blog chapters on matching](https://xiangao.github.io/blog_book/matching-part1.html)
and [treatment-effects in Stata](https://xiangao.github.io/blog_book/treatment-matching.html).

## Summary

- **Matching** estimates causal effects by pairing treated units to
  similar controls, restricting the comparison to the region of overlap
  in covariates.
- **Choose a distance**: propensity score for high-dimensional problems;
  Mahalanobis for low-dimensional; CEM when exact coarsened balance is
  desired.
- **Choose a method**: nearest-neighbour is simple but often imbalanced;
  full matching uses all observations and usually balances best.
- **Estimate with weighted regression** on the matched dataset, clustering
  SEs by subclass.
- **Diagnose with balance plots** (`love.plot`) before reporting the
  effect.
- For more advanced treatment of `MatchIt` and `WeightIt`, see the
  [companion blog chapter](https://xiangao.github.io/blog_book/matching-part1.html).

## References

- Dehejia, R. H., & Wahba, S. (1999). Causal effects in nonexperimental studies: reevaluating the evaluation of training programs. *JASA*, 94(448), 1053–1062.
- Stuart, E. A. (2010). Matching methods for causal inference: a review and a look forward. *Statistical Science*, 25(1), 1–21.
- Iacus, S. M., King, G., & Porro, G. (2012). Causal inference without balance checking: coarsened exact matching. *Political Analysis*, 20(1), 1–24.
- Greifer, N. (2024). *MatchIt: Getting Started.* CRAN vignette.
- Ho, D. E., Imai, K., King, G., & Stuart, E. A. (2007). Matching as nonparametric preprocessing for reducing model dependence in parametric causal inference. *Political Analysis*, 15(3), 199–236.
