# Matching Estimators

```{r}
#| include: false
library(tidyverse)
library(MatchIt)
library(cobalt)
library(WeightIt)
library(marginaleffects)
library(broom)
```

```{r}
#| eval: false
library(tidyverse)
library(MatchIt)
library(cobalt)
library(WeightIt)
library(marginaleffects)
```

Matching is a way to make treated and control observations more comparable.
For each treated observation, we look for controls with similar covariates.
Then we compare outcomes in this matched sample.

This does not solve endogeneity. Matching only helps with selection on
observables. If the treatment is selected on unobserved variables, matching
will not fix the problem. The value of matching is that it makes the overlap
problem visible. A regression can extrapolate quietly; matching often shows
that there are no good controls for part of the treated sample.

In R, the main workflow is `MatchIt` for matching and `cobalt` for balance
checking.

> **Related reading**: A longer treatment is in
> [Matching and Weighting Part 1](https://xiangao.github.io/blog_book/matching-part1.html)
> of *Topics on Econometrics and Causal Inference*, which credits Noah
> Greifer and coauthors (the `MatchIt`/`WeightIt` package authors).

## Assumptions

Matching needs the same assumptions as other selection-on-observables
methods:

1. **SUTVA** — no interference, no hidden treatment versions.
2. **Ignorability** (unconfoundedness) — conditional on $X$, the
   treatment $D$ is independent of the potential outcomes $(Y(0), Y(1))$.
3. **Overlap** (positivity) — every value of $X$ has positive probability
   of being both treated and untreated.

Even if ignorability is true, the covariate distribution can be very
different in the treated and control groups. Matching tries to repair that
before estimating the effect. If there is no overlap, it should not pretend
there is overlap.

## The Lalonde example

I use the Lalonde job-training data because it is the standard example for
matching. The treatment is job training. The outcome is 1978 earnings. The
data set in `MatchIt` is the observational version, so the treated and
control groups are not balanced at the start.

```{r}
#| label: data
#| cache: true

data("lalonde", package = "MatchIt")
glimpse(lalonde)
```

There are 614 observations, 185 treated and 429 control. The treatment is
`treat`; the outcome is `re78`, 1978 earnings in dollars; the covariates are
`age`, `educ` (years of schooling), `race` (black, hispanic, white), `married`,
`nodegree` (no high-school degree), and `re74` and `re75`, earnings in 1974 and
1975. The earnings variables matter most here, because they pile up at zero and
differ sharply between the groups.

Before matching anything we check balance. `bal.tab` reports the standardized
mean difference for each covariate and flags any above 0.1:

```{r}
#| label: pre-balance
#| cache: true

# Construct a pre-match MatchIt object for balance reporting
m.pre <- matchit(treat ~ age + educ + race + married + nodegree + re74 + re75,
                 data = lalonde, method = NULL,
                 distance = "glm")
bal.tab(m.pre, thresholds = c(m = 0.1))
```

Seven of the nine contrasts exceed 0.1. Earnings in 1974 differ by $-0.72$
standard deviations, the black indicator by 0.64, the white indicator by
$-0.56$, marital status by $-0.32$, age by $-0.31$, earnings in 1975 by
$-0.29$. Only `educ` (0.055) and the hispanic indicator ($-0.083$) pass. The
propensity score itself differs by 1.79 standard deviations. This sample is
nowhere near balanced.

## Distance measures

Matching needs a distance metric. The usual choices are:

1. **Propensity score**: estimate $\hat e(x)=P(D=1 \mid X=x)$ and match on
   the fitted probability. This turns many covariates into one score.
2. **Mahalanobis distance**: match on the raw covariates using the squared
   distance $(x_i-x_j)'\Sigma^{-1}(x_i-x_j)$ (rankings are unchanged by the
   square root). This works better in low dimension.
3. **Hybrid**: impose a propensity-score caliper, then use Mahalanobis
   distance inside the caliper.

## Matching methods

### Nearest-neighbour matching on a propensity score

We fit the propensity score by logistic regression, take its linear index, and
match each treated unit to the single closest control, without replacement.

```{r}
#| label: nn-match
#| cache: true

m.nn <- matchit(treat ~ age + educ + race + married + nodegree + re74 + re75,
                data = lalonde,
                method = "nearest",
                distance = "glm",
                link = "linear.logit",
                ratio = 1)

m.nn
```

All 185 treated units are matched, so 370 of the 614 observations are used and
244 controls are discarded. The target estimand is the ATT.

Now check balance again:

```{r}
#| label: nn-balance
#| cache: true

bal.tab(m.nn, thresholds = c(m = 0.1))
```

Five of nine now pass, up from two. The earnings histories are repaired: `re74`
goes from $-0.72$ to $-0.05$ and `re75` from $-0.29$ to $-0.03$. Race is not:
the black indicator is still 0.373 and the white indicator $-0.216$. And `educ`
has got worse, from 0.055 to $-0.129$. This is what matching on a scalar score
does: it balances the score, not the covariates. Two units with the same score
can differ on individual covariates in offsetting ways, and nothing in the
procedure prevents that.

### Full matching

Full matching forms subclasses with treated and control observations inside
each subclass. It can put one treated unit with several controls, or one
control with several treated units. It uses all observations and then assigns
weights.

```{r}
#| label: full-match
#| cache: true

m.full <- matchit(treat ~ age + educ + race + married + nodegree + re74 + re75,
                  data = lalonde,
                  method = "full",
                  distance = "glm",
                  link = "probit")

bal.tab(m.full, thresholds = c(m = 0.1))
```

All ten contrasts now pass, the largest being `educ` at $-0.096$, and no
observation is discarded. The cost is in the weights: the control effective
sample size falls from 429 to 50.8, because full matching concentrates weight
on the controls that resemble treated units.

The effective sample size is the number of equally weighted observations that
would carry the same precision. An unweighted mean of $m$ observations has
variance $\sigma^2/m$; a weighted mean with weights summing to one has variance
$\sigma^2 \sum_i w_i^2$. Setting the two equal gives $m = 1/\sum_i w_i^2$. So
50.8 means these 429 controls carry about as much information as 51 equally
weighted ones. Balance improved and precision did not.

### Mahalanobis matching

When the number of covariates is small, it is reasonable to match directly
on the covariates with Mahalanobis distance.

```{r}
#| label: maha-match
#| cache: true

m.maha <- matchit(treat ~ age + educ + race + married + nodegree + re74 + re75,
                  data = lalonde,
                  distance = "mahalanobis")
bal.tab(m.maha, thresholds = c(m = 0.1))
```

Mahalanobis matching balances only four of nine. The black and white indicators
sit at $\pm 0.378$, `re74` at $-0.248$, `re75` at $-0.132$, and `age` at 0.127.
It does worse than the propensity score did on earnings. With a three-level
race factor and two earnings variables that pile up at zero, the covariance
matrix is a poor description of the covariate space, and the metric it defines
is a poor guide to who resembles whom.

### Coarsened Exact Matching (CEM)

CEM coarsens covariates into bins and then exact-matches on the binned
values. This is very transparent: if a treated observation has no comparable
control in the binned covariate space, it is dropped.

```{r}
#| label: cem-match
#| cache: true

m.cem <- matchit(treat ~ age + educ + race + married + nodegree + re74 + re75,
                 data = lalonde,
                 method = "cem")
bal.tab(m.cem, thresholds = c(m = 0.1))
```

All nine contrasts pass, and the largest is `age` at 0.049. Read the sample
sizes before celebrating: only 65 of the 185 treated units are matched, and 120
are dropped, along with 354 of the 429 controls. The control effective sample
size is 41.3.

CEM can drop many observations. That is not necessarily a problem. It is
often telling us that the original sample does not support the target
comparison. But it does change the question. The estimand is now the average
effect for the 65 treated units that had a comparable control, not the ATT for
all 185, and those two need not be the same number.

## Estimation after matching

After matching, we estimate the effect on the matched data. The matching
weights come from `MatchIt`. With subclass matching, standard errors should
be clustered by subclass: units within the same matched subclass are not
independent draws -- they were selected together specifically because they
resemble each other on the matching covariates, and (for many-to-one or
full matching) a single unit's outcome can appear, re-weighted, in the
"observation" for more than one comparison. Treating them as independent
(ordinary or heteroskedasticity-robust SEs) understates the true sampling
variability; clustering by subclass accounts for that within-subclass
correlation.

```{r}
#| label: estimate
#| cache: true

m.data <- match_data(m.full)

fit <- lm(re78 ~ treat * (age + educ + race + married + nodegree + re74 + re75),
          data    = m.data,
          weights = weights)

# Use marginaleffects for the average treatment effect with cluster-by-subclass SEs
avg_comparisons(fit,
                variables = "treat",
                vcov = ~subclass,
                newdata = subset(m.data, treat == 1))   # ATT
```

The full-matching ATT is \$1,977 with a subclass-clustered standard error of
\$704, a 95% interval of $[596, 3357]$, and $p = 0.005$. Training raised 1978
earnings for the trained.

The point is not that matching mechanically produces the "right" answer.
The point is that after improving balance, the estimate is much less driven
by obvious covariate differences.

## Balance plots

`cobalt::love.plot` is the easiest way to see the balance change. It plots
standardized mean differences before and after matching.

```{r}
#| label: love-plot
#| cache: true
#| fig-cap: "Balance before and after full matching, as absolute standardized mean differences."
#| fig-width: 8
#| fig-height: 5

love.plot(m.full,
          stats = "mean.diffs",
          thresholds = c(m = 0.1),
          abs = TRUE,
          var.order = "unadjusted",
          title = "Balance plot — Full matching")
```

Each variable appears twice, unadjusted and adjusted, with the variables
ordered by their unadjusted difference and differences plotted in absolute
value. Every point moves toward zero except `educ`, which starts at 0.055 and
ends at 0.096 --- it was already balanced, and full matching traded a little of
it for the rest. `re74`, the worst offender before matching at $-0.72$, moves
furthest, ending at $-0.0009$.

A useful rule is that standardized mean differences should be below 0.1,
or below 0.05 if we want to be stricter.

## Matching with replacement vs without

By default a control observation can be used only once. With replacement,
the same good control can be used several times. This often improves balance
when overlap is weak, but the effective sample size becomes smaller.

```{r}
#| label: replace-match
#| cache: true

m.repl <- matchit(treat ~ age + educ + race + married + nodegree + re74 + re75,
                  data = lalonde,
                  method = "nearest",
                  distance = "glm",
                  replace = TRUE)
bal.tab(m.repl, thresholds = c(m = 0.1))
```

With replacement, nine of ten contrasts pass, against five of nine without.
`age` is the exception, at 0.240. Only 82 distinct controls are used, and their
effective sample size is 46.3. Compared with the 1:1 match without replacement,
which used 185 distinct controls, the same good controls are being reused.

If matching without replacement cannot balance the sample, matching with
replacement is a reasonable next step.

## ATE, ATT, ATC — choose carefully

`MatchIt` usually targets the ATT. That means we are asking what treatment
did for the treated units. For ATE, use `estimand = "ATE"` and a method
that keeps the whole sample, such as full matching or weighting. For ATC,
use `estimand = "ATC"`. These are not just software options; they are
different causal questions.

## When matching fails

Matching fails in predictable ways:

- High-dimensional matching is hard. Propensity scores reduce dimension,
  but they do not create overlap.
- If some values of $X$ have no treated or no control units, no matching
  method can recover the missing comparison.
- Propensity-score matching depends on the propensity-score model. If that
  model is bad, the matching can be bad too.
- Matching does not address hidden confounding. Rosenbaum bounds, discussed
  in the [Sensitivity Analysis chapter](sensitivity-analysis.qmd), ask how
  strong hidden bias would have to be to change the conclusion.

## Matching vs weighting

Matching and weighting do similar jobs. Matching chooses comparable
observations. Weighting keeps observations but changes their contribution so
the treated and control covariate distributions become closer.

When overlap is good, both approaches often give similar answers. When
overlap is bad, both approaches become fragile. Then the honest choices are
to change the target population, report the lack of overlap, or use a
doubly-robust estimator such as [AIPW](estimation.qmd) or
[TMLE](nonparametric.qmd) while still being clear about the support problem.

## Weighting Alternatives

`WeightIt` implements several weighting methods. The companion
[blog chapter on weighting](https://xiangao.github.io/blog_book/weighting-part2.html)
has more detail. Here I use the same Lalonde data so the comparison is easy.

### Inverse probability of treatment weighting (IPW)

The baseline method is IPW. Estimate the propensity score, then weight by
the inverse probability of receiving the observed treatment.

```{r}
#| label: ipw-weighting
#| cache: true

w_ipw <- weightit(treat ~ age + educ + race + married + nodegree + re74 + re75,
                  data = lalonde, estimand = "ATT", method = "glm")
bal.tab(w_ipw, stats = c("m"), thresholds = c(m = 0.05))
```

Note the stricter 0.05 threshold from here on. IPW balances nine of ten;
`age` fails at 0.119. The control effective sample size is 99.8 out of 429,
better than any of the matching methods above.

### Covariate balancing propensity score (CBPS)

CBPS estimates the propensity score while also trying to balance covariate
means. So the propensity-score model is not judged only by likelihood; it is
also judged by balance.

```{r}
#| label: cbps
#| cache: true

w_cbps <- weightit(treat ~ age + educ + race + married + nodegree + re74 + re75,
                   data = lalonde, estimand = "ATT", method = "cbps")
bal.tab(w_cbps, stats = c("m"), thresholds = c(m = 0.05))
```

Every covariate mean difference is zero to three decimals, and the control
effective sample size is 98.5.

### Entropy balancing

Entropy balancing chooses weights so that the weighted control group matches
the treated group on covariate means. It is useful because the balance
constraint is explicit.

```{r}
#| label: ebal
#| cache: true

w_ebal <- weightit(treat ~ age + educ + race + married + nodegree + re74 + re75,
                   data = lalonde, estimand = "ATT", method = "ebal")
bal.tab(w_ebal, stats = c("m"), thresholds = c(m = 0.05))
```

The standardized mean differences are essentially zero because entropy
balancing imposes mean balance directly. The effective sample size is 98.5.

That is the same table CBPS produced, and it is not a coincidence: the two sets
of control weights are numerically identical here, agreeing to twelve decimal
places. Just-identified CBPS for the ATT solves the exact mean-balance moment
conditions with a logistic link, which makes the control weights proportional
to $\exp(x'\lambda)$. Entropy balancing minimises the Kullback-Leibler
divergence from uniform weights subject to the same exact mean-balance
constraints, and its solution has the same $\exp(x'\lambda)$ form. Same
constraints, same functional form, same weights. The two methods are motivated
differently and arrive at one answer.

### Energy balancing

Energy balancing targets the whole covariate distribution, not just means.
It is more ambitious than entropy balancing, and usually more computationally
expensive.

```{r}
#| label: energy
#| cache: true

w_energy <- weightit(treat ~ age + educ + race + married + nodegree + re74 + re75,
                     data = lalonde, estimand = "ATT", method = "energy")
bal.tab(w_energy, stats = c("m"), thresholds = c(m = 0.05))
```

All nine contrasts pass, the largest being `re75` at 0.023 --- close to zero
but not exactly zero, because energy balancing does not impose mean balance as
a constraint. The control effective sample size is 41.8, less than half of
entropy balancing's 98.5. That is the price of targeting the whole covariate
distribution rather than its means.

### Estimation with weighted-aware regression

After computing weights, use `lm_weightit()` instead of plain `lm` if we
want standard errors that account for the estimated weights.

```{r}
#| label: lm-weightit
#| cache: true

fit_ebal <- lm_weightit(
  re78 ~ treat * (age + educ + race + married + nodegree + re74 + re75),
  data     = lalonde,
  weightit = w_ebal
)

avg_comparisons(fit_ebal, variables = "treat",
                newdata = subset(lalonde, treat == 1))   # ATT
```

The entropy-balanced ATT is \$1,273 with a sandwich standard error of \$770,
a 95% interval of $[-236, 2783]$, and $p = 0.098$. The interval includes zero.

### When to use each weighting method

| Method | Strengths | Weaknesses |
|---|---|---|
| **IPW** (glm) | Familiar; well-studied | Sensitive to PS misspecification; extreme weights |
| **CBPS** | Balance-constrained PS; robust to model misspecification | Slower; can fail with many covariates |
| **Entropy balancing** | Exact mean balance; doubly robust for ATT | Balances means only, not distributions |
| **Energy balancing** | Balances entire distribution | Computationally heavier; less mature theory |

For applied work, entropy balancing is often a good default because it is
simple and the balance diagnostics are easy to explain. Energy balancing is
useful when matching the whole covariate distribution is important.

### Comparing matching to weighting

The matching and weighting examples here target the same ATT, so put the two
estimates next to each other rather than reporting whichever came last:

```{r}
#| label: compare-methods
#| cache: true

cmp <- rbind(
  transform(as.data.frame(
    avg_comparisons(fit, variables = "treat", vcov = ~subclass,
                    newdata = subset(m.data, treat == 1))),
    method = "Full matching (subclass-clustered SE)"),
  transform(as.data.frame(
    avg_comparisons(fit_ebal, variables = "treat",
                    newdata = subset(lalonde, treat == 1))),
    method = "Entropy balancing (sandwich SE)")
)

cmp[, c("method", "estimate", "std.error", "conf.low", "conf.high", "p.value")] |>
  knitr::kable(digits = c(0, 0, 0, 0, 0, 3),
               caption = "Two routes to the same ATT on the Lalonde data")
```

The gap is instructive. The two point estimates differ by roughly one standard
error, and one interval excludes zero while the other does not --- a reader who
sees only one of these tables would draw a different conclusion about whether
the programme worked. Note also that the ranking on balance does not settle it:
entropy balancing imposes *exact* mean balance (every standardized difference
above is at most 0.023), whereas full matching leaves `educ` at $-0.096$, yet it
is entropy balancing that gives the smaller estimate and the wider interval.
Better mean balance is not the same thing as a more precise or more credible
answer.

This is model dependence, not a bug in either method. In an applied paper I would
report balance for each method, not just the final coefficient, and I would report
both estimates. If several reasonable methods give similar balance and similar
estimates, the result is more credible. If they differ sharply, as here, the
problem is usually weak overlap or model dependence, and the honest summary says
so instead of picking the number that crosses the significance line.

For more details on `MatchIt` and the Stata `teffects` commands, see the
[matching blog chapter](https://xiangao.github.io/blog_book/matching-part1.html)
and [treatment-effects in Stata](https://xiangao.github.io/blog_book/treatment-matching.html).

## Matching, weighting, and direct estimation

The estimation chapter estimated a treatment effect by regression adjustment,
IPW, AIPW and IPWRA. This chapter estimated one by matching and by four
weighting methods. Those are not two families of estimators. They are one
estimand and several ways of building it, and it is worth saying how they line
up.

The ATT is

$$
\tau_{ATT} = E[Y \mid W = 1] - E_{X \mid W = 1}\bigl[\, E[Y \mid W = 0, X] \,\bigr].
$$ {#eq-matching-att-routes}

The first term is observed. It is the mean outcome among the treated. The
second term is not observed. It is the mean outcome the treated would have had
untreated, averaged over the covariate distribution of the treated. Every
method in both chapters is a way of forming that second term, and every one of
them ends up as a weighted average of control outcomes,
$\sum_{i: W_i = 0} w_i Y_i$. What differs is where the weights come from.

| Route | What is modelled | Weight on control $i$ |
|---|---|---|
| Regression adjustment | the outcome, $\mu_0(x) = E[Y \mid W=0, X=x]$ | implicit in the fitted values used to predict at treated $x$ |
| IPW | the propensity score $\pi(x)$ | the odds $\pi(x_i)/(1-\pi(x_i))$ |
| Matching | neither; a distance rule | 0 if unmatched, else 1, the number of times reused, or the subclass ratio |
| Entropy, CBPS, energy balancing | the weights themselves | whatever satisfies the balance constraints |
| AIPW | both | the odds, applied to the residual $Y - \hat\mu_0(X)$ |

Weighting is IPW. That is the second row, and `weightit` with
`estimand = "ATT"` and `method = "glm"` returns exactly those odds. CBPS,
entropy balancing and energy balancing do not change the estimator. They
change how the weights are found. IPW fits $\pi$ by maximum likelihood and
then transforms it into weights; the balancing methods skip the likelihood and
solve for the weights directly. For entropy balancing the link is exact rather
than an analogy: the control weights are proportional to $\exp(x'\lambda)$, as
the CBPS comparison above showed, and that is the odds of a logistic
propensity score. Entropy balancing is IPW for a logit model whose
coefficients are set by exact mean balance instead of by likelihood.

Matching is weighting too, with the weights restricted to zero and small
integers and chosen by a distance rule rather than by a model. That is what
lets `match_data` hand a weights column to `lm`. The restriction is where the
trade-off lives. Matching sets some weights to zero, so it discards
observations, and when it discards treated units --- as CEM did above, keeping
65 of 185 --- the estimand becomes the ATT among the matched treated. Weighting
keeps everyone, so the estimand survives, and pays for it with extreme weights
when overlap is poor, which is what the effective sample sizes in this chapter
have been reporting. Matching buys stability with the estimand. Weighting buys
the estimand with variance.

AIPW uses both columns. Start from regression adjustment, then add the
odds-weighted average of the control residuals $Y - \hat\mu_0(X)$. It is
consistent if either model is right. The section on estimation after matching
did an informal version of the same thing: match first, then run a regression
with treatment-covariate interactions on the matched sample. That is
@abadie-imbens-2011 bias-corrected matching, and the regression plays the part of
the augmentation term.

Five routes, one data set. Each row below is built by hand from the weight
representation so the arithmetic is visible.

```{r}
#| label: compare-routes
#| cache: true

covs <- c("age", "educ", "race", "married", "nodegree", "re74", "re75")
trt  <- lalonde$treat == 1
y    <- lalonde$re78

# Outcome model, fitted on controls only, then predicted everywhere
mu0_fit <- lm(reformulate(covs, response = "re78"),
              data = subset(lalonde, treat == 0))
mu0 <- predict(mu0_fit, newdata = lalonde)

# Odds weights from a logistic propensity score
ps     <- predict(glm(reformulate(covs, response = "treat"),
                      data = lalonde, family = binomial), type = "response")
odds   <- ps / (1 - ps)
wt_ipw <- ifelse(trt, 1, odds)

# Weighted difference in means: the common shape of rows 2 to 4
hajek <- function(w) {
  weighted.mean(y[trt], w[trt]) - weighted.mean(y[!trt], w[!trt])
}

att_ra   <- mean(y[trt] - mu0[trt])
att_ipw  <- hajek(wt_ipw)
att_eb   <- hajek(w_ebal$weights)
att_fm   <- hajek(m.full$weights)
att_aipw <- att_ra - weighted.mean(y[!trt] - mu0[!trt], odds[!trt])

data.frame(
  route = c("Regression adjustment (outcome model)",
            "IPW (propensity model)",
            "Entropy balancing (balance constraints)",
            "Full matching (distance rule)",
            "AIPW (both models)"),
  att = c(att_ra, att_ipw, att_eb, att_fm, att_aipw)
) |>
  knitr::kable(digits = 0, col.names = c("Route", "ATT"),
               caption = "Five routes to the same ATT on the Lalonde data.")
```

Two of these numbers confirm the algebra directly. The hand-built odds weights
agree with the `weightit` propensity-score weights from the IPW section above to
machine precision, so the claim that weighting is IPW is an identity here and not
a resemblance. And entropy balancing gives \$1,273 --- the same figure that
`lm_weightit` produced earlier with the full set of treatment-covariate
interactions, agreeing to nine decimal places. Exact mean balance leaves nothing for a linear
outcome model to correct, so adding one changes the estimate not at all. That is
the double robustness of entropy balancing [@zhao-percival-2017] appearing as
arithmetic: the same weights deliver the IPW estimator for a logit propensity
model and the regression-adjustment estimator for a linear outcome model.

The spread across routes is the part to read carefully. Regression adjustment
gives \$1,648 and IPW gives \$1,214. AIPW starts from the regression-adjustment
figure and the augmentation term subtracts \$417, landing at \$1,231 --- next to
IPW, not next to regression adjustment. The propensity side is doing the work
here, which is what we should expect when the treated and control covariate
distributions are as far apart as the balance tables in this chapter have shown.
Full matching sits highest at \$1,855. It fits a propensity score too --- a
probit, used as the distance for forming subclasses --- so the gap between it and
IPW's \$1,214 is not about whether a propensity model was estimated. It is about
how the score is turned into weights: coarse subclass weights that throw away the
score's fine gradations, against the odds transformation that keeps them.

These are point estimates. Standard errors are not comparable across the rows
and are deliberately left out: matching is not a smooth functional of the data,
so the bootstrap is invalid for it [@abadie-imbens-2008] and the subclass
clustering used earlier is needed, while IPW and AIPW have influence-function
variances and AIPW attains the semiparametric efficiency bound that fixed-$M$
matching does not. The lesson from the column of estimates is the same one the
matching-versus-weighting comparison gave: with overlap this weak, the choice
of route moves the answer by more than one standard error, so report several
routes rather than the one that reads best.

## Summary

- Matching is for selection on observables. It does not fix unobserved
  confounding.
- The main diagnostic is covariate balance, not the treatment-effect
  coefficient.
- Propensity-score matching is simple, but full matching or weighting often
  balances better.
- If there is no overlap, change the estimand or restrict the sample. Do not
  hide the support problem.
- Matching, weighting, and the direct estimators of the previous chapter are
  one estimand built several ways. Weighting with `method = "glm"` is IPW, and
  matching is weighting with the weights restricted to zero and small integers.
