# Matching Estimators

```{r}
#| include: false
library(tidyverse)
library(MatchIt)
library(cobalt)
library(WeightIt)
library(marginaleffects)
library(broom)
```

```{r}
#| eval: false
library(tidyverse)
library(MatchIt)
library(cobalt)
library(WeightIt)
library(marginaleffects)
```

Matching is a way to make treated and control observations more comparable.
For each treated observation, we look for controls with similar covariates.
Then we compare outcomes in this matched sample.

This does not solve endogeneity. Matching only helps with selection on
observables. If the treatment is selected on unobserved variables, matching
will not fix the problem. The value of matching is that it makes the overlap
problem visible. A regression can extrapolate quietly; matching often shows
that there are no good controls for part of the treated sample.

In R, the main workflow is `MatchIt` for matching and `cobalt` for balance
checking.

> **Related reading**: A longer treatment is in
> [Matching and Weighting Part 1](https://xiangao.github.io/blog_book/matching-part1.html)
> of *Topics on Econometrics and Causal Inference*, which credits Noah
> Greifer and coauthors (the `MatchIt`/`WeightIt` package authors).

## Assumptions

Matching needs the same assumptions as other selection-on-observables
methods:

1. **SUTVA** — no interference, no hidden treatment versions.
2. **Ignorability** (unconfoundedness) — conditional on $X$, the
   treatment $D$ is independent of the potential outcomes $(Y(0), Y(1))$.
3. **Overlap** (positivity) — every value of $X$ has positive probability
   of being both treated and untreated.

Even if ignorability is true, the covariate distribution can be very
different in the treated and control groups. Matching tries to repair that
before estimating the effect. If there is no overlap, it should not pretend
there is overlap.

## The Lalonde example

I use the Lalonde job-training data because it is the standard example for
matching. The treatment is job training. The outcome is 1978 earnings. The
data set in `MatchIt` is the observational version, so the treated and
control groups are not balanced at the start.

```{r}
#| label: data
#| cache: true

data("lalonde", package = "MatchIt")
glimpse(lalonde)
```

The treatment is `treat`; the outcome is `re78`; the covariates are `age`,
`educ`, `race`, `married`, `nodegree`, `re74`, and `re75`.

First check balance before doing any matching:

```{r}
#| label: pre-balance
#| cache: true

# Construct a pre-match MatchIt object for balance reporting
m.pre <- matchit(treat ~ age + educ + race + married + nodegree + re74 + re75,
                 data = lalonde, method = NULL,
                 distance = "glm")
bal.tab(m.pre, thresholds = c(m = 0.1))
```

Several standardized mean differences are larger than 0.1. That is usually
too large to call the sample balanced.

## Distance measures

Matching needs a distance metric. The usual choices are:

1. **Propensity score**: estimate $\hat e(x)=P(D=1 \mid X=x)$ and match on
   the fitted probability. This turns many covariates into one score.
2. **Mahalanobis distance**: match on the raw covariates using the squared
   distance $(x_i-x_j)'\Sigma^{-1}(x_i-x_j)$ (rankings are unchanged by the
   square root). This works better in low dimension.
3. **Hybrid**: impose a propensity-score caliper, then use Mahalanobis
   distance inside the caliper.

## Matching methods

### Nearest-neighbour matching on a propensity score

```{r}
#| label: nn-match
#| cache: true

m.nn <- matchit(treat ~ age + educ + race + married + nodegree + re74 + re75,
                data = lalonde,
                method = "nearest",
                distance = "glm",
                link = "linear.logit",
                ratio = 1)

m.nn
```

Now check balance again:

```{r}
#| label: nn-balance
#| cache: true

bal.tab(m.nn, thresholds = c(m = 0.1))
```

Some covariates are still not balanced. Nearest-neighbor matching on a
logit propensity score is simple, but it is not automatically good.

### Full matching

Full matching forms subclasses with treated and control observations inside
each subclass. It can put one treated unit with several controls, or one
control with several treated units. It uses all observations and then assigns
weights.

```{r}
#| label: full-match
#| cache: true

m.full <- matchit(treat ~ age + educ + race + married + nodegree + re74 + re75,
                  data = lalonde,
                  method = "full",
                  distance = "glm",
                  link = "probit")

bal.tab(m.full, thresholds = c(m = 0.1))
```

In this example full matching gives better balance than simple 1:1 matching.

### Mahalanobis matching

When the number of covariates is small, it is reasonable to match directly
on the covariates with Mahalanobis distance.

```{r}
#| label: maha-match
#| cache: true

m.maha <- matchit(treat ~ age + educ + race + married + nodegree + re74 + re75,
                  data = lalonde,
                  distance = "mahalanobis")
bal.tab(m.maha, thresholds = c(m = 0.1))
```

### Coarsened Exact Matching (CEM)

CEM coarsens covariates into bins and then exact-matches on the binned
values. This is very transparent: if a treated observation has no comparable
control in the binned covariate space, it is dropped.

```{r}
#| label: cem-match
#| cache: true

m.cem <- matchit(treat ~ age + educ + race + married + nodegree + re74 + re75,
                 data = lalonde,
                 method = "cem")
bal.tab(m.cem, thresholds = c(m = 0.1))
```

CEM can drop many observations. That is not necessarily a problem. It is
often telling us that the original sample does not support the target
comparison.

## Estimation after matching

After matching, we estimate the effect on the matched data. The matching
weights come from `MatchIt`. With subclass matching, standard errors should
be clustered by subclass.

```{r}
#| label: estimate
#| cache: true

m.data <- match_data(m.full)

fit <- lm(re78 ~ treat * (age + educ + race + married + nodegree + re74 + re75),
          data    = m.data,
          weights = weights)

# Use marginaleffects for the average treatment effect with cluster-by-subclass SEs
avg_comparisons(fit,
                variables = "treat",
                vcov = ~subclass,
                newdata = subset(m.data, treat == 1))   # ATT
```

The point is not that matching mechanically produces the "right" answer.
The point is that after improving balance, the estimate is much less driven
by obvious covariate differences.

## Balance plots

`cobalt::love.plot` is the easiest way to see the balance change. It plots
standardized mean differences before and after matching.

```{r}
#| label: love-plot
#| cache: true
#| fig-cap: "Balance improvement from nearest-neighbour and full matching."
#| fig-width: 8
#| fig-height: 5

love.plot(m.full,
          stats = "mean.diffs",
          thresholds = c(m = 0.1),
          abs = TRUE,
          var.order = "unadjusted",
          title = "Balance plot — Full matching")
```

A useful rule is that standardized mean differences should be below 0.1,
or below 0.05 if we want to be stricter.

## Matching with replacement vs without

By default a control observation can be used only once. With replacement,
the same good control can be used several times. This often improves balance
when overlap is weak, but the effective sample size becomes smaller.

```{r}
#| label: replace-match
#| cache: true

m.repl <- matchit(treat ~ age + educ + race + married + nodegree + re74 + re75,
                  data = lalonde,
                  method = "nearest",
                  distance = "glm",
                  replace = TRUE)
bal.tab(m.repl, thresholds = c(m = 0.1))
```

If matching without replacement cannot balance the sample, matching with
replacement is a reasonable next step.

## ATE, ATT, ATC — choose carefully

`MatchIt` usually targets the ATT. That means we are asking what treatment
did for the treated units. For ATE, use `estimand = "ATE"` and a method
that keeps the whole sample, such as full matching or weighting. For ATC,
use `estimand = "ATC"`. These are not just software options; they are
different causal questions.

## When matching fails

Matching fails in predictable ways:

- High-dimensional matching is hard. Propensity scores reduce dimension,
  but they do not create overlap.
- If some values of $X$ have no treated or no control units, no matching
  method can recover the missing comparison.
- Propensity-score matching depends on the propensity-score model. If that
  model is bad, the matching can be bad too.
- Matching does not address hidden confounding. Rosenbaum bounds, discussed
  in the [Sensitivity Analysis chapter](sensitivity-analysis.qmd), ask how
  strong hidden bias would have to be to change the conclusion.

## Matching vs weighting

Matching and weighting do similar jobs. Matching chooses comparable
observations. Weighting keeps observations but changes their contribution so
the treated and control covariate distributions become closer.

When overlap is good, both approaches often give similar answers. When
overlap is bad, both approaches become fragile. Then the honest choices are
to change the target population, report the lack of overlap, or use a
doubly-robust estimator such as [AIPW](estimation.qmd) or
[TMLE](nonparametric.qmd) while still being clear about the support problem.

## Weighting Alternatives

`WeightIt` implements several weighting methods. The companion
[blog chapter on weighting](https://xiangao.github.io/blog_book/weighting-part2.html)
has more detail. Here I use the same Lalonde data so the comparison is easy.

### Inverse probability of treatment weighting (IPW)

The baseline method is IPW. Estimate the propensity score, then weight by
the inverse probability of receiving the observed treatment.

```{r}
#| label: ipw-weighting
#| cache: true

w_ipw <- weightit(treat ~ age + educ + race + married + nodegree + re74 + re75,
                  data = lalonde, estimand = "ATT", method = "glm")
bal.tab(w_ipw, stats = c("m"), thresholds = c(m = 0.05))
```

### Covariate balancing propensity score (CBPS)

CBPS estimates the propensity score while also trying to balance covariate
means. So the propensity-score model is not judged only by likelihood; it is
also judged by balance.

```{r}
#| label: cbps
#| cache: true

w_cbps <- weightit(treat ~ age + educ + race + married + nodegree + re74 + re75,
                   data = lalonde, estimand = "ATT", method = "cbps")
bal.tab(w_cbps, stats = c("m"), thresholds = c(m = 0.05))
```

### Entropy balancing

Entropy balancing chooses weights so that the weighted control group matches
the treated group on covariate means. It is useful because the balance
constraint is explicit.

```{r}
#| label: ebal
#| cache: true

w_ebal <- weightit(treat ~ age + educ + race + married + nodegree + re74 + re75,
                   data = lalonde, estimand = "ATT", method = "ebal")
bal.tab(w_ebal, stats = c("m"), thresholds = c(m = 0.05))
```

The standardized mean differences are essentially zero because entropy
balancing imposes mean balance directly.

### Energy balancing

Energy balancing targets the whole covariate distribution, not just means.
It is more ambitious than entropy balancing, and usually more computationally
expensive.

```{r}
#| label: energy
#| cache: true

w_energy <- weightit(treat ~ age + educ + race + married + nodegree + re74 + re75,
                     data = lalonde, estimand = "ATT", method = "energy")
bal.tab(w_energy, stats = c("m"), thresholds = c(m = 0.05))
```

### Estimation with weighted-aware regression

After computing weights, use `lm_weightit()` instead of plain `lm` if we
want standard errors that account for the estimated weights.

```{r}
#| label: lm-weightit
#| cache: true

fit_ebal <- lm_weightit(
  re78 ~ treat * (age + educ + race + married + nodegree + re74 + re75),
  data     = lalonde,
  weightit = w_ebal
)

avg_comparisons(fit_ebal, variables = "treat",
                newdata = subset(lalonde, treat == 1))   # ATT
```

This gives the entropy-balanced ATT and a sandwich-style standard error.

### When to use each weighting method

| Method | Strengths | Weaknesses |
|---|---|---|
| **IPW** (glm) | Familiar; well-studied | Sensitive to PS misspecification; extreme weights |
| **CBPS** | Balance-constrained PS; robust to model misspecification | Slower; can fail with many covariates |
| **Entropy balancing** | Exact mean balance; doubly robust for ATT | Balances means only, not distributions |
| **Energy balancing** | Balances entire distribution | Computationally heavier; less mature theory |

For applied work, entropy balancing is often a good default because it is
simple and the balance diagnostics are easy to explain. Energy balancing is
useful when matching the whole covariate distribution is important.

### Comparing matching to weighting

The matching and weighting examples here target the same ATT. In an applied
paper I would report balance for each method, not just the final coefficient.
If several reasonable methods give similar balance and similar estimates, the
result is more credible. If they differ sharply, the problem is usually weak
overlap or model dependence.

For more details on `MatchIt` and the Stata `teffects` commands, see the
[matching blog chapter](https://xiangao.github.io/blog_book/matching-part1.html)
and [treatment-effects in Stata](https://xiangao.github.io/blog_book/treatment-matching.html).

## Summary

- Matching is for selection on observables. It does not fix unobserved
  confounding.
- The main diagnostic is covariate balance, not the treatment-effect
  coefficient.
- Propensity-score matching is simple, but full matching or weighting often
  balances better.
- If there is no overlap, change the estimand or restrict the sample. Do not
  hide the support problem.
