# Causal Mediation Analysis

```{r}
#| include: false
library(ggplot2)
library(ggdag)
library(lavaan)
library(tidyverse)
library(data.table)
library(medoutcon)
```

```{r dag4, warning=FALSE, cache=TRUE, message=FALSE, echo=FALSE}
g <- dagify(
  y ~ m,
  w ~ c,
  y ~ c,
  m ~ w,
  m ~ c,
  exposure = "w",
  outcome = "y",
  coords = list(x = c(m = 2, c = 2, y = 4, w=0),
                y = c(m = 2, c = 3, y = 2, w=2))
)
ggdag(g)
```

## Classical Mediation

Traditionally mediation model can be represented in the following equations:

$$ Y = a W + b M + \epsilon_1 $$ $$ M = c W + \epsilon_2 $$

That is, we'd like to study the effect of $W$ on $Y$, and we see the effect can be a direct effect, and an indirect effect, through $M$.

Baron and Kenny's (http://davidakenny.net/cm/mediate.htm) method is done in four steps. Modern approach tends to use SEM (structural equation modeling) to model these two equations directly.

```{r med1, warning=FALSE, cache=TRUE, message=FALSE, echo=TRUE}
library(lavaan)
set.seed(1234)
n <- 10000
X <- rnorm(n)
M <- 0.5*X + rnorm(n)
Y <- 0.7*M + 0.3*X + rnorm(n)
Data <- data.frame(X = X, Y = Y, M = M)
model <- '   Y ~ a*X + b*M
             M ~ c*X
           # direct effect (a)
           # indirect effect (b*c)
             bc := b*c
           # total effect
             total := a + (b*c)
         '
```

```{r med2, warning=FALSE, cache=TRUE, message=FALSE, echo=TRUE}
fit <- sem(model, data = Data)
summary(fit)
```

### Problems with Classical Mediation

-   Lack of causal claim. We have to assume that there is no unmeasured confounder between $M$ and $Y$. This is a strong assumption.
-   Assumption of homogeneous effect.

## Causal Mediation

### CDE

Suppose we can set $W$ and $M$ at will to any $(w, m)$, then we have the potential outcome $Y(w,m)$.

Conditional direct effect (CDE) is defined as $$ CDE(m) = E[Y(1,m)-Y(0,m)] $$ That is, setting $M$ to $m$, what is the effect of $W$ on $Y$?

### Assumptions

-   Conditional treatment randomization. Suppose we observe confounders X, which can be a joint set of confounders for the W to Y pathway, or the M to Y pathway.

$$  Y(w,m) \perp W|X  $$

This is the usual conditional independence (unconfoundedness, ignorability, etc.) assumption. This is saying the assignment of treatment, given covariates $X$, has nothing to do with potential outcomes.

-   Conditional mediator randomization.

$$  Y(a,m) \perp M|X, W=w  $$

This is to say, within each strata of X, given treatment status, the assignment of mediator gives no information about potential outcome. This randomization is usually not implemented during many experiments (random trials). This is the assumption that makes a lot of mediation hard to make a causal claim.

-   Positivity (overlap). There are two positivity assumptions:

$$  P(W=w | X=x) > 0  $$ for all $x$. This is the usual positivity assumption.

$$  P(M=m | X=x, W=w) > 0   $$ for all $w$. This is the mediator positivity.

### Estimation

CDE can be estimated using G-computation method, or IPW, or the doubly-robust methods, one of them is AIPW (augmented IPW).

### G-computation

G-computation is to model the outcome equation: $$ \small
\begin{eqnarray*}
CDE_G(m) &=& E[ E[Y(W=1, M=m, X=x)] - \\
&&  E[Y(W=0, M=m, X=x)]] \\
&=& \sum_x [ E[Y(W=1, M=m, X=x)] - \\
&&  E[Y(W=0, M=m, X=x)]]  P(X=x)
\end{eqnarray*}
$$

### IPW

IPW is to model the treatment assignment equation and the mediator assignment equation:

$$ \small
E[Y(w,m)] = E[{\frac{I(W=w, M=m)}{P(W=w, M=m | X)}} Y]
$$

Therefore,

$$ \small
CDE_{ipw}(m) = E[{\frac{I(W=1,M=m)}{g_M(m|1,X)g_W(1,X)} Y - \frac{I(W=0,M=m)}{g_M(m|0,X)g_W(0,X)}} Y]
$$

where $g_M$ is the probability of mediator, $g_W$ is the probability of treatment.

### AIPW

$$ CDE_{AIPW} = CDE_G(m) + B(\bar Q, g_M, g_W) $$ where $\bar Q$ is the mean outcome function.

$$ \small
\begin{eqnarray*}
B(\bar Q, g_M, g_W) &=& \frac{1}{n} \sum_n {\frac{I(M=m, W=1)}{g_m(m|1, X) g_W(1|X)}[Y-\bar Q(m,1,X)]} \\
&  &  - \frac{1}{n} \sum_n {\frac{I(M=m, W=0)}{g_m(m|0, X) g_W(0|X)}[Y-\bar Q(m,0,X)]}
\end{eqnarray*}
$$

### Example: G-computation

```{r med3, warning=FALSE, cache=TRUE, message=FALSE, echo=TRUE}
set.seed(1234)
n <- 5000
# confounder of A/Y
W1 <- rnorm(n)
# confounder of M/Y
W2 <- rnorm(n)
# treatment
A <- rbinom(n, 1, plogis(-1 + W1 / 2))
# binary mediator
M <- rbinom(n, 1, plogis(-2 + A / 2 + W2 / 3))
# binary outcome
Y <- rbinom(n, 1, plogis(-1 + A - M / 2 + W1 / 3 + W2 / 3))
full_data <- data.frame(W1 = W1, W2 = W2, A = A, M = M, Y = Y)

# fit outcome regression
or_fit <- glm(Y ~ A + M + W1 + W2, family = binomial(), data = full_data)
# new data setting A and M
data_A1_M0 <- data_A0_M0 <- full_data
data_A1_M0$A <- 1; data_A1_M0$M <- 0
data_A0_M0$A <- 0; data_A0_M0$M <- 0
# predict on new data
Qbar_A1_M0 <- predict(or_fit, newdata = data_A1_M0, type = "response")
Qbar_A0_M0 <- predict(or_fit, newdata = data_A0_M0, type = "response")
# gcomp estimate of CDE(0)
mean(Qbar_A1_M0 - Qbar_A0_M0)
```

### Example: IPW

```{r med4, warning=FALSE, cache=TRUE, message=FALSE, echo=TRUE}
# model for P(A = 1 | W)
ps_fit1 <- glm(A ~ W1 + W2, family = binomial(), data = full_data)
P_A1_W <- predict(ps_fit1, type = "response")
P_A0_W <- 1 - P_A1_W
# model for P(M = 0 | A, W)
ps_fit2 <- glm(M ~ A + W1 + W2, family = binomial(), data = full_data)
# P(M = 0 | A = 1, W)
data_A1 <- full_data; data_A1$A <- 1
P_M0_A1_W <- 1 - predict(ps_fit2, newdata = data_A1, type = "response")
# P(M = 0 | A = 0, W)
data_A0 <- full_data; data_A0$A <- 0
P_M0_A0_W <- 1 - predict(ps_fit2, newdata = data_A0, type = "response")
# ipw estimate of CDE(0)
mean( (A == 1) / P_A1_W * (M == 0) / P_M0_A1_W * Y ) -
mean( (A == 0) / P_A0_W * (M == 0) / P_M0_A0_W * Y )
```

### Example: AIPW

```{r med5, warning=FALSE, cache=TRUE, message=FALSE, echo=TRUE}
# aipw estimate of E[Y(1,0)]
aiptw_EY_A1_M0 <- mean(Qbar_A1_M0) +
mean( (A == 1) / P_A1_W * (M == 0) / P_M0_A1_W * (Y - Qbar_A1_M0) )
# aipw estimate of E[Y(0,0)]
aiptw_EY_A0_M0 <- mean(Qbar_A0_M0) +
mean( (A == 0) / P_A0_W * (M == 0) / P_M0_A0_W * (Y - Qbar_A0_M0) )
# aipw estimate of CDE(0)
aiptw_EY_A1_M0 - aiptw_EY_A0_M0
```

## NIE and NDE: Natural Direct and Indirect Effects

CDE is to study the effect of treatment, given the level of mediator. Instead, Natural Effect is to set mediator to its natural value with the value of treatment, that is, $M=M(w)$.

$$
\begin{eqnarray*}
ATE &=& NIE + NDE \\
&=&  (E[ Y(1,M(1))] - E[Y(1,M(0))]) + \\
&& (E[Y(1, M(0))] - E[Y(0,M(0))])
\end{eqnarray*}
$$

The advantage of NDE and NIE comparing to CDE is that it's more "natural"; that is, you don't set the level of mediator deterministically. And it can decompose the ATE into direct and indirect effects.

However, there is an additional assumption for NDE and NIE identified.

### Additional Assumption

$$ Y(w, m) \perp M(w^*) | X$$

This is the "cross-world" condition: the outcome under $(w,m)$ is independent of $M$ under $w^*$. These two situations cannot happen in the same world; you cannot set $W$ to both $w$ and $w^*$. There is no experiment can implement it.

### Estimation

```{r med6, warning=FALSE, cache=TRUE, message=FALSE, echo=TRUE}
# fit outcome regression (include interaction because we can)
or_fit <- glm(Y ~ A + M + W1 + W2 + A*M + M*W1,
family = binomial(), data = full_data)
# need E(Y | A = 0/1, M = 0/1, W1 = W1i, W2 = W2i)
get_EY_a_m_Wi <- function(full_data, or_fit, a, m){
data_Aa_Mm_Wi <- full_data
data_Aa_Mm_Wi$A <- a; data_Aa_Mm_Wi$M <- m
predict(or_fit, newdata = data_Aa_Mm_Wi, type = "response")
}
EY_A0_M0_Wi <- get_EY_a_m_Wi(full_data, or_fit, a = 0, m = 0)
EY_A0_M1_Wi <- get_EY_a_m_Wi(full_data, or_fit, a = 0, m = 1)
EY_A1_M0_Wi <- get_EY_a_m_Wi(full_data, or_fit, a = 1, m = 0)
EY_A1_M1_Wi <- get_EY_a_m_Wi(full_data, or_fit, a = 1, m = 1)

# include interactions -- why not?
med_fit <- glm(M ~ A*W1 + W1*W2, family = binomial(), data = full_data)
# estimates of P(M = m | A = a, W = W_i)
get_Pm_a_Wi <- function(full_data, med_fit, a, m){
data_Aa_Wi <- full_data; data_Aa_Wi$A <- a
p <- predict(med_fit, newdata = data_Aa_Wi, type = "response")
if(m == 1){
p
}else{
1 - p
}
}
PM0_A0_Wi <- get_Pm_a_Wi(full_data, med_fit, a = 0, m = 0)
PM1_A0_Wi <- get_Pm_a_Wi(full_data, med_fit, a = 0, m = 1)
PM0_A1_Wi <- get_Pm_a_Wi(full_data, med_fit, a = 1, m = 0)
PM1_A1_Wi <- get_Pm_a_Wi(full_data, med_fit, a = 1, m = 1)

# E(E(Y | A = 1, M, W) | A = 1, W)
EY1M1_Wi <- EY_A1_M1_Wi * PM1_A1_Wi + EY_A1_M0_Wi * PM0_A1_Wi
# E(E(Y | A = 0, M, W) | A = 1, W)
EY0M1_Wi <- EY_A0_M1_Wi * PM1_A1_Wi + EY_A0_M0_Wi * PM0_A1_Wi
# E(E(Y | A = 1, M, W) | A = 0, W)
EY1M0_Wi <- EY_A1_M1_Wi * PM1_A0_Wi + EY_A1_M0_Wi * PM0_A0_Wi
# E(E(Y | A = 0, M, W) | A = 0, W)
EY0M0_Wi <- EY_A0_M1_Wi * PM1_A0_Wi + EY_A0_M0_Wi * PM0_A0_Wi

# estimate of E[Y(1, M(1))]
E_Y1M1 <- mean(EY1M1_Wi)
# estimate of E[Y(0, M(1))]
E_Y0M1 <- mean(EY0M1_Wi)
# estimate of E[Y(1, M(0))]
E_Y1M0 <- mean(EY1M0_Wi)
# estimate of E[Y(0, M(0))]
E_Y0M0 <- mean(EY0M0_Wi)
```

## IIE and IDE: Interventional Direct and Indirect Effects

People are not happy with the cross-world assumption in general. Interventional direct and indirect effects are introduced to avoid this assumption and still be able to decompose the $ATE$.

$$
\begin{eqnarray*}
ATE &=& IIE + IDE \\
&=& (E[ Y(1,M(1))] - E[Y(1,M^*)]) \\
&& + (E[Y(1,M^*)]-E[Y(0,M(0))])
\end{eqnarray*}
$$

The different point here is to set $M=M^*$, where $M^*$ is a random draw from $M(w^*) | X=x$. That is, there is a distribution of $M$ in the strata that $X=x$. We take a random draw, instead of setting to a specific value.

The advantage of this is that it does not need cross-world assumption to identify $IIE$ and $IDE$.

### Example

```{r med7, warning=FALSE, cache=TRUE, message=FALSE, echo=TRUE}
library(data.table)
library(tidyverse)
library(medoutcon)
set.seed(1584)

# produces a simple data set based on ca causal model with mediation
make_example_data <- function(n_obs = 1000) {
  ## baseline covariates
  w_1 <- rbinom(n_obs, 1, prob = 0.6)
  w_2 <- rbinom(n_obs, 1, prob = 0.3)
  w_3 <- rbinom(n_obs, 1, prob = pmin(0.2 + (w_1 + w_2) / 3, 1))
  w <- cbind(w_1, w_2, w_3)
  w_names <- paste("W", seq_len(ncol(w)), sep = "_")

  ## exposure
  a <- as.numeric(rbinom(n_obs, 1, plogis(rowSums(w) - 2)))

  ## mediator-outcome confounder affected by treatment
  z <- rbinom(n_obs, 1, plogis(rowMeans(-log(2) + w - a) + 0.2))

  ## mediator -- could be multivariate
  m <- rbinom(n_obs, 1, plogis(rowSums(log(3) * w[, -3] + a - z)))
  m_names <- "M"

  ## outcome
  y <- rbinom(n_obs, 1, plogis(1 / (rowSums(w) - z + a + m)))

  ## construct output
  dat <- as.data.table(cbind(w = w, a = a, z = z, m = m, y = y))
  setnames(dat, c(w_names, "A", "Z", m_names, "Y"))
  return(dat)
}
```

```{r med8, warning=FALSE, cache=TRUE, message=FALSE, echo=TRUE}
# set seed and simulate example data
example_data <- make_example_data()
w_names <- str_subset(colnames(example_data), "W")
m_names <- str_subset(colnames(example_data), "M")

# quick look at the data
head(example_data)
```

```{r med9, warning=FALSE, cache=TRUE, message=FALSE, echo=TRUE}
# compute one-step estimate of the interventional direct effect
os_de <- medoutcon(W = example_data[, ..w_names],
                   A = example_data$A,
                   Z = example_data$Z,
                   M = example_data[, ..m_names],
                   Y = example_data$Y,
                   effect = "direct",
                   estimator = "onestep")
summary(os_de)
```

```{r med10, warning=FALSE, cache=TRUE, message=FALSE, echo=TRUE}
# compute targeted minimum loss estimate of the interventional direct effect
tmle_de <- medoutcon(W = example_data[, ..w_names],
                     A = example_data$A,
                     Z = example_data$Z,
                     M = example_data[, ..m_names],
                     Y = example_data$Y,
                     effect = "direct",
                     estimator = "tmle")

summary(tmle_de)
```
