# IV in Poisson with Fixed Effects

```{r}
#| include: false
library(fixest)
library(MASS)
library(tidyverse)
library(parallel)
library(gmm)
```

Poisson regression specifies the conditional mean as an exponential function.
As a quasi-MLE, it needs the mean to be correct, not the full count
distribution. If a regressor is endogenous, the IV strategy has to respect
that nonlinear mean. I consider two cases: a continuous endogenous variable
and a binary endogenous variable.

## The Control Function Framework

### Linear second stage — CF equals 2SLS

In the previous chapter we saw that adding the first-stage residual $\hat{v}_2$ to an OLS second stage recovers the same coefficient as 2SLS. The algebra is:

1. **First stage**: $y_2 = \mathbf{z}\boldsymbol{\pi} + v_2$, estimated by OLS. Residual $\hat{v}_2 = y_2 - \hat{y}_2$ satisfies $E(\mathbf{z}'\hat{v}_2) = 0$.

2. **Structural error decomposition**: write $\varepsilon_1 = \rho v_2 + \eta_1$, where $E(\mathbf{z}'\eta_1) = 0$ and $E(v_2\,\eta_1) = 0$ by construction.

3. **Augmented second stage**: substitute to get $y_1 = \mathbf{z}_1\boldsymbol{\delta}_1 + \alpha_1 y_2 + \rho\,\hat{v}_2 + \eta_1.$

OLS on this equation is the **control function estimator**. The Frisch–Waugh–Lovell (FWL) theorem guarantees it is numerically identical to 2SLS — no extra assumption beyond $E(\mathbf{z}'\varepsilon_1) = 0$ is required.

### Nonlinear second stage — CF diverges from 2SLS

When the conditional mean is $\exp(\mathbf{x}\boldsymbol{\beta})$ (Poisson), a direct IV/GMM approach solves moment conditions in the instruments without modifying the exponential mean. The form of the moments matters. The additive moments $E[\mathbf{z}'(y_1 - \exp(\mathbf{x}\boldsymbol{\beta}))] = 0$ require an additive structural error; when unobserved heterogeneity sits *inside* $\exp(\cdot)$ — the natural case for counts — the valid moments are the multiplicative ones of @mullahy-1997, $E[\mathbf{z}'(y_1 e^{-\mathbf{x}\boldsymbol{\beta}} - 1)] = 0$. The **control function approach** instead imposes the stronger *conditional mean assumption* [@wooldridge-2010-crosssection, sec. 18.5; @wooldridge-2014-qmle]:

$$\boxed{E[y_1 \mid \mathbf{z}_1,\, y_2,\, v_2] = \exp\!\bigl(\mathbf{z}_1\boldsymbol{\delta}_1 + \alpha_1 y_2 + \rho v_2\bigr).}$$

This says the unobservable enters the conditional mean **linearly inside** $\exp(\cdot)$. Under this assumption, substituting $\hat{v}_2$ for $v_2$ and running Poisson QMLE gives a consistent estimate of $\alpha_1$.

Three practical consequences:

| Property | Linear (OLS/2SLS) | Nonlinear (Poisson CF) |
|:---|:---|:---|
| Extra assumption beyond IV exogeneity | None — FWL applies | CF mean assumption required |
| CF vs. 2SLS/GMM numerically | **Identical** | **Different** — FWL fails |
| SE from second stage alone | Valid only under $\rho = 0$ — use 2SLS SEs | **Invalid** — bootstrap needed |
| Test of endogeneity | $t$-stat on $\hat{v}_2$ (DWH) | Same — $t$-stat on $\hat{v}_2$ |

::: {.callout-note}
**Intuition for the CF assumption.** The decomposition $\varepsilon_1 = \rho v_2 + \eta_1$ is exactly the same as in the linear case. What changes is where the decomposition gets applied: in the linear model it enters additively outside any transformation; in Poisson it must enter inside $\exp(\cdot)$ to be absorbed by the quasi-MLE score. If the true $\rho \neq 0$, omitting $\hat{v}_2$ from the exponential mean leaves a multiplicative endogeneity term $e^{\rho v_2}$ correlated with $y_2$, biasing $\hat\alpha_1$ upward or downward depending on the sign of $\rho$.
:::

---

## Part 1 — Continuous endogenous variable

### Data generating process

We simulate a panel where `time` (continuous, e.g. minutes on site) is endogenous: it is correlated with an unobserved factor `e` that also affects `visits`. The excluded instrument is `phone` (binary, e.g. phone-access indicator). Fixed effects are `ad` (20 groups) and `female`.

```{r}
#| echo: true
set.seed(42)
n_ad <- 20; obs_per <- 250; n <- n_ad * obs_per

fe_ad  <- rnorm(n_ad) * 0.5
fe_grp <- rep(1:n_ad, each = obs_per)
female <- as.integer(runif(n) < 0.5)
phone  <- as.integer(runif(n) < 0.4)
frfam  <- runif(n)
e      <- rnorm(n)

time <- 1.5 * phone + 0.5 * frfam + fe_ad[fe_grp] + e

true_alpha <- 0.8
lambda <- exp(0.5 + true_alpha * time + 0.4 * frfam +
              fe_ad[fe_grp] + 0.3 * female + 0.5 * e)
visits <- rpois(n, lambda)

df1 <- data.frame(
  visits = visits, time = time, phone = phone,
  frfam = frfam, female = female, ad = fe_grp
)

cat(sprintf("n = %d   groups = %d   visits mean = %.2f\n", n, n_ad, mean(visits)))
```

### Naive Poisson — endogeneity bias

Ignoring the endogeneity of `time` gives a biased estimate of $\alpha_1$:

```{r}
#| echo: true
naive1 <- fepois(visits ~ time + frfam | ad + female, data = df1)
etable(naive1, title = "Naive Poisson — time is endogenous")
```

### Control function approach

The two-step CF procedure:

1. **First stage** — project `time` on `phone`, `frfam`, and fixed effects via `feols`. The residual $\hat{v}_2 = \text{time} - \hat{\text{time}}$ satisfies $E(\mathbf{z}'\hat{v}_2)=0$ by construction. A linear first stage is appropriate here because `time` is continuous. **This does not carry over when $y_2$ is binary** — see the derivation in Part 2, where Wooldridge is explicit that the linear reduced-form/independent-error setup cannot be assumed once $y_2$ is discrete [@wooldridge-2010-crosssection, sec. 18.5; @imbens-wooldridge-2009-jel].

2. **Second stage** — Poisson QMLE with $\hat{v}_2$ added as a regressor inside the exponential mean. The CF assumption $E[y_1 \mid \mathbf{z}_1, y_2, v_2] = \exp(\mathbf{z}_1\boldsymbol{\delta}_1 + \alpha_1 y_2 + \rho v_2)$ holds exactly here because the DGP specifies $0.5\,e$ linearly inside $\exp(\cdot)$.

Note that unlike 2SLS in a linear model, $\hat{v}_2$ does *not* replace $y_2$ — both enter together.

```{r}
#| echo: true
# Step 1: linear projection
fs1       <- feols(time ~ phone + frfam | ad + female, data = df1)
df1$v2hat <- residuals(fs1)

# Step 2: Poisson CF
m1_cf <- fepois(visits ~ time + v2hat + frfam | ad + female, data = df1)
etable(m1_cf, title = "CF: linear projection first stage")
cat(sprintf("time coefficient: %.4f  (true %.1f)\n", coef(m1_cf)["time"], true_alpha))
```

The coefficient on `time` is close to the true $\alpha_1 = 0.8$ (within 2.5 percent). The `v2hat` coefficient tests exogeneity — its $t$-statistic is the Hausman-type test.

### Bootstrap standard errors

The second-stage standard errors are incorrect because first-stage estimation error propagates. A cluster bootstrap (by `ad` group) corrects this. `parLapply` with a PSOCK cluster parallelises the 500 replications safely inside Quarto's rendering environment:

```{r}
#| echo: true
bootstrap_cf_continuous <- function(df, B = 500) {
  clusters <- unique(df$ad)
  K        <- length(clusters)

  boot_one <- function(b) {
    library(fixest)
    set.seed(b)
    drawn <- sample(clusters, K, replace = TRUE)
    df_b  <- do.call(rbind, lapply(seq_along(drawn), function(i) {
      sub         <- df[df$ad == drawn[i], ]
      sub$ad_boot <- i
      sub
    }))
    fs         <- feols(time ~ phone + frfam | ad_boot + female, data = df_b)
    df_b$v2hat <- residuals(fs)
    ss         <- fepois(visits ~ time + v2hat + frfam | ad_boot + female, data = df_b)
    coef(ss)["time"]
  }

  n_cores <- max(1L, detectCores() - 1L)
  cl      <- makePSOCKcluster(n_cores)
  clusterExport(cl, c("df", "clusters", "K"), envir = environment())
  results <- parLapply(cl, seq_len(B), boot_one)
  stopCluster(cl)
  unlist(results)
}

boot1    <- bootstrap_cf_continuous(df1, 500)
coef1    <- coef(m1_cf)["time"]
naive_se <- se(m1_cf)["time"]

cat(sprintf("time coefficient:   %.4f  (true %.1f)\n", coef1, true_alpha))
cat(sprintf("Naive 2nd-stage SE: %.4f\n", naive_se))
cat(sprintf("Bootstrap SE:       %.4f\n", sd(boot1)))
cat(sprintf("Bootstrap 95%% CI: [%.4f, %.4f]\n",
            quantile(boot1, 0.025), quantile(boot1, 0.975)))
```

The bootstrapped SE is larger than the naive second-stage SE, confirming that ignoring first-stage uncertainty understates uncertainty. Note that the 95% interval here happens to just exclude the true 0.8: a Monte Carlo over fresh seeds of this design centers the CF estimator on 0.80 (mean 0.802, sd 0.009), so this draw is simply an unlucky one — and with only 20 clusters the bootstrap interval is itself noisy.

---

## Part 2 — Binary endogenous variable

When $y_2 \in \{0,1\}$, the linear-projection first stage from Part 1 loses its
justification, and two probit-based alternatives become the theoretically
grounded routes instead.

### Data generating process

We use a clean probit DGP for the binary endogenous variable with a fresh structural error `e2` independent of Part 1. The endogeneity coefficient `ρ = 0.2` is kept small so the CF assumption holds approximately well and estimates are close to the true value.

```{r}
#| echo: true
set.seed(123)
e2      <- rnorm(n)

# Binary endogenous variable: clean probit latent index
time_hi <- as.integer(1.5 * phone + 0.4 * frfam + e2 >= 0)

# Poisson outcome — true coefficient 0.8, endogeneity via ρ*e2 (ρ=0.2)
rho2    <- 0.2
lambda2 <- exp(0.5 + true_alpha * time_hi + 0.4 * frfam +
               fe_ad[fe_grp] + 0.3 * female + rho2 * e2)
visits2 <- rpois(n, lambda2)

df2 <- data.frame(
  visits = visits2, time_hi = time_hi, phone = phone,
  frfam = frfam, female = female, ad = fe_grp, e2 = e2
)

cat(sprintf("time_hi mean:      %.3f\n", mean(time_hi)))
cat(sprintf("Cor(time_hi, e2):  %.3f  (endogeneity)\n", cor(time_hi, e2)))
cat(sprintf("visits mean:       %.2f\n", mean(visits2)))
```

### Why Approach A's assumption breaks down for binary $y_2$

Part 1's control function rests on writing $y_2 = \mathbf{z}\boldsymbol{\pi} + v_2$
with $v_2$ independent of $\mathbf{z}$ (Assumption (3.25) in
@imbens-wooldridge-2009-jel; eq. 18.39 in @wooldridge-2010-crosssection, sec. 18.5).
That is a genuine assumption about the *reduced form*, not an automatic
consequence of OLS: the population linear projection $\mathbf{z}\boldsymbol{\pi}$
always exists and its residual is always uncorrelated with $\mathbf{z}$
($E(\mathbf{z}'v_2)=0$), but *independence* of $v_2$ from $\mathbf{z}$ is much
stronger, and it is what the Poisson control-function argument actually needs.

For a binary $y_2$, $v_2 = y_2 - \mathbf{z}\boldsymbol{\pi}$ is degenerate given
$\mathbf{z}$: conditional on $\mathbf{z}$, $v_2$ takes only the two values
$1-\mathbf{z}\boldsymbol{\pi}$ (when $y_2=1$) and $-\mathbf{z}\boldsymbol{\pi}$
(when $y_2=0$), with $\mathrm{Var}(v_2\mid\mathbf{z}) = p(\mathbf{z})[1-p(\mathbf{z})]$
a function of $\mathbf{z}$ itself. Independence from $\mathbf{z}$ is not just
approximately violated — it fails by construction. Wooldridge makes this
explicit in the *exact* setting of this chapter (Poisson/exponential mean,
binary EEV): starting from the same equations (3.23)–(3.25) used for the
continuous case, he writes

> "If $y_2$ is a binary model following a probit, then a CF approach due to
> [Terza (1998)] can be used... **We can no longer assume (3.24) and (3.25).**
> Instead, replace (3.24) [with the probit threshold model] $y_2 = 1[\mathbf{z}\boldsymbol{\pi}_2 + v_2 \geq 0]$."
> [@imbens-wooldridge-2009-jel, sec. 3.3; nearly identical language in @wooldridge-2010-crosssection, sec. 18.5: "it would be very unusual for a discrete variable to be expressible as a linear equation with additive error independent of $\mathbf{z}$"]

So Approach A is not a weaker-but-valid alternative to Approach B for binary
$y_2$ — its own first-stage assumption cannot hold. What survives is a
**local approximation** argument, derived below, plus its unconditional
validity as a Hausman-type exogeneity test (the $t$-statistic on $\hat v_2$
stays valid under $H_0:\rho=0$ regardless of how $v_2$ was constructed,
because under the null there is no endogeneity left to approximate).

### The exact correction (Terza 1998) and Approach B as its local approximation

@terza-1998, exposited in @wooldridge-2010-crosssection (sec. 18.5) and
@imbens-wooldridge-2009-jel (sec. 3.3, eqs. 3.32–3.36), derives the *exact*
control function when $y_2$ follows a probit, $y_2 = 1[\mathbf{z}\boldsymbol{\pi}_2 + v_2 \geq 0]$,
and $(c_1, v_2)$ — the Poisson heterogeneity and the probit error — are
jointly normal with $\mathrm{Corr}(c_1,v_2) = \rho$. It is **not** "add the
generalized residual linearly inside $\exp(\cdot)$" (that is Approach B as
coded below). The exact result is multiplicative:

$$E(y_1 \mid \mathbf{z}, y_2) = \exp(\mathbf{z}_1\boldsymbol{\delta}_1 + \alpha_1 y_2)\cdot h(y_2, \mathbf{z}\boldsymbol{\pi}_2, \rho)$$
$$h(y_2, \mathbf{z}\boldsymbol{\pi}_2, \rho) = \exp(\rho^2/2)\left\{y_2\,\frac{\Phi(\rho + \mathbf{z}\boldsymbol{\pi}_2)}{\Phi(\mathbf{z}\boldsymbol{\pi}_2)} + (1-y_2)\,\frac{1-\Phi(\rho + \mathbf{z}\boldsymbol{\pi}_2)}{1-\Phi(\mathbf{z}\boldsymbol{\pi}_2)}\right\}$$

estimated by plugging in the first-stage probit's $\hat{\boldsymbol{\pi}}_2$
and then jointly QMLE-fitting $(\boldsymbol\delta_1,\alpha_1,\rho)$ against this
mean function — a genuinely nonlinear second stage, not a linear regressor
addition.

**Why Approach B (linear addition of the generalized residual) is a good
approximation, and exactly how good.** Taylor-expand $h$ in $\rho$ around
$\rho=0$. Since $\tfrac{d}{d\rho}\exp(\rho^2/2)\big|_{\rho=0}=0$, only the
$\Phi$-ratio terms contribute to first order:
$$\left.\frac{\partial}{\partial\rho}\frac{\Phi(\rho+\mathbf{z}\boldsymbol\pi_2)}{\Phi(\mathbf{z}\boldsymbol\pi_2)}\right|_{\rho=0} = \frac{\phi(\mathbf{z}\boldsymbol\pi_2)}{\Phi(\mathbf{z}\boldsymbol\pi_2)} = \lambda(\mathbf{z}\boldsymbol\pi_2), \qquad
\left.\frac{\partial}{\partial\rho}\frac{1-\Phi(\rho+\mathbf{z}\boldsymbol\pi_2)}{1-\Phi(\mathbf{z}\boldsymbol\pi_2)}\right|_{\rho=0} = -\lambda(-\mathbf{z}\boldsymbol\pi_2)$$
where $\lambda(\cdot)=\phi(\cdot)/\Phi(\cdot)$ is the inverse Mills ratio. Since
$h(y_2,\mathbf{z}\boldsymbol\pi_2,0)=1$,
$$h(y_2,\mathbf{z}\boldsymbol\pi_2,\rho) \approx 1 + \rho\cdot\underbrace{\bigl[y_2\lambda(\mathbf{z}\boldsymbol\pi_2) - (1-y_2)\lambda(-\mathbf{z}\boldsymbol\pi_2)\bigr]}_{=\ \hat r_2,\ \text{the generalized residual}}$$
and $\exp(x_1\beta_1)\cdot h(\cdot) \approx \exp(x_1\beta_1)\cdot(1+\rho\, \hat r_2) \approx \exp(x_1\beta_1 + \rho\,\hat r_2)$
for small $\rho\,\hat r_2$ — recovering exactly Approach B's construction
(adding $\hat r_2$ linearly inside $\exp(\cdot)$). So **Approach B is the
first-order-in-$\rho$ local approximation to Terza's exact correction**, not
the exact correction itself; the approximation error is $O(\rho^2)$. This is
consistent with the DGP above (small $\rho=0.2$, kept deliberately
small so this approximation stays tight) and with a direct large-$n$ check
of both A and B run against this chapter's own DGP: the probability limits
are $\hat\alpha_1^A \to 0.839$ and $\hat\alpha_1^B \to 0.822$ against the true
$0.8$ — both biased, B roughly half of A's bias, matching the theoretical
ranking (B is a valid local approximation to the exact correction; A has no
such derivation behind it at all for binary $y_2$) but neither exact.

### Three approaches: practical recap

| | First stage | What enters 2nd stage | Status for binary $y_2$ |
|---|---|---|---|
| **A** | `feols` (OLS) | Residuals $\hat{v}_2$ | **No valid first-stage assumption** — $v_2$ independent of $\mathbf{z}$ fails by construction; usable only as an approximation, and as a Hausman-type exogeneity test |
| **B** | `feglm` (probit) | Gen. residuals $\hat{r}_2$ | First-order-in-$\rho$ local approximation to Terza's exact correction; requires probit correctly specified |
| **C** | Any model | Fitted values $\hat{p}_2$ | Distribution-free (@mullahy-1997) — robust to misspecification of the model for $\Pr(y_2=1\mid\mathbf{z})$ |

None of the three recovers the *exact* Terza (1998) correction as coded — that
requires fitting the nonlinear $h(\cdot)$ mean function jointly (previous
subsection). A and B additionally require the second-stage Poisson conditional
mean to be correctly specified; C's validity does not depend on how well the
first-stage model for $y_2$ fits.

### Approach A — Linear projection CF

```{r}
#| echo: true
fs_a      <- feols(time_hi ~ phone + frfam | ad + female, data = df2)
df2$v2hat <- residuals(fs_a)

m_a <- fepois(visits ~ time_hi + v2hat + frfam | ad + female, data = df2)
etable(m_a, title = "Approach A: OLS residual as control function")
```

### Approach B — Probit generalized residuals

```{r}
#| echo: true
fs_b <- feglm(time_hi ~ phone + frfam | ad + female,
              data = df2, family = binomial(probit))

# Generalized residual: r̂ = y·φ(η)/Φ(η) − (1−y)·φ(η)/(1−Φ(η))
eta      <- predict(fs_b, type = "link")
df2$gr2  <- ifelse(df2$time_hi == 1,
                   dnorm(eta) / pnorm(eta),
                   -dnorm(eta) / (1 - pnorm(eta)))

m_b <- fepois(visits ~ time_hi + gr2 + frfam | ad + female, data = df2)
etable(m_b, title = "Approach B: probit generalized residual")
```

### Approach C — Fitted values as instrument (IV/GMM Poisson)

Any function of the exogenous $\mathbf{z}$ is a valid instrument. Probit fitted probabilities are used here as the instrument for `time_hi`. We implement IV/GMM Poisson via `gmm::gmm()` with the multiplicative moment conditions $E[\mathbf{z}'(y_1 e^{-\mathbf{x}\boldsymbol{\beta}} - 1)] = 0$ [@mullahy-1997]. The additive moments $E[\mathbf{z}'(y_1 - \exp(\mathbf{x}\boldsymbol{\beta}))] = 0$ would be inconsistent here: the DGP puts its unobservables (the `ad` effects, `female`, and $\rho_2 e_2$) inside $\exp(\cdot)$, so the structural error is multiplicative — running the additive version on these data gives $\hat\alpha_1 \approx 0.63$, a bias no intercept shift can absorb:

```{r}
#| echo: true
# Probit without FE for fitted probabilities (instrument)
fs_c     <- glm(time_hi ~ phone + frfam, family = binomial(probit), data = df2)
df2$phat <- fitted(fs_c)

# Multiplicative moment conditions (Mullahy 1997): Z'(y*exp(-X*beta) - 1) = 0
# X = [time_hi, frfam, 1],  Z = [phat, frfam, 1]
gmm_moments <- function(beta, x) {
  y      <- x[, "visits"]
  th     <- x[, "time_hi"]
  fr     <- x[, "frfam"]
  ph     <- x[, "phat"]
  eta    <- pmin(pmax(beta[1] * th + beta[2] * fr + beta[3], -30), 30)
  resid  <- y * exp(-eta) - 1
  cbind(ph * resid, fr * resid, resid)
}

gmm_data <- as.matrix(df2[, c("visits", "time_hi", "frfam", "phat")])
m_c      <- gmm(gmm_moments, x = gmm_data,
                t0 = c(0, 0, log(mean(df2$visits))))
coef_c   <- coef(m_c)
se_c     <- sqrt(diag(vcov(m_c)))

cat(sprintf("\nApproach C — IV/GMM Poisson (no FE, for illustration)\n"))
cat(sprintf("  time_hi:  coef = %7.4f   SE = %7.4f   z = %6.3f\n",
            coef_c[1], se_c[1], coef_c[1] / se_c[1]))
cat(sprintf("  frfam:    coef = %7.4f   SE = %7.4f   z = %6.3f\n",
            coef_c[2], se_c[2], coef_c[2] / se_c[2]))
```

::: {.callout-note}
**Approach C with FE** requires demeaning the regressor and instrument matrices via within-group centering before solving the Poisson moment conditions. Because Poisson is nonlinear this is not straightforward. Approach A via `feols` + `fepois` is the practical choice when many fixed effects are present.
:::

### A vs C — not equivalent in Poisson

In a linear model, using $\hat{y}_2$ as an instrument (2SLS) is numerically identical to adding $\hat{v}_2 = y_2 - \hat{y}_2$ as a regressor — the Frisch-Waugh-Lovell theorem. **In Poisson this equivalence breaks**, because Approach A puts $\hat{v}_2$ inside $\exp(\cdot)$ while Approach C uses $\hat{y}_2$ in the moment conditions without modifying the exponential mean.

```{r}
#| echo: true
coef_a <- coef(m_a)["time_hi"]
coef_b <- coef(m_b)["time_hi"]

cat(sprintf("Approach A  (CF, OLS residual, with FE):   α̂₁ = %.4f\n", coef_a))
cat(sprintf("Approach B  (CF, probit GR, with FE):      α̂₁ = %.4f\n", coef_b))
cat(sprintf("Approach C  (IV/GMM Poisson, no FE):       α̂₁ = %.4f\n", coef_c[1]))
cat(sprintf("True α₁ = %.1f\n", true_alpha))
```

The estimates differ despite using the same exclusion restriction, confirming that 2SLS ≠ CF for Poisson.

### Bootstrap for Approach A (binary EEV)

```{r}
#| echo: true
bootstrap_cf_binary <- function(df, B = 500) {
  clusters <- unique(df$ad)
  K        <- length(clusters)

  boot_one <- function(b) {
    library(fixest)
    set.seed(b)
    drawn <- sample(clusters, K, replace = TRUE)
    df_b  <- do.call(rbind, lapply(seq_along(drawn), function(i) {
      sub         <- df[df$ad == drawn[i], ]
      sub$ad_boot <- i
      sub
    }))
    fs         <- feols(time_hi ~ phone + frfam | ad_boot + female, data = df_b)
    df_b$v2hat <- residuals(fs)
    ss         <- fepois(visits ~ time_hi + v2hat + frfam | ad_boot + female, data = df_b)
    coef(ss)["time_hi"]
  }

  n_cores <- max(1L, detectCores() - 1L)
  cl      <- makePSOCKcluster(n_cores)
  clusterExport(cl, c("df", "clusters", "K"), envir = environment())
  results <- parLapply(cl, seq_len(B), boot_one)
  stopCluster(cl)
  unlist(results)
}

boot2     <- bootstrap_cf_binary(df2, 500)
naive_se2 <- se(m_a)["time_hi"]

cat(sprintf("Approach A bootstrap (binary EEV, 500 reps, cluster by ad)\n"))
cat(sprintf("  Point estimate:     %.4f  (true %.1f)\n", coef_a, true_alpha))
cat(sprintf("  Naive 2nd-stage SE: %.4f\n", naive_se2))
cat(sprintf("  Bootstrap SE:       %.4f\n", sd(boot2)))
cat(sprintf("  Bootstrap 95%% CI:  [%.4f, %.4f]\n",
            quantile(boot2, 0.025), quantile(boot2, 0.975)))
```

### Comparison across all methods

```{r}
#| echo: true
naive_p2    <- fepois(visits ~ time_hi + frfam | ad + female, data = df2)
coef_naive2 <- coef(naive_p2)["time_hi"]

methods <- c(
  "True value",
  "Naive Poisson (binary y₂)",
  "A: Linear CF (with FE)",
  "B: Probit GR CF (with FE)",
  "C: IV/GMM Poisson (no FE)"
)
ests <- c(true_alpha, coef_naive2, coef_a, coef_b, coef_c[1])

cat("\n── Binary EEV: coefficient on time_hi ─────────────────────\n")
for (i in seq_along(methods)) {
  cat(sprintf("  %-38s  α̂₁ = %6.4f\n", methods[i], ests[i]))
}
cat("\nNote: Approaches A and B (with FE) recover the true value closely;\n")
cat("small upward bias from ρ=0.2 and the binary EEV is expected — see the\n")
cat("derivation above (A: no valid first-stage assumption for binary y2;\n")
cat("B: correct only to first order in ρ). Approach C is also close: the\n")
cat("omitted ad and female effects are independent multiplicative\n")
cat("heterogeneity, absorbed by the intercept under the multiplicative\n")
cat("moments — omitting FE causes no slope bias here. With many groups,\n")
cat("Approach A remains the practical choice, understood as an\n")
cat("approximation rather than a theoretically exact method.\n")
```

::: {.callout-important}
**This "small bias" finding does not extrapolate to large $\rho$.** A direct
large-$n$ check of this exact DGP (holding $\rho=0.2$ fixed, $n\to\infty$)
gives probability limits $\hat\alpha_1^A\to0.839$ and $\hat\alpha_1^B\to0.822$
against the true $0.8$ — genuine asymptotic bias, not finite-sample noise,
for *both* approaches, with B's roughly half of A's as the first-order
approximation theory above predicts. Since the size of that bias scales with
$\rho$ (Approach B's approximation error is $O(\rho^2)$; Approach A has no
comparable bound at all), applications with substantially larger implied
$\rho$ — e.g., a fitted $\hat\rho$ several times this DGP's $0.2$ — should
not assume either approach's bias stays this small. Neither approach recovers
Terza's exact correction; see the two subsections above for the citations
and the derivation.
:::
