# IV in Poisson with Fixed Effects

Poisson regression specifies the conditional mean as an exponential function.
As a quasi-MLE it needs that mean to be correct, not the full count
distribution. If a regressor is endogenous, the IV strategy has to respect the
nonlinear mean — and this is where intuition carried over from 2SLS starts to
mislead. This chapter works through what goes wrong, what to do instead, and how
much the choice costs.

```{r}
#| include: false
library(fixest)
library(MASS)
library(tidyverse)
library(parallel)
library(gmm)
```

## The problem, and three routes to it

Throughout, $y_1$ is a count, $y_2$ is the endogenous explanatory variable, and
$\mathbf{z} = (\mathbf{z}_1, \mathbf{z}_2)$ collects the exogenous variables —
$\mathbf{z}_1$ appearing in the structural equation and $\mathbf{z}_2$ excluded
from it. The structural conditional mean is exponential,

$$E(y_1 \mid \mathbf{z}_1, y_2, c_1) = \exp(\mathbf{z}_1\boldsymbol{\delta}_1 + \alpha_1 y_2 + c_1),$$ {#eq-poisson-iv-1}

with $c_1$ an unobservable correlated with $y_2$. Everything below is about how
to remove that correlation.

### A note on the notation: the subscript indexes *equations*, not variables

This chapter follows @wooldridge-2010-crosssection (sec. 18.5) rather than the
$Y, W, Z, X$ convention of the [IV chapter](iv-rdd.qmd), because it quotes his
equations (18.37)–(18.45) directly and the symbols need to match the source.

The translation:

| Role | IV chapter | Here |
|:---|:---|:---|
| Outcome (the count) | $Y$ | $y_1$ |
| Endogenous regressor | $W$ | $y_2$ |
| Excluded instrument | $Z$ | $\mathbf{z}_2$ |
| Included exogenous | $X$ | $\mathbf{z}_1$ |
| First-stage error | $V$ | $v_2$ |

The subscript is the source of most of the confusion, so it is worth saying what
it means: **it indexes the equation, not the variable.** $y_1$ is the variable
explained by equation 1, the structural equation; $y_2$ is the variable explained
by equation 2, the reduced form. The same convention governs $c_1$, $v_2$,
$\boldsymbol{\delta}_1$ and $\boldsymbol{\pi}_2$ — each subscript says which
equation the object belongs to. This is the simultaneous-equations convention,
and it scales: a second endogenous regressor is $y_3$, with no renaming.

Read instead as "$y$-one and $y$-two, two outcomes," it means nothing — which is
the misreading the notation invites.

One deliberate deviation: Wooldridge writes the coefficient on $y_2$ as
$\gamma_1$; this chapter calls it $\alpha_1$ throughout, matching the rest of the
book. Quoted equations are left exactly as he writes them.

Three estimators recur, and it is worth naming them by what they *do* rather
than by letters, since the names then carry the argument:

| Name | First stage | What enters the second stage | Family |
|:---|:---|:---|:---|
| **Linear-residual CF** | OLS projection of $y_2$ on $\mathbf{z}$ | the residual $\hat v_2$, inside $\exp(\cdot)$ | control function |
| **Generalized-residual CF** | probit or logit for $y_2$ | the generalized residual $\hat r_2$, inside $\exp(\cdot)$ | control function |
| **Multiplicative-moment GMM** | any model for $\Pr(y_2=1 \mid \mathbf{z})$ | the fitted value $\hat p_2$, as an *instrument* | GMM / IV |

**Earlier drafts of this chapter called these Approach A, B and C.** The letters are gone. Linear-residual CF was A, generalized-residual CF was B,
and multiplicative-moment GMM was C.

The third is not a control function at all. It does not modify the exponential
mean; it changes the moment conditions. Grouping all three as "approaches"
obscured that, and the distinction matters for when each is valid.

The plan: the theory first, in one place; then the three compared side by side;
then simulations, continuous case and binary case.

## Theory

### The control function idea

#### A linear second stage: the CF *is* 2SLS

Add the first-stage residual $\hat v_2$ to an OLS second stage and you recover
the 2SLS coefficient exactly. The algebra:

1. **First stage.** $y_2 = \mathbf{z}\boldsymbol{\pi} + v_2$, by OLS. The
   residual satisfies $E(\mathbf{z}'\hat v_2) = 0$ by construction.
2. **Error decomposition.** Write $\varepsilon_1 = \rho v_2 + \eta_1$, with
   $E(\mathbf{z}'\eta_1) = 0$ and $E(v_2\eta_1) = 0$.
3. **Augmented second stage.**
   $y_1 = \mathbf{z}_1\boldsymbol{\delta}_1 + \alpha_1 y_2 + \rho\hat v_2 + \eta_1$.

OLS on step 3 is the **control function estimator**, and Frisch–Waugh–Lovell
makes it numerically identical to 2SLS. **No assumption beyond
$E(\mathbf{z}'\varepsilon_1) = 0$ is required.** This matters more than it looks:
in the linear model the control-function story is a *reinterpretation* of an
estimator justified by a moment condition. If the story is wrong, 2SLS is still
consistent.

#### A nonlinear second stage: the CF is a genuine assumption

With an exponential mean, that safety net disappears. A direct IV/GMM approach
solves moment conditions in the instruments without touching the exponential
mean, and the form of those moments matters: the additive moments
$E[\mathbf{z}'(y_1 - \exp(\mathbf{x}\boldsymbol{\beta}))] = 0$ require an
additive structural error, whereas when the unobservable sits *inside*
$\exp(\cdot)$ — the natural case for counts — the valid moments are the
multiplicative ones of @mullahy-1997.

The control function does something different. It imposes the stronger
*conditional mean assumption* [@wooldridge-2010-crosssection, sec. 18.5;
@wooldridge-2014-qmle], Wooldridge's eq. (18.40):

$$E[y_1 \mid \mathbf{z}_1,\, y_2,\, v_2] = \exp\!\bigl(\mathbf{z}_1\boldsymbol{\delta}_1 + \alpha_1 y_2 + \rho v_2\bigr).$$ {#eq-poisson-iv-2}

The unobservable enters the mean **linearly inside** $\exp(\cdot)$. Under this,
substituting $\hat v_2$ for $v_2$ and running Poisson QMLE is consistent for
$\alpha_1$. Without it, nothing rescues the procedure.

| Property | Linear (OLS/2SLS) | Nonlinear (Poisson CF) |
|:---|:---|:---|
| Extra assumption beyond IV exogeneity | None — FWL applies | CF mean assumption required |
| CF vs. 2SLS/GMM numerically | **Identical** | **Different** — FWL fails |
| SE from second stage alone | Valid only under $\rho = 0$ — use 2SLS SEs | **Invalid** — bootstrap needed |
| Test of endogeneity | $t$-stat on $\hat v_2$ (DWH) | Same — $t$-stat on $\hat v_2$ |

**Where the decomposition gets applied.** The decomposition $\varepsilon_1 = \rho v_2 + \eta_1$ is the same in both cases.
What changes is *where* it enters: additively outside any transformation in the
linear model, inside $\exp(\cdot)$ in Poisson, so that the quasi-MLE score can
absorb it. If $\rho \neq 0$ and $\hat v_2$ is omitted from the exponential mean,
a multiplicative $e^{\rho v_2}$ term correlated with $y_2$ survives, biasing
$\hat\alpha_1$ in the direction of $\rho$.

### A continuous $y_2$: the linear reduced form is enough

For continuous $y_2$ the assumptions are the natural ones. Wooldridge's
eqs. (18.38)–(18.39) are

$$y_2 = \mathbf{z}\boldsymbol{\pi}_2 + v_2, \qquad c_1 = \rho_1 v_2 + e_1,$$ {#eq-poisson-iv-3}

with $v_2$ **independent** of $\mathbf{z}$ and $e_1$ independent of
$(\mathbf{z}, v_2)$. Those two independence conditions deliver @eq-poisson-iv-2
directly, and the estimator follows: regress, take residuals, add them to the
index.

Note which condition is doing the work. It is **independence**, not
uncorrelatedness — and Wooldridge flags this at the point he states the
assumption, before discreteness enters the discussion at all
[@wooldridge-2010-crosssection, p. 743]:

> "(We could relax the independence assumptions to some degree, but **we cannot
> just assume that $v_2$ is uncorrelated with $\mathbf{z}$ and that $e_1$ is
> uncorrelated with $v_2$**.)"

The distinction is invisible for continuous $y_2$ — a linear model with a normal
additive error satisfies both — which is exactly why it is easy to carry the
procedure across to a binary $y_2$ without noticing that it has stopped holding.

### Why independence, when the linear model needs only uncorrelatedness?

Because of the exponential. Substituting @eq-poisson-iv-3 into
@eq-poisson-iv-1 gives

$$E(y_1 \mid \mathbf{z}, y_2, v_2) = \exp(\mathbf{z}_1\boldsymbol{\delta}_1 + \alpha_1 y_2)\cdot\exp(\rho_1 v_2)\cdot E[\exp(e_1) \mid \mathbf{z}, y_2, v_2],$$ {#eq-poisson-iv-17}

and @eq-poisson-iv-2 follows only if that final factor is a **constant** — which
is why Wooldridge normalises $E[\exp(e_1)] = 1$ rather than $E(e_1) = 0$.

That is a condition on the *moment generating function* of $e_1$, that is, on
every moment at once, which is essentially full independence. Compare the linear
model, where the requirement is $E[\mathbf{z}'\varepsilon_1] = 0$ — a statement
about one moment, delivered free by the OLS projection.

**This is the same fact as "CF $=$ 2SLS in a linear model, CF $\neq$ 2SLS in
Poisson," seen from the assumption side rather than the algebra side.**
$\exp(\cdot)$ is what turns a moment condition into a distributional one, and
everything difficult in this chapter follows from it.

### A binary $y_2$: why the linear reduced form fails

OLS always produces a residual uncorrelated with $\mathbf{z}$. That is a
property of the population linear projection and it holds whatever $y_2$ looks
like. But the control-function derivation needs $v_2$ *independent* of
$\mathbf{z}$, which is a substantive assumption about the reduced form and not
something OLS supplies.

For binary $y_2$ it cannot hold. Conditional on $\mathbf{z}$, the residual
$v_2 = y_2 - \mathbf{z}\boldsymbol{\pi}$ takes exactly two values,

$$v_2 \mid \mathbf{z} = \begin{cases} 1 - \mathbf{z}\boldsymbol{\pi} & \text{with probability } p(\mathbf{z}) \\ -\mathbf{z}\boldsymbol{\pi} & \text{with probability } 1 - p(\mathbf{z}),\end{cases}$$ {#eq-poisson-iv-4}

so both the two support points *and* the probabilities of landing on them are
functions of $\mathbf{z}$. Its variance,
$\mathrm{Var}(v_2 \mid \mathbf{z}) = p(\mathbf{z})[1 - p(\mathbf{z})]$, is a
function of $\mathbf{z}$ too. Independence fails by construction, not by
approximation.

Wooldridge states this outright, at [@wooldridge-2010-crosssection, p. 746]:

> "The assumption that $e_1$ in (18.39) is independent of $v_2$ and $\mathbf{z}$
> **rules out most cases where $y_2$ has some discreteness, such as with binary,
> count, or corner responses.** (In particular, the independence assumption
> likewise fails unless $v_2$ is independent of $\mathbf{z}$, and it would be
> very unusual for a discrete variable to be expressible as a linear equation
> with additive error independent of $\mathbf{z}$.)"

That sentence is compressed, so it is worth unpacking. It names one assumption,
but the failure runs through three links:

1. **What the correction needs.** From @eq-poisson-iv-17, $e_1$ must be
   independent of $(\mathbf{z}, y_2, v_2)$, so that $E[\exp(e_1)\mid\cdot]$ is
   constant.
2. **That silently requires $v_2 \perp \mathbf{z}$.** $e_1$ is defined as what
   is left of $c_1$ after projecting on $v_2$, so if $v_2$ depends on
   $\mathbf{z}$, $e_1$ inherits the dependence. This is what "the independence
   assumption likewise fails unless" refers to.
3. **And $v_2 \perp \mathbf{z}$ cannot hold for discrete $y_2$.** This is the
   step he leaves as "very unusual", and it is in fact airtight. Independence
   means the conditional distribution of $v_2$ given $\mathbf{z}$ is *the same
   for every* $\mathbf{z}$. But by @eq-poisson-iv-4 the support is
   $\{-\mathbf{z}\boldsymbol{\pi},\ 1-\mathbf{z}\boldsymbol{\pi}\}$, and both
   points **slide as $\mathbf{z}$ changes**. A distribution whose support moves
   with $\mathbf{z}$ cannot be independent of it — unless
   $\boldsymbol{\pi} = \mathbf{0}$, in which case there is no first stage at all.
   The same argument covers count and corner responses: the mass point sits at
   $-\mathbf{z}\boldsymbol{\pi}$ and travels.

So the sentence reads: *the correction needs $e_1 \perp (v_2, \mathbf{z})$; that
needs $v_2 \perp \mathbf{z}$; and $v_2 \perp \mathbf{z}$ cannot hold when $y_2$
is discrete, because the residual's support is pinned to
$\mathbf{z}\boldsymbol{\pi}$.*

Note what does **not** break. $E[\mathbf{z}'v_2] = 0$ still holds exactly — the
projection is fine, and OLS has done nothing wrong. It is independence that
fails, and only the exponential mean cares about the difference.

Half a page earlier he concedes the discomfort in his own voice, even for the
continuous case: "we may not wish to think $v_2$ is independent of $\mathbf{z}$
in $y_2 = \mathbf{z}\boldsymbol{\pi}_{o2} + v_2$ (even though obtaining the form
of $E(y_1 \mid \mathbf{z}, y_2)$ uses something like it)."

#### The subtlety: discreteness of the residual is not the problem

It is tempting to summarise this as "the residual is binary, so it cannot be a
control function." That argument proves too much, because **the generalized
residual of the next section is two-valued given $\mathbf{z}$ as well.**

The difference is not the discreteness of the derived residual. It is *which
object independence is asserted about*. The linear-residual CF asserts it of the
**observed** $v_2$, where both the support points and their probabilities move
with $\mathbf{z}$. Terza's setup asserts it of the **continuous latent** probit
error, $v_2 \mid \mathbf{z} \sim \mathrm{Normal}(0,1)$ — which is the assumption
every probit already makes, and costs nothing extra.

So the fix is not to find a continuous control function. It is to relocate the
independence assumption to a latent variable where it is plausible.

#### What survives: the test, but not the correction

The linear-residual construction is not worthless for binary $y_2$. It remains
**fully valid as a Hausman-type exogeneity test**, and Wooldridge is explicit
about why [@wooldridge-2010-crosssection, p. 744]:

> "For the purposes of getting a limiting chi-square distribution, it does not
> matter where the linear combination $\hat v_2$ comes from. In other words,
> **under the null hypothesis none of the assumptions we made about
> $(c_1, v_2)$ need to hold**: $v_2$ need not be independent of $\mathbf{z}$,
> and $e_1$ in equation (18.39) need not be independent of $v_2$. Therefore, as
> a test, this procedure is very robust, and it can be applied when $y_2$
> contains binary, count, or other discrete variables. **Unfortunately, if $y_2$
> is endogenous, the correction does not work without something like the
> assumptions made previously.**"

That is the cleanest statement of the position: under $H_0: \rho = 0$ there is
no endogeneity left to approximate, so the test needs no reduced-form
assumption. Once $\rho \neq 0$, the correction needs one, and for binary $y_2$
it is unavailable.

### Terza's exact correction

@terza-1998, exposited in @wooldridge-2010-crosssection (sec. 18.5) and
@imbens-wooldridge-2007-cf (sec. 3.3, eqs. 3.32–3.36), replaces the linear
reduced form with a **threshold model** — Wooldridge's eq. (18.43):

$$y_2 = \mathbb{1}[\mathbf{z}\boldsymbol{\pi}_2 + v_2 \geq 0], \qquad v_2 \mid \mathbf{z} \sim \mathrm{Normal}(0,1),$$ {#eq-poisson-iv-5}

and *keeps* the independence assumption, now about the latent $v_2$. The
derivation then runs in two steps. With $(c_1, v_2)$ jointly normal, mean zero
and independent of $\mathbf{z}$, write $c_1 = \rho_1 v_2 + e_1$ where
$e_1 \mid \mathbf{z}, v_2 \sim \mathrm{Normal}(0, \tau_1^2 - \rho_1^2)$.
Conditioning on the latent error gives eq. (18.44),

$$E(y_1 \mid \mathbf{z}, v_2) = E[\exp(e_1)]\exp(\mathbf{x}_1\boldsymbol{\beta}_1 + \rho_1 v_2) = \exp\!\bigl((\tau_1^2 - \rho_1^2)/2 + \mathbf{x}_1\boldsymbol{\beta}_1\bigr)\exp(\rho_1 v_2),$$ {#eq-poisson-iv-6}

using the lognormal mean. But $v_2$ is not observed — only $y_2$ is. Averaging
over the truncated normal implied by $y_2$ gives eq. (18.45),
$E(y_1 \mid \mathbf{z}, y_2) = \exp((\tau_1^2-\rho_1^2)/2)\exp(\mathbf{x}_1\boldsymbol{\beta}_1)\,E[\exp(\rho_1 v_2) \mid \mathbf{z}, y_2]$,
and evaluating that last expectation yields the exact result:

$$E(y_1 \mid \mathbf{z}, y_2) = \exp(\mathbf{z}_1\boldsymbol{\delta}_1 + \alpha_1 y_2)\cdot h(y_2, \mathbf{z}\boldsymbol{\pi}_2, \rho)$$ {#eq-poisson-iv-7}
$$h(y_2, \mathbf{z}\boldsymbol{\pi}_2, \rho) = \exp(\rho^2/2)\left\{y_2\,\frac{\Phi(\rho + \mathbf{z}\boldsymbol{\pi}_2)}{\Phi(\mathbf{z}\boldsymbol{\pi}_2)} + (1-y_2)\,\frac{1-\Phi(\rho + \mathbf{z}\boldsymbol{\pi}_2)}{1-\Phi(\mathbf{z}\boldsymbol{\pi}_2)}\right\}$$ {#eq-poisson-iv-8}

This is estimated by plugging in the first-stage probit's
$\hat{\boldsymbol{\pi}}_2$ and jointly QMLE-fitting
$(\boldsymbol\delta_1, \alpha_1, \rho)$ against this mean — a genuinely
nonlinear second stage, **not** a linear regressor addition.

#### $\rho$ here is a coefficient, not a correlation

$\rho$ is the coefficient in the linear projection of $c_1$ on $v_2$, i.e.
$E[\exp(c_1) \mid v_2] = \exp(\rho v_2)$. With $v_2$ standard normal the two
differ by a factor of $\sigma_{c_1}$, since
$\rho = \mathrm{Corr}(c_1, v_2)\,\sigma_{c_1}$.

The distinction is not cosmetic. This chapter's binary DGP sets
$c_1 = 0.2\,e_2$ with $v_2 = e_2$, so the *correlation* is exactly $1$ while the
$\rho$ governing the approximation below is $0.2$. Reading $\rho$ as a
correlation is also what $h(\cdot)$ forbids — it is $\rho$ in the coefficient
sense that makes $\exp(\rho^2/2)$ the lognormal correction and puts $\rho$ on
the same scale as the probit index.

### The generalized-residual CF as a first-order approximation

Taylor-expand $h$ in $\rho$ around $\rho = 0$. Since
$\tfrac{d}{d\rho}\exp(\rho^2/2)\big|_{\rho=0} = 0$, only the $\Phi$-ratios
contribute to first order:

$$\left.\frac{\partial}{\partial\rho}\frac{\Phi(\rho+\mathbf{z}\boldsymbol\pi_2)}{\Phi(\mathbf{z}\boldsymbol\pi_2)}\right|_{\rho=0} = \frac{\phi(\mathbf{z}\boldsymbol\pi_2)}{\Phi(\mathbf{z}\boldsymbol\pi_2)} = \lambda(\mathbf{z}\boldsymbol\pi_2), \qquad
\left.\frac{\partial}{\partial\rho}\frac{1-\Phi(\rho+\mathbf{z}\boldsymbol\pi_2)}{1-\Phi(\mathbf{z}\boldsymbol\pi_2)}\right|_{\rho=0} = -\lambda(-\mathbf{z}\boldsymbol\pi_2)$$ {#eq-poisson-iv-9}

where $\lambda(\cdot) = \phi(\cdot)/\Phi(\cdot)$ is the inverse Mills ratio.
Since $h(y_2, \mathbf{z}\boldsymbol\pi_2, 0) = 1$,

$$h(y_2,\mathbf{z}\boldsymbol\pi_2,\rho) \approx 1 + \rho\cdot\underbrace{\bigl[y_2\lambda(\mathbf{z}\boldsymbol\pi_2) - (1-y_2)\lambda(-\mathbf{z}\boldsymbol\pi_2)\bigr]}_{=\ \hat r_2,\ \text{the generalized residual}}$$ {#eq-poisson-iv-10}

and $\exp(\mathbf{x}_1\boldsymbol\beta_1)\cdot h(\cdot) \approx \exp(\mathbf{x}_1\boldsymbol\beta_1)(1+\rho\hat r_2) \approx \exp(\mathbf{x}_1\boldsymbol\beta_1 + \rho\hat r_2)$
for small $\rho\hat r_2$ — which is exactly "add $\hat r_2$ linearly inside
$\exp(\cdot)$."

So the **generalized-residual CF is the first-order-in-$\rho$ local
approximation to Terza's exact correction**, with error $O(\rho^2)$. A direct
large-$n$ check against this chapter's own DGP gives probability limits of
$0.839$ for the linear-residual CF and $0.822$ for the generalized-residual CF
against a true $0.8$ — both biased, the latter roughly half the former, matching
the theoretical ranking, and neither exact.

#### Why use the approximation rather than the exact correction?

Because $h(\cdot)$ does not reduce to $\exp(\text{linear index})$. The
$\Phi$-ratios and the $\exp(\rho^2/2)$ term make the second stage genuinely
nonlinear in $\rho$, so it needs custom optimisation code, and combining that
with the within-group demeaning that makes FE-Poisson tractable is not
straightforward. The cost scales badly precisely where FE-Poisson is most needed.

The approximation's second stage, by contrast, has the same functional form as
ordinary Poisson with one extra regressor in the index, so it runs in `fepois`,
`ppmlhdfe` or `xtpoisson` unmodified. Joint nonlinear estimation would be more
efficient, but Wooldridge notes for the closely analogous binary-$y_1$ case that
the iterations can be hard to converge, with $\hat\rho_1$ drifting toward $\pm 1$
[@wooldridge-2010-crosssection, sec. 15.7.3]. At small $\rho$ the approximation
error is minor next to sampling noise, and the coefficient on $\hat r_2$ doubles
as a Hausman-type $t$-test sitting next to the coefficient of interest.

#### Probit or logit — and why logit is simpler than it looks

Nothing in the local-approximation argument is specific to the normal. The
generalized residual is defined for *any* correctly specified binary-response
GLM [@gourieroux-monfort-renault-trognon-1987], and for a **logit** it collapses.
With $\Lambda(\cdot)$ the logistic CDF and
$\lambda(\eta) = \Lambda(\eta)[1-\Lambda(\eta)]$ its density,

$$\hat r_2 = y_2\frac{\lambda(\hat\eta)}{\Lambda(\hat\eta)} - (1-y_2)\frac{\lambda(\hat\eta)}{1-\Lambda(\hat\eta)} = y_2[1-\Lambda(\hat\eta)] - (1-y_2)\Lambda(\hat\eta) = y_2 - \Lambda(\hat\eta) = y_2 - \hat p_2,$$ {#eq-poisson-iv-11}

i.e. **the logit generalized residual is just the plain response residual**
$y_2 - \hat p_2$. No Mills-ratio machinery: `logit`, then
`y2 - predict(fit, type="response")`, *is* the generalized residual — not a
shortcut to it.

Two limits. Terza's *exact* correction is probit-specific: $h(\cdot)$ comes from
the truncated moments of the bivariate normal, and no analogue for a logistic
$v_2$ appears in @terza-1998, @wooldridge-2010-crosssection or
@imbens-wooldridge-2007-cf. And the link-robustness that the GMM route enjoys
belongs to using the fitted value as an *instrument*, not as a *residual*; it
does not carry over to either control function, where the second-stage
conditional mean still has to be right.

### A different family: multiplicative-moment GMM

The third route abandons the control function. Any function of the exogenous
$\mathbf{z}$ is a valid instrument, so the fitted probability $\hat p_2$ can
instrument $y_2$ directly in a GMM problem that leaves the exponential mean
untouched. The moments are Mullahy's multiplicative ones,

$$E[\mathbf{z}'(y_1 e^{-\mathbf{x}\boldsymbol{\beta}} - 1)] = 0,$$ {#eq-poisson-iv-12}

which is the right choice when the unobserved heterogeneity sits inside
$\exp(\cdot)$, as it does for counts. The additive alternative
$E[\mathbf{z}'(y_1 - \exp(\mathbf{x}\boldsymbol{\beta}))] = 0$ assumes an
additive structural error and is inconsistent here.

Two consequences follow, and they are the whole tradeoff. **The validity of this
route does not depend on the first-stage model being right** — $\hat p_2$ need
only be a function of $\mathbf{z}$. But **it does not accommodate
high-dimensional fixed effects**: absorbing them would require within-group
demeaning of the regressor and instrument matrices before solving nonlinear
moment conditions, which is not straightforward.

## The three estimators compared

Two questions separate these estimators, and everything else they share. **Is it
valid when $y_2$ is binary, and can it absorb high-dimensional fixed effects?**

| | Linear-residual CF | Generalized-residual CF | Multiplicative-moment GMM |
|:---|:---|:---|:---|
| **How it works** | OLS residual $\hat v_2$ in the index | probit/logit generalized residual $\hat r_2$ in the index | fitted $\hat p_2$ as an *instrument* |
| **Valid for binary $y_2$?** | **No** — reduced-form assumption cannot hold | **Approximately** — error $O(\rho^2)$, derived | **Yes** — consistent |
| **First stage must be correct?** | moot | yes | **no** |
| **Takes high-dimensional FE?** | yes | yes | **no** |

**No estimator gets both.** The generalized-residual CF comes closest, buying
fixed effects at the price of an approximation whose error is at least *derived*
and shrinks with $\rho$. The GMM route is the only genuinely consistent one, and
pays for it by giving up fixed effects — and, as the simulations show, by roughly
tripling the variance. The linear-residual CF is the fallback: for binary $y_2$
it has no valid reduced form behind it, though it remains a fully valid
*exogeneity test* (§ above) and is accurate when $\rho$ is small.

Common to all three: the second-stage exponential mean must be correctly
specified, and the two control functions need bootstrapped standard errors
because their residual is generated. Only the GMM route reports usable standard
errors directly, from the usual sandwich.

## Simulation 1 — continuous endogenous variable

### Data generating process

We simulate a panel where `time` (continuous, e.g. minutes on site) is
endogenous: it is correlated with an unobserved factor `e` that also affects
`visits`. The excluded instrument is `phone`. Fixed effects are `ad` (20 groups)
and `female`.

There are 20 ad groups of 250 observations each, $n = 5000$. Each group draws a
fixed effect $\alpha_g \sim N(0, 0.5^2)$. At the individual level,
$\text{female} \sim \text{Bernoulli}(0.5)$,
$\text{phone} \sim \text{Bernoulli}(0.4)$, $\text{frfam} \sim U(0,1)$, and the
unobservable is $e \sim N(0,1)$. The endogenous regressor is

$$
\text{time} = 1.5\,\text{phone} + 0.5\,\text{frfam} + \alpha_g + e,
$$ {#eq-poisson-iv-13}

and the count is $\text{visits} \sim \text{Poisson}(\lambda)$ with

$$
\lambda = \exp(0.5 + 0.8\,\text{time} + 0.4\,\text{frfam} + \alpha_g
          + 0.3\,\text{female} + 0.5\,e).
$$ {#eq-poisson-iv-14}

The true $\alpha_1$ is 0.8. The same $e$ appears in `time` and in the log mean,
which is the endogeneity, and `phone` appears only in `time`, which is the
exclusion restriction. Note that $e$ enters the log mean *linearly inside*
$\exp(\cdot)$: @eq-poisson-iv-2 holds here by construction, so this simulation
tests the estimator rather than the assumption.

```{r}
#| echo: true
set.seed(42)
n_ad <- 20; obs_per <- 250; n <- n_ad * obs_per

fe_ad  <- rnorm(n_ad) * 0.5
fe_grp <- rep(1:n_ad, each = obs_per)
female <- as.integer(runif(n) < 0.5)
phone  <- as.integer(runif(n) < 0.4)
frfam  <- runif(n)
e      <- rnorm(n)

time <- 1.5 * phone + 0.5 * frfam + fe_ad[fe_grp] + e

true_alpha <- 0.8
lambda <- exp(0.5 + true_alpha * time + 0.4 * frfam +
              fe_ad[fe_grp] + 0.3 * female + 0.5 * e)
visits <- rpois(n, lambda)

df1 <- data.frame(
  visits = visits, time = time, phone = phone,
  frfam = frfam, female = female, ad = fe_grp
)

cat(sprintf("n = %d   groups = %d   visits mean = %.2f\n", n, n_ad, mean(visits)))
```

### Naive Poisson — the endogeneity bias

```{r}
#| echo: true
naive1 <- fepois(visits ~ time + frfam | ad + female, data = df1)
etable(naive1, title = "Naive Poisson — time is endogenous")
```

The naive estimate is 1.139 against a true 0.8, an upward bias of 0.34. The sign
follows the callout above: $\rho = 0.5 > 0$, so omitting the control function
leaves a multiplicative $e^{0.5e}$ term positively correlated with `time`. The
magnitude is roughly predictable too. Conditioning on `frfam`, `ad` and `female`
but not `phone`, the surviving variation in `time` is $1.5\,\text{phone} + e$
with variance $1.5^2(0.4)(0.6) + 1 = 1.54$, so a linear approximation puts the
bias near $0.5 \times 1/1.54 = 0.32$, close to the 0.34 observed.

### The linear-residual control function

Project `time` on `phone`, `frfam` and the fixed effects, then add the residual
to a Poisson QMLE. A linear first stage is appropriate here because `time` is
continuous — the condition that fails for binary $y_2$ holds in this design.

Note that unlike 2SLS, $\hat v_2$ does *not* replace $y_2$; both enter together.

```{r}
#| echo: true
# Step 1: linear projection
fs1       <- feols(time ~ phone + frfam | ad + female, data = df1)
df1$v2hat <- residuals(fs1)

# Step 2: Poisson CF
m1_cf <- fepois(visits ~ time + v2hat + frfam | ad + female, data = df1)
etable(m1_cf, title = "CF: linear projection first stage")
cat(sprintf("time coefficient: %.4f  (true %.1f)\n", coef(m1_cf)["time"], true_alpha))
```

The coefficient on `time` is 0.7801, within 2.5 percent of the true 0.8, against
the naive 1.139. The coefficient on `v2hat` is 0.5150, which is $\rho$, and the
DGP sets $\rho = 0.5$ — the control function recovers the endogeneity parameter
as well as the structural one. Its $t$-statistic of 103 is the Hausman-type test,
and it rejects exogeneity decisively.

### Bootstrap standard errors

The second-stage standard errors are wrong because first-stage estimation error
propagates. A cluster bootstrap by `ad` corrects this. `parLapply` with a PSOCK
cluster parallelises the 500 replications safely inside Quarto's rendering
environment:

```{r}
#| echo: true
bootstrap_cf_continuous <- function(df, B = 500) {
  clusters <- unique(df$ad)
  K        <- length(clusters)

  boot_one <- function(b) {
    library(fixest)
    set.seed(b)
    drawn <- sample(clusters, K, replace = TRUE)
    df_b  <- do.call(rbind, lapply(seq_along(drawn), function(i) {
      sub         <- df[df$ad == drawn[i], ]
      sub$ad_boot <- i
      sub
    }))
    fs         <- feols(time ~ phone + frfam | ad_boot + female, data = df_b)
    df_b$v2hat <- residuals(fs)
    ss         <- fepois(visits ~ time + v2hat + frfam | ad_boot + female, data = df_b)
    coef(ss)["time"]
  }

  n_cores <- max(1L, detectCores() - 1L)
  cl      <- makePSOCKcluster(n_cores)
  clusterExport(cl, c("df", "clusters", "K"), envir = environment())
  results <- parLapply(cl, seq_len(B), boot_one)
  stopCluster(cl)
  unlist(results)
}

boot1    <- bootstrap_cf_continuous(df1, 500)
coef1    <- coef(m1_cf)["time"]
naive_se <- se(m1_cf)["time"]

cat(sprintf("time coefficient:   %.4f  (true %.1f)\n", coef1, true_alpha))
cat(sprintf("Naive 2nd-stage SE: %.4f\n", naive_se))
cat(sprintf("Bootstrap SE:       %.4f\n", sd(boot1)))
cat(sprintf("Bootstrap 95%% CI: [%.4f, %.4f]\n",
            quantile(boot1, 0.025), quantile(boot1, 0.975)))
```

The bootstrap standard error is 0.0079 against the naive second-stage 0.0041,
almost twice as large, confirming that ignoring first-stage uncertainty
understates uncertainty by about half here. The 95% interval happens to just
exclude the true 0.8: a Monte Carlo over fresh seeds centres the CF estimator on
0.80 (mean 0.802, sd 0.009), so this draw is simply unlucky — and with only 20
clusters the bootstrap interval is itself noisy.

## Simulation 2 — binary endogenous variable

### Data generating process

A clean probit DGP, with a fresh structural error `e2` independent of
Simulation 1. The endogeneity coefficient $\rho = 0.2$ is kept small so the
first-order approximation stays tight and the estimators can be compared without
that error dominating.

The covariates `phone`, `frfam`, `female` and the group effects $\alpha_g$ carry
over. The binary endogenous variable is a probit threshold,

$$
\text{time\_hi} = \mathbb{1}\{1.5\,\text{phone} + 0.4\,\text{frfam} + e_2 \ge 0\},
$$ {#eq-poisson-iv-15}

and the count is $\text{visits} \sim \text{Poisson}(\lambda_2)$ with

$$
\lambda_2 = \exp(0.5 + 0.8\,\text{time\_hi} + 0.4\,\text{frfam} + \alpha_g
            + 0.3\,\text{female} + 0.2\,e_2).
$$ {#eq-poisson-iv-16}

The true $\alpha_1$ is again 0.8, and $\rho = 0.2$. Because the latent index is
exactly a probit, the generalized-residual CF's first stage is correctly
specified by construction, which isolates the approximation error from any
misspecification.

```{r}
#| echo: true
set.seed(123)
e2      <- rnorm(n)

# Binary endogenous variable: clean probit latent index
time_hi <- as.integer(1.5 * phone + 0.4 * frfam + e2 >= 0)

# Poisson outcome — true coefficient 0.8, endogeneity via ρ*e2 (ρ=0.2)
rho2    <- 0.2
lambda2 <- exp(0.5 + true_alpha * time_hi + 0.4 * frfam +
               fe_ad[fe_grp] + 0.3 * female + rho2 * e2)
visits2 <- rpois(n, lambda2)

df2 <- data.frame(
  visits = visits2, time_hi = time_hi, phone = phone,
  frfam = frfam, female = female, ad = fe_grp, e2 = e2
)

cat(sprintf("time_hi mean:      %.3f\n", mean(time_hi)))
cat(sprintf("Cor(time_hi, e2):  %.3f  (endogeneity)\n", cor(time_hi, e2)))
cat(sprintf("visits mean:       %.2f\n", mean(visits2)))
```

73% of observations have `time_hi = 1`, the correlation between `time_hi` and
$e_2$ is 0.622, and mean visits are 6.18, well below Simulation 1's 26.4 — a
binary regressor cannot generate the exponential variation that a continuous,
unbounded `time` does.

### Linear-residual CF

The estimator with no valid reduced-form assumption here, included to see how
far wrong it goes.

```{r}
#| echo: true
fs_a      <- feols(time_hi ~ phone + frfam | ad + female, data = df2)
df2$v2hat <- residuals(fs_a)

m_a <- fepois(visits ~ time_hi + v2hat + frfam | ad + female, data = df2)
etable(m_a, title = "Approach A: OLS residual as control function")
```

It gives 0.8613, with a coefficient of 0.2674 on `v2hat`.

### Generalized-residual CF

```{r}
#| echo: true
fs_b <- feglm(time_hi ~ phone + frfam | ad + female,
              data = df2, family = binomial(probit))

# Generalized residual: r̂ = y·φ(η)/Φ(η) − (1−y)·φ(η)/(1−Φ(η))
eta      <- predict(fs_b, type = "link")
df2$gr2  <- ifelse(df2$time_hi == 1,
                   dnorm(eta) / pnorm(eta),
                   -dnorm(eta) / (1 - pnorm(eta)))

m_b <- fepois(visits ~ time_hi + gr2 + frfam | ad + female, data = df2)
etable(m_b, title = "Approach B: probit generalized residual")
```

It gives 0.8544, slightly closer to the truth than 0.8613, with a coefficient of
0.1664 on the generalized residual. That coefficient estimates $\rho$, whose true
value is 0.2; it is not exact because the estimator is itself only a first-order
approximation.

### Probit or logit

Confirming @eq-poisson-iv-11 numerically — the logit generalized residual is
`time_hi - phat`:

```{r}
#| echo: true
fs_b_logit <- feglm(time_hi ~ phone + frfam | ad + female,
                     data = df2, family = binomial(logit))
df2$phat_logit <- predict(fs_b_logit, type = "response")
df2$gr2_logit  <- df2$time_hi - df2$phat_logit   # algebraically = generalized residual

m_b_logit <- fepois(visits ~ time_hi + gr2_logit + frfam | ad + female, data = df2)
etable(m_b_logit, title = "Approach B with a logit first stage")
cat(sprintf("Probit-based Approach B: alpha1 = %.4f\n", coef(m_b)["time_hi"]))
cat(sprintf("Logit-based  Approach B: alpha1 = %.4f\n", coef(m_b_logit)["time_hi"]))
```

The logit first stage gives 0.8625 against the probit's 0.8544, a difference of
0.008 on estimates whose standard errors are 0.032. The choice of link is not
what matters here.

### Multiplicative-moment GMM

Probit fitted probabilities as the instrument, via `gmm::gmm()` with
@eq-poisson-iv-12. The additive moments would be inconsistent on these data,
because the DGP puts its unobservables (the `ad` effects, `female`, and
$\rho_2 e_2$) inside $\exp(\cdot)$ — running the additive version gives
$\hat\alpha_1 \approx 0.63$, a bias no intercept shift can absorb:

```{r}
#| echo: true
# Probit without FE for fitted probabilities (instrument)
fs_c     <- glm(time_hi ~ phone + frfam, family = binomial(probit), data = df2)
df2$phat <- fitted(fs_c)

# Multiplicative moment conditions (Mullahy 1997): Z'(y*exp(-X*beta) - 1) = 0
# X = [time_hi, frfam, 1],  Z = [phat, frfam, 1]
gmm_moments <- function(beta, x) {
  y      <- x[, "visits"]
  th     <- x[, "time_hi"]
  fr     <- x[, "frfam"]
  ph     <- x[, "phat"]
  eta    <- pmin(pmax(beta[1] * th + beta[2] * fr + beta[3], -30), 30)
  resid  <- y * exp(-eta) - 1
  cbind(ph * resid, fr * resid, resid)
}

gmm_data <- as.matrix(df2[, c("visits", "time_hi", "frfam", "phat")])
m_c      <- gmm(gmm_moments, x = gmm_data,
                t0 = c(0, 0, log(mean(df2$visits))))
coef_c   <- coef(m_c)
se_c     <- sqrt(diag(vcov(m_c)))

cat(sprintf("\nApproach C — IV/GMM Poisson (no FE, for illustration)\n"))
cat(sprintf("  time_hi:  coef = %7.4f   SE = %7.4f   z = %6.3f\n",
            coef_c[1], se_c[1], coef_c[1] / se_c[1]))
cat(sprintf("  frfam:    coef = %7.4f   SE = %7.4f   z = %6.3f\n",
            coef_c[2], se_c[2], coef_c[2] / se_c[2]))
```

GMM gives 0.7844 with a standard error of 0.0656 on `time_hi`, and 0.4576 on
`frfam` against a true 0.4. The point estimate is the closest of the three to
0.8, and the standard error is twice the others' — the trade-off the comparison
below takes apart.

**Why this one has no fixed effects.** Adding them would require demeaning the regressor and instrument matrices within
group before solving the Poisson moment conditions, and because Poisson is
nonlinear that is not straightforward. Both control functions run unmodified in
`feols`/`feglm` + `fepois`, so either is practical when many fixed effects are
present.

### Control function and GMM are not equivalent in Poisson

In a linear model, using $\hat y_2$ as an instrument (2SLS) is numerically
identical to adding $\hat v_2 = y_2 - \hat y_2$ as a regressor, by
Frisch–Waugh–Lovell. **In Poisson this equivalence breaks**, because the control
function puts $\hat v_2$ inside $\exp(\cdot)$ while the GMM route uses $\hat p_2$
in the moment conditions without modifying the exponential mean.

```{r}
#| echo: true
coef_a <- coef(m_a)["time_hi"]
coef_b <- coef(m_b)["time_hi"]

cat(sprintf("Approach A  (CF, OLS residual, with FE):   α̂₁ = %.4f\n", coef_a))
cat(sprintf("Approach B  (CF, probit GR, with FE):      α̂₁ = %.4f\n", coef_b))
cat(sprintf("Approach C  (IV/GMM Poisson, no FE):       α̂₁ = %.4f\n", coef_c[1]))
cat(sprintf("True α₁ = %.1f\n", true_alpha))
```

The three land at 0.8613, 0.8544 and 0.7844 against a true 0.8. They differ
despite using the same exclusion restriction, which is the point: in a linear
model FWL would force the first and third to agree exactly, and here they are
0.077 apart. **2SLS $\neq$ CF for Poisson.**

### Bootstrap for the binary case

```{r}
#| echo: true
bootstrap_cf_binary <- function(df, B = 500) {
  clusters <- unique(df$ad)
  K        <- length(clusters)

  boot_one <- function(b) {
    library(fixest)
    set.seed(b)
    drawn <- sample(clusters, K, replace = TRUE)
    df_b  <- do.call(rbind, lapply(seq_along(drawn), function(i) {
      sub         <- df[df$ad == drawn[i], ]
      sub$ad_boot <- i
      sub
    }))
    fs         <- feols(time_hi ~ phone + frfam | ad_boot + female, data = df_b)
    df_b$v2hat <- residuals(fs)
    ss         <- fepois(visits ~ time_hi + v2hat + frfam | ad_boot + female, data = df_b)
    coef(ss)["time_hi"]
  }

  n_cores <- max(1L, detectCores() - 1L)
  cl      <- makePSOCKcluster(n_cores)
  clusterExport(cl, c("df", "clusters", "K"), envir = environment())
  results <- parLapply(cl, seq_len(B), boot_one)
  stopCluster(cl)
  unlist(results)
}

boot2     <- bootstrap_cf_binary(df2, 500)
naive_se2 <- se(m_a)["time_hi"]

cat(sprintf("Approach A bootstrap (binary EEV, 500 reps, cluster by ad)\n"))
cat(sprintf("  Point estimate:     %.4f  (true %.1f)\n", coef_a, true_alpha))
cat(sprintf("  Naive 2nd-stage SE: %.4f\n", naive_se2))
cat(sprintf("  Bootstrap SE:       %.4f\n", sd(boot2)))
cat(sprintf("  Bootstrap 95%% CI:  [%.4f, %.4f]\n",
            quantile(boot2, 0.025), quantile(boot2, 0.975)))
```

The bootstrap standard error is 0.0331 against the naive 0.0321 — almost
identical here, unlike Simulation 1 where the bootstrap was twice the naive
figure. The binary first stage carries much less estimation uncertainty into the
second stage than the continuous one did. The 95% interval $[0.803, 0.931]$
excludes the true 0.8, which is the approximation bias showing up in inference,
not a coverage failure of the bootstrap.

### All methods side by side

```{r}
#| echo: true
naive_p2    <- fepois(visits ~ time_hi + frfam | ad + female, data = df2)
coef_naive2 <- coef(naive_p2)["time_hi"]

methods <- c(
  "True value",
  "Naive Poisson (binary y₂)",
  "A: Linear CF (with FE)",
  "B: Probit GR CF (with FE)",
  "C: IV/GMM Poisson (no FE)"
)
ests <- c(true_alpha, coef_naive2, coef_a, coef_b, coef_c[1])

cat("\n── Binary EEV: coefficient on time_hi ─────────────────────\n")
for (i in seq_along(methods)) {
  cat(sprintf("  %-38s  α̂₁ = %6.4f\n", methods[i], ests[i]))
}
```

The naive Poisson gives 1.0832 against a true 0.8, so ignoring the endogeneity
overstates the effect by 35%. Both control functions recover the true value
closely, at 0.8613 and 0.8544, with the small upward bias expected from
$\rho = 0.2$ and the binary $y_2$.

The GMM estimate lands closest in this draw, and its near-centring is not luck:
the omitted `ad` and `female` effects are independent multiplicative
heterogeneity, absorbed by the intercept under the multiplicative moments
($\beta_0 \to \beta_0 + \log E[e^{c}]$), so omitting the fixed effects costs no
slope *bias* at all. Mullahy's moments are *consistent* for this DGP, where the
control functions are only approximately unbiased: over 10 seeds at this $n$, GMM
averages $0.7985$ against $0.8323$ and $0.8185$, and at $n = 10^6$ it sits at
$0.8015$ (sd $0.0013$) against plims of $0.839$ and $0.822$.

The catch is variance rather than bias. Without the fixed effects GMM is about
three times as noisy (sd $0.068$ at $n = 5000$, against $0.024$ and $0.026$), so
at this sample size its RMSE is the **worst** of the three — $0.065$, against
$0.039$ and $0.031$ — and a single draw can easily leave it further from $0.8$
than either control function.

## Practical guidance

- **With fixed effects and a binary $y_2$, use the generalized-residual CF.**
  Among the estimators that drop into FE-Poisson unmodified it is the
  theoretically motivated choice, and its error is *derived*, not assumed:
  $O(\rho^2)$, shrinking as the endogeneity shrinks.

- **Multiplicative-moment GMM trades bias for variance — decide which binds.**
  Mullahy's moments are *consistent* here, not merely accurate to first order,
  whenever the multiplicative heterogeneity is independent of $\mathbf{z}$. But
  consistency is not accuracy at a given sample size: over 10 seeds at $n=5000$
  it is nearly unbiased (mean $0.7985$) and yet almost three times as noisy
  (sd $0.068$ against $0.026$), giving it the worst RMSE of the three. Prefer it
  once $n$ is large enough that variance is not the binding constraint — and note
  that it does not accommodate high-dimensional fixed effects either way.

- **The linear-residual CF is a fallback, not a first choice.** It lands close to
  the truth at $\rho = 0.2$ here, but for binary $y_2$ it has no valid
  reduced-form assumption behind it. **Do not lean on it when $\rho$ is not known
  to be small** — and note that $\rho$ is estimated by the coefficient on the
  control function, so this is checkable rather than a matter of faith.

- **It remains fully valid as an exogeneity test.** Under $H_0: \rho = 0$ none of
  the reduced-form assumptions are needed
  [@wooldridge-2010-crosssection, p. 744]. Use it to test; do not use it to
  correct.

- **Probit or logit both work for the generalized-residual CF.** Only Terza's
  exact correction needs normality.

- **Always bootstrap.** In every control-function specification above the
  second-stage standard errors are wrong, because the residual is generated.
