# Recentered Formula Instruments

```{r}
#| include: false
library(tidyverse)
library(fixest)
library(parallel)
library(knitr)
setFixest_nthreads(1)
```

Many treatments are computed from a formula that combines a random shock with
non-random exposure. Medicaid eligibility is a state's rules applied to a
person's income. Import exposure is national industry shocks applied to a
region's industry mix. The Bartik instrument of the previous chapter is one
such formula. This chapter covers the general version of Borusyak and Hull
[@borusyak-hull-2023], their result on which formula is best
[@borusyak-hull-2026], and what both mean for shift-share.

## Formula instruments

Unit $i$ has outcome $y_i$ and treatment $x_i$, with a constant effect $\beta$:

$$
y_i = \beta x_i + \varepsilon_i .
$$ {#eq-ri-model}

The treatment is $x_i = h_i(g, w, u)$, where $h_i$ is a known function, $g$
collects the random shocks, $w$ collects predetermined variables such as
income or industry shares, and $u$ collects anything else. The identifying
assumption puts all the randomness on the shocks [@borusyak-hull-2026,
Assumption 1]:

$$
E[\varepsilon_i \mid g, w] = E[\varepsilon_i \mid w] .
$$ {#eq-ri-exog}

The error may depend on $w$ in any way. Poorer people may have worse
outcomes, and regions with more manufacturing may have been declining anyway.

A formula instrument is $z_i = f_i(g, w)$ for a fixed function $f_i$. Its
expected value over the shock draws, holding $w$ fixed, is the expected
instrument $\mu_i = E[z_i \mid w]$. The recentered instrument
$\tilde z_i = z_i - \mu_i$ is valid, because by iterated expectations and
@eq-ri-exog,

$$
E[\tilde z_i \varepsilon_i]
= E\big[\tilde z_i\, E[\varepsilon_i \mid w]\big]
= E\big[E[\tilde z_i \mid w]\, E[\varepsilon_i \mid w]\big] = 0 .
$$

Without recentering the same steps give $E[z_i \varepsilon_i] = E[\mu_i
E[\varepsilon_i \mid w]]$, which is not zero when the expected instrument is
correlated with the predictable part of the error. Subtracting $\mu_i$ and
controlling for $\mu_i$ are equivalent by Frisch–Waugh–Lovell. Computing
$\mu_i$ requires the design: how the shocks could have been drawn. We redraw
them from that design, or permute them if they are exchangeable, and average
the formula over the draws.

Any recentered formula is valid, so we can pick the most precise.
@borusyak-hull-2026 [Theorem 1] show it starts from the best prediction of the
treatment, recentered:

$$
\tilde z = E[x \mid g, w] - E[x \mid w] .
$$ {#eq-ri-optimal}

The full optimal instrument adds two refinements, partly controlling for the
predictable part of the error and GLS weighting, which need a model of the
error. The authors drop them. Their algorithm predicts the treatment from its
formula, recenters the prediction by redrawing the shocks, and uses the result
as the instrument. When $x_i$ depends only on $(g, w)$, the prediction is
$x_i$ itself.

## Shift-share as a formula instrument

The Bartik instrument $B_\ell = \sum_k s_{\ell k} g_k$ of @eq-shift-share-iv-1
is the linear formula, shares times shocks. It differs from
@eq-ri-optimal in two ways.

The first is recentering. Suppose the industry shocks are exchangeable within
period $t$, with mean $\bar g_t$, and let $S_\ell = \sum_k s_{\ell k}$ be
region $\ell$'s sum of shares. Then

$$
\mu_\ell = S_\ell\, \bar g_t .
$$ {#eq-ri-ss-mu}

If the shares sum to one, $\mu_\ell = \bar g_t$ is common to all regions and
period effects absorb it, so Bartik is already recentered. If they are
incomplete, as with manufacturing shares, heavily manufacturing regions get a
large instrument in every draw, and Bartik is biased unless we control for
$S_\ell$ interacted with period [@borusyak-hull-jaravel-2022].

The second is the formula. Take two regions with half their employment in
electronics, and an import shock to electronics. Region A's plants compete
with Chinese imports. Region B's make military components and do not. Bartik
gives both half the shock, but only A's import competition rises, so the
instrument varies where the treatment does not. The best predictor gives A
half the shock and B nothing, then recenters each by its own expected value.
Both instruments are valid, and the second has the stronger first stage. When
the treatment really is shares times shocks, the best predictor is Bartik
itself and the two differ only by recentering. The formula matters when the
treatment has more structure than Bartik uses: eligibility rules, export
exposure, transport networks, nonlinear indices.

|                     | Shift-share (Bartik)      | Recentered formula instrument          |
|---------------------|---------------------------|----------------------------------------|
| Formula             | Shares times shocks       | Best prediction of the treatment       |
| Expected instrument | Usually not subtracted    | Subtracted, or controlled for          |
| Shock design        | Implicit                  | Written down to compute $\mu_\ell$     |
| Inputs              | Shares and shocks         | Treatment formula and shock design     |

: Shift-share and recentered formula instruments. {#tbl-ri-compare}

The formula instrument needs more inputs: the treatment's formula and an explicit
shock design. With heterogeneous effects, different instruments weight units
differently, so a gain in precision is a gain for a different weighted
average [@borusyak-hull-2026, Proposition D1].

### Why recenter

Bartik needs one of two assumptions. If the shares are as good as random given
the controls [@goldsmith-pinkham-sorkin-swift-2020], Bartik is valid as it
stands and recentering is not needed. If only the shocks are random
[@borusyak-hull-jaravel-2022], the shares are not, and with incomplete shares
and a positive mean shock a region with more exposure gets a larger instrument
in every draw. That part of the instrument comes from the shares, not from the
shocks, and it is what recentering removes. What remains is the variation from
which industries happened to receive larger shocks than expected.

This gives the formula instrument four advantages. Its validity rests only on the shocks, such as a policy choice or a foreign supply shock, which are often easier to defend than the shares. The
design has to be written down, so the assumption is explicit and can be
varied, as the two ADH designs below show. The instrument can use exposure to
gain power, because the best predictor tracks the treatment more closely than
a shock-only index. And it is not tied to a linear formula or to independent
units: it covers eligibility rules, networks and nonlinear indices, with
shocks shared by many units.

## Simulation

The aim is to compare a shock-only instrument with the recentered treatment.
There are 50 states and 10,000 people. Income $\text{inc}_i$, in multiples of
the poverty line, is log-normal with median 1.4. Each state draws an
eligibility threshold $T_k$ from $\{0.5, 1, 1.38, 2\}$ with equal
probabilities, and

$$
x_i = 1[\text{inc}_i < T_{s(i)}], \qquad
y_i = 0.3\, x_i + \varepsilon_i, \qquad
\varepsilon_i = -0.5 \log \text{inc}_i + \alpha_{s(i)} + \nu_i ,
$$ {#eq-ri-dgp}

where $s(i)$ is $i$'s state, $\alpha_k \sim N(0, 0.3^2)$ is a state effect
and $\nu_i \sim N(0, 1)$. Poorer people are more often eligible and have
higher errors, so OLS is biased. Expected eligibility $\mu_i$ is the share of
the four thresholds above $\text{inc}_i$.

We compare OLS (the same as instrumenting $x_i$ with itself); the simulated
instrument of @currie-gruber-1996, the share of all 10,000 people eligible
under their state's threshold; the recentered treatment $x_i - \mu_i$; and
OLS of $y$ on $x$ and $\mu$, which is the controlled form of the recentered
treatment [@borusyak-hull-2026, footnote 8]. Holding the people fixed, we
redraw the 50 thresholds 1,000 times, so only the shocks change. The people
and draws are stored as CSVs so the Julia companion computes the same numbers.

```{r}
#| label: ri-mc
#| cache: true

people <- read_csv("data/recenter_people.csv", show_col_types = FALSE)
draws  <- as.matrix(read_csv("data/recenter_policy_draws.csv",
                             show_col_types = FALSE)[, -1])
pool <- c(0.5, 1.0, 1.38, 2.0)
beta_true <- 0.3
mu    <- sapply(people$income, function(v) mean(v < pool))  # expected eligibility
gener <- sapply(pool, function(t) mean(people$income < t))  # state generosity

coef_x <- function(f) {
  nm <- intersect(c("fit_x", "x"), names(coef(f)))[1]
  c(est = unname(coef(f)[nm]), se = unname(se(f)[nm]))
}

one_draw <- function(r) {
  thr <- draws[r, people$state]
  d <- mutate(people, x = as.numeric(income < thr), y = beta_true * x + eps,
              z_sim = gener[match(thr, pool)], z_rc = x - mu, mu = mu)
  fits <- list(feols(y ~ x, d, cluster = ~state),
               feols(y ~ 1 | x ~ z_sim, d, cluster = ~state),
               feols(y ~ 1 | x ~ z_rc, d, cluster = ~state),
               feols(y ~ x + mu, d, cluster = ~state))
  rbind(sapply(fits, coef_x),
        r2 = c(NA, cor(d$x, d$z_sim)^2, cor(d$x, d$z_rc)^2, NA))
}

mc <- simplify2array(mclapply(seq_len(nrow(draws)), one_draw, mc.cores = 6))

tibble(estimator = c("OLS", "Simulated instrument", "Recentered treatment",
                     "OLS controlling for mu"),
       mean     = apply(mc["est", , ], 1, mean),
       sd       = apply(mc["est", , ], 1, sd),
       coverage = apply(abs(mc["est", , ] - beta_true) <= 1.96 * mc["se", , ], 1, mean),
       first_stage_R2 = apply(mc["r2", , ], 1, mean)) |>
  kable(digits = 3)
```

OLS averages 0.637 against a true 0.3. The three other estimators are centred
on 0.3, with coverage of 92.7% to 93.7% for clustered 95% intervals with 50
clusters. The simulated instrument has a standard deviation of 0.167 across
draws, the recentered treatment 0.086, and its controlled form 0.080. The
simulated instrument varies only across states, so its first-stage $R^2$ is
0.26. The recentered treatment varies within states, among people whose
income lies between the lowest and highest thresholds, and its $R^2$ is 0.59.

## The China shock

In ADH the shocks are at the industry-period level: the growth of Chinese
imports into eight other high-income countries, per US worker, in 397
manufacturing industries over 1990–2000 and 2000–2007. ADH argue they are
exogenous because the growth came from China's side, through reform, rising
productivity and WTO accession, and because imports into other countries do
not respond to US demand. The argument is weaker for demand or technology
shocks shared by all rich countries, and the identifying variation is
concentrated in a few industries such as computers, toys and audio equipment
[@goldsmith-pinkham-sorkin-swift-2020].

The formula instrument follows the three steps of the algorithm. The
treatment, US import exposure per worker, is shares times US import growth,
and US imports also reflect US demand. Replacing them with the other-country
shocks gives the prediction, which is the Bartik instrument itself, so the
first step changes nothing. The second step does matter, because the shares,
lagged employment shares in the 722 commuting zones, are incomplete. If the
397 shocks are exchangeable within period, @eq-ri-ss-mu gives the expected
instrument $S_\ell \bar g_t$, with a mean shock of 4.8 in the first period and
17.3 in the second. A stricter design treats shocks as exchangeable only
within the 20 two-digit SIC sectors in each period, giving
$\mu_\ell = \sum_m S_{\ell m} \bar g_{mt}$, where $S_{\ell m}$ is the zone's
share in sector $m$ and $\bar g_{mt}$ the sector's mean shock. The third step
controls for the expected instrument, which is equivalent to subtracting it.
ADH themselves control for the start-of-period manufacturing share, with one
coefficient for both periods. All fits use ADH's controls, census-division
effects, population weights and standard errors clustered by state, with the
GPSS replication files.

```{r}
#| label: ri-adh
#| cache: true

adh <- read_csv("data/adh_master.csv", show_col_types = FALSE)
tr  <- read_csv("data/adh_shares_triplets.csv", show_col_types = FALSE)
shk <- read_csv("data/adh_shocks.csv", show_col_types = FALSE)

by_row <- function(v) {                        # sum over industries, by zone
  out <- numeric(nrow(adh)); s <- rowsum(v, tr$row)
  out[as.integer(rownames(s))] <- s[, 1]; out
}
sector  <- paste(shk$sic87dd %/% 100, shk$year)
g_bar_m <- tapply(shk$g, sector, mean)

adh <- mutate(adh, B = by_row(tr$share * shk$g[tr$col]), S = by_row(tr$share),
              mu2 = by_row(tr$share * g_bar_m[sector[tr$col]]))

controls <- "l_sh_popedu_c + l_sh_popfborn + l_sh_empl_f + l_sh_routine33 +
             l_task_outsource + t2 | division"
iv <- function(extra) {
  feols(as.formula(paste("d_sh_empl_mfg ~", extra, controls, "| shock ~ B")),
        adh, weights = ~weights, cluster = ~statefip)
}
fits <- list("ADH controls"                     = iv("l_shind_manuf_cbp +"),
             "Without the manufacturing share"  = iv(""),
             "Design 1: S x period"             = iv("S + S:t2 +"),
             "Design 2: sector expected instrument" = iv("mu2 +"))
tibble(specification = names(fits),
       estimate = sapply(fits, \(f) coef(f)["fit_shock"]),
       se       = sapply(fits, \(f) se(f)["fit_shock"])) |>
  kable(digits = 3)
```

The first row reproduces ADH's −0.596: an increase in import exposure of
$1,000 per worker over a decade lowers the manufacturing employment share by
about 0.6 percentage points. Controlling for the expected instrument period by
period gives −0.267 (0.087). ADH's control is close to $S_\ell$, with a
correlation of 0.92, but one coefficient cannot absorb a term whose slope more
than triples between periods. The sector design gives −0.144 (0.117), because
it no longer uses the fact that electronics and toys received larger shocks
than food or paper.

Each row is right under its own design, and the first row is also right if
the shares themselves are exogenous given the controls. The data cannot tell
us which assumption holds. Recentering requires the design to be stated, and here the design changes the estimate by more than half.

## The Medicaid application

@borusyak-hull-2026 instrument Medicaid eligibility in the 2014 ACA expansion
with actual minus expected eligibility, treating expansion as random within
the governor's party. The first-stage partial $R^2$ rises from 0.01 to 0.80
over the simulated instrument and the standard error on crowd-out falls by
44%. Most of the gain comes from the 18% of the sample whose eligibility the
expansion could change; outside it, the recentered instrument is zero, and in
the full sample without an exposure control it is less precise than the
simulated one [@borusyak-hull-2026, Table A4]. Their 2012–2013 placebo on
private insurance is −0.011, half the main estimate of −0.023 (Table A2).

## Summary

Shift-share is one formula instrument, with the formula and the design fixed
by convention. Any formula of shocks and exposure is valid once its expected
value over the shock design is subtracted, and the most efficient formula is
the recentered best prediction of the treatment. For Bartik this gives two
checks: whether incomplete shares require controlling for the sum of shares
times the mean shock, and whether the treatment depends on the shocks through
more than shares times shocks. In the ADH data the first check alone moves the
estimate from −0.60 to −0.27.

For the shift-share construction, the share and shock views, and inference,
see the previous chapter, [Shift-Share Instrumental Variables](shift-share-iv.qmd).
