# Sensitivity Analysis

```{r}
#| include: false
library(tidyverse)
library(sensemakr)
library(EValue)
library(causaldata)
library(broom)
```

```{r}
#| eval: false
library(tidyverse)
library(sensemakr)
library(EValue)
library(causaldata)
library(broom)
```

The previous chapters often treat identification as yes or no. Applied work
is usually less clean. We may believe the adjustment set is good, but still
worry about some remaining confounding. Sensitivity analysis asks a practical
question: how strong would unmeasured confounding have to be to change the
conclusion?

> **Related reading**: An extended treatment of the Cinelli-Hazlett
> robustness value and benchmark interpretations appears in the
> [companion blog chapter on sensitivity analysis](https://xiangao.github.io/blog_book/sensitivity-analysis.html)
> of *Topics on Econometrics and Causal Inference*.

Here I use three methods:

1. **Cinelli-Hazlett (2020) omitted-variable bias bounds** — for any
   regression-based estimator, quantifies how strong an omitted variable
   would need to be (in terms of partial $R^2$ with treatment and outcome)
   to bring the estimate to zero.
2. **E-values (VanderWeele-Ding 2017)** — for risk ratios, the minimum
   strength of association (on the risk-ratio scale) that an unmeasured
   confounder would need with both treatment and outcome to fully explain
   the observed association.
3. **Rosenbaum bounds** — for matched estimators, the magnitude of
   non-random treatment assignment within matched pairs that would render
   the test result insignificant.

## The NHEFS smoking-cessation example

We use the National Health and Nutrition Examination Survey I Epidemiologic
Follow-up Study (NHEFS) data — the same dataset used in the
[applied DAG-workflow chapter](smoking-cessation-graphs.qmd). The question:
what is the effect of quitting smoking (`qsmk`) on weight gain over
10 years (`wt82_71`)?

```{r}
#| label: data
#| cache: true

nhefs <- causaldata::nhefs_complete |>
  filter(!is.na(wt82_71))

cat(sprintf("n = %d\n", nrow(nhefs)))
cat(sprintf("Treated (quit smoking) = %d\n", sum(nhefs$qsmk)))
cat(sprintf("Mean weight gain (treated):   %.2f kg\n",
            mean(nhefs$wt82_71[nhefs$qsmk == 1])))
cat(sprintf("Mean weight gain (untreated): %.2f kg\n",
            mean(nhefs$wt82_71[nhefs$qsmk == 0])))
```

A naive comparison says quitters gained $\approx$ 2.5 kg more than non-quitters,
but smokers and quitters differ on many baseline characteristics. Adjust for
the standard set of confounders (age, sex, race, education, baseline weight,
smoking intensity, exercise, activity):

```{r}
#| label: ols
#| cache: true

ols <- lm(wt82_71 ~ qsmk + sex + age + race + education + smokeintensity +
                    smokeyrs + exercise + active + wt71,
          data = nhefs)

tidy(ols)[1:3, ] |> knitr::kable(digits = 3)
cat(sprintf("\nAdjusted ACE estimate: %.3f kg  (SE = %.3f)\n",
            coef(ols)["qsmk"], summary(ols)$coefficients["qsmk", 2]))
```

The adjusted estimate is around 3.4 kg, larger than the unadjusted difference.
But the identification assumption requires that we have controlled for *all*
relevant confounders. What if we missed one — say, a propensity to gain
weight that also correlates with quitting decisions? Sensitivity analysis
addresses this directly.

## Cinelli-Hazlett OVB bounds

`sensemakr` implements the Cinelli-Hazlett (2020) approach. The key inputs
are:

- a fitted regression
- the treatment variable name
- one or more *benchmark covariates* whose strength serves as a reference
  for how strong a hypothetical unobserved confounder might be

```{r}
#| label: sensemakr
#| cache: true

sens <- sensemakr(
  model = ols,
  treatment = "qsmk",
  benchmark_covariates = c("smokeintensity", "smokeyrs"),
  kd = 1:3,                 # benchmark: 1×, 2×, 3× as strong as the benchmark
  ky = 1:3,
  q  = 1,                   # the "bias proportion" at which to flag a problem
  alpha = 0.05,
  reduce = TRUE
)

summary(sens)
```

The summary tells us:

- **Robustness value (RV)**: the minimum strength of association (in terms of
  partial $R^2$) that an unmeasured confounder would need with both the
  treatment and the outcome to reduce the estimate to zero. Higher = more
  robust.
- **RV_q**: the strength needed to "explain away" a fraction $q$ of the
  estimate (here $q = 1$ means complete explanation).
- **Benchmark comparisons**: how the RV compares to multiples of the
  observed covariates.

```{r}
#| label: sens-plot
#| cache: true
#| fig-cap: "Cinelli-Hazlett sensitivity contour. The dashed lines show benchmarks based on smokeintensity and smokeyrs at 1x, 2x, and 3x strength. Confounders falling in the shaded region would reduce the estimate to zero."
#| fig-width: 7
#| fig-height: 5

plot(sens)
```

The contour plot shows how strong an unobserved confounder would have to be
with the treatment and outcome to reduce the estimate to zero. The benchmark
dots compare that hypothetical confounder to observed covariates.

## E-values for risk ratios

For binary outcomes and risk-ratio-type estimands, VanderWeele and Ding
(2017) propose the **E-value**: the minimum strength of association on the
risk-ratio scale that an unmeasured confounder would need to have with both
the treatment and the outcome to fully explain the observed risk ratio.

To illustrate, dichotomize weight gain ("any" gain vs no gain) and fit a
risk-ratio model:

```{r}
#| label: evalue
#| cache: true

nhefs$gain <- as.numeric(nhefs$wt82_71 > 0)

rr_fit <- glm(gain ~ qsmk + sex + age + race + education + smokeintensity +
                     smokeyrs + exercise + active + wt71,
              data = nhefs, family = poisson(link = "log"))
rr_summary <- tidy(rr_fit, exp = TRUE, conf.int = TRUE)

rr_est <- exp(coef(rr_fit)["qsmk"])
# The Poisson likelihood is misspecified for a binary outcome (the Zou 2004
# log-binomial-via-Poisson trick), so the model-based variance is wrong.
# Use robust (sandwich/HC) standard errors for the RR confidence interval.
rr_ci  <- lmtest::coefci(rr_fit, parm = "qsmk",
                         vcov. = sandwich::vcovHC(rr_fit, type = "HC0"))
rr_lo  <- exp(rr_ci[1])
rr_hi  <- exp(rr_ci[2])

cat(sprintf("Adjusted RR (quit vs not): %.3f  [%.3f, %.3f]\n",
            rr_est, rr_lo, rr_hi))

ev <- evalues.RR(est = rr_est, lo = rr_lo, hi = rr_hi)
print(ev)
```

The E-value reports two numbers:

- **E-value (point estimate)**: an unmeasured confounder would need to be
  associated with both treatment and outcome by risk ratios of at least
  this value, above and beyond the measured confounders, to fully explain
  away the observed effect.
- **E-value (confidence interval)**: the minimum strength needed to shift
  the lower confidence bound to 1.

An E-value much greater than 1 suggests the result is hard to explain by
unmeasured confounding. An E-value close to 1 means a modest unmeasured
confounder could overturn the conclusion.

The useful part of the E-value is that it is on the risk-ratio scale. It can
be explained without specifying a model for the unobserved confounder.

## Rosenbaum bounds (matched estimators)

For matching estimators, **Rosenbaum bounds** (Rosenbaum 2002) provide a
sensitivity analysis based on the maximum deviation from random assignment
within matched pairs.

Suppose units in a matched pair are observationally identical, but one
might still be more likely than the other to receive treatment due to
unobserved factors. Let $\Gamma$ be the odds ratio of treatment within a
matched pair due to hidden bias: $\Gamma = 1$ means matched pairs are
perfectly randomised; $\Gamma = 2$ means within a pair, the treated unit
was twice as likely to get treatment due to unobservables.

Rosenbaum bounds compute the range of $p$-values for the matched test
under each value of $\Gamma$. The **critical $\Gamma$** is the smallest
value at which the test becomes insignificant.

Here is a worked example using simple pair matching on propensity score and
a Wilcoxon signed-rank test (the original Rosenbaum bound calculation):

```{r}
#| label: rosenbaum
#| cache: true

# Estimate propensity score
ps_fit <- glm(qsmk ~ sex + age + race + education + smokeintensity +
                     smokeyrs + exercise + active + wt71,
              data = nhefs, family = binomial)
nhefs$ps <- predict(ps_fit, type = "response")

# Simple 1:1 nearest-neighbour matching on propensity score
treated_idx <- which(nhefs$qsmk == 1)
control_idx <- which(nhefs$qsmk == 0)

set.seed(1)
matched_pairs <- data.frame(t_idx = integer(), c_idx = integer())
available_controls <- control_idx
for (ti in treated_idx) {
  if (length(available_controls) == 0) break
  dists <- abs(nhefs$ps[ti] - nhefs$ps[available_controls])
  best  <- which.min(dists)
  matched_pairs <- rbind(matched_pairs,
                         data.frame(t_idx = ti,
                                    c_idx = available_controls[best]))
  available_controls <- available_controls[-best]
}

# Compute paired differences in outcome
paired_diff <- nhefs$wt82_71[matched_pairs$t_idx] -
               nhefs$wt82_71[matched_pairs$c_idx]
cat(sprintf("N matched pairs: %d\n", nrow(matched_pairs)))
cat(sprintf("Mean paired difference (matched ATT): %.3f kg\n",
            mean(paired_diff)))

# Rosenbaum bound calculation: range of p-values for Wilcoxon signed-rank
# under each Gamma. We compute this directly rather than using a package
# to keep the dependency footprint small.
rosenbaum_pvalue <- function(diffs, gamma) {
  abs_d <- abs(diffs)
  signs <- sign(diffs)
  ranks <- rank(abs_d)
  # Under Gamma, P(positive sign) is bounded between Γ/(1+Γ) and 1/(1+Γ).
  # Upper-bound p-value: assume P(positive) = Γ/(1+Γ) (against null of zero)
  p_pos_upper <- gamma / (1 + gamma)
  # Test statistic: sum of positive ranks
  T_obs <- sum(ranks[signs > 0])
  n <- length(diffs)
  mu_upper  <- p_pos_upper * sum(ranks)
  var_upper <- p_pos_upper * (1 - p_pos_upper) * sum(ranks^2)
  z <- (T_obs - mu_upper) / sqrt(var_upper)
  1 - pnorm(z)
}

gammas <- seq(1, 4, by = 0.25)
pvals  <- sapply(gammas, function(g) rosenbaum_pvalue(paired_diff, g))

result <- tibble(Gamma = gammas, p_value_upper = pvals)
knitr::kable(result, digits = 4,
             caption = "Upper-bound p-values from Rosenbaum sensitivity analysis")
```

The **critical $\Gamma$** is the value at which the upper-bound $p$-value
crosses 0.05. In the table above, locate where this happens — if the
critical $\Gamma$ is around 2 or higher, the result is robust to moderate
hidden bias. If it is close to 1, even small unmeasured confounding could
overturn the conclusion.

```{r}
#| label: critical-gamma
#| cache: true

critical_gamma <- result$Gamma[min(which(result$p_value_upper > 0.05))]
cat(sprintf("Critical Gamma (p > 0.05): %.2f\n",
            critical_gamma))
```

## What about partial identification?

When sensitivity analysis shows the point estimate is fragile, partial
identification is often the next step. Instead of one estimate, report bounds
under weaker assumptions.

The Manski (1990) "no-assumption" bounds say:

$$
\text{ATE} \in \Big[\; \mathbb{E}[Y|D{=}1]P(D{=}1) - \big(\mathbb{E}[Y|D{=}0]P(D{=}0) + K\,P(D{=}1)\big),\;\;
\big(\mathbb{E}[Y|D{=}1]P(D{=}1) + K\,P(D{=}0)\big) - \mathbb{E}[Y|D{=}0]P(D{=}0) \;\Big].
$$

Each potential-outcome mean is split by the law of total expectation, filling
in the unobserved arm with its worst/best case. With outcomes bounded in
$Y \in [0, K]$, this is a single interval of width exactly $K$ — bounded, but
often too wide to be useful. Lee (2009) bounds exploit selection on observables
plus monotonicity to get tighter bounds. Mourifié-Wan provide LATE bounds
when monotonicity may fail.

These bounds usually complement a point estimate. They are useful when the
point estimate depends too much on an untestable assumption.

## Summary

- Sensitivity analysis asks how strong hidden confounding must be to change
  the result.
- Cinelli-Hazlett bounds are useful for regression estimates.
- E-values are useful for risk ratios.
- Rosenbaum bounds are useful for matched pairs.
- Observational causal estimates should usually include at least one
  sensitivity analysis.
