# Sensitivity Analysis

```{r}
#| include: false
library(tidyverse)
library(sensemakr)
library(EValue)
library(causaldata)
library(broom)
```

```{r}
#| eval: false
library(tidyverse)
library(sensemakr)
library(EValue)
library(causaldata)
library(broom)
```

The previous chapters often treat identification as yes or no. Applied work
is usually less clean. We may believe the adjustment set is good, but still
worry about some remaining confounding. Sensitivity analysis asks a practical
question: how strong would unmeasured confounding have to be to change the
conclusion?

> **Related reading**: An extended treatment of the Cinelli-Hazlett
> robustness value and benchmark interpretations appears in the
> [companion blog chapter on sensitivity analysis](https://xiangao.github.io/blog_book/sensitivity-analysis.html)
> of *Topics on Econometrics and Causal Inference*.

Here I use three methods:

1. **Cinelli-Hazlett (2020) omitted-variable bias bounds** — for OLS
   regression estimates, quantifies how strong an omitted variable
   would need to be (in terms of partial $R^2$ with treatment and outcome)
   to bring the estimate to zero.
2. **E-values (VanderWeele-Ding 2017)** — for risk ratios, the minimum
   strength of association (on the risk-ratio scale) that an unmeasured
   confounder would need with both treatment and outcome to fully explain
   the observed association.
3. **Rosenbaum bounds** — for matched estimators, the magnitude of
   non-random treatment assignment within matched pairs that would render
   the test result insignificant.

## The NHEFS smoking-cessation example

We use the National Health and Nutrition Examination Survey I Epidemiologic
Follow-up Study (NHEFS) data — the same dataset used in the
[applied DAG-workflow chapter](smoking-cessation-graphs.qmd). The question:
what is the effect of quitting smoking (`qsmk`) on weight gain over
10 years (`wt82_71`)? We start from the raw comparison.

```{r}
#| label: data
#| cache: true

nhefs <- causaldata::nhefs_complete |>
  filter(!is.na(wt82_71))

cat(sprintf("n = %d\n", nrow(nhefs)))
cat(sprintf("Treated (quit smoking) = %d\n", sum(nhefs$qsmk)))
cat(sprintf("Mean weight gain (treated):   %.2f kg\n",
            mean(nhefs$wt82_71[nhefs$qsmk == 1])))
cat(sprintf("Mean weight gain (untreated): %.2f kg\n",
            mean(nhefs$wt82_71[nhefs$qsmk == 0])))
```

Of the 1,566 complete cases, 403 quit. Quitters gained 4.53 kg on average and
continuing smokers 1.98 kg, a naive difference of 2.55 kg. But quitters and
continuers differ on many baseline characteristics. We adjust for the standard
set by OLS --- sex, age, race, education, smoking intensity, years smoked,
exercise, activity and baseline weight --- and print the first three
coefficients along with the treatment estimate:

```{r}
#| label: ols
#| cache: true

ols <- lm(wt82_71 ~ qsmk + sex + age + race + education + smokeintensity +
                    smokeyrs + exercise + active + wt71,
          data = nhefs)

tidy(ols)[1:3, ] |> knitr::kable(digits = 3)
cat(sprintf("\nAdjusted ACE estimate: %.3f kg  (SE = %.3f)\n",
            coef(ols)["qsmk"], summary(ols)$coefficients["qsmk", 2]))
```

The adjusted estimate is 3.381 kg with a standard error of 0.441, larger than
the naive 2.55 kg. Adjustment moved the answer away from zero, not toward it.
But the identification assumption requires that we have controlled for *all*
relevant confounders. What if we missed one — say, a propensity to gain
weight that also correlates with quitting decisions? Sensitivity analysis
addresses this directly.

## Cinelli-Hazlett OVB bounds

The framework is Cinelli and Hazlett [-@cinelli-hazlett-2020].

`sensemakr` implements the Cinelli-Hazlett (2020) approach. The key inputs
are:

- a fitted regression
- the treatment variable name
- one or more *benchmark covariates* whose strength serves as a reference
  for how strong a hypothetical unobserved confounder might be

```{r}
#| label: sensemakr
#| cache: true

sens <- sensemakr(
  model = ols,
  treatment = "qsmk",
  benchmark_covariates = c("smokeintensity", "smokeyrs"),
  kd = 1:3,                 # benchmark: 1×, 2×, 3× as strong as the benchmark
  ky = 1:3,
  q  = 1,                   # the "bias proportion" at which to flag a problem
  alpha = 0.05,
  reduce = TRUE
)

summary(sens)
```

The summary tells us:

- **Robustness value (RV)**: the minimum strength of association (in terms of
  partial $R^2$) that an unmeasured confounder would need with both the
  treatment and the outcome to reduce the estimate to zero. Higher = more
  robust.
- **RV_q**: the strength needed to "explain away" a fraction $q$ of the
  estimate (here $q = 1$ means complete explanation).
- **Benchmark comparisons**: how the RV compares to multiples of the
  observed covariates.

For this regression the partial $R^2$ of quitting with weight change is 3.65%.
The robustness value at $q=1$ is 17.67%: a confounder orthogonal to the
covariates would have to explain more than 17.67% of the residual variance of
both quitting and weight change to bring the estimate to zero. To leave it
merely statistically indistinguishable from zero at the 5% level, 13.47% would
be enough.

The benchmark rows put those numbers on a scale. Smoking intensity explains
1.43% of the residual variance of quitting and 0.10% of the residual variance
of weight change. A confounder three times as strong on both margins would move
the estimate from 3.381 to 3.182, with a 95% interval of $[2.30, 4.07]$. None
of the measured covariates comes near the 17.67% that would be needed, so the
estimate is robust to a confounder resembling the ones we already control for.

The same information is easier to read as a contour plot. The horizontal axis
is the confounder's partial $R^2$ with the treatment, the vertical axis its
partial $R^2$ with the outcome, and the contours give the adjusted estimate at
each combination. The benchmark diamonds mark 1x, 2x and 3x smoking intensity
and years smoked. All six sit very close to the origin, so we cut the axes at
0.06 and 0.006 to keep their labels apart; the robustness value of 0.177 that
would drive the estimate to zero is far outside this window.

```{r}
#| label: sens-plot
#| cache: true
#| fig-cap: "Cinelli-Hazlett sensitivity contour, with the axes cut at 0.06 and 0.006 so the benchmark labels stay legible. Contours give the adjusted estimate at each pair of partial R-squared values. Diamonds mark confounders as strong as 1x, 2x and 3x smokeintensity and smokeyrs, with the implied estimate in parentheses. The contour reaching zero lies far outside this window."
#| fig-width: 7
#| fig-height: 5

plot(sens, lim = 0.06, lim.y = 0.006, cex.label.text = 0.6)
```

All six benchmark points sit below 0.05 on the treatment axis and below 0.005
on the outcome axis, in the region where the adjusted estimate stays above
3 kg. The contour that reaches zero is a long way out.

## E-values for risk ratios

The E-value is VanderWeele and Ding [-@vanderweele-ding-2017].

For binary outcomes and risk-ratio-type estimands, VanderWeele and Ding
(2017) propose the **E-value**: the minimum strength of association on the
risk-ratio scale that an unmeasured confounder would need to have with both
the treatment and the outcome to fully explain the observed risk ratio.

To illustrate, we dichotomize weight change into any gain versus none, fit the
risk ratio by Poisson regression with a log link on the same nine covariates,
and take the confidence interval from robust standard errors:

```{r}
#| label: evalue
#| cache: true

nhefs$gain <- as.numeric(nhefs$wt82_71 > 0)

rr_fit <- glm(gain ~ qsmk + sex + age + race + education + smokeintensity +
                     smokeyrs + exercise + active + wt71,
              data = nhefs, family = poisson(link = "log"))
rr_summary <- tidy(rr_fit, exp = TRUE, conf.int = TRUE)

rr_est <- exp(coef(rr_fit)["qsmk"])
# The Poisson likelihood is misspecified for a binary outcome (the Zou 2004
# log-binomial-via-Poisson trick), so the model-based variance is wrong.
# Use robust (sandwich/HC) standard errors for the RR confidence interval.
rr_ci  <- lmtest::coefci(rr_fit, parm = "qsmk",
                         vcov. = sandwich::vcovHC(rr_fit, type = "HC0"))
rr_lo  <- exp(rr_ci[1])
rr_hi  <- exp(rr_ci[2])

cat(sprintf("Adjusted RR (quit vs not): %.3f  [%.3f, %.3f]\n",
            rr_est, rr_lo, rr_hi))

ev <- evalues.RR(est = rr_est, lo = rr_lo, hi = rr_hi)
print(ev)
```

The E-value reports two numbers:

- **E-value (point estimate)**: an unmeasured confounder would need to be
  associated with both treatment and outcome by risk ratios of at least
  this value, above and beyond the measured confounders, to fully explain
  away the observed effect.
- **E-value (confidence interval)**: the minimum strength needed to shift
  the lower confidence bound to 1.

The adjusted risk ratio is 1.218, with a 95% interval of $[1.135, 1.307]$.
Quitters are about 22% more likely to gain any weight. The E-value for the
point estimate is 1.73: an unmeasured confounder would need risk-ratio
associations of at least 1.73 with both quitting and weight gain, beyond the
nine measured covariates, to account for the whole association. The E-value for
the confidence limit is 1.53, the strength that would push the lower bound to 1.

An E-value much greater than 1 suggests the result is hard to explain by
unmeasured confounding. An E-value close to 1 means a modest unmeasured
confounder could overturn the conclusion. At 1.73 this result sits in between:
a confounder as strong as that is not implausible, but it would have to be
unrelated to everything already in the model.

The useful part of the E-value is that it is on the risk-ratio scale. It can
be explained without specifying a model for the unobserved confounder.

## Rosenbaum bounds (matched estimators)

The sensitivity model and the bound computed below are Rosenbaum's [-@rosenbaum-2002].

For matching estimators, **Rosenbaum bounds** (Rosenbaum 2002) provide a
sensitivity analysis based on the maximum deviation from random assignment
within matched pairs.

Suppose units in a matched pair are observationally identical, but one
might still be more likely than the other to receive treatment due to
unobserved factors. Let $\Gamma$ bound the ratio of treatment odds between
the two units of a pair: $\Gamma = 1$ means matched pairs are perfectly
randomised (both units equally likely to be the treated one); $\Gamma = 2$
means hidden bias could make one unit's odds of treatment up to twice the
other's.

Rosenbaum bounds compute the range of $p$-values for the matched test
under each value of $\Gamma$. The **critical $\Gamma$** is the smallest
value at which the test becomes insignificant.

Here is a worked example on the same NHEFS data. We fit the propensity score by
logistic regression on the nine baseline covariates, match each of the 403
quitters to its nearest unmatched control on that score, take the paired
differences in weight change, and compute the upper-bound Wilcoxon
signed-rank $p$-value on a grid $\Gamma \in [1, 4]$ in steps of 0.25:

```{r}
#| label: rosenbaum
#| cache: true

# Estimate propensity score
ps_fit <- glm(qsmk ~ sex + age + race + education + smokeintensity +
                     smokeyrs + exercise + active + wt71,
              data = nhefs, family = binomial)
nhefs$ps <- predict(ps_fit, type = "response")

# Simple 1:1 nearest-neighbour matching on propensity score
treated_idx <- which(nhefs$qsmk == 1)
control_idx <- which(nhefs$qsmk == 0)

set.seed(1)
matched_pairs <- data.frame(t_idx = integer(), c_idx = integer())
available_controls <- control_idx
for (ti in treated_idx) {
  if (length(available_controls) == 0) break
  dists <- abs(nhefs$ps[ti] - nhefs$ps[available_controls])
  best  <- which.min(dists)
  matched_pairs <- rbind(matched_pairs,
                         data.frame(t_idx = ti,
                                    c_idx = available_controls[best]))
  available_controls <- available_controls[-best]
}

# Compute paired differences in outcome
paired_diff <- nhefs$wt82_71[matched_pairs$t_idx] -
               nhefs$wt82_71[matched_pairs$c_idx]
cat(sprintf("N matched pairs: %d\n", nrow(matched_pairs)))
cat(sprintf("Mean paired difference (matched ATT): %.3f kg\n",
            mean(paired_diff)))

# Rosenbaum bound calculation: range of p-values for Wilcoxon signed-rank
# under each Gamma. We compute a large-sample NORMAL APPROXIMATION to the
# weighted Bernoulli rank sum directly rather than using a package. With 403
# pairs it tracks the canonical bound closely, but it is an approximation:
# there is no exact Poisson-binomial tail and no continuity correction here.
# to keep the dependency footprint small.
rosenbaum_pvalue <- function(diffs, gamma) {
  # The Wilcoxon signed-rank protocol discards exact-zero paired differences
  # before ranking (a zero carries no information about the sign) and
  # reduces the effective sample size accordingly; including them would
  # skew both the rank sum and its null variance.
  diffs <- diffs[diffs != 0]
  abs_d <- abs(diffs)
  signs <- sign(diffs)
  ranks <- rank(abs_d)
  # Under Gamma, P(positive sign) is bounded between 1/(1+Γ) and Γ/(1+Γ).
  # Upper-bound p-value: assume P(positive) = Γ/(1+Γ) (against null of zero)
  p_pos_upper <- gamma / (1 + gamma)
  # Test statistic: sum of positive ranks
  T_obs <- sum(ranks[signs > 0])
  n <- length(diffs)
  mu_upper  <- p_pos_upper * sum(ranks)
  var_upper <- p_pos_upper * (1 - p_pos_upper) * sum(ranks^2)
  z <- (T_obs - mu_upper) / sqrt(var_upper)
  1 - pnorm(z)
}

gammas <- seq(1, 4, by = 0.25)
pvals  <- sapply(gammas, function(g) rosenbaum_pvalue(paired_diff, g))

result <- tibble(Gamma = gammas, p_value_upper = pvals)
knitr::kable(result, digits = 4,
             caption = "Upper-bound p-values from Rosenbaum sensitivity analysis")
```

Matching gives 403 pairs, one per quitter, with a mean paired difference of
2.926 kg. The upper-bound $p$-value is 0.0334 at $\Gamma = 1.50$ and 0.2905 at
$\Gamma = 1.75$, so the crossing of 0.05 falls between those two grid points.

The **critical $\Gamma$** is the value at which the upper-bound $p$-value
crosses 0.05. If it is around 2 or higher, the result is robust to moderate
hidden bias. If it is close to 1, even small unmeasured confounding could
overturn the conclusion. We read it off the grid:

```{r}
#| label: critical-gamma
#| cache: true

sig_idx <- which(result$p_value_upper > 0.05)
critical_gamma <- if (length(sig_idx) > 0) result$Gamma[min(sig_idx)] else NA
if (is.na(critical_gamma)) {
  cat(sprintf("No Gamma up to %.2f pushed the p-value above 0.05 -- the result is robust across the whole grid tested; extend `gammas` to find the true crossing.\n", max(result$Gamma)))
} else {
  cat(sprintf("Critical Gamma (p > 0.05): %.2f\n", critical_gamma))
}
```

The critical $\Gamma$ is 1.75. This is the first value *on the 0.25 grid* at
which significance is lost; the true crossing lies between 1.50 and 1.75, and a
finer grid would locate it more precisely. The convention in applied work is
often to report the largest $\Gamma$ at which significance is retained, which
here is 1.50.

So hidden bias that made one member of a matched pair up to about 1.5 times
more likely to quit than the other would leave the result significant, and bias
beyond that would not.

The three analyses do not share a scale. $\Gamma$ bounds an odds ratio within a
matched pair, the E-value bounds a risk ratio, and the robustness value is a
partial $R^2$. They also ask different questions: $\Gamma$ and the E-value ask
when the result stops being significant, the robustness value asks when the
point estimate reaches zero. Reporting one of them is normal practice; the
numbers should not be averaged or compared directly.

## What about partial identification?

The bounding tradition here starts with Manski [-@manski-1990-bounds]; Lee [-@lee-2009-bounds] gives the treatment-selection bounds used in applied work.

When sensitivity analysis shows the point estimate is fragile, partial
identification is often the next step. Instead of one estimate, report bounds
under weaker assumptions.

The Manski (1990) "no-assumption" bounds say:

$$
\begin{aligned}
\text{ATE} &\ge \mathbb{E}[Y|D{=}1]P(D{=}1) - \mathbb{E}[Y|D{=}0]P(D{=}0) - K\,P(D{=}1), \\
\text{ATE} &\le \mathbb{E}[Y|D{=}1]P(D{=}1) - \mathbb{E}[Y|D{=}0]P(D{=}0) + K\,P(D{=}0).
\end{aligned}
$$ {#eq-sensitivity-analysis-1}

Each potential-outcome mean is split by the law of total expectation, filling
in the unobserved arm with its worst/best case. With outcomes bounded in
$Y \in [0, K]$, this is a single interval of width exactly $K$ — bounded, but
often too wide to be useful. Tighter bounds come from adding structure to a
specific problem: Lee (2009) bounds address sample selection/attrition (the
outcome is observed only for selected units) by combining random assignment
with monotone selection and trimming the over-selected arm. For IV settings,
Mourifié and Wan (2017) provide a test of the LATE identifying assumptions
(instrument validity plus monotonicity).

These bounds usually complement a point estimate. They are useful when the
point estimate depends too much on an untestable assumption.

## Summary

- Sensitivity analysis asks how strong hidden confounding must be to change
  the result.
- Cinelli-Hazlett bounds are useful for regression estimates.
- E-values are useful for risk ratios.
- Rosenbaum bounds are useful for matched pairs.
- Observational causal estimates should usually include at least one
  sensitivity analysis.
