# Sensitivity Analysis

```{r}
#| include: false
library(tidyverse)
library(sensemakr)
library(EValue)
library(causaldata)
library(broom)
```

```{r}
#| eval: false
library(tidyverse)
library(sensemakr)
library(EValue)
library(causaldata)
library(broom)
```

The previous chapters proceeded as if identification is binary: either a graph
supports the effect (backdoor or front-door routes), or it doesn't (the effect
is not identified). In practice, applied researchers face a more nuanced
situation. They believe their identification strategy is *approximately*
correct — there might be some residual confounding, but probably not much.
The honest question is: **how much unmeasured confounding would be needed to
overturn the conclusion?**

> **Related reading**: An extended treatment of the Cinelli-Hazlett
> robustness value and benchmark interpretations appears in the
> [companion blog chapter on sensitivity analysis](https://xiangao.github.io/blog_book/sensitivity-analysis.html)
> of *Topics on Econometrics and Causal Inference*.

This is the question sensitivity analysis answers. Rather than claiming
unconfoundedness as a binary assumption, it asks how strong the unmeasured
confounding would have to be — relative to observed covariates — to change
the sign or significance of the estimate.

We cover three complementary methods:

1. **Cinelli-Hazlett (2020) omitted-variable bias bounds** — for any
   regression-based estimator, quantifies how strong an omitted variable
   would need to be (in terms of partial $R^2$ with treatment and outcome)
   to bring the estimate to zero.
2. **E-values (VanderWeele-Ding 2017)** — for risk ratios, the minimum
   strength of association (on the risk-ratio scale) that an unmeasured
   confounder would need with both treatment and outcome to fully explain
   the observed association.
3. **Rosenbaum bounds** — for matched estimators, the magnitude of
   non-random treatment assignment within matched pairs that would render
   the test result insignificant.

## The NHEFS smoking-cessation example

We use the National Health and Nutrition Examination Survey I Epidemiologic
Follow-up Study (NHEFS) data — the same dataset used in the
[applied DAG-workflow chapter](smoking-cessation-graphs.qmd). The question:
what is the effect of quitting smoking (`qsmk`) on weight gain over
10 years (`wt82_71`)?

```{r}
#| label: data
#| cache: true

nhefs <- causaldata::nhefs_complete |>
  filter(!is.na(wt82_71))

cat(sprintf("n = %d\n", nrow(nhefs)))
cat(sprintf("Treated (quit smoking) = %d\n", sum(nhefs$qsmk)))
cat(sprintf("Mean weight gain (treated):   %.2f kg\n",
            mean(nhefs$wt82_71[nhefs$qsmk == 1])))
cat(sprintf("Mean weight gain (untreated): %.2f kg\n",
            mean(nhefs$wt82_71[nhefs$qsmk == 0])))
```

A naive comparison says quitters gained $\approx$ 2.5 kg more than non-quitters,
but smokers and quitters differ on many baseline characteristics. Adjust for
the standard set of confounders (age, sex, race, education, baseline weight,
smoking intensity, exercise, activity):

```{r}
#| label: ols
#| cache: true

ols <- lm(wt82_71 ~ qsmk + sex + age + race + education + smokeintensity +
                    smokeyrs + exercise + active + wt71,
          data = nhefs)

tidy(ols)[1:3, ] |> knitr::kable(digits = 3)
cat(sprintf("\nAdjusted ACE estimate: %.3f kg  (SE = %.3f)\n",
            coef(ols)["qsmk"], summary(ols)$coefficients["qsmk", 2]))
```

The adjusted estimate is around 3.5 kg, larger than the unadjusted difference.
But the identification assumption requires that we have controlled for *all*
relevant confounders. What if we missed one — say, a propensity to gain
weight that also correlates with quitting decisions? Sensitivity analysis
addresses this directly.

## Cinelli-Hazlett OVB bounds

`sensemakr` implements the Cinelli-Hazlett (2020) approach. The key inputs
are:

- a fitted regression
- the treatment variable name
- one or more *benchmark covariates* whose strength serves as a reference
  for how strong a hypothetical unobserved confounder might be

```{r}
#| label: sensemakr
#| cache: true

sens <- sensemakr(
  model = ols,
  treatment = "qsmk",
  benchmark_covariates = c("smokeintensity", "smokeyrs"),
  kd = 1:3,                 # benchmark: 1×, 2×, 3× as strong as the benchmark
  ky = 1:3,
  q  = 1,                   # the "bias proportion" at which to flag a problem
  alpha = 0.05,
  reduce = TRUE
)

summary(sens)
```

The summary tells us:

- **Robustness value (RV)**: the minimum strength of association (in terms of
  partial $R^2$) that an unmeasured confounder would need with both the
  treatment and the outcome to reduce the estimate to zero. Higher = more
  robust.
- **RV_q**: the strength needed to "explain away" a fraction $q$ of the
  estimate (here $q = 1$ means complete explanation).
- **Benchmark comparisons**: how the RV compares to multiples of the
  observed covariates.

```{r}
#| label: sens-plot
#| cache: true
#| fig-cap: "Cinelli-Hazlett sensitivity contour. The dashed lines show benchmarks based on smokeintensity and smokeyrs at 1x, 2x, and 3x strength. Confounders falling in the shaded region would reduce the estimate to zero."
#| fig-width: 7
#| fig-height: 5

plot(sens)
```

The contour plot shows pairs of (partial $R^2$ with treatment, partial $R^2$
with outcome) that an unobserved confounder would need to have to render the
estimate zero. Points farther from the origin represent stronger confounders.
The benchmark dots show what strength corresponds to 1×, 2×, 3× the
observed strength of `smokeintensity` and `smokeyrs`.

For the NHEFS case, the robustness value is large — even a confounder several
times stronger than smoking intensity would not eliminate the estimated
effect. The conclusion is robust.

## E-values for risk ratios

For binary outcomes and risk-ratio-type estimands, VanderWeele and Ding
(2017) propose the **E-value**: the minimum strength of association on the
risk-ratio scale that an unmeasured confounder would need to have with both
the treatment and the outcome to fully explain the observed risk ratio.

To illustrate, dichotomize weight gain ("any" gain vs no gain) and fit a
risk-ratio model:

```{r}
#| label: evalue
#| cache: true

nhefs$gain <- as.numeric(nhefs$wt82_71 > 0)

rr_fit <- glm(gain ~ qsmk + sex + age + race + education + smokeintensity +
                     smokeyrs + exercise + active + wt71,
              data = nhefs, family = poisson(link = "log"))
rr_summary <- tidy(rr_fit, exp = TRUE, conf.int = TRUE)

rr_est <- exp(coef(rr_fit)["qsmk"])
rr_lo  <- exp(confint.default(rr_fit)["qsmk", 1])
rr_hi  <- exp(confint.default(rr_fit)["qsmk", 2])

cat(sprintf("Adjusted RR (quit vs not): %.3f  [%.3f, %.3f]\n",
            rr_est, rr_lo, rr_hi))

ev <- evalues.RR(est = rr_est, lo = rr_lo, hi = rr_hi)
print(ev)
```

The E-value reports two numbers:

- **E-value (point estimate)**: an unmeasured confounder would need to be
  associated with both treatment and outcome by risk ratios of at least
  this value, above and beyond the measured confounders, to fully explain
  away the observed effect.
- **E-value (confidence interval)**: the minimum strength needed to shift
  the lower confidence bound to 1.

An E-value much greater than 1 suggests the result is hard to explain by
unmeasured confounding. An E-value close to 1 means a modest unmeasured
confounder could overturn the conclusion.

A common rule of thumb: an E-value of 2 or more is considered moderate
robustness; values of 4 or more indicate strong robustness. The biggest
strength of the E-value is its model-free interpretation — it applies to
any binary-outcome risk-ratio estimate, regardless of the specific
estimator used.

## Rosenbaum bounds (matched estimators)

For matching estimators, **Rosenbaum bounds** (Rosenbaum 2002) provide a
sensitivity analysis based on the maximum deviation from random assignment
within matched pairs.

Suppose units in a matched pair are observationally identical, but one
might still be more likely than the other to receive treatment due to
unobserved factors. Let $\Gamma$ be the odds ratio of treatment within a
matched pair due to hidden bias: $\Gamma = 1$ means matched pairs are
perfectly randomised; $\Gamma = 2$ means within a pair, the treated unit
was twice as likely to get treatment due to unobservables.

Rosenbaum bounds compute the range of $p$-values for the matched test
under each value of $\Gamma$. The **critical $\Gamma$** is the smallest
value at which the test becomes insignificant.

Here is a worked example using simple pair matching on propensity score and
a Wilcoxon signed-rank test (the original Rosenbaum bound calculation):

```{r}
#| label: rosenbaum
#| cache: true

# Estimate propensity score
ps_fit <- glm(qsmk ~ sex + age + race + education + smokeintensity +
                     smokeyrs + exercise + active + wt71,
              data = nhefs, family = binomial)
nhefs$ps <- predict(ps_fit, type = "response")

# Simple 1:1 nearest-neighbour matching on propensity score
treated_idx <- which(nhefs$qsmk == 1)
control_idx <- which(nhefs$qsmk == 0)

set.seed(1)
matched_pairs <- data.frame(t_idx = integer(), c_idx = integer())
available_controls <- control_idx
for (ti in treated_idx) {
  if (length(available_controls) == 0) break
  dists <- abs(nhefs$ps[ti] - nhefs$ps[available_controls])
  best  <- which.min(dists)
  matched_pairs <- rbind(matched_pairs,
                         data.frame(t_idx = ti,
                                    c_idx = available_controls[best]))
  available_controls <- available_controls[-best]
}

# Compute paired differences in outcome
paired_diff <- nhefs$wt82_71[matched_pairs$t_idx] -
               nhefs$wt82_71[matched_pairs$c_idx]
cat(sprintf("N matched pairs: %d\n", nrow(matched_pairs)))
cat(sprintf("Mean paired difference (matched ATT): %.3f kg\n",
            mean(paired_diff)))

# Rosenbaum bound calculation: range of p-values for Wilcoxon signed-rank
# under each Gamma. We compute this directly rather than using a package
# to keep the dependency footprint small.
rosenbaum_pvalue <- function(diffs, gamma) {
  abs_d <- abs(diffs)
  signs <- sign(diffs)
  ranks <- rank(abs_d)
  # Under Gamma, P(positive sign) is bounded between Γ/(1+Γ) and 1/(1+Γ).
  # Upper-bound p-value: assume P(positive) = Γ/(1+Γ) (against null of zero)
  p_pos_upper <- gamma / (1 + gamma)
  # Test statistic: sum of positive ranks
  T_obs <- sum(ranks[signs > 0])
  n <- length(diffs)
  mu_upper  <- p_pos_upper * sum(ranks)
  var_upper <- p_pos_upper * (1 - p_pos_upper) * sum(ranks^2)
  z <- (T_obs - mu_upper) / sqrt(var_upper)
  1 - pnorm(z)
}

gammas <- seq(1, 4, by = 0.25)
pvals  <- sapply(gammas, function(g) rosenbaum_pvalue(paired_diff, g))

result <- tibble(Gamma = gammas, p_value_upper = pvals)
knitr::kable(result, digits = 4,
             caption = "Upper-bound p-values from Rosenbaum sensitivity analysis")
```

The **critical $\Gamma$** is the value at which the upper-bound $p$-value
crosses 0.05. In the table above, locate where this happens — if the
critical $\Gamma$ is around 2 or higher, the result is robust to moderate
hidden bias. If it is close to 1, even small unmeasured confounding could
overturn the conclusion.

```{r}
#| label: critical-gamma
#| cache: true

critical_gamma <- result$Gamma[min(which(result$p_value_upper > 0.05))]
cat(sprintf("Critical Gamma (p > 0.05): %.2f\n",
            critical_gamma))
```

## What about partial identification?

When sensitivity analysis shows the effect is fragile, the next step is
often **partial identification** — instead of point-estimating, derive
bounds on the effect under weaker assumptions.

The Manski (1990) "no-assumption" bounds say:

$$
\text{ATE} \in [\mathbb{E}[Y | D=1] - \max Y(1)] \cup [\mathbb{E}[Y | D=0] - \min Y(0)],
$$

i.e., when outcomes are bounded $Y \in [0, K]$, the ATE is bounded but the
width is often large. Lee (2009) bounds exploit selection on observables
plus monotonicity to get tighter bounds. Mourifié-Wan provide LATE bounds
when monotonicity may fail.

These are typically applied as a *complement* to a point estimate, not a
replacement. They are most useful when sensitivity analysis suggests the
point estimate is fragile and you want to characterise the full range of
effects consistent with the data and minimal assumptions.

## Summary

- **Sensitivity analysis** quantifies how strong unobserved confounding
  would need to be to overturn a conclusion.
- **Cinelli-Hazlett OVB bounds** (`sensemakr`): regression-based, gives
  robustness values and benchmark comparisons. The most flexible and
  widely applicable.
- **E-values** (`EValue`): risk-ratio-based, model-free interpretation,
  particularly useful for epidemiology and medicine.
- **Rosenbaum bounds**: for matched estimators; the critical $\Gamma$ is
  the threshold odds ratio of hidden bias that would render the result
  insignificant.
- **Reporting practice**: every observational causal estimate should be
  accompanied by *some* sensitivity analysis. A point estimate without
  one is a claim of certainty that the data cannot support.

