# Applied Graph Workflow: Smoking Cessation

```{r}
#| include: false
suppressPackageStartupMessages({
  library(dagitty)
  library(ggdag)
  library(ggplot2)
  library(causaldata)
  library(dplyr)
  library(knitr)
})
```

The previous chapter used small graph examples. Here I apply the same graph-identification-estimation workflow to a real data set: the complete-case National Health and Nutrition Examination Survey I Epidemiologic Follow-up Study (NHEFS) data distributed through the `causaldata` package.

The question is:

```text
What is the effect of quitting smoking on weight change from 1971 to 1982?
```

The treatment is `qsmk`, an indicator for quitting smoking between baseline and follow-up. The outcome is `wt82_71`, weight change over the follow-up period. This is observational data. The effect is not identified by the data alone. It is identified only after we state the adjustment assumptions in the graph.

## Data

For this example, keep the treatment, outcome, and baseline covariates that are commonly used in NHEFS smoking-cessation examples.

```{r}
#| warning: false
#| message: false
#| cache: true
nhefs_cols <- c(
  "qsmk", "wt82_71",
  "sex", "race", "age", "school",
  "smokeintensity", "smokeyrs",
  "exercise", "active", "wt71"
)
nhefs <- causaldata::nhefs_complete |>
  dplyr::select(all_of(nhefs_cols)) |>
  dplyr::mutate(across(everything(), as.numeric))

c(n = nrow(nhefs), variables = ncol(nhefs))
```

```{r}
head(nhefs, 5) |>
  dplyr::mutate(across(everything(), \(x) signif(x, 4))) |>
  kable()
```

## Adjustment DAG

Start with the grouped graph

```text
X -> A -> Y
X -----> Y
```

where `X` is the baseline covariate set, `A` is quitting smoking, and `Y` is later weight change. This graph says the baseline variables may affect both quitting and later weight change, and that there is no unmeasured common cause of quitting and later weight change after conditioning on those baseline variables.

```{r}
conceptual_graph <- dagitty("dag {
  X -> A
  X -> Y
  A -> Y
}")
coordinates(conceptual_graph) <- list(
  x = c(X = 1, A = 0, Y = 2),
  y = c(X = 1, A = 0, Y = 0)
)
```

```{r}
#| label: fig-conceptual
#| fig-cap: "Conceptual adjustment DAG: baseline X confounds A → Y."
#| fig-width: 5
#| fig-height: 3
ggdag(conceptual_graph) + theme_dag_blank()
```

For estimation, expand `X` into the observed NHEFS columns. The expanded graph is still an adjustment DAG: each baseline variable is allowed to point into quitting and into later weight change.

```{r}
baseline_vars <- setdiff(nhefs_cols, c("qsmk", "wt82_71"))

build_grouped_graph <- function(baseline) {
  edges_X_to_A <- paste0(baseline, " -> qsmk")
  edges_X_to_Y <- paste0(baseline, " -> wt82_71")
  body <- paste(c(edges_X_to_A, edges_X_to_Y, "qsmk -> wt82_71"),
                collapse = "\n  ")
  dagitty(paste0("dag {\n  ", body, "\n}"))
}

nhefs_graph <- build_grouped_graph(baseline_vars)
c(vertices = length(names(nhefs_graph)),
  directed_edges = length(edges(nhefs_graph)$v))
```

## Identification

`dagitty::adjustmentSets` checks the graph before estimation. In this adjustment DAG, the minimal sufficient adjustment set is the full baseline:

```{r}
adj_sets <- adjustmentSets(nhefs_graph, exposure = "qsmk",
                           outcome = "wt82_71", type = "minimal")
adj_sets
```

The graph therefore identifies the causal mean by the adjustment formula

$$
E[Y(a)] = E_X\{E(Y \mid A=a, X)\}.
$$ {#eq-smoking-cessation-graphs-1}

This formula is not a modeling claim by itself. It is the observed-data functional implied by the graph. Estimation still requires nuisance models for the outcome regression and treatment mechanism.

## Estimation with AIPW

The book keeps identification and estimation as separate steps. The graph supplies the adjustment set; the estimator uses that set.

```{r}
aipw_backdoor <- function(data, treatment, outcome, adjust) {
  fmla_y <- as.formula(paste0(outcome, " ~ ", treatment, " * (",
                              paste(adjust, collapse = " + "), ")"))
  fmla_a <- as.formula(paste0(treatment, " ~ ",
                              paste(adjust, collapse = " + ")))

  mu_fit <- lm(fmla_y, data = data)
  pi_fit <- glm(fmla_a, data = data, family = binomial())

  d1 <- data; d1[[treatment]] <- 1
  d0 <- data; d0[[treatment]] <- 0
  mu1 <- predict(mu_fit, newdata = d1)
  mu0 <- predict(mu_fit, newdata = d0)
  pi1 <- predict(pi_fit, newdata = data, type = "response")
  # Truncate (winsorize) the propensity score away from 0/1. AIPW divides by
  # pi1 and (1-pi1), so an estimated propensity near the boundary produces
  # huge inverse-probability weights and blows up the estimator's variance.
  # Bounding at [0.01, 0.99] trades a small amount of bias (we no longer use
  # the true estimated propensity for the most extreme units) for much
  # lower variance -- a standard and usually favorable trade-off, but worth
  # flagging explicitly since it is a bias-variance choice, not a free lunch.
  pi1 <- pmin(pmax(pi1, 0.01), 0.99)

  w  <- data[[treatment]]; y <- data[[outcome]]
  if1 <- w     * (y - mu1) / pi1       + mu1
  if0 <- (1-w) * (y - mu0) / (1 - pi1) + mu0
  psi  <- if1 - if0

  est <- mean(psi); se <- sd(psi) / sqrt(length(psi))
  ci  <- est + c(-1, 1) * qnorm(0.975) * se
  data.frame(estimand = "Quit smoking vs continued smoking",
             estimate = signif(est, 4),
             lower_95 = signif(ci[1], 4),
             upper_95 = signif(ci[2], 4),
             se       = signif(se, 4))
}

aipw_backdoor(nhefs, "qsmk", "wt82_71", adjust = baseline_vars) |> kable()
```

Under the DAG, the estimate is the average effect of quitting smoking versus continuing smoking on weight change in the complete-case NHEFS population. The positive estimate is consistent with weight gain after quitting. The causal interpretation depends on the graph, especially the assumption that the measured baseline variables block the relevant backdoor paths.

The `sd(psi) / sqrt(n)` standard error is the AIPW influence-function
SE and is valid here because the nuisances are parametric logistic/linear
fits. If you replace them with flexible machine learners, use
cross-fitting and an influence-function (or bootstrap) variance instead;
the plug-in SE on full-sample ML fits is not generally valid.

## Sensitivity Graph

Now encode a different assumption: quitting and later weight change share an unobserved common cause even after conditioning on the measured baseline variables. In an ADMG this is a bidirected edge, `qsmk <-> wt82_71`.

```{r}
build_sensitivity_graph <- function(baseline) {
  edges_X_to_A <- paste0(baseline, " -> qsmk")
  edges_X_to_Y <- paste0(baseline, " -> wt82_71")
  body <- paste(c(edges_X_to_A, edges_X_to_Y,
                  "qsmk -> wt82_71", "qsmk <-> wt82_71"),
                collapse = "\n  ")
  dagitty(paste0("dag {\n  ", body, "\n}"))
}

sensitivity_graph <- build_sensitivity_graph(baseline_vars)
sens_adj <- adjustmentSets(sensitivity_graph, exposure = "qsmk",
                           outcome = "wt82_71", type = "minimal")
cat("Number of sufficient adjustment sets:", length(sens_adj), "\n")
```

The identifying conclusion changes. Under this graph, the observed NHEFS variables are not enough to identify the effect: `adjustmentSets` returns an empty list. This is why the graph has to come before the estimator. The same data can imply different causal answers under different assumptions.

## Structured Baseline Graph

The simple adjustment graph is often the clearest starting point. A more detailed DAG can record baseline ordering: demographics may precede education, smoking history, activity, and baseline weight; those variables may then affect quitting and later weight change.

```{r}
structured_graph <- dagitty("dag {
  age -> school
  age -> smokeyrs
  age -> wt71
  age -> active
  age -> exercise
  age -> qsmk
  age -> wt82_71

  sex -> smokeintensity
  sex -> smokeyrs
  sex -> wt71
  sex -> qsmk
  sex -> wt82_71

  race -> school
  race -> smokeintensity
  race -> qsmk
  race -> wt82_71

  school -> smokeintensity
  school -> qsmk
  school -> wt82_71

  smokeyrs -> smokeintensity
  smokeyrs -> qsmk
  smokeyrs -> wt82_71

  smokeintensity -> qsmk
  smokeintensity -> wt82_71

  exercise -> wt71
  exercise -> qsmk
  exercise -> wt82_71

  active -> wt71
  active -> qsmk
  active -> wt82_71

  wt71 -> qsmk
  wt71 -> wt82_71

  qsmk -> wt82_71
}")

struct_adj <- adjustmentSets(structured_graph, exposure = "qsmk",
                             outcome = "wt82_71", type = "minimal")
struct_adj
data.frame(
  matches_full_baseline = setequal(unlist(struct_adj[[1]]), baseline_vars)
) |> kable()
```

For this treatment effect, the detailed baseline DAG is adjustment-equivalent to the simpler expanded DAG. It implies the same sufficient adjustment set and the same identifying functional. The extra arrows among baseline variables may be useful for documentation, but they are also extra causal assumptions. If those assumptions are not needed for the estimand, the grouped adjustment graph is usually easier to defend.
