# Survival Analysis with Causal Inference

```{r}
#| include: false
library(tidyverse)
library(survival)
library(survminer)
library(riskRegression)
library(broom)
```

```{r}
#| eval: false
library(tidyverse)
library(survival)
library(survminer)
library(riskRegression)
```

When the outcome is time-to-event, such as death, readmission, job exit, or
machine failure, ordinary regression is not enough. Some observations are
censored. We know the event had not happened by the end of follow-up, but we
do not know the event time. Dropping those observations or treating the
observed follow-up time as the event time creates bias.

I cover:

1. Cox proportional hazards models with covariate adjustment.
2. Inverse probability of censoring weighting (IPCW).
3. Restricted mean survival time (RMST) differences.
4. Doubly robust survival estimators in `riskRegression`.

## The challenge of censored outcomes

Censoring means we observe $\min(T_i, C_i)$ and $\delta_i = \mathbb{1}\{T_i \le C_i\}$,
where $T_i$ is the event time and $C_i$ is the censoring time. The
main quantity of interest is the survival function

$$
S(t) = P(T > t)
$$

the probability of surviving past time $t$. Treatment effects are usually
reported as:

- **Hazard ratio (HR)**: $HR = h_1(t) / h_0(t)$ (the ratio of hazards
  under treatment vs control). Easy to estimate but hard to interpret
  causally (Hernán 2010).
- **Restricted mean survival difference (RMST)**: $\int_0^\tau [S_1(t) - S_0(t)] dt$
  — the difference in mean survival up to time $\tau$. Causally
  interpretable as a "years gained" quantity.
- **Survival probability difference**: $S_1(t) - S_0(t)$ at a specified
  time. Reports a direct probability difference.

## Setup: simulated survival data

```{r}
#| label: dgp
#| cache: true

df <- read_csv("data/survival_sim.csv", show_col_types = FALSE)
n <- nrow(df)
cat(sprintf("n = %d, events = %d (%.1f%%), censored = %d\n",
            n, sum(df$event), 100 * mean(df$event), sum(1 - df$event)))
```

## Kaplan-Meier survival curves

The Kaplan-Meier estimator is the basic nonparametric estimate of $S(t)$:

```{r}
#| label: km
#| cache: true
#| fig-cap: "Kaplan-Meier survival curves by treatment arm. Treatment shifts the survival curve upward, but the gap reflects both the causal effect AND any baseline imbalance in age/sex."
#| fig-width: 8
#| fig-height: 5

km_fit <- survfit(Surv(time, event) ~ trt, data = df)
ggsurvplot(km_fit, data = df,
           pval = TRUE,
           risk.table = TRUE,
           xlab = "Time (years)",
           legend.labs = c("Control", "Treatment"))
```

The KM curve is descriptive. It is not enough for causal inference here
because treatment is related to baseline covariates.

## Cox proportional hazards with adjustment

The familiar regression approach is a Cox proportional hazards model with
covariate adjustment:

```{r}
#| label: cox
#| cache: true

cox_fit <- coxph(Surv(time, event) ~ trt + age + sex, data = df)
tidy(cox_fit, exp = TRUE, conf.int = TRUE) |>
  knitr::kable(digits = 3,
               caption = "Cox PH regression coefficients (exponentiated = hazard ratios)")
```

The estimated hazard ratio for treatment should be close to the true
$\exp(-0.4) \approx 0.67$.

### The interpretation problem with hazard ratios

The Cox hazard ratio is not a simple mean treatment effect. It compares
hazards among those still at risk at time $t$, and that risk set is itself
affected by treatment. I would report it, but not rely on it as the only
causal summary.

## Restricted Mean Survival Time (RMST)

The **restricted mean survival time** is the expected survival time up
to a cutoff $\tau$:

$$
\text{RMST}(\tau) = E[\min(T, \tau)] = \int_0^\tau S(t)\, dt.
$$

The RMST difference is easier to interpret:

$$
\Delta_{\text{RMST}}(\tau) = \text{RMST}_1(\tau) - \text{RMST}_0(\tau)
= \int_0^\tau [S_1(t) - S_0(t)]\, dt.
$$

It also does not require proportional hazards.

```{r}
#| label: rmst
#| cache: true

# Use survRM2 via riskRegression / manually computing from KM
tau <- 5  # restrict to 5 years

# Manual RMST: integrate the KM curve up to tau
rmst_arm <- function(time, event, tau) {
  fit <- survfit(Surv(time, event) ~ 1)
  s    <- summary(fit, times = c(0, sort(unique(time[event == 1])), tau),
                  extend = TRUE)
  surv_at_t <- s$surv
  t_at_t    <- s$time
  # Trapezoidal integration of S(t) from 0 to tau
  use      <- t_at_t <= tau
  t_vals   <- c(t_at_t[use], tau)
  s_vals   <- c(surv_at_t[use], surv_at_t[max(which(use))])
  sum(diff(t_vals) * head(s_vals, -1))
}

rmst_trt  <- rmst_arm(df$time[df$trt == 1], df$event[df$trt == 1], tau)
rmst_ctrl <- rmst_arm(df$time[df$trt == 0], df$event[df$trt == 0], tau)

cat(sprintf("RMST(τ=5) under treatment: %.3f years\n", rmst_trt))
cat(sprintf("RMST(τ=5) under control:   %.3f years\n", rmst_ctrl))
cat(sprintf("Difference:                %.3f years\n", rmst_trt - rmst_ctrl))
```

The RMST difference says how many survival years are gained up to the cutoff
$\tau$.

### Adjusted RMST via IPW

To adjust for confounding, weight the KM curve by the inverse propensity
score:

```{r}
#| label: rmst-ipw
#| cache: true

# Propensity score
ps_fit <- glm(trt ~ age + sex, data = df, family = binomial)
df$ps  <- predict(ps_fit, type = "response")
df$ipw <- ifelse(df$trt == 1, 1 / df$ps, 1 / (1 - df$ps))
df$ipw <- pmin(df$ipw, quantile(df$ipw, 0.99))  # trim

km_ipw <- survfit(Surv(time, event) ~ trt, data = df, weights = ipw)
print(km_ipw, rmean = tau)
```

The IPW-adjusted RMST removes the baseline imbalance.

## Inverse Probability of Censoring Weighting (IPCW)

If censoring is informative, for example if sicker patients drop out earlier,
treating censoring as random is biased. IPCW weights observations by the
inverse probability of remaining uncensored, conditional on covariates.

```{r}
#| label: ipcw
#| cache: true

# Model the censoring hazard
# Use 1 - event as the "censoring" indicator
cens_fit <- coxph(Surv(time, 1 - event) ~ trt + age + sex, data = df)

# Predict survival of censoring at each event time for each unit
df_sub <- df |>
  arrange(time)
times_at_risk <- df_sub$time
# G(t | X): probability of being uncensored at time t given X
# Use survfit on cens_fit to get S_C(t | X)
sc_fit <- survfit(cens_fit, newdata = df_sub)
sc_at_t <- diag(summary(sc_fit, times = df_sub$time, extend = TRUE)$surv)

# IPCW weight: 1 / P(C > T_i | X_i) for event times, 0 for censoring times
df_sub$ipcw <- ifelse(df_sub$event == 1, 1 / pmax(sc_at_t, 0.01), 0)
summary(df_sub$ipcw[df_sub$event == 1])
```

IPCW upweights observations that remain uncensored despite having a high
predicted chance of censoring.

## Doubly-robust survival estimation

The `riskRegression` package combines the outcome model, propensity score,
and censoring model:

```{r}
#| label: dr-survival
#| cache: true
#| warning: false

# Average treatment effect on survival probability at t = 5
# ATE(t) = P(T(1) > t) - P(T(0) > t)
df$trt_f <- factor(df$trt)
cox_for_ate <- coxph(Surv(time, event) ~ trt_f + age + sex, data = df,
                     x = TRUE, y = TRUE)
ps_for_ate  <- glm(trt_f ~ age + sex, data = df, family = binomial, x = TRUE)

ate_surv <- ate(
  event       = cox_for_ate,
  treatment   = ps_for_ate,
  data        = df,
  times       = 5,
  estimator   = c("Gformula", "IPTW", "AIPTW"),
  cause       = 1,
  se          = TRUE,
  B           = 0,
  verbose     = FALSE
)

summary(ate_surv)
```

The output reports three estimators side-by-side:

- **Gformula** — parametric g-computation using the Cox model only.
- **IPTW** — inverse probability of treatment weighting.
- **AIPTW** — augmented IPTW (doubly robust); the recommended default.

Agreement across these estimates is reassuring. Disagreement is a warning
about model specification.

## Hazard ratio vs RMST: which to report?

For causal reporting, RMST is often easier to defend than the hazard ratio:

| Quantity | Pro | Con |
|---|---|---|
| **Hazard ratio** | Compact summary; familiar | Requires PH; no causal interpretation; conditional on survivors |
| **RMST difference** | Directly interpretable as "years gained"; no PH assumption | Depends on cutoff $\tau$; less familiar |
| **Survival probability difference at t** | Direct probability statement | Depends on choice of $t$ |

In practice, report at least:

1. The KM curves with risk tables (description).
2. The Cox HR with adjustment (familiar comparison to literature).
3. The RMST difference at a clinically meaningful $\tau$ (interpretable
   causal contrast).
4. The DR ATE on survival probability at one or two follow-up times.

## Connections

- The [G-Methods chapter](g-methods.qmd) covers IPTW with time-varying
  treatments — directly applicable when treatment evolves during
  follow-up. The IPCW machinery here is the censoring analogue.
- The [Heterogeneous Effects chapter](heterogeneous-effects.qmd) ML
  methods extend to survival via causal survival forests
  (`grf::causal_survival_forest`).
- The [Bayesian Causal Inference chapter](bayesian-causal.qmd) extends
  to survival via Bayesian additive regression trees with a Weibull or
  piecewise-exponential likelihood (`BART::wbart` with the appropriate
  configuration).

## Summary

- Survival outcomes require methods that handle censoring.
- Cox models are familiar, but hazard ratios are hard to interpret causally.
- RMST differences are often the clearest causal summary.
- IPCW handles informative censoring.
- `riskRegression::ate` reports g-formula, IPTW, and AIPTW estimates for
  survival probabilities.
