# Survival Analysis with Causal Inference

```{r}
#| include: false
library(tidyverse)
library(survival)
library(survminer)
library(riskRegression)
library(broom)
```

```{r}
#| eval: false
library(tidyverse)
library(survival)
library(survminer)
library(riskRegression)
```

When the outcome is time-to-event, such as death, readmission, job exit, or
machine failure, ordinary regression is not enough. Some observations are
censored. We know the event had not happened by the end of follow-up, but we
do not know the event time. Dropping those observations or treating the
observed follow-up time as the event time creates bias.

I cover:

1. Cox proportional hazards models with covariate adjustment.
2. Inverse probability of censoring weighting (IPCW).
3. Restricted mean survival time (RMST) differences.
4. Doubly robust survival estimators in `riskRegression`.

## The challenge of censored outcomes

Censoring means we observe $\min(T_i, C_i)$ and $\delta_i = \mathbb{1}\{T_i \le C_i\}$,
where $T_i$ is the event time and $C_i$ is the censoring time. The
main quantity of interest is the survival function

$$
S(t) = P(T > t)
$$ {#eq-survival-causal-1}

the probability of surviving past time $t$. Treatment effects are usually
reported as:

- **Hazard ratio (HR)**: $HR = h_1(t) / h_0(t)$ (the ratio of hazards
  under treatment vs control). Easy to estimate but hard to interpret
  causally [@hernan-2010-hazards].
- **Restricted mean survival difference (RMST)**: $\int_0^\tau [S_1(t) - S_0(t)] dt$
  — the difference in mean survival up to time $\tau$. Causally
  interpretable as a "years gained" quantity.
- **Survival probability difference**: $S_1(t) - S_0(t)$ at a specified
  time. Reports a direct probability difference.

## Setup: simulated survival data

The data are simulated once and shared with the Julia companion, so both books
report the same numbers. The generator is `data/gen_survival.R`.

We draw $n = 1500$ people with $\text{age} \sim N(60, 10^2)$ and
$\text{sex} \sim \text{Bernoulli}(0.5)$. Treatment is
$\text{trt} \sim \text{Bernoulli}(\Lambda(-1.0 + 0.03(\text{age}-60) + 0.4\,\text{sex}))$,
so older people and one sex are more likely to be treated. Event times follow a
Weibull proportional-hazards model with shape $k = 1.4$ and scale
$\lambda_0 = 0.11$,

$$
S(t \mid X) = \exp\!\bigl(-(\lambda_0 t)^k e^{\eta}\bigr), \qquad
\eta = -0.4\,\text{trt} + 0.03(\text{age}-60) + 0.3\,\text{sex},
$$ {#eq-survival-causal-4}

so the true log hazard ratio for treatment is $-0.4$ and the true hazard ratio
is $e^{-0.4} = 0.670$. Censoring is exponential dropout with mean 40 years,
capped by administrative end of follow-up at 12 years, and we observe
$\min(T, C)$ with an event indicator.

Because the model is fully specified, the other estimands follow by
integration. Over the population distribution of age and sex, the true
five-year risk is 0.295 under treatment and 0.403 under control, a **risk
difference of $-0.108$**, and the true restricted mean survival to five years
is 4.341 years under treatment against 4.070 under control, an **RMST
difference of 0.271 years**. Note that treatment is confounded in the direction
that hides its benefit: age raises both the chance of treatment and the hazard,
so the treated are sicker at baseline.

```{r}
#| label: dgp
#| cache: true

df <- read_csv("data/survival_sim.csv", show_col_types = FALSE)
n <- nrow(df)
cat(sprintf("n = %d, events = %d (%.1f%%), censored = %d\n",
            n, sum(df$event), 100 * mean(df$event), sum(1 - df$event)))
```

Of the 1,500 people, 1,010 have an observed event (67.3%) and 490 are censored.

## Kaplan-Meier survival curves

The Kaplan-Meier estimator is the basic nonparametric estimate of $S(t)$:

```{r}
#| label: km
#| cache: true
#| fig-cap: "Kaplan-Meier survival curves by treatment arm. Treatment shifts the survival curve upward, but the gap reflects both the causal effect AND any baseline imbalance in age/sex."
#| fig-width: 8
#| fig-height: 5

km_fit <- survfit(Surv(time, event) ~ trt, data = df)
ggsurvplot(km_fit, data = df,
           pval = TRUE,
           risk.table = TRUE,
           xlab = "Time (years)",
           legend.labs = c("Control", "Treatment"))
```

The KM curve is descriptive. It is not enough for causal inference here
because treatment is related to baseline covariates.

## Cox proportional hazards with adjustment

The familiar regression approach is a Cox proportional hazards model with
covariate adjustment:

```{r}
#| label: cox
#| cache: true

cox_fit <- coxph(Surv(time, event) ~ trt + age + sex, data = df)
tidy(cox_fit, exp = TRUE, conf.int = TRUE) |>
  knitr::kable(digits = 3,
               caption = "Cox PH regression coefficients (exponentiated = hazard ratios)")
```

The estimated hazard ratio for treatment is 0.669 with a 95% interval of
$[0.582, 0.770]$, against the true $e^{-0.4} = 0.670$. The Cox model is
correctly specified here --- the DGP is proportional hazards with linear
covariate effects --- so this is a check that the machinery works, not evidence
about robustness. Age carries a hazard ratio of 1.035 per year and sex 1.347,
against the true $e^{0.03} = 1.030$ and $e^{0.3} = 1.350$.

### The interpretation problem with hazard ratios

The Cox hazard ratio is not a simple mean treatment effect. It compares
hazards among those still at risk at time $t$, and that risk set is itself
affected by treatment. I would report it, but not rely on it as the only
causal summary.

## Restricted Mean Survival Time (RMST)

The **restricted mean survival time** is the expected survival time up
to a cutoff $\tau$:

$$
\text{RMST}(\tau) = E[\min(T, \tau)] = \int_0^\tau S(t)\, dt.
$$ {#eq-survival-causal-2}

The RMST difference is easier to interpret:

$$
\Delta_{\text{RMST}}(\tau) = \text{RMST}_1(\tau) - \text{RMST}_0(\tau)
= \int_0^\tau [S_1(t) - S_0(t)]\, dt.
$$ {#eq-survival-causal-3}

It also does not require proportional hazards.

```{r}
#| label: rmst
#| cache: true

# Use survRM2 via riskRegression / manually computing from KM
tau <- 5  # restrict to 5 years

# Manual RMST: integrate the KM curve up to tau.
# (For production use, prefer a tested implementation such as
#  survRM2::rmst2; the helper below is for transparency.)
# Read the step function directly from fit$time / fit$surv, anchor
# S(0) = 1 explicitly, and integrate as a left-endpoint rectangle sum
# over the grid (0, event times <= tau, tau). This avoids relying on
# how summary.survfit emits a row at time 0 and never duplicates tau in
# a way that drops the final [last event, tau] interval.
rmst_arm <- function(time, event, tau) {
  fit  <- survfit(Surv(time, event) ~ 1)
  keep <- fit$time <= tau
  t_grid <- c(0, fit$time[keep], tau)
  # Survival height on each interval [t_j, t_{j+1}): S(0) = 1, then the
  # KM value just after each event; the trailing value is a placeholder
  # dropped by head(., -1) below.
  last_s <- if (any(keep)) fit$surv[keep][sum(keep)] else 1
  s_grid <- c(1, fit$surv[keep], last_s)
  sum(diff(t_grid) * head(s_grid, -1))
}

rmst_trt  <- rmst_arm(df$time[df$trt == 1], df$event[df$trt == 1], tau)
rmst_ctrl <- rmst_arm(df$time[df$trt == 0], df$event[df$trt == 0], tau)

cat(sprintf("RMST(τ=5) under treatment: %.3f years\n", rmst_trt))
cat(sprintf("RMST(τ=5) under control:   %.3f years\n", rmst_ctrl))
cat(sprintf("Difference:                %.3f years\n", rmst_trt - rmst_ctrl))
```

The unadjusted RMST is 4.351 years under treatment and 4.147 under control, a
difference of 0.203 years, against a true 0.271. The RMST difference says how
many survival years are gained up to the cutoff $\tau$, and this one understates
the gain by a quarter, because the treated are older and therefore at higher
baseline hazard.

### Adjusted RMST via IPW

To adjust for confounding, weight the KM curve by the inverse propensity
score:

```{r}
#| label: rmst-ipw
#| cache: true

# Propensity score
ps_fit <- glm(trt ~ age + sex, data = df, family = binomial)
df$ps  <- predict(ps_fit, type = "response")
df$ipw <- ifelse(df$trt == 1, 1 / df$ps, 1 / (1 - df$ps))
df$ipw <- pmin(df$ipw, quantile(df$ipw, 0.99))  # trim

km_ipw <- survfit(Surv(time, event) ~ trt, data = df, weights = ipw)
print(km_ipw, rmean = tau)
```

The weighted restricted means are 4.40 years under treatment and 4.12 under
control, a difference of 0.28 against the true 0.271. Weighting moved the
estimate from 0.203 to 0.28 and closed almost the whole gap. Note that the
`n` column now reports the weighted totals, 1,494 and 1,499, rather than the
461 and 1,039 actual records --- that is the pseudo-population the weights
construct.

## Inverse Probability of Censoring Weighting (IPCW)

If censoring is informative, for example if sicker patients drop out earlier,
treating censoring as random is biased. IPCW weights observations by the
inverse probability of remaining uncensored, conditional on covariates.

```{r}
#| label: ipcw
#| cache: true

# Model the censoring hazard
# Use 1 - event as the "censoring" indicator
cens_fit <- coxph(Surv(time, 1 - event) ~ trt + age + sex, data = df)

# Predict survival of censoring at each event time for each unit
df_sub <- df |>
  arrange(time)
times_at_risk <- df_sub$time
# G(t | X): probability of being uncensored at time t given X
# Use survfit on cens_fit to get S_C(t | X)
sc_fit <- survfit(cens_fit, newdata = df_sub)
# We only need each subject's OWN survival at their OWN event time, not the
# full cross product of all N subjects' curves at all N event times.
# summary(sc_fit, times=df_sub$time, extend=TRUE)$surv followed by diag()
# builds that full N x N matrix just to throw away everything off the
# diagonal -- for N in the tens of thousands this is a multi-GB allocation
# and can OOM. Look each subject's own time up in sc_fit's existing
# per-subject step function instead (sc_fit$surv is already an
# n_times x N matrix that survfit(newdata=) computes regardless; this just
# avoids building a second, larger one on top of it). For very large N,
# packages like `riskRegression`/`pec` avoid materializing even that
# per-subject curve matrix and would scale further.
sc_idx  <- findInterval(df_sub$time, sc_fit$time)
sc_at_t <- sc_fit$surv[cbind(pmax(sc_idx, 1), seq_len(nrow(df_sub)))]
sc_at_t[sc_idx == 0] <- 1  # before the first censoring event, S_C = 1

# IPCW weight: 1 / P(C > T_i | X_i) for event times, 0 for censoring times
df_sub$ipcw <- ifelse(df_sub$event == 1, 1 / pmax(sc_at_t, 0.01), 0)
summary(df_sub$ipcw[df_sub$event == 1])
```

IPCW upweights observations that remain uncensored despite having a high
predicted chance of censoring. The weights here run from 1.001 to 1.353 with a
mean of 1.149, which is mild. That is expected in this simulation: censoring is
exponential dropout with a 40-year mean plus an administrative cutoff, and
neither depends on the covariates in a way that would make censoring strongly
informative. The correction is available and small; in a study where sicker
patients drop out, it would not be.

## Doubly-robust survival estimation

The `riskRegression` package combines the outcome model, propensity score,
and censoring model:

```{r}
#| label: dr-survival
#| cache: true
#| warning: false

# Average treatment effect on survival probability at t = 5
# ATE(t) = P(T(1) > t) - P(T(0) > t)
df$trt_f <- factor(df$trt)
cox_for_ate <- coxph(Surv(time, event) ~ trt_f + age + sex, data = df,
                     x = TRUE, y = TRUE)
ps_for_ate  <- glm(trt_f ~ age + sex, data = df, family = binomial, x = TRUE)
# With right-censoring, the IPTW/AIPTW survival estimator needs a model for
# the censoring mechanism (a Cox model for time-to-censoring).
# Treatment belongs in the censoring model unless design knowledge rules it out.
# Censoring is treatment-independent in this simulation, so including trt_f
# changes little here -- but the assumption that has to hold is independent
# censoring *conditional on treatment and covariates*, plus censoring
# positivity, and the code should say so to be reusable.
cens_for_ate <- coxph(Surv(time, event == 0) ~ trt_f + age + sex, data = df,
                      x = TRUE, y = TRUE)

ate_surv <- ate(
  event       = cox_for_ate,
  treatment   = ps_for_ate,
  censor      = cens_for_ate,
  data        = df,
  times       = 5,
  estimator   = c("Gformula", "IPTW", "AIPTW"),
  cause       = 1,
  se          = TRUE,
  B           = 0,
  verbose     = FALSE
)

summary(ate_surv)

# The summary displays one estimator; put all three side by side
ate_tab <- as.data.frame(data.table::as.data.table(ate_surv, type = "diffRisk"))
knitr::kable(ate_tab[, c("estimator", "estimate", "se", "lower", "upper")],
             digits = 3,
             caption = "Risk difference at t = 5 under each estimator.")
```

The object contains three estimators, shown side-by-side in the table:

- **Gformula** — parametric g-computation using the Cox model only.
- **IPTW** — inverse probability of treatment weighting.
- **AIPTW** — augmented IPTW (doubly robust); usually the best default, though
  not unconditionally. It earns that status when the nuisance models are
  flexible and cross-fitted, treatment and censoring probabilities stay away
  from zero, and the censoring assumption holds conditional on treatment and
  covariates. Where positivity is thin it can be less stable than a simpler
  estimator.

All three agree, and all three are close to the truth: $-0.106$, $-0.095$ and
$-0.099$ against a true risk difference of $-0.108$. Treatment cuts the
five-year risk of the event by about ten percentage points. The g-formula has
the smallest standard error, 0.018 against 0.027 for the other two, because it
leans entirely on the Cox model, which is correctly specified here. That is an
efficiency gain bought with an assumption, and AIPTW's wider interval is the
price of not needing it.

Agreement across these estimates is reassuring. Disagreement is a warning
about model specification.

## Hazard ratio vs RMST: which to report?

For causal reporting, RMST is often easier to defend than the hazard ratio:

| Quantity | Pro | Con |
|---|---|---|
| **Hazard ratio** | Compact summary; familiar | Requires PH; no causal interpretation; conditional on survivors |
| **RMST difference** | Directly interpretable as "years gained"; no PH assumption | Depends on cutoff $\tau$; less familiar |
| **Survival probability difference at t** | Direct probability statement | Depends on choice of $t$ |

In practice, report at least:

1. The KM curves with risk tables (description).
2. The Cox HR with adjustment (familiar comparison to literature).
3. The RMST difference at a clinically meaningful $\tau$ (interpretable
   causal contrast).
4. The DR ATE on survival probability at one or two follow-up times.

## Connections

- The [G-Methods chapter](g-methods.qmd) covers IPTW with time-varying
  treatments — directly applicable when treatment evolves during
  follow-up. The IPCW machinery here is the censoring analogue.
- The [Heterogeneous Effects chapter](heterogeneous-effects.qmd) ML
  methods extend to survival via causal survival forests
  (`grf::causal_survival_forest`).
- The [Bayesian Causal Inference chapter](bayesian-causal.qmd) extends
  to survival via Bayesian additive regression trees for time-to-event
  outcomes (`BART::surv.bart` for discrete-time survival, or `BART::abart`
  for an accelerated-failure-time model with right-censoring).

## Summary

- Survival outcomes require methods that handle censoring.
- Cox models are familiar, but hazard ratios are hard to interpret causally.
- RMST differences are often the clearest causal summary.
- IPCW handles informative censoring.
- `riskRegression::ate` reports g-formula, IPTW, and AIPTW estimates for
  survival probabilities.
