# Survival Analysis with Causal Inference

```{r}
#| include: false
library(tidyverse)
library(survival)
library(survminer)
library(riskRegression)
library(broom)
```

```{r}
#| eval: false
library(tidyverse)
library(survival)
library(survminer)
library(riskRegression)
```

When the outcome is **time-to-event** — death, hospital re-admission,
employment termination, machine failure — standard causal inference
methods need to be adapted. Some observations are *censored*: we know
the event has not yet occurred at the end of follow-up, but we do not
know when it will. Standard regression on the event time would discard
or mishandle these observations and produce biased estimates.

This chapter covers the workhorses of survival-causal inference:

1. **The Cox proportional hazards model** with treatment adjustment —
   the classical regression approach to causal effects on time-to-event
   outcomes.
2. **Inverse probability of censoring weighting (IPCW)** — extends IPTW
   to handle informative dropout.
3. **Restricted mean survival time (RMST) differences** — a more
   interpretable, assumption-light alternative to hazard ratios.
4. **Doubly-robust survival estimators** — combine the outcome model
   and the censoring weights for efficiency.

## The challenge of censored outcomes

Censoring means we observe $\min(T_i, C_i)$ and $\delta_i = \mathbb{1}\{T_i \le C_i\}$,
where $T_i$ is the event time and $C_i$ is the censoring time. The
fundamental quantity of interest is the **survival function**

$$
S(t) = P(T > t)
$$

— the probability of surviving past time $t$. Treatment effects in
survival analysis are typically expressed as:

- **Hazard ratio (HR)**: $HR = h_1(t) / h_0(t)$ (the ratio of hazards
  under treatment vs control). Easy to estimate but hard to interpret
  causally (Hernán 2010).
- **Restricted mean survival difference (RMST)**: $\int_0^\tau [S_1(t) - S_0(t)] dt$
  — the difference in mean survival up to time $\tau$. Causally
  interpretable as a "years gained" quantity.
- **Survival probability difference**: $S_1(t) - S_0(t)$ at a specified
  time. Reports a direct probability difference.

## Setup: simulated survival data

```{r}
#| label: dgp
#| cache: true

df <- read_csv("data/survival_sim.csv", show_col_types = FALSE)
n <- nrow(df)
cat(sprintf("n = %d, events = %d (%.1f%%), censored = %d\n",
            n, sum(df$event), 100 * mean(df$event), sum(1 - df$event)))
```

## Kaplan-Meier survival curves

The starting point of any survival analysis is the **Kaplan-Meier (KM)
estimator** — a nonparametric estimate of $S(t)$:

```{r}
#| label: km
#| cache: true
#| fig-cap: "Kaplan-Meier survival curves by treatment arm. Treatment shifts the survival curve upward, but the gap reflects both the causal effect AND any baseline imbalance in age/sex."
#| fig-width: 8
#| fig-height: 5

km_fit <- survfit(Surv(time, event) ~ trt, data = df)
ggsurvplot(km_fit, data = df,
           pval = TRUE,
           risk.table = TRUE,
           xlab = "Time (years)",
           legend.labs = c("Control", "Treatment"))
```

The KM curve shows the *unadjusted* survival difference — useful for
description but not for causal inference, because the treatment was
assigned based on age.

## Cox proportional hazards with adjustment

The standard approach is the Cox proportional hazards regression with
covariate adjustment:

```{r}
#| label: cox
#| cache: true

cox_fit <- coxph(Surv(time, event) ~ trt + age + sex, data = df)
tidy(cox_fit, exp = TRUE, conf.int = TRUE) |>
  knitr::kable(digits = 3,
               caption = "Cox PH regression coefficients (exponentiated = hazard ratios)")
```

The estimated hazard ratio for treatment should be close to the true
$\exp(-0.4) \approx 0.67$.

### The interpretation problem with hazard ratios

The Cox HR is *not* a causal contrast in the usual potential-outcomes
sense — it averages over the distribution of survivors at each time,
which itself depends on the treatment effect (Hernán 2010). HR estimates
are useful for ranking effects but should be reported alongside more
interpretable quantities such as RMST differences.

## Restricted Mean Survival Time (RMST)

The **restricted mean survival time** is the expected survival time up
to a cutoff $\tau$:

$$
\text{RMST}(\tau) = E[\min(T, \tau)] = \int_0^\tau S(t)\, dt.
$$

The RMST difference between treatment arms is a clean causal contrast
when treatment is randomised (or when the conditioning set is right):

$$
\Delta_{\text{RMST}}(\tau) = \text{RMST}_1(\tau) - \text{RMST}_0(\tau)
= \int_0^\tau [S_1(t) - S_0(t)]\, dt.
$$

Crucially, it does not require the proportional hazards assumption.

```{r}
#| label: rmst
#| cache: true

# Use survRM2 via riskRegression / manually computing from KM
tau <- 5  # restrict to 5 years

# Manual RMST: integrate the KM curve up to tau
rmst_arm <- function(time, event, tau) {
  fit <- survfit(Surv(time, event) ~ 1)
  s    <- summary(fit, times = c(0, sort(unique(time[event == 1])), tau),
                  extend = TRUE)
  surv_at_t <- s$surv
  t_at_t    <- s$time
  # Trapezoidal integration of S(t) from 0 to tau
  use      <- t_at_t <= tau
  t_vals   <- c(t_at_t[use], tau)
  s_vals   <- c(surv_at_t[use], surv_at_t[max(which(use))])
  sum(diff(t_vals) * head(s_vals, -1))
}

rmst_trt  <- rmst_arm(df$time[df$trt == 1], df$event[df$trt == 1], tau)
rmst_ctrl <- rmst_arm(df$time[df$trt == 0], df$event[df$trt == 0], tau)

cat(sprintf("RMST(τ=5) under treatment: %.3f years\n", rmst_trt))
cat(sprintf("RMST(τ=5) under control:   %.3f years\n", rmst_ctrl))
cat(sprintf("Difference:                %.3f years\n", rmst_trt - rmst_ctrl))
```

The RMST difference says "treated patients gain on average X.X years of
survival up to 5 years" — a directly interpretable quantity for policy
or clinical reporting.

### Adjusted RMST via IPW

For an *adjusted* RMST that handles confounding, weight the KM curve by
the inverse propensity score:

```{r}
#| label: rmst-ipw
#| cache: true

# Propensity score
ps_fit <- glm(trt ~ age + sex, data = df, family = binomial)
df$ps  <- predict(ps_fit, type = "response")
df$ipw <- ifelse(df$trt == 1, 1 / df$ps, 1 / (1 - df$ps))
df$ipw <- pmin(df$ipw, quantile(df$ipw, 0.99))  # trim

km_ipw <- survfit(Surv(time, event) ~ trt, data = df, weights = ipw)
print(km_ipw, rmean = tau)
```

The IPW-adjusted RMST removes the baseline imbalance.

## Inverse Probability of Censoring Weighting (IPCW)

When censoring is *informative* — e.g. sicker patients drop out earlier
— treating censoring as random produces bias. The fix is to weight each
unit by the inverse of their probability of remaining uncensored,
conditional on covariates.

```{r}
#| label: ipcw
#| cache: true

# Model the censoring hazard
# Use 1 - event as the "censoring" indicator
cens_fit <- coxph(Surv(time, 1 - event) ~ trt + age + sex, data = df)

# Predict survival of censoring at each event time for each unit
df_sub <- df |>
  arrange(time)
times_at_risk <- df_sub$time
# G(t | X): probability of being uncensored at time t given X
# Use survfit on cens_fit to get S_C(t | X)
sc_fit <- survfit(cens_fit, newdata = df_sub)
sc_at_t <- diag(summary(sc_fit, times = df_sub$time, extend = TRUE)$surv)

# IPCW weight: 1 / P(C > T_i | X_i) for event times, 0 for censoring times
df_sub$ipcw <- ifelse(df_sub$event == 1, 1 / pmax(sc_at_t, 0.01), 0)
summary(df_sub$ipcw[df_sub$event == 1])
```

The IPCW weights upweight observations that survived "improbably long
without being censored" — replacing the dropouts they implicitly represent.

## Doubly-robust survival estimation

The `riskRegression` package implements doubly-robust estimators that
combine the outcome model (Cox) with the propensity score and the
censoring model. The result is robust to misspecification of either:

```{r}
#| label: dr-survival
#| cache: true
#| warning: false

# Average treatment effect on survival probability at t = 5
# ATE(t) = P(T(1) > t) - P(T(0) > t)
df$trt_f <- factor(df$trt)
cox_for_ate <- coxph(Surv(time, event) ~ trt_f + age + sex, data = df,
                     x = TRUE, y = TRUE)
ps_for_ate  <- glm(trt_f ~ age + sex, data = df, family = binomial, x = TRUE)

ate_surv <- ate(
  event       = cox_for_ate,
  treatment   = ps_for_ate,
  data        = df,
  times       = 5,
  estimator   = c("Gformula", "IPTW", "AIPTW"),
  cause       = 1,
  se          = TRUE,
  B           = 0,
  verbose     = FALSE
)

summary(ate_surv)
```

The output reports three estimators side-by-side:

- **Gformula** — parametric g-computation using the Cox model only.
- **IPTW** — inverse probability of treatment weighting.
- **AIPTW** — augmented IPTW (doubly robust); the recommended default.

Their agreement is reassuring; disagreement suggests model misspecification.

## Hazard ratio vs RMST: which to report?

Modern survival-causal reporting guidance (Royston-Parmar 2013,
Hernán 2010) increasingly favours RMST over HR for several reasons:

| Quantity | Pro | Con |
|---|---|---|
| **Hazard ratio** | Compact summary; familiar | Requires PH; no causal interpretation; conditional on survivors |
| **RMST difference** | Directly interpretable as "years gained"; no PH assumption | Depends on cutoff $\tau$; less familiar |
| **Survival probability difference at t** | Direct probability statement | Depends on choice of $t$ |

In practice, report at least:

1. The KM curves with risk tables (description).
2. The Cox HR with adjustment (familiar comparison to literature).
3. The RMST difference at a clinically meaningful $\tau$ (interpretable
   causal contrast).
4. The DR ATE on survival probability at one or two follow-up times
   (the doubly-robust modern default).

## Connections

- The [G-Methods chapter](g-methods.qmd) covers IPTW with time-varying
  treatments — directly applicable when treatment evolves during
  follow-up. The IPCW machinery here is the censoring analogue.
- The [Heterogeneous Effects chapter](heterogeneous-effects.qmd) ML
  methods extend to survival via causal survival forests
  (`grf::causal_survival_forest`).
- The [Bayesian Causal Inference chapter](bayesian-causal.qmd) extends
  to survival via Bayesian additive regression trees with a Weibull or
  piecewise-exponential likelihood (`BART::wbart` with the appropriate
  configuration).

## Summary

- **Survival outcomes** require methods that handle censoring (some
  units' event times are unobserved).
- **Cox proportional hazards** is the workhorse but the HR is not a
  causal contrast in the usual sense.
- **RMST differences** are interpretable and avoid the PH assumption —
  prefer for causal reporting.
- **IPCW** extends IPTW to handle informative dropout.
- **Doubly-robust survival** (`riskRegression::ate`) combines outcome
  models, treatment weights, and censoring weights for efficient and
  robust estimation.
- **Reporting**: combine KM curves (description), adjusted Cox (familiar),
  RMST difference (interpretable), and DR ATE on survival probability
  (modern default).

